---
title: "CD45Neg_exploratory_analysis"
author: "Dillon Corvino"
date: "25/01/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Built with R version `r getRversion()`

Exploratory analyses on the annotated CD45- dataset: cluster overview/stats,
cell cycle scoring, cluster-wise DEGs, melanocyte gene panel, chemokine/
cytokine expression (RQ1), and treatment-wise DEG scans (RQ2/RQ3). None of
this feeds the manuscript panels except where noted -- see
`scripts/manuscript_figures.R` for those (Sup Fig 4c reuses the
`chemokine_cytokine_expression` chunk's Custom_genes panel below). Split out
of the original combined `03_CD45Neg_data_analysis.Rmd` on 2026-08-14 and
archived here; the annotation-producing code that used to precede this
content now lives in `scripts/03_CD45Neg_data_annotation.Rmd`.

## Setup {.tabset}

### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)


# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- FALSE
do.plot <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())

# function found in Setup.R script
ggplot2_switch_version(desired_version = "3.4.4")

library("Seurat")
library("SeuratObject")
library("SeuratWrappers")
library("dplyr")
library("scCustomize")
library("tibble")
library("ggplot2")


# load annotated dataset (produced by scripts/03_CD45Neg_data_annotation.Rmd)
CD45neg.seurat <- LoadSeuratRds("saves/CD45neg_seurat.rds")
DefaultAssay(CD45neg.seurat) <- "SCT"

# Create new metadata column that groups all tumor together
CD45neg.seurat@meta.data <- CD45neg.seurat@meta.data %>%
  dplyr::mutate(celltype = case_when(Disease == "tumor" ~ "tumor",
                                     Disease == "normal" ~ cluster_anno))

```





## Overview of dataset {.tabset}

### Broad overview of dataset
```{r broad_overview}

if(do.plot){
  
  
  # create output dir
  output.dir <- paste0("results/CD45Neg_Overview/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  
  # set assay
  DefaultAssay(CD45neg.seurat) <- "SCT"
  head(CD45neg.seurat)
  
  vars.to.plot <- c("Treatment", "SampleID", "seurat_clusters", "cluster_anno", "Disease", "subclone")
  
  
  for(j in seq_along(vars.to.plot)){
    name.var <- gsub("\\.", "_", vars.to.plot[j])
    
    print(
      scCustomize::DimPlot_scCustom(CD45neg.seurat,
                                    reduction = "umap", 
                                    label = TRUE,
                                    figure_plot = TRUE,
                                    group.by = vars.to.plot[j]) 
    )
    
    
    dev.copy(pdf, paste0(output.dir, "CD45Neg_UMAP_", name.var, ".pdf"))
    dev.off()
    
  }
  
  
  # split umap plots
  print(
    scCustomize::DimPlot_scCustom(CD45neg.seurat,
                                  reduction = "umap", 
                                  label = TRUE,
                                  figure_plot = TRUE,
                                  group.by = "cluster_anno", 
                                  split.by = "Treatment") 
  )
  
  
  dev.copy(pdf, paste0(output.dir, "CD45Neg_UMAP_cluster_anno_split_by_treatment.pdf"))
  dev.off()
  

    print(
    scCustomize::DimPlot_scCustom(CD45neg.seurat,
                                  reduction = "umap", 
                                  label = TRUE,
                                  figure_plot = TRUE,
                                  group.by = "Disease", 
                                  split.by = "Treatment") 
  )
  
  
  dev.copy(pdf, paste0(output.dir, "CD45Neg_UMAP_Disease_split_by_treatment.pdf"))
  dev.off()
  
  
  }


```

### Cluster stats
```{r cluster_stats}

# create output dir
output.dir <- paste0("results/CD45Neg_Overview/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

#############################################
# Number of cells per cluster
#############################################

cluster.n <- table(CD45neg.seurat@meta.data$cluster_anno)

cluster.n <- as.data.frame(cluster.n)
cluster.n$Var1 <- factor(cluster.n$Var1, levels = cluster.n$Var1)


ggplot(cluster.n) +
  aes(x = Var1, y = Freq, fill = Var1) +
  geom_col() +
  scale_fill_viridis_d(option = "viridis", 
                       direction = -1) +
  labs(x = "Cluster ID", y = "Number of cells", title = "Number of cells per cluster") +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

ggsave(paste0(output.dir, "Barplot_number_of_cells_per_cluster.pdf"))


#############################################
# Freq of clusters across treatment 
#############################################

dist.table <- table(CD45neg.seurat@meta.data$cluster_anno, CD45neg.seurat@meta.data$Treatment)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Vehicle", "PDS"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Treatment", y = "% of treatment", title = "Frequency of clusters per treatment") +
  scale_fill_manual(values = c(Vehicle = "grey", PDS = "darkred")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_clusters_per_treatment.pdf"))


#############################################
# Freq of disease across treatment 
#############################################

dist.table <- table(CD45neg.seurat@meta.data$Disease, CD45neg.seurat@meta.data$Treatment)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Vehicle", "PDS"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Treatment", y = "% of treatment", title = "Frequency of disease per treatment") +
  scale_fill_manual(values = c(Vehicle = "grey", PDS = "darkred")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_disease_per_treatment.pdf"))




#############################################
# Freq of tumor subclones across treatment
#############################################

dist.table <- table(CD45neg.seurat@meta.data$subclone, CD45neg.seurat@meta.data$Treatment)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Vehicle", "PDS"))



ggplot(freq.table) +
 aes(x = Var1, y = Freq, fill = Var2) +
 geom_col() +
 scale_fill_hue(direction = 1) +
 theme_minimal()

ggsave(paste0(output.dir, "Barplot_freq_of_subclone_per_treatment.pdf"))



```

### Cell Cycle scoring
```{r cell_cycle}

# create output dir
output.dir <- paste0("results/CD45Neg_cell_cycle/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


################################
# Get cell cycle genes 
################################
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

# pseudo-map to mouse gene (captures most except Mlf1ip, Fam64a and Hn1)
s.genes <- stringr::str_to_title(s.genes)
g2m.genes <- stringr::str_to_title(g2m.genes)

#####################
# score cell cycle
#####################

CD45neg.seurat <- CellCycleScoring(CD45neg.seurat,
                                   s.features = s.genes,
                                   g2m.features = g2m.genes,
                                   set.ident = TRUE)


#####################
# Visualise score
#####################

head(CD45neg.seurat@meta.data)

# Feature plots 

# S-score 
scCustomize::FeaturePlot_scCustom(CD45neg.seurat, 
                                  feature = "S.Score", 
                                  reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Neg_S_score.pdf"))
dev.off()

# G2M-score 
scCustomize::FeaturePlot_scCustom(CD45neg.seurat, 
                                  feature = "G2M.Score", 
                                  reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Neg_G2M_score.pdf"))
dev.off()

# Dim plot split by phase
scCustomize::DimPlot_scCustom(CD45neg.seurat, 
                              group.by = "cluster_anno",
                              split.by = "Phase",
                              label = TRUE,
                              reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_CD45Neg_cluster_by_phase.pdf"))
dev.off()

scCustomize::DimPlot_scCustom(CD45neg.seurat, 
                              group.by = "celltype",
                              split.by = "Phase",
                              label = TRUE,
                              reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_CD45Neg_celltype_by_phase.pdf"))
dev.off()

# Stacked VlnPlot
scCustomize::Stacked_VlnPlot(CD45neg.seurat,
                             features = c("S.Score", "G2M.Score"), 
                             group.by = "cluster_anno")

dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_cell_cycle.pdf"))
dev.off()

# Stacked VlnPlot
scCustomize::Stacked_VlnPlot(CD45neg.seurat,
                             features = c("S.Score", "G2M.Score"), 
                             group.by = "celltype")

dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_cell_cycle_by_celltype.pdf"))
dev.off()



###################
# Quantification 
###################


# Frequency of phase per cluster
dist.table <- table(CD45neg.seurat@meta.data$cluster_anno, CD45neg.seurat@meta.data$Phase)

cluster.total <- rowSums(dist.table)

freq.table <- sweep(dist.table, 1, cluster.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin
rowSums(freq.table)

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("G1", "G2M", "S"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Predicted phase", y = "% of cluster", title = "Frequency of phase per cluster") +
  scale_fill_manual(values = c(G1 = "lightblue", G2M = "orange", S = "red")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_phase_per_cluster.pdf"))




# Frequency of phase per celltype

dist.table <- table(CD45neg.seurat@meta.data$celltype, CD45neg.seurat@meta.data$Phase)

cluster.total <- rowSums(dist.table)

freq.table <- sweep(dist.table, 1, cluster.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin
rowSums(freq.table)

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("G1", "G2M", "S"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Predicted phase", y = "% of celltype", title = "Frequency of phase per celltype") +
  scale_fill_manual(values = c(G1 = "lightblue", G2M = "orange", S = "red")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_phase_per_celltype.pdf"))




# Frequency of phase per treatment

dist.table <- table(CD45neg.seurat@meta.data$Treatment, CD45neg.seurat@meta.data$Phase)

treatment.total <- rowSums(dist.table)

freq.table <- sweep(dist.table, 1, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin
rowSums(freq.table)

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("G1", "G2M", "S"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Predicted phase", y = "% of treatment", title = "Frequency of phase per treatment") +
  scale_fill_manual(values = c(G1 = "lightblue", G2M = "orange", S = "red")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_phase_per_treatment.pdf"))


```


## DEG analysis {.tabset}

### Cluster-wise DEGs - calculate
```{r Clusterwise_DEGs_calculation}


# create output dir
output.dir <- paste0("results/CD45Neg_cluster_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  
  CD45neg.seurat <- PrepSCTFindMarkers(CD45neg.seurat)
  
  # set assay
  DefaultAssay(CD45neg.seurat) <- "SCT"
  
  
  # set idents
  Idents(CD45neg.seurat) <- "cluster_anno"
  
  CD45Neg.de <- FindAllMarkers(CD45neg.seurat,
                               assay = "SCT",
                               only.pos = TRUE, 
                               verbose = TRUE)
  
  
  # save to file 
  write.csv(CD45Neg.de, paste0(output.dir, "CD45Neg_DEGs.csv"))
  
  
  
}else{
  
  # load DEG table
  CD45Neg.de <- read.csv(paste0(output.dir, "CD45Neg_DEGs.csv"))
  
}



```

### Cluster-wise DEGs - heatmaps
```{r clusterwise_DEGs_heatmaps}

# create output dir
output.dir <- paste0("results/CD45Neg_cluster_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

if(do.plot){
  # aggregate expression for heatmap visualisation
  
  pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("cluster_anno"))
  
  
  
  goi <- CD45Neg.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::group_by(cluster) %>%
    dplyr::top_n(10, avg_log2FC) %>%
    dplyr::pull(gene)
  
  
  DoHeatmap(pseudo.cd45neg, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_top_10_DEGs_pseudobulk.pdf"))
  dev.off()
  
  
  
  
  # iterate over the top degs for each cluster 
  cluster.id <- unique(CD45Neg.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- CD45Neg.de %>%
      dplyr::filter(p_val_adj < 0.05) %>%
      dplyr::filter(cluster == cluster.id[i]) %>%
      dplyr::top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.cd45neg, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE) + 
        scico::scale_fill_scico(palette = "batlow", 
                                direction = 1, 
                                na.value = "white")
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
}

```

## Melanocyte gene visualisation {.tabset}

### Melanocyte-specific gene expression
```{r melanocyte_genes}

# create output dir
output.dir <- paste0("results/CD45Neg_Melanocyte_genes/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# list melanocyte genes 
goi <- c("Mitf", "Tyr", "Pmel", "Mlana", "Tyrp1", "Sox10")


# Plot Feature and Density plots iteratively over goi 
for(i in seq_along(goi)){
  
  # feature plots
  print(
    scCustomize::FeaturePlot_scCustom(CD45neg.seurat, 
                                      reduction = "umap", 
                                      features = goi[i])
    
  )
  
  dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Neg_", goi[i], ".pdf"))
  dev.off()
  
  
  # density plots
  
  print(
    scCustomize::Plot_Density_Custom(CD45neg.seurat, 
                                     reduction = "umap",
                                     features = goi[i], 
                                     custom_palette = batlow.pal)
  )
  dev.copy(pdf, paste0(output.dir, "UMAP_DensityPlot_CD45Neg_", goi[i], ".pdf"))
  dev.off()
  
  
  
}


##############################
# Stacked vln plot
##############################

scCustomize::Stacked_VlnPlot(CD45neg.seurat, 
                             features = goi,
                             group.by = "cluster_anno",
                             pt.size = 0)

dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_melanocyte_genes.pdf"))
dev.off()


###########
# heatmap
###########

# aggregate expression for heatmap visualisation
pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("cluster_anno"))

DoHeatmap(pseudo.cd45neg, 
          features = goi, 
          draw.lines = FALSE, 
          angle = 90,
          raster = FALSE) + 
  scico::scale_fill_scico(palette = "batlow", 
                          direction = 1, 
                          na.value = "white")

dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_melanocyte_genes.pdf"))
dev.off()


```


## RQ1: Which CD45Neg clusters express chemokines or cytokines {.tabset}

### Chemokine and Cytokine expression visualisation 
```{r chemokine_cytokine_expression}

# create output dir
output.dir <- paste0("results/CD45Neg_chemokine_cytokines/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# aggregate expression for heatmap visualisation
pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("cluster_anno"))


########################################
# List gene sets of interest to plot
########################################

cxc.genes <- grep("^Cxc.*", rownames(pseudo.cd45neg), value = T)
ccr.genes <- grep("^Ccr.*", rownames(pseudo.cd45neg), value = T)
il.genes <- grep("^Il.*", rownames(pseudo.cd45neg), value = T)
gzm.genes <- grep("^Gzm.*", rownames(pseudo.cd45neg), value = T)
ifn.genes <- grep("^Ifn.*", rownames(pseudo.cd45neg), value = T)
custom.genes <- c("Il1b", "Il6", "Il12a", "Ifng", "Tnf", "Ccl2", "Ccl3", "Ccl4", "Ccl5", "Cxcl1", "Cxcl9", "Cxcl10", "Cxcl13") # Il12b not present in CD45Neg dataset

goi <- list(Cxcr_genes = cxc.genes, 
            Ccr_genes = ccr.genes,
            Interleukin_genes = il.genes,
            Gzm_genes = gzm.genes,
            Ifn_genes = ifn.genes,
            Custom_genes = custom.genes)


###################
# Plot heatmaps
###################

for(i in seq_along(names(goi))){
  
  print(
    DoHeatmap(pseudo.cd45neg, 
              features = goi[[i]], 
              draw.lines = FALSE, 
              angle = 90,
              raster = FALSE) + 
      scico::scale_fill_scico(palette = "batlow", 
                              direction = 1, 
                              na.value = "white")
  )
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_", names(goi)[i], ".pdf"))
  dev.off()
  
}


##############################
# Stacked vln plot
##############################

for(i in seq_along(names(goi))){
  
  if(length(goi[[i]]) <= 15){
    
    
    print(
      scCustomize::Stacked_VlnPlot(CD45neg.seurat, 
                                   features = goi[[i]],
                                   group.by = "cluster_anno",
                                   pt.size = 0)
    )
    
    dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_", names(goi)[i], ".pdf"))
    dev.off()
    
  }else{
    
    temp.goi <- goi[[i]]
    
    # Split the vector into multiple vectors
    split_indices <- seq(1, length(temp.goi), by = 15)
    split_list <- lapply(split_indices, function(i) temp.goi[i:min(i + 14, length(temp.goi))])
    
    
    for(j in seq_along(length(split_list))){
      
      print(
        scCustomize::Stacked_VlnPlot(CD45neg.seurat, 
                                     features = split_list[[j]],
                                     group.by = "cluster_anno",
                                     pt.size = 0)
      )
      
      dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_", names(goi)[i], "_", j, ".pdf"))
      dev.off()
      
    }
    
  }
  
}


```

## RQ2: DEGs within clusters across treatment {.tabset}

### Treatment-wise DEGs - calculate
```{r Treatmentwise_DEGs_calculation}

# create output dir
output.dir <- paste0("results/CD45Neg_treatment_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  
  CD45neg.seurat <- PrepSCTFindMarkers(CD45neg.seurat)
  
  # set assay
  DefaultAssay(CD45neg.seurat) <- "SCT"
  
  
  
  # add annotation for treatment 
  CD45neg.seurat@meta.data$Treatment_cluster <- paste0(CD45neg.seurat@meta.data$Treatment, "_", CD45neg.seurat@meta.data$cluster_anno)
  
  # set Idents
  Idents(CD45neg.seurat) <- "Treatment_cluster"
  
  
  cluster.ids <- unique(CD45neg.seurat@meta.data$cluster_anno)
  
  # create list to store output
  DEG.list <- list()
  
  # run through DEG specifically between vehicle and treatment for each cluster 
  
  for(i in seq_along(cluster.ids)){
    temp.de <- FindMarkers(CD45neg.seurat,
                           assay = "SCT",
                           ident.1 = paste0("PDS_", cluster.ids[i]),
                           ident.2 = paste0("Vehicle_", cluster.ids[i]),
                           only.pos = FALSE, 
                           recorrect_umi = FALSE,
                           verbose = TRUE)
    
    DEG.list[[i]] <- temp.de
    
    # save to file 
    write.csv(temp.de, paste0(output.dir, "DEGs_PDS_", cluster.ids[i], "_vs_Vehicle_", cluster.ids[i], ".csv"))
    
    
  }
  # add names to DEG.list
  names(DEG.list) <- cluster.ids
  
  # save DEG list to file 
  saveRDS(DEG.list, paste0(output.dir, "DEG_list_object.rds"))
  
  
}else{
  
  DEG.list <- readRDS(paste0(output.dir, "DEG_list_object.rds"))
  
}


```

### Treatment-wise DEGs - Volcano
```{r treatmentwise_DEGs_volcano}

# visualise treatment-wise DEGs as volcano plots

# create output dir
output.dir <- paste0("results/CD45Neg_treatment_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(do.plot){
  
  for(i in 1:length(DEG.list)){
    
    FDR.theshold <- 0.01
    logFC.threshold <- 1
    
    # filter data
    input.data <- DEG.list[[i]] %>%
      dplyr::rename(logFC = avg_log2FC) %>%
      dplyr::rename(FDR = p_val_adj) %>%
      filter(FDR < FDR.theshold)
    
    # color points
    volcano.col <- usefulfunctions::colour.points.volcano(input.data,
                                                          increase.col = "Red",
                                                          decreased.col = "Blue",
                                                          FDR.cutoff = FDR.theshold, 
                                                          logFC.cutoff = logFC.threshold)
    # plot heatmap
    print(
      usefulfunctions::custom.enhanced.volcano(input.data, 
                                               pCutoff = FDR.theshold,
                                               FCcutoff = logFC.threshold,
                                               title = paste0(names(DEG.list)[i], " PDS vs Vehicle"),
                                               subtitle = paste0("adjPval threshold = ", FDR.theshold, " & log2FC threshold = ", logFC.threshold),
                                               colCustom = volcano.col)
    )
    dev.copy(pdf, paste0(output.dir, "VolcanoPlot_PDS_vs_Vehicle_cluster_", names(DEG.list)[i], ".pdf"))
    dev.off()
    
    
  }
  
}

```

### Export aggregated counts table
```{r export_aggregated_counts}

# create output dir
output.dir <- paste0("results/tables/CD45Neg/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  # Aggregate expression by cluster 
  pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("cluster_anno"))
  
  # Write to file
  temp.out <- GetAssayData(pseudo.cd45neg,
                           layer = "data",
                           assay = "SCT")
  
  write.csv(temp.out, paste0(output.dir, "CD45Neg_Aggregate_SCT_data_by_cluster.csv"))
  
  # Aggregate expression by cluster and treatment
  pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("cluster_anno", "Treatment"))
  # write to file
  temp.out <- GetAssayData(pseudo.cd45neg,
                           layer = "data",
                           assay = "SCT")
  
  write.csv(temp.out, paste0(output.dir, "CD45Neg_Aggregate_SCT_data_by_cluster_and_treatment.csv"))
  
  # clean environment
  rm(pseudo.cd45neg)
  rm(temp.out)
  
}



```


## RQ3: DEGs tumour clusters across treatment {.tabset}


```{r RQ3_DEG_tumor_clusters_across_treatment}


# create output dir
output.dir <- paste0("results/CD45Neg_tumor_clusters_DEGs_across_treatment/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

table(CD45neg.seurat@meta.data$cluster_anno, CD45neg.seurat@meta.data$Disease)

# create new metadata disease and treatment 
CD45neg.seurat@meta.data$disease_treat <- paste0(CD45neg.seurat@meta.data$Disease, "_", CD45neg.seurat@meta.data$Treatment)

Idents(CD45neg.seurat) <- CD45neg.seurat@meta.data$disease_treat



if(long.compute){
  
  
  CD45neg.seurat <- PrepSCTFindMarkers(CD45neg.seurat)
  
  # set assay
  DefaultAssay(CD45neg.seurat) <- "SCT"
  

    
    tumor.de <- FindMarkers(CD45neg.seurat,
                           assay = "SCT",
                           ident.1 = "tumor_PDS",
                           ident.2 = "tumor_Vehicle",
                           only.pos = FALSE, 
                           recorrect_umi = FALSE,
                           verbose = TRUE)
    
    
    
    # save to file 
    write.csv(tumor.de, paste0(output.dir, "DEGs_Tumor_PDS_vs_Tumor_Vehicle.csv"))
    
    
}



# plot DEGs 
if(do.plot){
  
    FDR.theshold <- 0.01
    logFC.threshold <- 1
    
    # filter data
    input.data <- tumor.de %>%
      dplyr::rename(logFC = avg_log2FC) %>%
      dplyr::rename(FDR = p_val_adj) %>%
      filter(FDR < FDR.theshold)
    
    # color points
    volcano.col <- usefulfunctions::colour.points.volcano(input.data,
                                                          increase.col = "Red",
                                                          decreased.col = "Blue",
                                                          FDR.cutoff = FDR.theshold, 
                                                          logFC.cutoff = logFC.threshold)
    # plot heatmap
    print(
      usefulfunctions::custom.enhanced.volcano(input.data, 
                                               pCutoff = FDR.theshold,
                                               FCcutoff = logFC.threshold,
                                               title = paste0("Tumor PDS vs Tumor vehicle"),
                                               subtitle = paste0("adjPval threshold = ", FDR.theshold, " & log2FC threshold = ", logFC.threshold),
                                               colCustom = volcano.col)
    )
    dev.copy(pdf, paste0(output.dir, "VolcanoPlot_Tumor_PDS_vs_Tumor_Vehicle.pdf"))
    dev.off()
    
    
  
    
  #####################
  # heatmaps of top DEGs
  #####################
  
  
  
  Idents(CD45neg.seurat) <- CD45neg.seurat@meta.data$disease_treat
  
  

  pseudo.tumor <- AggregateExpression(CD45neg.seurat,
                                           assays = "SCT",
                                           return.seurat = TRUE, 
                                           group.by = c("disease_treat"))
  
  
  # Extract top fold change DEGs 
  
  goi <- tumor.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(abs(avg_log2FC) > 2) %>%
    dplyr::arrange(desc(avg_log2FC)) %>%
    rownames()
  
  
  # plot heatmap
  DoHeatmap(pseudo.tumor, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "CD45Neg_Heatmap_Tumor_PDS_vs_Tumor_Vehicle_logfold_2.pdf"))
  dev.off()
  
  
  # extract just top 10 up and down degs
  goi.up <- tumor.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::top_n(10, avg_log2FC) %>%
    rownames()
  
  goi.dn <- tumor.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::top_n(-10, avg_log2FC) %>%
    rownames()
  
  goi <- unique(c(goi.up, goi.dn))
  
  DoHeatmap(pseudo.tumor, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "CD45Neg_Heatmap_Tumor_PDS_vs_Tumor_Vehicle_top_10_DEGs_up_down.pdf"))
  dev.off()
  
}
  


# clean environment 
  rm(pseudo.tumor)
  gc()


```



