---
title: "Nils_dataset_preprocessing"
author: "Dillon Corvino"
date: "25/01/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# Data 
# B16 - OVA 

# PDS = I.T - bind Gquadraplex

# Dataset 1 = day 4 post-treatment 

# E4 = vehicle CD45Neg
# F4 = vehicle CD45Pos
# G4 = PDS CD45Neg
# H4 = PDS CD45Pos

# Dataset 2 = day 9 post-treatment 
# E6 = vehicle CD45Pos
# F6 = PDS CD45Pos

# Research questions 
# How does PDS induces an anti-tumor immune response?
# Known that after PDS treatment, certain DAMPs are induced, intratumoral chemokine levels go up drastically and subsequntly immune cells are recruited into the tumor.
# Main research questions are:
#1. Which cell subsets are producing the chemokines
#2. Are there differences between macrophages and DCs between PDS and Vehicle (because there are DAMPs that should activate them)
#3. How does the treatment effect the tumor cells (is it possible to discriminate them from other tissue cells, if not they express ovalbumin...this should make it at least possible to identify them.)




# key genes
# IL1b, IL6, IL12, IFNG, TNFA, 

# CCL2,3,4,5 
# CXCL1, CXCL9,10,13 





```

# TO DO:
# re-calculate umap after removing contamination immune cells from CD45neg dataset
# clean pipeline
# clean annotation - cluster ID plus anno for cd45Pos at least
# clean metadata on objects 
# do DEG between treat and vehcile for both datasets 
# vis these differences. 
# clean pre-process script and move to subscripts for deg etc analysis. keep this just pre-processing and annotation etc etc. 

# General ToDo 
# add SoupX code to scRNAseq analysis template
# Add dublet detection code to scRNAseq analysis template



### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)


# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- TRUE
do.plot <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())

# script uses Seurat v5
#install.packages('remotes')
#remotes::install_github(repo = 'satijalab/seurat')


#setRepositories(ind = 1:3, addURLs = c('https://satijalab.r-universe.dev', 'https://bnprks.r-universe.dev/'))
#install.packages(c("BPCells", "presto", "glmGamPoi"))

#remotes::install_github("mojaveazure/seurat-object", quiet = TRUE)
#install.packages("Matrix")

#remotes::install_github("satijalab/seurat-data", quiet = TRUE)
#remotes::install_github("satijalab/azimuth", "seurat5", quiet = TRUE)
#remotes::install_github("satijalab/seurat-wrappers", "seurat5", quiet = TRUE)


# function found in Setup.R script
ggplot2_switch_version(desired_version = "3.4.4")



library("Seurat")
library("SeuratObject")
library("SeuratData")
library("SeuratWrappers")
library("dplyr")
library("scCustomize")

# perhaps these packages are not needed
#library("BPCells")
#library("presto")


# if older version is needed
# remotes::install_version("Seurat", "4.4.0", repos = c("https://satijalab.r-universe.dev", getOption("repos")))

########################################
# Parallelization of seurat functions
########################################

parallelization <- TRUE


if(parallelization){
  library(future)
  # check the current active plan
  plan()
  
  # detect cores
  ncores <- parallel::detectCores()
  ncores <- ncores-5
  # set plan
  plan("multisession", workers = ncores)
  plan()
  
  # set max limit for global variables
  # note this will drastically increase RAM usage, be aware of machine limits
  # set value is in bites therefore 1000*1024^2 = 1GB Therefore set to 10GBs 
  options(future.globals.maxSize = 10000 * 1024^2, 
          future.rng.onMisuse="ignore") 
}


```


## Pre-processing {.tabset}

### Ambient RNA correction
```{r ambient_RNA_correction}

if(long.compute){
  
  # Create output dir
  
  output.dir <- paste0("results/QC/Ambien_RNA_correction/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  
  # install packages if not already installed
  #install.packages("SoupX")
  #BiocManager::install("DropletUtils")
  
  # load package
  library("SoupX")
  
  
  
  # list all directories in data folder
  samples <- list.dirs(path = "data", recursive = FALSE)
  # keep only _outs dirs 
  samples <- grep(".*_outs", samples, value = T)
  
  for(i in seq_along(samples)){
    
    
    
    ##################
    # define input path
    ##################
    input.path <- paste0(samples[i])
    
    ##################
    # extract sample info
    ##################
    
    sample.id <- gsub("data/", "", samples[i])
    sample.id <- gsub("_outs", "", sample.id)
    
    
    # Only proceed if output dir does not exist
    output.dir.name <- paste0("data/", sample.id, "_strainedCounts")
    
    if(!dir.exists(output.dir.name)){
      print(paste0("Directory ", output.dir.name, " does not exists, preceeding with this sample"))
      
      ##################
      # load data 
      ##################
      # filtered and raw matrix as well as cluster info from cellranger analysis
      sc.temp <- SoupX::load10X(input.path)
      
      ##################
      # Estimate rho
      ##################
      # note default params are optimised for fresh (not nuclear) single cell experiments 
      # if processing nuclear scRNAseq, consider adjusting prior Rho default estimates
      
      sc.temp <- autoEstCont(sc.temp)
      
      dev.copy(pdf, paste0(output.dir, sample.id, "_Rho_estimation.pdf"))
      dev.off()
      
      ##################
      # Clean the data
      ##################
      sc.out <- adjustCounts(sc.temp)
      
      ##################
      # save data to file
      ##################
      DropletUtils:::write10xCounts(paste0(output.dir.name), sc.out)
      
      ##################
      # clean environment 
      ##################
      rm(list = c("sc.temp", "sc.out"))
      
    }else{
      print(paste0("Directory ", output.dir.name, " already exists, not recomputing this sample"))
      
    }
  }
  
}


# Misc info 
  # various tools to visualise and inspect ambient RNA removal and detection
  # plotMarkerDistribution(sc.temp)
  # 
  # # identify most zeroed genes
  # cntSoggy = rowSums(sc.temp$toc > 0)
  # cntStrained = rowSums(sc.out > 0)
  # mostZeroed = tail(sort((cntSoggy - cntStrained)/cntSoggy), n = 20)
  # mostZeroed
  # 
  # tail(sort(rowSums(sc.temp$toc > sc.out)/rowSums(sc.temp$toc > 0)), n = 20)
  # 
  # 
  # plotChangeMap(sc.temp, sc.out, "Il1b")
  # 
```


### create seurat object | merge and annotate
```{r create_seurat_object}

if(long.compute){
  
  # list all directories in data folder
  samples <- list.dirs(path = "data", recursive = FALSE)
  
  # keep only "strainedCounts" directories
  samples <- grep(".*_strainedCounts", samples, value = T)
  
  
  
  # create list to store data 
  seurat.list <- list()
  
  for(i in seq_along(samples)){
    
    # define input path
    input.path <- paste0(samples[i])
    
    # read in data (GEX and ADT/hashtag)
    sequence.data <- Seurat::Read10X(data.dir = paste0(input.path))
    
    
    # extract sequence run info 
    sample.id <- gsub("data/", "", samples[i])
    sample.id <- gsub("_strainedCounts", "", sample.id)
    
    # create seurat object with gene expression 
    seurat.temp <- CreateSeuratObject(counts = sequence.data,
                                      project = paste0(sample.id))
    
    # store data in a seurat list
    seurat.list[[paste0(sample.id)]] <- seurat.temp
    
    # clean environment 
    rm(list = c("seurat.temp", "sequence.data"))
    gc()
    
  }
  
  # for some reason creating these objects sometimes has a glitch where nfeatures etc of metadata is not generated 
  # if this happens, restart R session and rerun from scratch
  #seurat.list$E4@meta.data
  
  seurat.list$E4@meta.data
  
  
  
  ########################################
  # Merge and annotate object
  ########################################
  
  # change names of objects to more user friendly names 
  
  # Day 4
  # E4 = vehicle CD45Neg
  # F4 = vehicle CD45Pos
  # G4 = PDS CD45Neg
  # H4 = PDS CD45Pos
  
  # Day 9 
  # E6 = vehicle CD45Pos
  # F6 = PDS CD45Pos
  
  names(seurat.list)
  
  names(seurat.list)[names(seurat.list) == "E4"] <- "Day4_Vehicle_CD45Neg"
  names(seurat.list)[names(seurat.list) == "F4"] <- "Day4_Vehicle_CD45Pos"
  names(seurat.list)[names(seurat.list) == "G4"] <- "Day4_PDS_CD45Neg"
  names(seurat.list)[names(seurat.list) == "H4"] <- "Day4_PDS_CD45Pos"
  
  names(seurat.list)[names(seurat.list) == "E6"] <- "Day9_Vehicle_CD45Pos"
  names(seurat.list)[names(seurat.list) == "F6"] <- "Day9_PDS_CD45Pos"
  
  names(seurat.list)
  
  # create a vector of sequencing samples for later use in loops 
  sample.id <- names(seurat.list)
  sample.id
  
  
  # merge object for easier QC visualisation etc
  Nils.seurat <- scCustomize::Merge_Seurat_List(seurat.list, add.cell.ids = names(seurat.list))
  
  Nils.seurat # 32,285 genes across 55,127 cells
  unique(Nils.seurat@meta.data$orig.ident)
  
  # Use the sub function to replace the portion after the last underscore in each row name.
  Nils.seurat@meta.data$SampleID <- sub("^(.*)_.*$", "\\1", rownames(Nils.seurat@meta.data))
  Nils.seurat@meta.data$Time <- gsub("_.*", "", Nils.seurat@meta.data$SampleID)
  Nils.seurat@meta.data$Treatment <- gsub("^[^_]+_([^_]+)_.*$", "\\1", Nils.seurat@meta.data$SampleID)
  Nils.seurat@meta.data$Compartment <- gsub(".*_", "", Nils.seurat@meta.data$SampleID)
  
  
  unique(Nils.seurat@meta.data$SampleID)
  unique(Nils.seurat@meta.data$Time)
  unique(Nils.seurat@meta.data$Treatment)
  unique(Nils.seurat@meta.data$Compartment)
  
  # CD45Neg and CD45Pos datasets are very different and need to be QC'd differently and integrated differently 
  # split object to process compartments seperately 
  Nils.seurat <- SplitObject(Nils.seurat, split.by = "Compartment")
  
  # fully merge data by also combining layers for all assays, for clarity, thus downstream will explicitly split layers when required
  # layers can be combined for visualisation and subsetting 
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = JoinLayers, 
                        assay = "RNA")
  Nils.seurat
  
  rm(seurat.list)
  
}


```


## QC and Filter {.tabset}

### QC & filter

```{r QC_and_filter}

if(long.compute){
  
  # Create output dir
  
  output.dir <- paste0("results/QC/filtering/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  for(i in seq_along(names(Nils.seurat))){
    
    # set idents to sampleID
    Idents(Nils.seurat[[i]]) <- Nils.seurat[[i]]$SampleID
    
    # set default assay 
    DefaultAssay(Nils.seurat[[i]]) <- "RNA"
    
  }
  
  
  # Add mitochondrial and ribosomal gene percentages
  
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = Add_Mito_Ribo, 
                        species = "Mouse") 
  
  # Add cell complexity/novelty QC metric
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = Add_Cell_Complexity) 
  
  
  
  
  # create cutoff dataframe for easy plotting and interation 
  
  cutoff.df <- as.data.frame(matrix(nrow = 2, ncol = 7))
  
  colnames(cutoff.df) <- c("feature_low", "feature_high", "umi_low", "umi_low", "mito_high", "ribo_low", "complexity_low")
  
  rownames(cutoff.df) <- c("CD45Neg", "CD45Pos")
  # Setting cutoff values 
  cutoff.df$feature_low <- c(200, 200)
  cutoff.df$feature_high <- c(5500, 5500)
  cutoff.df$umi_low <- c(1000, 1000)
  cutoff.df$umi_high <- c(45000, 45000)
  cutoff.df$mito_high <- c(10, 10)
  cutoff.df$ribo_low <- c(10, 6)
  cutoff.df$complexity_low <- c(0.75, 0.8)
  
  
  ########################
  # QC plots
  ########################
  if(do.plot){
    
    for(i in seq_along(names(Nils.seurat))){
      # nFeatures
      print(
        QC_Plots_Genes(seurat_object = Nils.seurat[[i]], 
                       low_cutoff = cutoff.df$feature_low[i], 
                       high_cutoff = cutoff.df$feature_high[i],
                       pt.size = 0.05) #how many genes per cell? See where to set the cutoffs later
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_genes.pdf"))
      dev.off()
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "nFeature_RNA",
                     low_cutoff = cutoff.df$feature_low[i], 
                     high_cutoff = cutoff.df$feature_high[i],
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_genes_histo.pdf"))
      dev.off()
      
      
      # UMI
      print(
        QC_Plots_UMIs(seurat_object = Nils.seurat[[i]], 
                      low_cutoff = cutoff.df$umi_low[i], 
                      high_cutoff = cutoff.df$umi_high[i], 
                      y_axis_log = T,
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_UMI.pdf"))
      dev.off()
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "nCount_RNA",
                     low_cutoff = cutoff.df$umi_low[i], 
                     high_cutoff = cutoff.df$umi_high[i],
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_UMI_histo.pdf"))
      dev.off()
      
      
      
      
      # Mito percentage
      print(
        QC_Plots_Mito(seurat_object = Nils.seurat[[i]], 
                      high_cutoff = cutoff.df$mito_high[i],
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_mito.pdf"))
      dev.off()
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "percent_mito",
                     high_cutoff = cutoff.df$mito_high[i],
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_mito_histo.pdf"))
      dev.off()
      
      
      
      
      # Ribo percentage
      print(
        QC_Plots_Feature(seurat_object = Nils.seurat[[i]],
                         feature = "percent_ribo",
                         low_cutoff = cutoff.df$ribo_low[i],
                         pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_Ribo.pdf"))
      dev.off()
      
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "percent_ribo",
                     low_cutoff = cutoff.df$ribo_low[i], 
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_Ribo_histo.pdf"))
      dev.off()
      
      
      
      
      # Complexity
      print(
        QC_Plots_Complexity(seurat_object = Nils.seurat[[i]], 
                            low_cutoff = cutoff.df$complexity_low[i],
                            pt.size = 0.05)
      ) 
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_complexity.pdf"))
      dev.off()
      
      
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "log10GenesPerUMI",
                     low_cutoff = cutoff.df$complexity_low[i], 
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_complexity_histo.pdf"))
      dev.off()
      
      
    }
    
  }
  
  
  # Note, there are a large amount of cells with low ribosomal gene content. 
  # these are more preventant in PDS treated CD45Pos sample 
  # it is unclear what this means, but as filtering on ribosomal gene content is not a recommended step in fabian theis best scRNAseq practices, i will omit this
  # therefore, cells regardless of ribosomal content will be kept 
  
  
  
  # Filter dataset
  
  for(i in seq_along(names(Nils.seurat))){
    
    Nils.seurat[[i]] <- subset(Nils.seurat[[i]], 
                               subset = nFeature_RNA > cutoff.df$feature_low[i] & 
                                 nFeature_RNA < cutoff.df$feature_high[i] & 
                                 percent_mito < cutoff.df$mito_high[i] & 
                                 #percent_ribo > cutoff.df$ribo_low[i] & # do not filter for ribosomal content
                                 log10GenesPerUMI > cutoff.df$complexity_low[i])
    
  }
  
  
  
  if(do.plot){
    
    for(i in seq_along(names(Nils.seurat))){
      # nFeatures
      print(
        QC_Plots_Genes(seurat_object = Nils.seurat[[i]], 
                       low_cutoff = cutoff.df$feature_low[i], 
                       high_cutoff = cutoff.df$feature_high[i],
                       pt.size = 0.05) #how many genes per cell? See where to set the cutoffs later
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_genes_filtered.pdf"))
      dev.off()
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "nFeature_RNA",
                     low_cutoff = cutoff.df$feature_low[i], 
                     high_cutoff = cutoff.df$feature_high[i],
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_genes_filtered_histo.pdf"))
      dev.off()
      
      
      
      # UMI
      print(
        QC_Plots_UMIs(seurat_object = Nils.seurat[[i]], 
                      low_cutoff = cutoff.df$umi_low[i], 
                      high_cutoff = cutoff.df$umi_high[i], 
                      y_axis_log = T,
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_UMI_filtered.pdf"))
      dev.off()
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "nCount_RNA",
                     low_cutoff = cutoff.df$umi_low[i], 
                     high_cutoff = cutoff.df$umi_high[i],
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_UMI_filtered_histo.pdf"))
      dev.off()
      
      # Mito percentage
      print(
        QC_Plots_Mito(seurat_object = Nils.seurat[[i]], 
                      high_cutoff = cutoff.df$mito_high[i],
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_mito_filtered.pdf"))
      dev.off()
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "percent_mito",
                     high_cutoff = cutoff.df$mito_high[i],
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_mito_filtered_histo.pdf"))
      dev.off()
      
      
      # Ribo percentage
      print(
        QC_Plots_Feature(seurat_object = Nils.seurat[[i]],
                         feature = "percent_ribo",
                         high_cutoff = cutoff.df$ribo_high[i],
                         pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_Ribo_filtered.pdf"))
      dev.off()
      
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "percent_ribo",
                     low_cutoff = cutoff.df$ribo_low[i], 
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_Ribo_filtered_histo.pdf"))
      dev.off()
      
      
      # Complexity
      print(
        QC_Plots_Complexity(seurat_object = Nils.seurat[[i]], 
                            high_cutoff = cutoff.df$complexity_high[i],
                            pt.size = 0.05)
      ) 
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_complexity_filtered.pdf"))
      dev.off()
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "log10GenesPerUMI",
                     low_cutoff = cutoff.df$complexity_low[i], 
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_complexity_filtered_histo.pdf"))
      dev.off()
      
    }
    
  }
  
  Nils.seurat$CD45Neg  # 32,285 features by 13,562 cells
  Nils.seurat$CD45Pos # 32,285 features by 34,557 cells
  
  table(Nils.seurat$CD45Neg@meta.data$Treatment) # 7,698 PDS and 5,864 Vehicle
  
  table(Nils.seurat$CD45Pos@meta.data$Treatment) # 17351 PDS and 17206 Vehicle
  table(Nils.seurat$CD45Pos@meta.data$Time) # 18,032 Day 4 and 16,525 Day 9
  table(Nils.seurat$CD45Pos@meta.data$Time, Nils.seurat$CD45Pos@meta.data$Treatment)
  
  #        PDS Vehicle
  #  Day4 8082    9950
  # Day9 9269    7256
  
}


```


## Normalisation and integration {.tabset}

## Normalise & integrate data
```{r Normalise_and_integrate_data}

if(long.compute){
  
  # Note, 
  # I compared CCA integration and rpca integration, looks cca is better fit for this dataset 
  # also see documentation on cca vs rpca integration, cca more aligns with this dataset 
  
  # perform sct and default/standard normalisation before cca integration to compare both normalisation methods 
  
  
  
  # split object up based on sequencing run / in seurat V5 these are splitting the layers
  Nils.seurat$CD45Neg[["RNA"]] <- split(Nils.seurat$CD45Neg[["RNA"]],  f = Nils.seurat$CD45Neg$SampleID)
  Nils.seurat$CD45Pos[["RNA"]] <- split(Nils.seurat$CD45Pos[["RNA"]],  f = Nils.seurat$CD45Pos$SampleID)
  
  
  
  #####################################################
  # Normalise with scTransform and integrate data
  #####################################################
  # note error with more recent version of matrixStats resulting in "useNames = NA is defunct" error. downgrade to matrixStats version 1.1.0 as a workaround
  # https://github.com/satijalab/seurat/issues/7501#issuecomment-1854571904
  #remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts  
  
  DefaultAssay(Nils.seurat[[1]]) <- "RNA"
  DefaultAssay(Nils.seurat[[2]]) <- "RNA"
  
  # note RNA assay is split into run level layers, below script will apply normalisation etc layer-wise
  
  # Normalise and scale with SCTransform v2
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = SCTransform, 
                        method = "glmGamPoi", 
                        vars.to.regress = "percent_mito", 
                        vst.flavor = "v2",
                        verbose = TRUE)
  
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = RunPCA, 
                        reduction.name = "pca_sct")
  
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = RunUMAP, 
                        reduction = "pca_sct", 
                        dims = 1:30, 
                        reduction.name = "umap.unintegrated.sct")
  
  
  DefaultAssay(Nils.seurat[[1]]) <- "SCT"
  DefaultAssay(Nils.seurat[[2]]) <- "SCT"
  
  
  # integrate layers using cca integration
  
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = IntegrateLayers, 
                        orig.reduction = "pca_sct",
                        new.reduction = "integrated.cca.sct",
                        method = CCAIntegration,
                        normalization.method = "SCT",
                        verbose = F)
  
  
  
  # find clusters and dim reduction using integrated data
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = FindNeighbors, 
                        dims = 1:30,
                        reduction = "integrated.cca.sct")
  
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = FindClusters, 
                        resolution = 1.5, 
                        cluster.name = "clusters_cca_sct")
  
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = RunUMAP, 
                        reduction = "integrated.cca.sct",
                        dims = 1:30, 
                        reduction.name = "umap.cca.sct")
  
  
}


```

## evaluate integration 
```{r eval_integration}

if(long.compute){
  
  if(do.plot){
    
    # create output dir
    output.dir <- paste0("results/QC/integration/")
    
    if(!dir.exists(output.dir)){
      dir.create(output.dir, 
                 recursive = T)
      
    }
    
    
    # set up variables
    unintegrated.reduction <- "umap.unintegrated.sct"
    integrated.reduction <- "umap.cca.sct"
    title.var <- "SCT"
    
    
    # evaluate integration and dim reduction
    
    # set default assay
    DefaultAssay(Nils.seurat[[1]]) <- "SCT"
    DefaultAssay(Nils.seurat[[2]]) <- "SCT"
    
    Nils.seurat$CD45Pos@meta.data
    
    
    
    for(j in seq_along(names(Nils.seurat))){
      
      vars.plot <- c("Treatment", "SampleID", "seurat_clusters", "Time")
      
      for(i in seq_along(vars.plot)){
        
        
        # remove "." from name
        name.var <- gsub("\\.", "_", vars.plot[i])
        
        before.int <- DimPlot(Nils.seurat[[j]],
                              reduction = unintegrated.reduction, 
                              label = T,
                              group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("Unintegrated ", title.var, " Normalisation"))
        
        after.int <- DimPlot(Nils.seurat[[j]],
                             reduction = integrated.reduction, 
                             label = T,
                             group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("CCA Integration using ", title.var, " Normalisation"))
        
        print(before.int | after.int)
        dev.copy(pdf, paste0(output.dir, "UMAP_before_after_integration_", name.var, "_", names(Nils.seurat)[j], "_dataset.pdf"))
        dev.off()
        
        
        
      }
    }
    
    
  }
  
}
```


## Cluster identification and UMAP optimisation {.tabset}

### Cluster optimisation
```{r Cluster_optimisation}


if(long.compute){
  
  
  # create output dir
  output.dir <- paste0("results/QC/Cluster_optimisation/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  # inspect object
  Graphs(Nils.seurat$CD45Neg, slot = NULL)
  Neighbors(Nils.seurat$CD45Neg, slot = NULL)
  Reductions(Nils.seurat$CD45Neg, slot = NULL)
  
  # set assay 
  DefaultAssay(Nils.seurat[[1]]) <- "SCT"
  DefaultAssay(Nils.seurat[[2]]) <- "SCT"
  
  
  # Set seed for reproducibility
  set.seed(42)
  
  # Calculate neighbours graph using integrated data reduction 
  # this was previously calculated as part of normalisation and integration code chunk, but put here for clarity and to ensure neighbour graph is calculated on correct integrated dataset
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = FindNeighbors, 
                        dims = 1:30,
                        reduction = "integrated.cca.sct")
  
  
  # explicitly use the SCT_snn graph calculated for finding clusters 
  
  ########################################################
  # Find clusters over an iteration of resolutions
  ########################################################
  res.var <- 0
  res.max <- 2
  while (res.var <= res.max) {
    
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = FindClusters, 
                          resolution = res.var, 
                          graph.name = "SCT_snn")  
    res.var = res.var + 0.1
  }
  
  
  ############################
  # Visualise cluster tree
  ############################
  
  # Use clustree package to visualise impact of cluster resolution on clusters determined
  
  # load package
  # install.packages("clustree")
  
  ggplot2_switch_version(desired_version = "3.5")

  library(clustree) # clusttree needs to be loaded, error occurs if just call function from namespace
  
  head(Nils.seurat[[2]][[]])
  
  for(i in seq_along(names(Nils.seurat))){
    
    print(
      clustree::clustree(Nils.seurat[[i]], 
                         prefix = "SCT_snn_res.",
                         layout = "sugiyama")
    )
    dev.copy(pdf, paste0(output.dir, "dataset_", names(Nils.seurat)[i], "_clusttree.pdf"))
    dev.off()
    
    
  }
  
  # switch back to 3.4.4
    ggplot2_switch_version(desired_version = "3.4.4")

  ###############################
  # Visualise clusters on UMAP
  ###############################
  
  for(i in seq_along(names(Nils.seurat))){
    
    res.var <- 0
    
    while (res.var <= res.max) {
      print(
        DimPlot(Nils.seurat[[i]],
                reduction = "umap.cca.sct", 
                label = T,
                group.by = paste0("SCT_snn_res.", res.var)) + 
          NoLegend() + 
          ggtitle(paste0("SCT_snn_res.", res.var))
      )
      
      dev.copy(pdf, paste0(output.dir, "dataset_", names(Nils.seurat)[i], "_umap_cluster_resolution_", res.var, ".pdf"))
      dev.off()
      
      res.var = res.var + 0.1
      
    }
    
  }
  
  
  ##################################################
  # Set resolution level | and set seurat_clusters
  ##################################################
  
  # emperically picked resolution of: 
  # CD45Pos = 0.4
  # CD45Neg = 0.4
  
  # set seurat clusters to this level 
  Nils.seurat$CD45Pos@meta.data$seurat_clusters <- Nils.seurat$CD45Pos@meta.data$SCT_snn_res.0.4
  Nils.seurat$CD45Neg@meta.data$seurat_clusters <- Nils.seurat$CD45Neg@meta.data$SCT_snn_res.0.4
  
  length(unique(Nils.seurat$CD45Pos@meta.data$seurat_clusters)) # 17 clusters
  length(unique(Nils.seurat$CD45Neg@meta.data$seurat_clusters)) # 12 clusters
  
  
  
  # Visualise clusters on UMAP projection 
  DimPlot(Nils.seurat$CD45Pos,
          reduction = "umap.cca.sct", 
          label = T,
          group.by = c("seurat_clusters")) + 
    NoLegend() + 
    ggtitle("SCT_snn_res.0.4")
  
  dev.copy(pdf, paste0(output.dir, "UMAP_CD45Pos_dataset_seurat_clusters.pdf"))
  dev.off()
  
  
  DimPlot(Nils.seurat$CD45Neg,
          reduction = "umap.cca.sct", 
          label = T,
          group.by = c("seurat_clusters")) + 
    NoLegend() + 
    ggtitle("SCT_snn_res.0.4")
  
  dev.copy(pdf, paste0(output.dir, "UMAP_CD45Neg_dataset_seurat_clusters.pdf"))
  dev.off()
  
}


```

### Optimise UMAP dim reduction params
```{r umap_optimisation}

if(long.compute){
  
  
  # create output dir
  output.dir <- paste0("results/QC/UMAP_optimisation/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  # Optimise UMAP projection 
  
  # set init conditions and step sizes
  init.min.dist = 0.001
  init.neigh = 5
  min.dist.step = 2 
  neigh.step = 20
  
  
  for(i in seq_along(names(Nils.seurat))){
    
    grouping.var <- "seurat_clusters"
    
    min.dist.val <- init.min.dist
    
    # start loop 
    while (min.dist.val < 1) {
      
      n.neigh.val <- init.neigh
      
      while (n.neigh.val <= 50) {
        
        print(paste0("Calculating UMAP for min.dist = ", 
                     min.dist.val, " & n.neighbor = ", n.neigh.val))
        
        # caculate UMAP
        temp <- Seurat::RunUMAP(object = Nils.seurat[[i]],
                                reduction = "integrated.cca.sct", 
                                dims = 1:30,
                                n.neighbors = n.neigh.val, 
                                min.dist = min.dist.val,
                                reduction.name = "umap.temp",
                                seed.use = 42)
        
        # print UMAP
        p1 <- Seurat::DimPlot(object = temp, 
                              reduction = "umap.temp", 
                              label = TRUE, 
                              label.size = 4, 
                              group.by = grouping.var) + 
          ggtitle(paste0(grouping.var, " UMAP Min dist = ", 
                         min.dist.val, " n.val = ", n.neigh.val)) +
          NoLegend()
        
        
        
        print(p1)
        
        dev.copy(pdf, 
                 paste0(output.dir,  names(Nils.seurat)[i], 
                        "_UMAP_Min_dist_", min.dist.val, "_neighval_", n.neigh.val, ".pdf"))
        dev.off()
        
        n.neigh.val <- n.neigh.val + neigh.step
      }
      min.dist.val <- min.dist.val * min.dist.step
    }
    
    
  }
  
  ###########################################
  # Calculate UMAP using optimised params
  ###########################################
  
  # just use default params
  
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = RunUMAP,
                        reduction = "integrated.cca.sct",
                        dims = 1:30)
  
  Seurat::DimPlot(object = Nils.seurat$CD45Neg, 
                  reduction = "umap", 
                  label = TRUE, 
                  label.size = 4, 
                  group.by = "seurat_clusters") +
    NoLegend()
  
  Seurat::DimPlot(object = Nils.seurat$CD45Pos, 
                  reduction = "umap", 
                  label = TRUE, 
                  label.size = 4, 
                  group.by = "seurat_clusters") +
    NoLegend()
}

```

### Save seurat object
```{r save_seurat_object_2}

if(long.compute){
  
  library("SeuratObject")
  
  SaveSeuratRds(Nils.seurat$CD45Pos, 
                file = "saves/Preprocessed_CD45Pos.rds")
  
  SaveSeuratRds(Nils.seurat$CD45Neg, 
                file = "saves/Preprocessed_CD45Neg.rds")
  
}

```



