---
title: "Nils_dataset_preprocessing"
author: "Dillon Corvino"
date: "25/01/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# Data 
# B16 - OVA 

# PDS = I.T - bind Gquadraplex
# day 4 post-treatment 




# E4 = vehicle CD45Neg
# F4 = vehicle CD45Pos
# G4 = PDS CD45Neg
# H4 = PDS CD45Pos


# Research questions 
# How does PDS induces an anti-tumor immune response?
# Known that after PDS treatment, certain DAMPs are induced, intratumoral chemokine levels go up drastically and subsequntly immune cells are recruited into the tumor.
# Main research questions are:
#1. Which cell subsets are producing the chemokines
#2. Are there differences between macrophages and DCs between PDS and Vehicle (because there are DAMPs that should activate them)
#3. How does the treatment effect the tumor cells (is it possible to discriminate them from other tissue cells, if not they express ovalbumin...this should make it at least possible to identify them.)




# key genes
# IL1b, IL6, IL12, IFNG, TNFA, 

# CCL2,3,4,5 
# CXCL1, CXCL9,10,13 





```

# TO DO:
# re-calculate umap after removing contamination immune cells from CD45neg dataset
# clean pipeline
# clean annotation - cluster ID plus anno for cd45Pos at least
# clean metadata on objects 
# do DEG between treat and vehcile for both datasets 
# vis these differences. 
# clean pre-process script and move to subscripts for deg etc analysis. keep this just pre-processing and annotation etc etc. 

# General ToDo 
# add SoupX code to scRNAseq analysis template
# Add dublet detection code to scRNAseq analysis template



### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)


# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- TRUE
do.plot <- TRUE
quick.load <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())

# script uses Seurat v5
#install.packages('remotes')
#remotes::install_github(repo = 'satijalab/seurat')


#setRepositories(ind = 1:3, addURLs = c('https://satijalab.r-universe.dev', 'https://bnprks.r-universe.dev/'))
#install.packages(c("BPCells", "presto", "glmGamPoi"))

#remotes::install_github("mojaveazure/seurat-object", quiet = TRUE)
#install.packages("Matrix")

#remotes::install_github("satijalab/seurat-data", quiet = TRUE)
#remotes::install_github("satijalab/azimuth", "seurat5", quiet = TRUE)
#remotes::install_github("satijalab/seurat-wrappers", "seurat5", quiet = TRUE)


library("Seurat")
library("SeuratObject")
library("SeuratData")
library("SeuratWrappers")
library("Azimuth")
library("dplyr")
library("scCustomize")

# perhaps these packages are not needed
#library("BPCells")
#library("presto")


# if older version is needed
# remotes::install_version("Seurat", "4.4.0", repos = c("https://satijalab.r-universe.dev", getOption("repos")))

########################################
# Parallelization of seurat functions
########################################

parallelization <- TRUE


if(parallelization){
  library(future)
  # check the current active plan
  plan()
  
  # detect cores
  ncores <- parallel::detectCores()
  ncores <- ncores-5
  # set plan
  plan("multisession", workers = ncores)
  plan()
  
  # set max limit for global variables
  # note this will drastically increase RAM usage, be aware of machine limits
  # set value is in bites therefore 1000*1024^2 = 1GB Therefore set to 10GBs 
  options(future.globals.maxSize = 10000 * 1024^2, 
          future.rng.onMisuse="ignore") 
}



if(quick.load){
  
  Nils.seurat <- list(CD45Pos = LoadSeuratRds("saves/Nils_seurat_CD45Pos.rds"), CD45Neg = LoadSeuratRds("saves/Nils_seurat_CD45Neg.rds"))
  
}
```


## Pre-processing {.tabset}

### Ambient RNA correction
```{r ambient_RNA_correction}

if(long.compute){
  
  # Create output dir
  
  output.dir <- paste0("results/QC/Ambien_RNA_correction/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  
  # install packages if not already installed
  #install.packages("SoupX")
  #BiocManager::install("DropletUtils")
  
  # load package
  library("SoupX")
  
  
  
  # list all directories in data folder
  samples <- list.dirs(path = "data", recursive = FALSE)
  
  
  for(i in seq_along(samples)){
    
    ##################
    # define input path
    ##################
    input.path <- paste0(samples[i])
    
    ##################
    # extract sample info
    ##################
    
    sample.id <- gsub("data/", "", samples[i])
    sample.id <- gsub("_outs", "", sample.id)
    
    ##################
    # load data 
    ##################
    # filtered and raw matrix as well as cluster info from cellranger analysis
    sc.temp <- SoupX::load10X(input.path)
    
    ##################
    # Estimate rho
    ##################
    # note default params are optimised for fresh (not nuclear) single cell experiments 
    # if processing nuclear scRNAseq, consider adjusting prior Rho default estimates
    
    sc.temp <- autoEstCont(sc.temp)
    
    dev.copy(pdf, paste0(output.dir, sample.id, "_Rho_estimation.pdf"))
    dev.off()
    
    ##################
    # Clean the data
    ##################
    sc.out <- adjustCounts(sc.temp)
    
    ##################
    # save data to file
    ##################
    DropletUtils:::write10xCounts(paste0("data/", sample.id, "_strainedCounts"), sc.out)
    
    ##################
    # clean environment 
    ##################
    rm(list = c("sc.temp", "sc.out"))
    
  }
  
  
  
  # various tools to visualise and inspect ambient RNA removal and detection
  # plotMarkerDistribution(sc.temp)
  # 
  # # identify most zeroed genes
  # cntSoggy = rowSums(sc.temp$toc > 0)
  # cntStrained = rowSums(sc.out > 0)
  # mostZeroed = tail(sort((cntSoggy - cntStrained)/cntSoggy), n = 20)
  # mostZeroed
  # 
  # tail(sort(rowSums(sc.temp$toc > sc.out)/rowSums(sc.temp$toc > 0)), n = 20)
  # 
  # 
  # plotChangeMap(sc.temp, sc.out, "Il1b")
  # 
  
  
}



```


### create seurat object | merge and annotate
```{r create_seurat_object}

if(long.compute){
  
  # list all directories in data folder
  samples <- list.dirs(path = "data", recursive = FALSE)
  
  # keep only "strainedCounts" directories
  samples <- grep(".*_strainedCounts", samples, value = T)
  
  
  
  # create list to store data 
  seurat.list <- list()
  
  for(i in seq_along(samples)){
    
    # define input path
    input.path <- paste0(samples[i])
    
    # read in data (GEX and ADT/hashtag)
    sequence.data <- Seurat::Read10X(data.dir = paste0(input.path))
    
    
    # extract sequence run info 
    sample.id <- gsub("data/", "", samples[i])
    sample.id <- gsub("_strainedCounts", "", sample.id)
    
    # create seurat object with gene expression 
    seurat.temp <- CreateSeuratObject(counts = sequence.data,
                                      project = paste0(sample.id))
    
    # store data in a seurat list
    seurat.list[[paste0(sample.id)]] <- seurat.temp
    
    # clean environment 
    rm(list = c("seurat.temp", "sequence.data"))
    gc()
    
  }
  
  # for some reason creating these objects sometimes has a glitch where nfeatures etc of metadata is not generated 
  # if this happens, restart R session and rerun from scratch
  #seurat.list$E4@meta.data
  
  seurat.list$E4@meta.data
  
  
  
  ########################################
  # Merge and annotate object
  ########################################
  
  # change names of objects to more user friendly names 
  
  # E4 = vehicle CD45Neg
  # F4 = vehicle CD45Pos
  # G4 = PDS CD45Neg
  # H4 = PDS CD45Pos
  
  names(seurat.list)
  names(seurat.list) <- c("Vehicle_CD45Neg", "Vehicle_CD45Pos", "PDS_CD45Neg", "PDS_CD45Pos")
  names(seurat.list)
  
  # create a vector of sequencing samples for later use in loops 
  sample.id <- names(seurat.list)
  sample.id
  
  
  # merge object for easier QC visualisation etc
  Nils.seurat <- scCustomize::Merge_Seurat_List(seurat.list, add.cell.ids = names(seurat.list))
  
  Nils.seurat # 32,285 genes across 35,302 cells
  unique(Nils.seurat@meta.data$orig.ident)
  
  Nils.seurat@meta.data$SampleID <- sub("^(.*?_.*?)(_.*)$", "\\1", rownames(Nils.seurat@meta.data))
  Nils.seurat@meta.data$Treatment <- gsub("_.*", "", Nils.seurat@meta.data$SampleID)
  Nils.seurat@meta.data$Compartment <- gsub(".*_", "", Nils.seurat@meta.data$SampleID)
  
  unique(Nils.seurat@meta.data$SampleID)
  unique(Nils.seurat@meta.data$Treatment)
  unique(Nils.seurat@meta.data$Compartment)
  
  # CD45Neg and CD45Pos datasets are very different and need to be QC'd differently and integrated differently 
  # split object to process compartments seperately 
  Nils.seurat <- SplitObject(Nils.seurat, split.by = "Compartment")
  
  # fully merge data by also combining layers for all assays, for clarity, thus downstream will explicitly split layers when required
  # layers can be combined for visualisation and subsetting 
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = JoinLayers, 
                        assay = "RNA")
  Nils.seurat
  
  rm(seurat.list)
  
}


```



### Temp
# read bam files and inspect bam alignment to OVAL
```{r}


# 
#  setwd("/Users/ieo/Documents/Sequence_alignment/scRNAseq/Nils_10x_B16_ex_vivo_2024/SI-TT-E4_plus_OVAL_forward/outs")
#  
#  sequence.data <- Seurat::Read10X(data.dir = paste0("./raw_feature_bc_matrix"))
#  sequence.data
#  logic.var <- grepl("OVAL", rownames(sequence.data))
#  rownames(sequence.data)[logic.var]
#  x <- sequence.data[logic.var, ]
#  sum(x)
#  
# # 
#  
#  
#  seurat.temp <- CreateSeuratObject(counts = sequence.data,
#                                    min.cells = 0,
#                                    min.features = 0,
#                                    project = "test")
#      
# 
#grep("OVAL", rownames(seurat.temp), value = T)

#logic.var <- grepl("OVAL", rownames(seurat.temp))
#rownames(seurat.temp)[logic.var]

# seurat.temp@assays$RNA@layers$counts[logic.var, ]


#VlnPlot_scCustom(seurat.temp, "OVAL")




# library(Rsamtools)
# 
# # Path to BAM file
# bam_file <- "possorted_genome_bam.bam"
# 
# # Load BAM file and index
# bam <- Rsamtools::BamFile(bam_file)
# 
# seqinfo(bam)
# 
# 
# which <- GRanges(c(
#     "OVAL:1-1875"
# ))
# ## equivalent:
# ## GRanges(
# ##     seqnames = c("seq1", "seq2", "seq2"),
# ##     ranges = IRanges(
# ##         start = c(1000, 100, 1000),
# ##         end = c(2000, 1000, 2000)
# ##     )
# ## )
# what <- c("qname", "rname", "strand", "pos", "qwidth", "mapq", "qual", "cigar", "flag", "seq")
# param <- ScanBamParam(which=which, what=what)
# 
# aln <- scanBam(bam, param=param)
# aln
# count.vals <- countBam(bam, param=param)
# count.vals
# 
# unique(aln$`OVAL:1-1875`$flag)
# 
# table(aln$`OVAL:1-1875`$pos)
# 
# #quickBamFlagSummary(bam)
# 
# 
# 
# 
# 
# 
# #install.packages("rhdf5")
# library(rhdf5)
# 
# 
# # Specify the path to your HDF5 file
# hdf5_file <- "molecule_info.h5"
# 
# # Open the HDF5 file
# hdf5_data <- H5Fopen(hdf5_file)
# 
# # Read a dataset from the HDF5 file
# dataset_name <- "features"
# dataset <- h5read(hdf5_data, dataset_name)
# 
# dataset$name[dataset$name == "OVAL"]
# 
# # Close the HDF5 file when done
# H5Fclose(hdf5_data)
# 
# 
# 
# 
# 
# 
# 
# 
# 
# # Specify the path to your HDF5 file
# hdf5_file <- "raw_feature_bc_matrix.h5"
# 
# # Open the HDF5 file
# hdf5_data <- H5Fopen(hdf5_file)
# 
# # Read a dataset from the HDF5 file
# dataset_name <- "matrix"
# dataset <- h5read(hdf5_data, dataset_name)
# dataset
# dataset$data[dataset$features$name == "OVAL"]
# 
# 
# # Close the HDF5 file when done
# H5Fclose(hdf5_data)
# 
# 
# 
# 
# 
# 
# 
# 
# 
# 
# 
# #aln <- scanBam(bam)
# 
# length(aln)
# class(aln)
# 
# 
# 
# 
# 
# 
# 
# Rsamtools::getSeq(bam)
# 
# 
# # Extract alignments overlapping the gene
# gene_alignments <- subset(bam, which(  # Use 'subset' to filter alignments
#   seqnames(bam) == "chromosome_name" &  # Replace with chromosome name
#   (start(bam) >= gene_coordinates[1] & end(bam) <= gene_coordinates[2])  # Filter by gene coordinates
# ))


```






## QC and Filter {.tabset}

### QC & filter

```{r QC_and_filter}

if(long.compute){
  
  # Create output dir
  
  output.dir <- paste0("results/QC/filtering/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  for(i in seq_along(names(Nils.seurat))){
    
    # set idents to sampleID
    Idents(Nils.seurat[[i]]) <- Nils.seurat[[i]]$SampleID
    
    # set default assay 
    DefaultAssay(Nils.seurat[[i]]) <- "RNA"
    
  }
  
  
  # Add mitochondrial and ribosomal gene percentages
  
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = Add_Mito_Ribo, 
                        species = "Mouse") 
  
  # Add cell complexity/novelty QC metric
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = Add_Cell_Complexity) 
  
  
  
  
  # create cutoff dataframe for easy plotting and interation 
  
  cutoff.df <- as.data.frame(matrix(nrow = 2, ncol = 7))
  
  colnames(cutoff.df) <- c("feature_low", "feature_high", "umi_low", "umi_low", "mito_high", "ribo_low", "complexity_low")
  
  rownames(cutoff.df) <- c("CD45Neg", "CD45Pos")
  # Setting cutoff values 
  cutoff.df$feature_low <- c(200, 200)
  cutoff.df$feature_high <- c(5500, 5500)
  cutoff.df$umi_low <- c(1000, 1000)
  cutoff.df$umi_high <- c(45000, 45000)
  cutoff.df$mito_high <- c(10, 10)
  cutoff.df$ribo_low <- c(10, 6)
  cutoff.df$complexity_low <- c(0.75, 0.8)
  
  
  ########################
  # QC plots
  ########################
  if(do.plot){
    
    for(i in seq_along(names(Nils.seurat))){
      # nFeatures
      print(
        QC_Plots_Genes(seurat_object = Nils.seurat[[i]], 
                       low_cutoff = cutoff.df$feature_low[i], 
                       high_cutoff = cutoff.df$feature_high[i],
                       pt.size = 0.05) #how many genes per cell? See where to set the cutoffs later
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_genes.pdf"))
      dev.off()
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "nFeature_RNA",
                     low_cutoff = cutoff.df$feature_low[i], 
                     high_cutoff = cutoff.df$feature_high[i],
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_genes_histo.pdf"))
      dev.off()
      
      
      # UMI
      print(
        QC_Plots_UMIs(seurat_object = Nils.seurat[[i]], 
                      low_cutoff = cutoff.df$umi_low[i], 
                      high_cutoff = cutoff.df$umi_high[i], 
                      y_axis_log = T,
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_UMI.pdf"))
      dev.off()
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "nCount_RNA",
                     low_cutoff = cutoff.df$umi_low[i], 
                     high_cutoff = cutoff.df$umi_high[i],
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_UMI_histo.pdf"))
      dev.off()
      
      
      
      
      # Mito percentage
      print(
        QC_Plots_Mito(seurat_object = Nils.seurat[[i]], 
                      high_cutoff = cutoff.df$mito_high[i],
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_mito.pdf"))
      dev.off()
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "percent_mito",
                     high_cutoff = cutoff.df$mito_high[i],
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_mito_histo.pdf"))
      dev.off()
      
      
      
      
      # Ribo percentage
      print(
        QC_Plots_Feature(seurat_object = Nils.seurat[[i]],
                         feature = "percent_ribo",
                         low_cutoff = cutoff.df$ribo_low[i],
                         pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_Ribo.pdf"))
      dev.off()
      
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "percent_ribo",
                     low_cutoff = cutoff.df$ribo_low[i], 
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_Ribo_histo.pdf"))
      dev.off()
      
      
      
      
      # Complexity
      print(
        QC_Plots_Complexity(seurat_object = Nils.seurat[[i]], 
                            low_cutoff = cutoff.df$complexity_low[i],
                            pt.size = 0.05)
      ) 
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_complexity.pdf"))
      dev.off()
      
      
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "log10GenesPerUMI",
                     low_cutoff = cutoff.df$complexity_low[i], 
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_complexity_histo.pdf"))
      dev.off()
      
      
    }
    
  }
  
  
  # Note, there are a large amount of cells with low ribosomal gene content. 
  # these are more preventant in PDS treated CD45Pos sample 
  # it is unclear what this means, but as filtering on ribosomal gene content is not a recommended step in fabian theis best scRNAseq practices, i will omit this
  # therefore, cells regardless of ribosomal content will be kept 
  
  
  
  # Filter dataset
  
  for(i in seq_along(names(Nils.seurat))){
    
    Nils.seurat[[i]] <- subset(Nils.seurat[[i]], 
                               subset = nFeature_RNA > cutoff.df$feature_low[i] & 
                                 nFeature_RNA < cutoff.df$feature_high[i] & 
                                 percent_mito < cutoff.df$mito_high[i] & 
                                 #percent_ribo > cutoff.df$ribo_low[i] & # do not filter for ribosomal content
                                 log10GenesPerUMI > cutoff.df$complexity_low[i])
    
  }
  
  
  
  if(do.plot){
    
    for(i in seq_along(names(Nils.seurat))){
      # nFeatures
      print(
        QC_Plots_Genes(seurat_object = Nils.seurat[[i]], 
                       low_cutoff = cutoff.df$feature_low[i], 
                       high_cutoff = cutoff.df$feature_high[i],
                       pt.size = 0.05) #how many genes per cell? See where to set the cutoffs later
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_genes_filtered.pdf"))
      dev.off()
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "nFeature_RNA",
                     low_cutoff = cutoff.df$feature_low[i], 
                     high_cutoff = cutoff.df$feature_high[i],
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_genes_filtered_histo.pdf"))
      dev.off()
      
      
      
      # UMI
      print(
        QC_Plots_UMIs(seurat_object = Nils.seurat[[i]], 
                      low_cutoff = cutoff.df$umi_low[i], 
                      high_cutoff = cutoff.df$umi_high[i], 
                      y_axis_log = T,
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_UMI_filtered.pdf"))
      dev.off()
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "nCount_RNA",
                     low_cutoff = cutoff.df$umi_low[i], 
                     high_cutoff = cutoff.df$umi_high[i],
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_UMI_filtered_histo.pdf"))
      dev.off()
      
      # Mito percentage
      print(
        QC_Plots_Mito(seurat_object = Nils.seurat[[i]], 
                      high_cutoff = cutoff.df$mito_high[i],
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_mito_filtered.pdf"))
      dev.off()
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "percent_mito",
                     high_cutoff = cutoff.df$mito_high[i],
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_mito_filtered_histo.pdf"))
      dev.off()
      
      
      # Ribo percentage
      print(
        QC_Plots_Feature(seurat_object = Nils.seurat[[i]],
                         feature = "percent_ribo",
                         high_cutoff = cutoff.df$ribo_high[i],
                         pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_Ribo_filtered.pdf"))
      dev.off()
      
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "percent_ribo",
                     low_cutoff = cutoff.df$ribo_low[i], 
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_Ribo_filtered_histo.pdf"))
      dev.off()
      
      
      # Complexity
      print(
        QC_Plots_Complexity(seurat_object = Nils.seurat[[i]], 
                            high_cutoff = cutoff.df$complexity_high[i],
                            pt.size = 0.05)
      ) 
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_complexity_filtered.pdf"))
      dev.off()
      
      print(
        QC_Histogram(seurat_object =  Nils.seurat[[i]],
                     features = "log10GenesPerUMI",
                     low_cutoff = cutoff.df$complexity_low[i], 
                     split.by = "SampleID")
      )
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_complexity_filtered_histo.pdf"))
      dev.off()
      
    }
    
  }
  
  Nils.seurat$CD45Neg  # 32,285 features by 13,562 cells
  Nils.seurat$CD45Pos # 32,285 features by 18,032 cells | note that originally when filtering for ribosomal content only 13,511 cells were retained
  
  table(Nils.seurat$CD45Neg@meta.data$Treatment) # 7,698 PDS and 5,864 Vehicle
  
  table(Nils.seurat$CD45Pos@meta.data$Treatment) # 8,082 PDS and 9,950 Vehicle
  
  
  
}


```


## Normalisation and integration {.tabset}

## Normalise & integrate data
```{r Normalise_and_integrate_data}

if(long.compute){
  
  # Note, 
  # I compared CCA integration and rpca integration, looks cca is better fit for this dataset 
  # also see documentation on cca vs rpca integration, cca more aligns with this dataset 
  
  # perform sct and default/standard normalisation before cca integration to compare both normalisation methods 
  
  
  #################################################################################
  # quick prediction of cell annotations, useful to assess integration
  #################################################################################
  
  # run azimuth only for CD45 compartment 
  
  
  # load library
  library("Azimuth")
  
  # set default assay
  DefaultAssay(Nils.seurat$CD45Pos) <- "RNA"
  
  # Run azimuth
  Nils.seurat$CD45Pos <- RunAzimuth(Nils.seurat$CD45Pos, reference = "pbmcref")
  
  Nils.seurat$CD45Pos
  
  Nils.seurat$CD45Pos@meta.data$predicted.celltype.l1
  
  # visualise results 
  DimPlot(Nils.seurat$CD45Pos, reduction = "ref.umap", group.by = "predicted.celltype.l1")
  DimPlot(Nils.seurat$CD45Pos, reduction = "ref.umap", group.by = "predicted.celltype.l2")
  
  
  
  # split object up based on sequencing run / in seurat V5 these are splitting the layers
  Nils.seurat$CD45Neg[["RNA"]] <- split(Nils.seurat$CD45Neg[["RNA"]],  f = Nils.seurat$CD45Neg$SampleID)
  Nils.seurat$CD45Pos[["RNA"]] <- split(Nils.seurat$CD45Pos[["RNA"]],  f = Nils.seurat$CD45Pos$SampleID)
  
  
  
  use.sct <- TRUE
  
  
  if(use.sct){
    
    #####################################################
    # Normalise with scTransform and integrate data
    #####################################################
    # note error with more recent version of matrixStats resulting in "useNames = NA is defunct" error. downgrade to matrixStats version 1.1.0 as a workaround
    # https://github.com/satijalab/seurat/issues/7501#issuecomment-1854571904
    #remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts  
    
    DefaultAssay(Nils.seurat[[1]]) <- "RNA"
    DefaultAssay(Nils.seurat[[2]]) <- "RNA"
    
    # note RNA assay is split into run level layers, below script will apply normalisation etc layer-wise
    
    # Normalise and scale with SCTransform v2
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = SCTransform, 
                          method = "glmGamPoi", 
                          vars.to.regress = "percent_mito", 
                          vst.flavor = "v2",
                          verbose = TRUE)
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = RunPCA, 
                          reduction.name = "pca_sct")
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = RunUMAP, 
                          reduction = "pca_sct", 
                          dims = 1:30, 
                          reduction.name = "umap.unintegrated.sct")
    
    
    DefaultAssay(Nils.seurat[[1]]) <- "SCT"
    DefaultAssay(Nils.seurat[[2]]) <- "SCT"
    
    
    # integrate layers using cca integration
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = IntegrateLayers, 
                          orig.reduction = "pca_sct",
                          new.reduction = "integrated.cca.sct",
                          method = CCAIntegration,
                          normalization.method = "SCT",
                          verbose = F)
    
    
    
    # find clusters and dim reduction using integrated data
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = FindNeighbors, 
                          dims = 1:30,
                          reduction = "integrated.cca.sct")
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = FindClusters, 
                          resolution = 1.5, 
                          cluster.name = "clusters_cca_sct")
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = RunUMAP, 
                          reduction = "integrated.cca.sct",
                          dims = 1:30, 
                          reduction.name = "umap.cca.sct")
    
    
    
  }else{
    
    
    ############################
    # standard normalisation 
    ############################
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = NormalizeData)
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = FindVariableFeatures)
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = ScaleData)
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = RunPCA, 
                          reduction.name = "pca")
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = RunUMAP, 
                          reduction = "pca", 
                          dims = 1:30, 
                          reduction.name = "umap.unintegrated")
    
    
    
    
    # # integrate layers using cca integration
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = IntegrateLayers, 
                          method = CCAIntegration,
                          orig.reduction = "pca",
                          new.reduction = "integrated.cca",
                          verbose = FALSE)
    
    
    # # find clusters and dim reduction using integrated data
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = FindNeighbors, 
                          reduction = "integrated.cca",
                          dims = 1:30)
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = FindClusters, 
                          resolution = 1.5,
                          cluster.name = "clusters_cca")
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = RunUMAP, 
                          reduction = "integrated.cca",
                          dims = 1:30, 
                          reduction.name = "umap.cca")
    
    
    
    Nils.seurat$CD45Neg[["RNA"]] <- JoinLayers(Nils.seurat$CD45Neg[["RNA"]],  assay = "RNA")
    Nils.seurat$CD45Pos[["RNA"]] <- JoinLayers(Nils.seurat$CD45Pos[["RNA"]],  assay = "RNA")
    
  }
  
  
  
  
}


```

## evaluate integration 

```{r eval_integration}

if(long.compute){
  
  if(do.plot){
    
    # create output dir
    output.dir <- paste0("results/integration/")
    
    if(!dir.exists(output.dir)){
      dir.create(output.dir, 
                 recursive = T)
      
    }
    
    if(use.sct){
      
      unintegrated.reduction <- "umap.unintegrated.sct"
      integrated.reduction <- "umap.cca.sct"
      title.var <- "SCT"
    }else{
      
      unintegrated.reduction <- "umap.unintegrated"
      integrated.reduction <- "umap.cca"
      title.var <- "Standard"
      
    }
    
    
    # evaluate integration and dim reduction
    
    # set default assay
    DefaultAssay(Nils.seurat[[1]]) <- "SCT"
    DefaultAssay(Nils.seurat[[2]]) <- "SCT"
    
    Nils.seurat$CD45Pos@meta.data
    
    
    
    
    for(j in seq_along(names(Nils.seurat))){
      
      if(names(Nils.seurat)[j] == "CD45Pos"){
        vars.plot <- c("Treatment", "SampleID", "seurat_clusters", "predicted.celltype.l1", "predicted.celltype.l2")
        
      }else{
        
        vars.plot <- c("Treatment", "SampleID", "seurat_clusters")
        
      }
      
      
      for(i in seq_along(vars.plot)){
        
        
        # remove "." from name
        name.var <- gsub("\\.", "_", vars.plot[i])
        
        before.int <- DimPlot(Nils.seurat[[j]],
                              reduction = unintegrated.reduction, 
                              label = T,
                              group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("Unintegrated ", title.var, " Normalisation"))
        
        after.int <- DimPlot(Nils.seurat[[j]],
                             reduction = integrated.reduction, 
                             label = T,
                             group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("CCA Integration using ", title.var, " Normalisation"))
        
        print(before.int | after.int)
        dev.copy(pdf, paste0(output.dir, "UMAP_before_after_integration_", name.var, "_", names(Nils.seurat)[j], "_dataset.pdf"))
        dev.off()
        
        
        
      }
    }
    
    
  }
  
}
```


### Doublet detection and removal \\ currently not done as packages are broken or not updated for seuratV5 assay
```{r doublet_detection_and_removal}


# would like to use DoubletFinder as I think this is a more elegant package for doublet detection 
# however, seems DoubletFinder has not been updated to work with SeuratV5 and may have additional bugs preventing use by simply converting object to V3

# scDblFinder also has issues - error = "as_cholmod_sparse" not provided by package "Matrix" 
# seems to be an issue with Matrix and irlba packages. 
# tried work arounds suggested on github pages, which essentially remove these packages and reinstall 
# however error persists. 

# therefore for now, dont identify doublets  



```



### Save seurat object
```{r save_seurat_object}

if(long.compute){
  
  library("SeuratDisk")
  
  
  # unsure why but error in trying to save as h5Seurat object
  
  # export and save seurat object
  # SaveH5Seurat(Nils.seurat$CD45Pos,
  #              filename = "saves/Nils_seurat_CD45Pos.h5Seurat",
  #              overwrite = TRUE,
  #              verbose = TRUE)
  
  # SaveH5Seurat(Nils.seurat$CD45Neg,
  #              filename = "saves/Nils_seurat_CD45Neg.h5Seurat",
  #              overwrite = TRUE, 
  #              verbose = TRUE)
  # 
  
  
  SaveSeuratRds(Nils.seurat$CD45Pos, 
                file = "saves/Nils_seurat_CD45Pos.rds")
  
  SaveSeuratRds(Nils.seurat$CD45Neg, 
                file = "saves/Nils_seurat_CD45Neg.rds")
  
}

```


## Cluster identification and UMAP optimisation {.tabset}


### Cluster optimisation
```{r Cluster_optimisation}


if(long.compute){
  
  
  # create output dir
  output.dir <- paste0("results/Cluster_optimisation/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  Graphs(Nils.seurat$CD45Neg, slot = NULL)
  Neighbors(Nils.seurat$CD45Neg, slot = NULL)
  Reductions(Nils.seurat$CD45Neg, slot = NULL)
  
  # set assay 
  DefaultAssay(Nils.seurat[[1]]) <- "SCT"
  DefaultAssay(Nils.seurat[[2]]) <- "SCT"
  
  
  # Set seed for reproducibility
  set.seed(42)
  
  # Calculate neighbours graph using integrated data reduction 
  # this was previously calculated as part of normalisation and integration code chunk, but put here for clarity and to ensure neighbour graph is calculated on correct integrated dataset
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = FindNeighbors, 
                        dims = 1:30,
                        reduction = "integrated.cca.sct")
  
  
  # explicitly use the SCT_snn graph calculated for finding clusters 
  
  
  # Find clusters over an iteration of resolutions
  res.var <- 0
  res.max <- 2.5
  while (res.var <= res.max) {
    
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = FindClusters, 
                          resolution = res.var, 
                          graph.name = "SCT_snn")  
    res.var = res.var + 0.1
  }
  
  # Use clustree package to visualise impact of cluster resolution on clusters determined
  
  #install.packages("clustree")
  library(clustree) # clusttree needs to be loaded, error occurs if just call function from namespace
  
  head(Nils.seurat[[2]][[]])
  
  for(i in seq_along(names(Nils.seurat))){
    
    print(
      clustree::clustree(Nils.seurat[[i]], 
                         prefix = "SCT_snn_res.",
                         layout = "sugiyama")
    )
    dev.copy(pdf, paste0(output.dir, "dataset_", names(Nils.seurat)[i], "_clusttree.pdf"))
    dev.off()
    
    
  }
  
  
  # also visualise clusters using UMAP dims 
  for(i in seq_along(names(Nils.seurat))){
    
    res.var <- 0
    
    while (res.var <= res.max) {
      print(
        DimPlot(Nils.seurat[[i]],
                reduction = "umap.cca.sct", 
                label = T,
                group.by = paste0("SCT_snn_res.", res.var)) + 
          NoLegend() + 
          ggtitle(paste0("SCT_snn_res.", res.var))
      )
      
      dev.copy(pdf, paste0(output.dir, "dataset_", names(Nils.seurat)[i], "_umap_cluster_resolution_", res.var, ".pdf"))
      dev.off()
      
      res.var = res.var + 0.1
      
    }
    
  }
  
  # use annotation on CD45pos dataset as a way to roughly estimate clustering /overclustering 
  table(Nils.seurat$CD45Pos@meta.data$predicted.celltype.l1, Nils.seurat$CD45Pos@meta.data$SCT_snn_res.1.3)
  
  
  
  
  # note resolution of 1.3 looks relatively ok for both datasets 
  # will use this until further annotation of clusters and refinement is done
  # assuming this is overclustering and can reduce clustering later
  
  # set seurat clusters to this level 
  Nils.seurat[[1]]@meta.data$seurat_clusters <- Nils.seurat[[1]]@meta.data$SCT_snn_res.1.3
  Nils.seurat[[2]]@meta.data$seurat_clusters <- Nils.seurat[[2]]@meta.data$SCT_snn_res.1.3
  
  length(unique(Nils.seurat[[1]]@meta.data$seurat_clusters)) # 28 clusters
  length(unique(Nils.seurat[[2]]@meta.data$seurat_clusters)) # 23 clusters
  
  
  
  # Visualise clusters on UMAP projection 
  DimPlot(Nils.seurat$CD45Pos,
          reduction = "umap.cca.sct", 
          label = T,
          group.by = c("seurat_clusters")) + 
    NoLegend() + 
    ggtitle("SCT_snn_res.1.3")
  
  dev.copy(pdf, paste0(output.dir, "UMAP_CD45Pos_dataset_seurat_clusters.pdf"))
  dev.off()
  
  
  DimPlot(Nils.seurat$CD45Neg,
          reduction = "umap.cca.sct", 
          label = T,
          group.by = c("seurat_clusters")) + 
    NoLegend() + 
    ggtitle("SCT_snn_res.1.3")
  
  dev.copy(pdf, paste0(output.dir, "UMAP_CD45Neg_dataset_seurat_clusters.pdf"))
  dev.off()
  
}


```

### Optimise UMAP dim reduction params
```{r umap_optimisation}

if(long.compute){
  
  
  # create output dir
  output.dir <- paste0("results/UMAP_optimisation/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  
  
  # Optimise UMAP projection 
  
  # set init conditions and step sizes
  init.min.dist = 0.001
  init.neigh = 5
  min.dist.step = 2 
  neigh.step = 20
  
  
  
  for(i in seq_along(names(Nils.seurat))){
    
    if(names(Nils.seurat)[i] == "CD45Pos"){
      
      grouping.var <- "predicted.celltype.l2"
      
    }else{
      
      grouping.var <- "seurat_clusters"
      
    }
    min.dist.val <- init.min.dist
    
    # start loop 
    while (min.dist.val < 1) {
      
      n.neigh.val <- init.neigh
      
      while (n.neigh.val <= 50) {
        
        print(paste0("Calculating UMAP for min.dist = ", 
                     min.dist.val, " & n.neighbor = ", n.neigh.val))
        
        # caculate UMAP
        temp <- Seurat::RunUMAP(object = Nils.seurat[[i]],
                                reduction = "integrated.cca.sct", 
                                dims = 1:30,
                                n.neighbors = n.neigh.val, 
                                min.dist = min.dist.val,
                                reduction.name = "umap.temp",
                                seed.use = 42)
        
        # print UMAP
        p1 <- Seurat::DimPlot(object = temp, 
                              reduction = "umap.temp", 
                              label = TRUE, 
                              label.size = 4, 
                              group.by = grouping.var) + 
          ggtitle(paste0(grouping.var, " UMAP Min dist = ", 
                         min.dist.val, " n.val = ", n.neigh.val)) +
          NoLegend()
        
        
        
        print(p1)
        
        dev.copy(pdf, 
                 paste0(output.dir,  names(Nils.seurat)[i], 
                        "_UMAP_Min_dist_", min.dist.val, "_neighval_", n.neigh.val, ".pdf"))
        dev.off()
        
        n.neigh.val <- n.neigh.val + neigh.step
      }
      min.dist.val <- min.dist.val * min.dist.step
    }
    
    
  }
  
  
  
  # calculate UMAP using default params 
  
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = RunUMAP,
                        reduction = "integrated.cca.sct",
                        dims = 1:30)
  
}

```

### Save seurat object
```{r save_seurat_object_2}

if(long.compute){
  
  library("SeuratDisk")
  
  
  # unsure why but error in trying to save as h5Seurat object
  
  # export and save seurat object
  # SaveH5Seurat(Nils.seurat$CD45Pos,
  #              filename = "saves/Nils_seurat_CD45Pos.h5Seurat",
  #              overwrite = TRUE, 
  #              verbose = TRUE)
  # 
  # SaveH5Seurat(Nils.seurat$CD45Neg,
  #              filename = "saves/Nils_seurat_CD45Neg.h5Seurat",
  #              overwrite = TRUE, 
  #              verbose = TRUE)
  # 
  
  
  SaveSeuratRds(Nils.seurat$CD45Pos, 
                file = "saves/Nils_seurat_CD45Pos.rds")
  
  SaveSeuratRds(Nils.seurat$CD45Neg, 
                file = "saves/Nils_seurat_CD45Neg.rds")
  
}

```



## Overview of dataset {.tabset}

### Broad overview of dataset
```{r broad_overview}

if(long.compute){
  
  if(do.plot){
    
    
    # create output dir
    output.dir <- paste0("results/overview_visualisation/")
    
    if(!dir.exists(output.dir)){
      dir.create(output.dir, 
                 recursive = T)
      
    }
    
    
    for(i in seq_along(names(Nils.seurat))){
      
      
      
      # set assay
      DefaultAssay(Nils.seurat[[i]]) <- "SCT"
      
      if(names(Nils.seurat)[i] == "CD45Pos"){
        vars.to.plot <- c("Treatment", "SampleID", "seurat_clusters", "predicted.celltype.l1", "predicted.celltype.l2", "predicted.celltype.l3")
      }else{
        
        vars.to.plot <- c("Treatment", "SampleID", "seurat_clusters")
        
      }
      for(j in seq_along(vars.to.plot)){
        name.var <- gsub("\\.", "_", vars.to.plot[j])
        
        print(
          scCustomize::DimPlot_scCustom(Nils.seurat[[i]],
                                        reduction = "umap", 
                                        label = TRUE,
                                        figure_plot = TRUE,
                                        group.by = vars.to.plot[j]) 
        )
        
        
        dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_UMAP_", name.var, ".pdf"))
        dev.off()
        
      }
      
    }
    
    
  }
}



```



## DEG and cluster annotation {.tabset}

## Cluster annotation CD45Neg dataset
```{r CD45Neg_data_cluster_annotation}

if(long.compute){
  output.dir <- "results/CD45Neg_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # load required packages
  library("SingleR")
  library("celldex")
  
  ################################################################
  # Use a couple databased to get broad overview of clusters 
  ################################################################
  
  # error with collect() when trying to load celldex datasets
  
  # i think its an error with biocmanager and has been fixed with a very recent update. > 3.16 
  # but im running r 4.2.3 and this would require 4.3
  # not keen to upgrade right now 
  # so will try a work around of downgrading dbplyr to version 2.3.4 from 2.4 which im currently running 
  
  # need to downgrade to this version 
  #devtools::install_version("dbplyr", version = "2.3.4")
  #devtools::install_version("dbplyr", version = "2.4")
  
  # in theory if biocmanager is up to date, could just run below and should fix the problem
  
  #library(BiocManager) # 3.16
  #install("BiocFileCache")
  
  ##################################
  # Use MouseRNAseqData
  ##################################
  ?MouseRNAseqData()
  ref.mouse <-  celldex::MouseRNAseqData()
  
  # need to ensure matrixStats is at version 1.1.0 
  #remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts
  #BiocManager::install(version = "3.18")
  
  # Benayoun Lab that identified, downloaded and processed data sets on GEO that corresponded to sorted cell types (Benayoun et al., 2019).
  
  # The dataset contains 358 mouse RNA-seq samples annotated to 18 main cell types
  # further split into 28 subtypes under label.fine 
  
  
  # First calcualte cell ID using all available labels
  SingleR.pred <- SingleR(test = Nils.seurat$CD45Neg@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = Nils.seurat$CD45Neg@meta.data$seurat_clusters,
                          labels = ref.mouse$label.fine)
  
  # Plot 
  plotScoreHeatmap(SingleR.pred, 
                   show_colnames = TRUE)
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_MouseRNAseqData_db.pdf"))
  dev.off()
  
  
  SingleR::plotDeltaDistribution(SingleR.pred)
  
  SingleR::plotScoreDistribution(SingleR.pred)
  
  
  
  SingleR.pred <- SingleR(test = Nils.seurat$CD45Neg@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = Nils.seurat$CD45Neg@meta.data$seurat_clusters,
                          labels = ref.mouse$label.main)
  
  # Plot 
  plotScoreHeatmap(SingleR.pred, 
                   show_colnames = TRUE)
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_cluster_annotations_main_labels_MouseRNAseqData_db.pdf"))
  dev.off()
  
  
  ##################
  # annotations
  ##################
  
  # Cardiomyocytes = 21
  # Granulocytes = 10
  # Endothelial cells = 9, 11, 16, 17
  # Erythrocytes or endothelial or fibroblasts = 22
  # everything else looks fibroblast/adipocyte/NPC/microglial
  # cluster 18 has high T and NK and immune signatures maybe contamination
}

```


### DEG analysis CD45Neg dataset clusters
```{r DEG_CD45Neg_clusters}

if(long.compute){
  # create output dir
  output.dir <- paste0("results/CD45Neg_DEG_analysis/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  Nils.seurat$CD45Neg <- PrepSCTFindMarkers(Nils.seurat$CD45Neg)
  
  # set assay
  DefaultAssay(Nils.seurat$CD45Neg) <- "SCT"
  
  
  # set idents
  Idents(Nils.seurat$CD45Neg) <- "seurat_clusters"
  
  CD45Neg.de <- FindAllMarkers(Nils.seurat$CD45Neg,
                               assay = "SCT",
                               only.pos = TRUE, 
                               verbose = TRUE)
  
  
  # save to file 
  write.csv(CD45Neg.de, paste0(output.dir, "CD45Neg_DEGs.csv"))
  
  
  # aggregate expression for heatmap visualisation
  
  pseudo.cd45neg <- AggregateExpression(Nils.seurat$CD45Neg,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("seurat_clusters"))
  
  
  
  goi <- CD45Neg.de %>%
    filter(p_val_adj < 0.05) %>%
    group_by(cluster) %>%
    top_n(10, avg_log2FC) %>%
    pull(gene)
  
  
  DoHeatmap(pseudo.cd45neg, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE)
  
  dev.copy(pdf, paste0(output.dir, "Heatmaptop_10_DEGs_pseudobulk.pdf"))
  dev.off()
  
  
  
  
  # iterate over the top degs for each cluster 
  cluster.id <- unique(CD45Neg.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- CD45Neg.de %>%
      filter(p_val_adj < 0.05) %>%
      filter(cluster == cluster.id[i]) %>%
      top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.cd45neg, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE)
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
  
  # together with singleR prediction 
  # which suggested cluster 18 is immune cells (T/NK cells)
  # it does look like cluster 18 is immune cells, high for immune markers 
  table(Nils.seurat$CD45Neg@meta.data$seurat_clusters) # cluster 18 = 265 cells
  # cluster 10 is predicted to be granulocytes and is 619 cells
  
  # cluster 10 does indeed have many granulocyte type genes and is a strong unique signature 
  
  # consider removing cluster 10 and cluster 18
  
}

```


## Cluster annotation CD45Pos dataset
```{r CD45Pos_data_cluster_annotation}

if(long.compute){
  output.dir <- "results/CD45Pos_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # load required packages
  library("SingleR")
  library("celldex")
  
  ################################################################
  # Use a couple databased to get broad overview of clusters 
  ################################################################
  
  # error with collect() when trying to load celldex datasets
  
  # i think its an error with biocmanager and has been fixed with a very recent update. > 3.16 
  # but im running r 4.2.3 and this would require 4.3
  # not keen to upgrade right now 
  # so will try a work around of downgrading dbplyr to version 2.3.4 from 2.4 which im currently running 
  
  # need to downgrade to this version 
  #devtools::install_version("dbplyr", version = "2.3.4")
  #devtools::install_version("dbplyr", version = "2.4")
  
  # in theory if biocmanager is up to date, could just run below and should fix the problem
  
  #library(BiocManager) # 3.16
  #install("BiocFileCache")
  
  
  # need to ensure matrixStats is at version 1.1.0 
  #remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts
  #BiocManager::install(version = "3.18")
  
  #####################
  # Use ImmGenData
  #####################
  # Info: Mouse, Microarray, 830 samples, 20 main labels, 253 fine lables, Hematopoietic & Immune focus
  
  ImmGen.data <- celldex::ImmGenData()
  
  SingleR.pred <- SingleR(test = Nils.seurat$CD45Pos@assays$SCT$data, 
                          ref = ImmGen.data, 
                          method = "cluster",
                          clusters = Nils.seurat$CD45Pos@meta.data$seurat_clusters,
                          labels = ImmGen.data$label.fine)
  
  
  # Plot 
  plotScoreHeatmap(SingleR.pred, 
                   show_colnames = TRUE)
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_MouseRNAseqData_db.pdf"))
  dev.off()
  
  
  
  
  SingleR.pred <- SingleR(test = Nils.seurat$CD45Pos@assays$SCT$data, 
                          ref = ImmGen.data, 
                          method = "cluster",
                          clusters = Nils.seurat$CD45Pos@meta.data$seurat_clusters,
                          labels = ImmGen.data$label.main)
  
  
  # Plot 
  plotScoreHeatmap(SingleR.pred, 
                   show_colnames = TRUE)
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_cluster_annotations_main_labels_MouseRNAseqData_db.pdf"))
  dev.off()
  
}

```


### DEG analysis CD45Pos dataset clusters
```{r DEG_CD45Pos_clusters}

if(long.compute){
  # create output dir
  output.dir <- paste0("results/CD45Pos_DEG_analysis/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  
  Nils.seurat$CD45Pos <- PrepSCTFindMarkers(Nils.seurat$CD45Pos)
  
  # set assay
  DefaultAssay(Nils.seurat$CD45Pos) <- "SCT"
  
  
  # set idents
  Idents(Nils.seurat$CD45Pos) <- "seurat_clusters"
  
  CD45Pos.de <- FindAllMarkers(Nils.seurat$CD45Pos,
                               assay = "SCT",
                               only.pos = TRUE, 
                               verbose = TRUE)
  
  
  # save to file 
  write.csv(CD45Pos.de, paste0(output.dir, "CD45Pos_DEGs.csv"))
  
  
  # aggregate expression for heatmap visualisation
  
  pseudo.cd45pos <- AggregateExpression(Nils.seurat$CD45Pos,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("seurat_clusters"))
  
  
  
  goi <- CD45Pos.de %>%
    filter(p_val_adj < 0.05) %>%
    group_by(cluster) %>%
    top_n(10, avg_log2FC) %>%
    pull(gene)
  
  
  DoHeatmap(pseudo.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE)
  
  dev.copy(pdf, paste0(output.dir, "Heatmaptop_10_DEGs_pseudobulk.pdf"))
  dev.off()
  
  
  # iterate over the top degs for each cluster 
  cluster.id <- unique(CD45Pos.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- CD45Pos.de %>%
      filter(p_val_adj < 0.05) %>%
      filter(cluster == cluster.id[i]) %>%
      top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.cd45pos, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE)
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
  
  # compare with azimuth annotation 
  table(Nils.seurat$CD45Pos@meta.data$predicted.celltype.l1, Nils.seurat$CD45Pos@meta.data$seurat_clusters)
  
  table(Nils.seurat$CD45Pos@meta.data$predicted.celltype.l2, Nils.seurat$CD45Pos@meta.data$seurat_clusters)
  table(Nils.seurat$CD45Pos@meta.data$predicted.celltype.l3, Nils.seurat$CD45Pos@meta.data$seurat_clusters)
  
  # how big are clusters
  x <- table(Nils.seurat$CD45Pos@meta.data$seurat_clusters)
  barplot(x, las = 2)
  #write.csv(x, paste0())
  
  
}


```



## Clean datasets {.tabset}

### CD45Neg - remove contamination immune clusters
```{r CD45Neg_remove_immune_contamination}

if(long.compute){
  
  # create output dir
  output.dir <- paste0("results/CD45Neg_immune_contamination/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  # cluster 18 and cluster 10 look to be immune contaimination 
  
  head(Nils.seurat$CD45Neg@meta.data)
  
  
  
  # set assay
  DefaultAssay(Nils.seurat$CD45Neg) <- "SCT"
  
  
  # set idents
  Idents(Nils.seurat$CD45Neg) <- "seurat_clusters"
  
  # read deg file 
  input.dir <- "results/CD45Neg_DEG_analysis/"
  
  CD45Neg.de <- read.csv(paste0(input.dir, "CD45Neg_DEGs.csv"))
  
  
  # aggregate expression for heatmap visualisation
  
  pseudo.cd45neg <- AggregateExpression(Nils.seurat$CD45Neg,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("seurat_clusters"))
  
  
  
  # check cluster 18 as immune contamination 
  
  goi <- CD45Neg.de %>%
    filter(p_val_adj < 0.05) %>%
    filter(cluster == "18") %>%
    top_n(50, avg_log2FC) %>%
    pull(gene)
  
  goi
  
  
  DoHeatmap(pseudo.cd45neg, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE)
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_top_50_DEGs_pseudobulk_CD45Neg_cluster_18.pdf"))
  dev.off()
  
  
  # specifically immune markers
  goi <- c("Cd3e", "Cd3g", "Cd3d", "Cd8a", "Cd8b", "Cd4", "Ncam1", "Ncr1", "Cd19", "Cd20", "Il7r", "Cd14")
  
  DoHeatmap(pseudo.cd45neg, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE)
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_immune_genes_CD45Neg_cluster_18.pdf"))
  dev.off()
  
  
  
  scCustomize::Cluster_Highlight_Plot(Nils.seurat$CD45Neg,
                                      cluster_name = "18")
  
  dev.copy(pdf, paste0(output.dir, "UMAP_immune_contamination_cluster_18.pdf"))
  dev.off()
  
  
  # yes, cluster 18 is a small cluster that appears to be immune contamination 
  table(Nils.seurat$CD45Neg@meta.data$seurat_clusters) # cluster 18 = 265 cells
  
  # remove this cluster from dataset 
  Idents(Nils.seurat$CD45Neg) <- Nils.seurat$CD45Neg@meta.data$seurat_clusters
  Nils.seurat$CD45Neg <- subset(Nils.seurat$CD45Neg, idents = "18", invert = TRUE)
  unique(Idents(Nils.seurat$CD45Neg))
  
  
  
  # inspect cluster 10 - granulocytes
  
  
  scCustomize::Cluster_Highlight_Plot(Nils.seurat$CD45Neg,
                                      cluster_name = "10")
  
  dev.copy(pdf, paste0(output.dir, "UMAP_immune_contamination_cluster_10.pdf"))
  dev.off()
  
  # cluster 10 is a medium to large ish group 
  table(Nils.seurat$CD45Neg@meta.data$seurat_clusters) # cluster 10 = 619 cells
  
  
  
  
  
  # check cluster 10 top genes
  
  goi <- CD45Neg.de %>%
    filter(p_val_adj < 0.05) %>%
    filter(cluster == "10") %>%
    top_n(50, avg_log2FC) %>%
    pull(gene)
  
  goi
  
  # add cd14 to goi if not already there
  goi <- unique(c(goi, "Cd14"))
  DoHeatmap(pseudo.cd45neg, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE)
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_top_50_DEGs_pseudobulk_CD45Neg_cluster_10.pdf"))
  dev.off()
  
  
  
  
  # cluster 10 probs granulocyte contam - remove this cluster 
  
  # remove this cluster from dataset 
  Idents(Nils.seurat$CD45Neg) <- Nils.seurat$CD45Neg@meta.data$seurat_clusters
  Nils.seurat$CD45Neg <- subset(Nils.seurat$CD45Neg, idents = "10", invert = TRUE)
  unique(Idents(Nils.seurat$CD45Neg))
  
  
  
  
  # quick vis of endo/markers 
  
  # Endo = CLDN5
  # Epithelial = KRT17, KRT14
  
  
  goi <- grep("Krt.*", rownames(pseudo.cd45neg), value = T)
  goi <- grep("Cld.*", rownames(pseudo.cd45neg), value = T)
  goi <- c(goi, "Cd34", "Cd14")
  DoHeatmap(pseudo.cd45neg, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE)
  
}

```



## Renormalise / integrate / dim reduction | after removing contamination immune cells {.tabset}

## re-calculate normalisation and integration for CD45Neg dataset
```{r re_normalise_and_integrate_CD45Neg_dataset}

if(long.compute){
  
  
  # split object up based on sequencing run / in seurat V5 these are splitting the layers
  Nils.seurat$CD45Neg[["RNA"]] <- split(Nils.seurat$CD45Neg[["RNA"]],  f = Nils.seurat$CD45Neg$SampleID)
  
  DefaultAssay(Nils.seurat$CD45Neg) <- "RNA"
  
  
  
  
  #####################################################
  # Normalise with scTransform and integrate data
  #####################################################
  # note error with more recent version of matrixStats resulting in "useNames = NA is defunct" error. downgrade to matrixStats version 1.1.0 as a workaround
  # https://github.com/satijalab/seurat/issues/7501#issuecomment-1854571904
  #remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts  
  
  
  
  # note RNA assay is split into run level layers, below script will apply normalisation etc layer-wise
  
  Nils.seurat$CD45Neg <- Nils.seurat$CD45Neg %>%
    SCTransform(method = "glmGamPoi", 
                vars.to.regress = "percent_mito", 
                vst.flavor = "v2",
                verbose = TRUE) %>%
    RunPCA(reduction.name = "pca_sct") %>%
    RunUMAP(reduction = "pca_sct", 
            dims = 1:30, 
            reduction.name = "umap.unintegrated.sct")
  
  
  
  DefaultAssay(Nils.seurat$CD45Neg) <- "SCT"
  
  
  # integrate layers using cca integration and calculate dim redu and cluster identification
  
  # cluster resolution value based on previous estimation 
  clust.res <- 1.3
  
  
  Nils.seurat$CD45Neg <- Nils.seurat$CD45Neg %>%
    IntegrateLayers(orig.reduction = "pca_sct",
                    new.reduction = "integrated.cca.sct",
                    method = CCAIntegration,
                    normalization.method = "SCT",
                    verbose = F) %>%
    FindNeighbors(dims = 1:30,
                  reduction = "integrated.cca.sct") %>%
    FindClusters(resolution = clust.res, 
                 cluster.name = "clusters_cca_sct") %>%
    RunUMAP(reduction = "integrated.cca.sct",
            dims = 1:30, 
            reduction.name = "umap")
  
  
}



```

## plot integration after removing contamination immune cells from CD45Neg dataset 

```{r plot_integration_after_removing_contamination}

if(long.compute){
  
  if(do.plot){
    
    # create output dir
    output.dir <- paste0("results/CD45Neg_dataset_overview_after_removing_immune_contamination/")
    
    if(!dir.exists(output.dir)){
      dir.create(output.dir, 
                 recursive = T)
      
    }
    
    # set variables
    unintegrated.reduction <- "umap.unintegrated.sct"
    integrated.reduction <- "umap"
    title.var <- "SCT"
    
    
    
    # evaluate integration and dim reduction
    
    # set default assay
    DefaultAssay(Nils.seurat$CD45Neg) <- "SCT"
    
    
    
    
    vars.plot <- c("Treatment", "SampleID", "seurat_clusters")
    
    
    for(i in seq_along(vars.plot)){
      
      
      # remove "." from name
      name.var <- gsub("\\.", "_", vars.plot[i])
      
      before.int <- DimPlot(Nils.seurat$CD45Neg,
                            reduction = unintegrated.reduction, 
                            label = T,
                            group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("Unintegrated ", title.var, " Normalisation"))
      
      after.int <- DimPlot(Nils.seurat$CD45Neg,
                           reduction = integrated.reduction, 
                           label = T,
                           group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("CCA Integration using ", title.var, " Normalisation"))
      
      print(before.int | after.int)
      dev.copy(pdf, paste0(output.dir, "UMAP_before_after_integration_", name.var, "_", names(Nils.seurat)[j], "_dataset.pdf"))
      dev.off()
      
      
      
    }
  }
  
  
}



```


### Save final cleaned seurat objects
```{r save_seurat_object_final}

if(long.compute){
  
  library("SeuratDisk")
  
  SaveSeuratRds(Nils.seurat$CD45Pos, 
                file = "saves/Nils_seurat_CD45Pos.rds")
  
  SaveSeuratRds(Nils.seurat$CD45Neg, 
                file = "saves/Nils_seurat_CD45Neg.rds")
  
}

```







# move to cd45 pos and cd45neg specific scripts 


## Melanocyte gene expression {.tabset}

### investigate melanin specific genes
```{r melanocyte_genes}


# create output dir
output.dir <- paste0("results/CD45_Neg_Melanin_genes/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# Mitf, Tyr, Pmel, Mlana, Tyrp1, 
# maybe = Sox10, 

grep("Mit", rownames(Nils.seurat$CD45Neg), value = T)





goi <- c("Mitf", "Tyr", "Pmel", "Mlana", "Tyrp1", "Sox10")

for(i in seq_along(goi)){
  print(
    scCustomize::FeaturePlot_scCustom(Nils.seurat$CD45Neg, 
                                      reduction = "umap", 
                                      features = goi[i])
    
  )
  
  dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Neg_", goi[i], ".pdf"))
  dev.off()
  
}



scCustomize::FeaturePlot_scCustom(Nils.seurat$CD45Pos, 
                                  reduction = "umap", 
                                  features = "SOX10")



Nils.seurat$CD45Pos@meta.data$nCount_RNA

p1 <- DimPlot_scCustom(Nils.seurat$CD45Pos, reduction = "umap", group.by = "predicted.celltype.l2", label = T)
p2 <- scCustomize::Plot_Density_Custom(Nils.seurat$CD45Pos, reduction = "umap", features = "nCount_RNA")
p1 | p2


VlnPlot(Nils.seurat$CD45Neg, "Sox10")

DimPlot_scCustom(Nils.seurat$CD45Neg, reduction = "umap")


scCustomize::Plot_Density_Custom(Nils.seurat$CD45Neg, reduction = "umap", features = "Tyrp1")




```








## Which cell subsets are producing the chemokines {.tabset}

## WORK IN PROGRESS
### Vis chemokine/cytokine expression
```{r chemokine_cytokine_expression}

# create output dir
output.dir <- paste0("results/chemo_cyto_expression/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}




# aggregate expression for heatmap visualisation
pseudo.cd45pos <- AggregateExpression(Nils.seurat$CD45Pos,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("seurat_clusters"))


pseudo.cd45neg <- AggregateExpression(Nils.seurat$CD45Neg,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("seurat_clusters"))



# identify chemokines/cytokines of interest
goi <- grep("^IL.*", rownames(pseudo.cd45pos), value = T)
goi


goi <- grep("^IF.*", rownames(Nils.seurat$CD45Pos), value = T)
goi <- grep("^CX.*", rownames(Nils.seurat$CD45Pos), value = T)
goi <- grep("^GZ.*", rownames(Nils.seurat$CD45Pos), value = T)
goi



goi <- c("IL1B", "IL6", "IFNG", "TNF", "CCL2", "CCL3", "CCL4", "CCL5", "CXCL1", "CXCL9", "CXCL10", "CXCL13")

DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          raster = FALSE)


head(Nils.seurat$CD45Pos@meta.data)
scCustomize::DimPlot_scCustom(Nils.seurat$CD45Pos,
                              group.by = "predicted.celltype.l2", 
                              reduction = "umap",
                              label = T)

scCustomize::FeaturePlot_scCustom(Nils.seurat$CD45Pos,
                                  feature = "IL7R", 
                                  reduction = "umap")









goi <- grep("^Il1b.*", rownames(pseudo.cd45neg), value = T)
goi


goi <- c("Il1b", "Il6", "Ifng", "Tnf")
goi <- c("Ccl2", "Ccl3", "Ccl4", "Ccl5")
goi <- c("Cxcl1", "Cxcl9", "Cxcl10", "Cxcl13")
DoHeatmap(pseudo.cd45neg, 
          features = goi, 
          draw.lines = FALSE, 
          raster = FALSE)

#dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
#dev.off()


# key genes
# IL1b, IL6, IL12, IFNG, TNFA, 

# CCL2,3,4,5 
# CXCL1, CXCL9,10,13 




```











# NOT WORKING - check all code 
## Difference between treatment and vehicle {.tabset}
# Cant really do DA without replicate samples 

### Differential Abundance testing
```{r Differential_Abundance_testing}

#BiocManager::install("miloR")

library("miloR")
library("SingleCellExperiment")
library("scater")
library("scran")


#############################################
# Perform analysis for CD45Pos dataset
#############################################


# set assay
DefaultAssay(Nils.seurat$CD45Pos) <- "SCT"



# create output dir
output.dir <- paste0("results/DA_analysis_CD45Pos/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}





# Note: 
# milo object capatilises lots of variables, ie umap = UMAP
# below is method was used before i realised this, keeping for reference 
# some error in buildGraph() call prevents specifiying to use integrated.cca as reduced.dim, so it recalculates PCA, therefore rewrite PCA to be integrated.cca
#HL.seurat[["pca"]] <- CreateDimReducObject(embeddings = HL.seurat[["integrated.cca"]]@cell.embeddings, key = "PC_ ", assay = DefaultAssay(HL.seurat))
#names of dim reductions get capatilized for some reason, therefore reduced.dim = "INTEGRATED.CCA"


#######################
# Create milo object
#######################

# create single cell experiment object
CD45Pos.sce <- Seurat::as.SingleCellExperiment(Nils.seurat[["CD45Pos"]],
                                               assay = "SCT")

# convert object into a milo object
CD45Pos.sce <- Milo(CD45Pos.sce)
CD45Pos.sce



#######################
# Create KNN graph
#######################

# use nn graph generated as part of seurat pipeline
Graphs(CD45Pos.sce, slot = NULL)
Neighbors(CD45Pos.sce, slot = NULL)
Reductions(CD45Pos.sce, slot = NULL)


CD45Pos.sce@graph$SCT_nn <- Nils.seurat[["CD45Pos"]]@graphs$SCT_nn


# build from the adjacency matrix
milo.graph <- miloR::buildFromAdjacency(CD45Pos.sce@graph$SCT_nn)

# append graph to milo object
CD45Pos.sce@graph <- milo.graph@graph


# set some global-ish variables 
k.parm.use <- 20
d.use <- 50


# conversly, recaculate knn graph 
#HL.sce <- buildGraph(HL.sce,
#                 k = k.parm.use,
#                  d = 30, 
#            reduced.dim = "INTEGRATED.CCA")


########################
# make Neighbourhoods
########################

CD45Pos.sce <- makeNhoods(CD45Pos.sce,
                          prop = 0.1,
                          k = k.parm.use, # default for seurat findneighbours is k.parm = 20
                          d = d.use, # dim reduction integrated.cca is 50 dims
                          reduced_dims = "INTEGRATED.CCA.SCT",
                          refined = TRUE, 
                          refinement_scheme = "graph") # make nhoods using graph-only as this is faster

# Rule of thumb, aim for average neighbourhood size over 5*n_samples 
# therefore 5*10 = 50
#Empirically, we found it’s best to have a distribution peaking between 50 and 100


plotNhoodSizeHist(CD45Pos.sce, bins = 100)

dev.copy(pdf, paste0(output.dir, "Neighbourhood_size_histogram.pdf"))
dev.off()


########################
# count cells 
########################

CD45Pos.sce <- countCells(CD45Pos.sce, 
                          meta.data = data.frame(colData(CD45Pos.sce)),
                          samples = "Treatment") # make the nhood X sample counts matrix

head(nhoodCounts(CD45Pos.sce))

##############################
# calculate nhood distance
##############################

# longest compute time of this entire pipeline 
CD45Pos.sce <- calcNhoodDistance(CD45Pos.sce, 
                                 d = d.use, 
                                 reduced.dim = "INTEGRATED.CCA.SCT")


####################################
# Differential Abundance testing
####################################

# design model matrix for DA testing

DA.design <- data.frame(colData(CD45Pos.sce))[,c("SampleID", "Treatment")]
DA.design <- distinct(DA.design)
rownames(DA.design) <- DA.design$SampleID

## Reorder rownames to match columns of nhoodCounts(milo)
DA.design <- DA.design[colnames(nhoodCounts(CD45Pos.sce)), , drop=FALSE]
table(DA.design$SampleID)

table(DA.design$Treatment)

# create contrasts
#contrast.1 <- c("Age1wk - Age4wk") # the syntax is <VariableName><ConditionLevel> - <VariableName><ControlLevel>
contrast.status <- c("SampleIDPDS_CD45Pos - SampleIDVehicle_CD45Pos")


# this is the edgeR code called by `testNhoods`
# check it looks correct
model <- model.matrix(~ 0 + SampleID, data=DA.design)
mod.constrast <- makeContrasts(contrasts=contrast.status, levels=model)

mod.constrast


# perform DA test
da_results_status <- testNhoods(CD45Pos.sce, 
                                design = ~ SampleID,
                                design.df = DA.design, 
                                model.contrasts = contrast.status,
                                fdr.weighting= "graph-overlap", 
                                reduced.dim = "INTEGRATED.CCA.SCT")






# old single contrast code 

# 
matrix_design <- data.frame(colData(CD45Pos.sce))[,c("Treatment", "SampleID")]
matrix_design
# 
# ## factorise Treatment
matrix_design$Treatment <- as.factor(matrix_design$Treatment) 
matrix_design <- distinct(matrix_design)
rownames(matrix_design) <- matrix_design$Treatment

matrix_design
# 
# 
# # test nhoods for DA
da_results <- testNhoods(CD45Pos.sce,
                         design = ~ Treatment,
                         design.df = matrix_design, 
                         reduced.dim = "INTEGRATED.CCA.SCT")



#######################
# overview of results 
#######################


# how many nhoods are DA
da_results_status %>%
  filter(SpatialFDR < 0.05) %>%
  arrange(SpatialFDR) %>%
  head()

table(da_results_status$SpatialFDR < 0.05) # 513 DA nhoods
table(da_results_outcome$SpatialFDR < 0.05) # 3 DA nhoods

length(da_results_status$Nhood) # 2,833 nhoods
length(da_results_outcome$Nhood) # 2,833 nhoods

# distribution of pvalue

ggplot(da_results_status, aes(PValue)) + geom_histogram(bins=50)

dev.copy(pdf, paste0(output.dir, "DA_result_pvalue_histogram_status.pdf"))
dev.off()


ggplot(da_results_outcome, aes(PValue)) + geom_histogram(bins=50)

dev.copy(pdf, paste0(output.dir, "DA_result_pvalue_histogram_outcome.pdf"))
dev.off()


# volcano plots 


ggplot(da_results_status, aes(logFC, -log10(SpatialFDR))) + 
  geom_point() +
  geom_hline(yintercept = 1) ## Mark significance threshold (10% FDR

dev.copy(pdf, paste0(output.dir, "DA_result_volcano_status.pdf"))
dev.off()


ggplot(da_results_outcome, aes(logFC, -log10(SpatialFDR))) + 
  geom_point() +
  geom_hline(yintercept = 1) ## Mark significance threshold (10% FDR

dev.copy(pdf, paste0(output.dir, "DA_result_volcano_outcome.pdf"))
dev.off()




head(da_results)

######################################################
# visualise nhoods on umap projection
######################################################

# Build nhood graph
HL.sce <- buildNhoodGraph(HL.sce)



# Plot single-cell UMAP
umap_pl <- plotReducedDim(HL.sce, 
                          dimred = "UMAP",
                          colour_by="predicted.celltype.l1",
                          text_by = "predicted.celltype.l1", 
                          text_size = 3, 
                          point_size = 0.5) +
  guides(fill = "none")

## Plot neighbourhood graph
nh_graph_status <- plotNhoodGraphDA(HL.sce,
                                    da_results_status, 
                                    layout="UMAP",
                                    alpha=0.1) 


nh_graph_outcome <- plotNhoodGraphDA(HL.sce,
                                     da_results_outcome, 
                                     layout="UMAP",
                                     alpha=0.1) 

# print plots
umap_pl + nh_graph_status + 
  patchwork::plot_layout(guides = "collect")

dev.copy(pdf, paste0(output.dir, "UMAP_Nhood_DA_graph_status.pdf"))
dev.off()

umap_pl + nh_graph_outcome + 
  patchwork::plot_layout(guides = "collect")

dev.copy(pdf, paste0(output.dir, "UMAP_Nhood_DA_graph_outcome.pdf"))
dev.off()



# for context show umap with disease status highlighted
plotReducedDim(HL.sce, 
               dimred = "UMAP",
               colour_by="DiseaseStatus",
               text_by = "DiseaseStatus", 
               text_size = 3, 
               point_size = 0.5) +
  guides(fill = "none")

dev.copy(pdf, paste0(output.dir, "UMAP_DiseaseStatus.pdf"))
dev.off()


plotReducedDim(HL.sce, 
               dimred = "UMAP",
               colour_by="DiseaseOutcome",
               text_by = "DiseaseOutcome", 
               text_size = 3, 
               point_size = 0.5) +
  guides(fill = "none")

dev.copy(pdf, paste0(output.dir, "UMAP_DiseaseOutcome.pdf"))
dev.off()




# proceed with analysis for DiseaseStatus only, currently dataset seems insufficient for calling DiseaseOutcome 


table(HL.seurat@meta.data$predicted.celltype.l3)

#############################
# annotate neighbourhoods 
#############################

da_results_status_anno <- annotateNhoods(HL.sce, da_results_status, coldata_col = "predicted.celltype.l1")
da_results_status_anno <- annotateNhoods(HL.sce, da_results_status_anno, coldata_col = "predicted.celltype.l2")
da_results_status_anno <- annotateNhoods(HL.sce, da_results_status_anno, coldata_col = "predicted.celltype.l3")
da_results_status_anno <- annotateNhoods(HL.sce, da_results_status_anno, coldata_col = "seurat_clusters")

head(da_results_status_anno)


# Neighbourhoods tend to be homogeneous, we can define a threshold for celltype_fraction to exclude neighbourhoods that are a mix of cell types.

# visualise at each level of annotation 
anno.var <- c("predicted.celltype.l1", "predicted.celltype.l2", "predicted.celltype.l3", "seurat_clusters")


for(i in seq_along(anno.var)){
  
  plot.var <- paste0(anno.var[i], "_fraction")
  name.var <- gsub("\\.", "_", anno.var[i])
  print(ggplot(da_results_status_anno, aes(!!sym(plot.var))) + geom_histogram(bins=100))
  
  dev.copy(pdf, paste0(output.dir, "DA_results_DiseaseOutcome_histogram_nhood_", name.var, "fraction.pdf"))
  dev.off()
}



# use above histograms to define thresholds for annotating nhoods as mixed at each resolution of annotation 

threshold <- 0.9


da_results_status_anno$celltype.l1 <- ifelse(da_results_status_anno$predicted.celltype.l1_fraction < threshold, "Mixed", da_results_status_anno$predicted.celltype.l1)
da_results_status_anno$celltype.l2 <- ifelse(da_results_status_anno$predicted.celltype.l2_fraction < threshold, "Mixed", da_results_status_anno$predicted.celltype.l2)
da_results_status_anno$celltype.l3 <- ifelse(da_results_status_anno$predicted.celltype.l3_fraction < threshold, "Mixed", da_results_status_anno$predicted.celltype.l3)
da_results_status_anno$seurat_clusters <- ifelse(da_results_status_anno$seurat_clusters_fraction < threshold, "Mixed", da_results_status_anno$seurat_clusters)



length(da_results_status_anno$Nhood) # 2,833 nhoods
table(da_results_status_anno$celltype.l1 == "Mixed") # 700 nhoods became mixed
table(da_results_status_anno$celltype.l2 == "Mixed") # 1335 nhoods became mixed
table(da_results_status_anno$celltype.l3 == "Mixed") # 1891 nhoods became mixed
table(da_results_status_anno$seurat_clusters == "Mixed") # 1045 nhoods became mixed






# visualise at each level of annotation 
anno.var <- c("predicted.celltype.l1", "predicted.celltype.l2", "predicted.celltype.l3", "seurat_clusters")


for(i in seq_along(anno.var)){
  
  plot.var <- paste0(anno.var[i], "_fraction")
  name.var <- gsub("\\.", "_", anno.var[i])
  
  
  # Create a ggplot with histogram and line
  print(
    ggplot(da_results_status_anno, aes(x = !!sym(plot.var))) +
      geom_histogram(aes(fill = ifelse(!!sym(plot.var) > threshold, "Above", "Below")), bins = 100, color = "black") +
      geom_vline(xintercept = threshold, color = "green", linetype = "dashed") +
      scale_fill_manual(values = c("red", "blue"), name = "Category") +
      theme_minimal()  # Add additional themes as needed
  )
  
  dev.copy(pdf, paste0(output.dir, "DA_results_DiseaseOutcome_histogram_nhood_", name.var, "fraction_withthreshold.pdf"))
  dev.off()
}




# distribution of DA nhoods across annotation 
table(da_results_status_anno$SpatialFDR < 0.05, da_results_status_anno$celltype.l1)
table(da_results_status_anno$SpatialFDR < 0.05, da_results_status_anno$celltype.l2)
table(da_results_status_anno$SpatialFDR < 0.05, da_results_status_anno$celltype.l3)
table(da_results_status_anno$SpatialFDR < 0.05, da_results_status_anno$seurat_clusters)



# Now we can visualize the distribution of DA Fold Changes in different cell types

# iterate over annotation levels 

anno.var <- c("celltype.l1", "celltype.l2", "celltype.l3", "seurat_clusters")

# factorise for plotting
da_results_status_anno$seurat_clusters <- factor(da_results_status_anno$seurat_clusters, levels = c(levels(HL.seurat$seurat_clusters), "Mixed"))



for(i in seq_along(anno.var)){
  
  name.var <- gsub("\\.", "_", anno.var[i])
  
  print(
    plotDAbeeswarm(da_results_status_anno, 
                   group.by = anno.var[i])
  )
  dev.copy(pdf, paste0(output.dir, "Beeswarm_DA_DiseaseStatus_results_", name.var, ".pdf"))
  dev.off()
  
}

#####################################################################################
# Group nhoods into larger groups for later deg identification
#####################################################################################

# once neighbourhoods are identified that are sig between conditions
# we want to identify markers of these DA neighbourhoods

#The function findNhoodGroupMarkers runs a one-VS-all differential gene expression test to identify marker genes for a group of neighbourhoods of interest. 
# Before running this function you will need to define your neighbourhood groups depending on your biological question, that need to be stored as a NhoodGroup column in the da_results data.frame.

# In many cases, such as this example, DA neighbourhoods are found in different areas of the KNN graph, and grouping together all significant DA populations might not be ideal, as they might include cells of very different celltypes. 
# For this kind of scenario, we have implemented a neighbourhood function that uses community detection to partition neighbourhoods into groups on the basis of
# (1) the number of shared cells between 2 neighbourhoods; (2) the direction of fold-change for DA neighbourhoods; (3) the difference in fold change.

## Run buildNhoodGraph to store nhood adjacency matrix
HL.sce <- buildNhoodGraph(HL.sce)

# group DA neighbourhoods together using above metrics of shared cells, direction FC and difference FC. 


# test different metrics to get nice grouping of nhoods 
overlap.var <- 5
max.fc.delta.var <- 10

da_results_status_anno <- groupNhoods(HL.sce,
                                      da_results_status_anno, 
                                      da.fdr = 0.05,
                                      overlap = overlap.var,
                                      max.lfc.delta = max.fc.delta.var)


head(da_results_status_anno) # found 513 DA neighbourhoods at FDR 5%
length(unique(da_results_status_anno$NhoodGroup)) # number of groups




# plot beeswarm
plotDAbeeswarm(da_results_status_anno, group.by = "NhoodGroup") + 
  ggtitle(paste0("overlap = ", overlap.var, " max LFC delta = ", max.fc.delta.var))

dev.copy(pdf, paste0(output.dir, "Beeswarm_nhood_groups_DiseaseStatus.pdf"))
dev.off()

# unsure what is happening but looks like some nhood groups contain nhoods which are DA in different directions. this should not be occuring 


# plot on nhood groups on umap projection 
p1 <- plotNhoodGroups(HL.sce, da_results_status_anno, layout = "UMAP") + NoLegend()

nh_graph_status + p1 + patchwork::plot_layout(guides = "collect")

dev.copy(pdf, paste0(output.dir, "UMAP_nhood_groups_DiseaseStatus.pdf"))
dev.off()


# seurat clusters already offers a decent grouping of nhoods, perhaps use this or more refined clustering from seurat package to define nhood groups
plotDAbeeswarm(da_results_status_anno, "seurat_clusters")


# note, need at least 2 neighbourhoods assigned to each group for further processing 


##################################################
# finding gene signatures for neighbourhoods
##################################################

# restrict to only highly variable genes 

## Exclude zero counts genes
keep.rows <- rowSums(logcounts(HL.sce)) != 0
HL.sce <- HL.sce[keep.rows, ]

## Find HVGs
set.seed(101)
dec <- scran::modelGeneVar(HL.sce)
hvgs <- scran::getTopHVGs(dec, n = 2000)
head(hvgs)



# We run findNhoodGroupMarkers to test for one-vs-all differential gene expression for each neighbourhood group

# use seurat groups 
#da_results_status_anno$NhoodGroup <- da_results_status_anno$seurat_clusters

#da_results_status_anno <- da_results_status_anno %>%
# filter(NhoodGroup != "Mixed")

#da_results_status_anno$NhoodGroup <- droplevels(da_results_status_anno$NhoodGroup)

set.seed(42)

nhood_markers <- findNhoodGroupMarkers(HL.sce, 
                                       da_results_status_anno, 
                                       subset.row = hvgs, 
                                       aggregate.samples = TRUE, 
                                       sample_col = "SampleID")

head(nhood_markers)
table(da_results_status_anno$NhoodGroup, da_results_status_anno$celltype.l1)

# inspect degs of nhood group1
# which had many downregulated nhoods



## ANALYSIS WRITTEN UP TO HERE 









nhood_markers %>%
  filter(adj.P.Val_13 < 0.05) %>%
  arrange(desc(logFC_13)) %>%
  select(c(GeneID, logFC_13, adj.P.Val_13))

# visualise detexted markers 
# select markers for group 1 at FDR 1% and logFC > 1
table(nhood_markers$adj.P.Val_13 < 0.01) # 35 genes

ggplot(nhood_markers, aes(logFC_13, -log10(adj.P.Val_13))) + 
  geom_point(alpha=0.5, size=0.5) +
  geom_hline(yintercept = 3)


markers <- nhood_markers$GeneID[nhood_markers$adj.P.Val_13 < 0.01 & nhood_markers$logFC_13 > 0]


set.seed(42)
plotNhoodExpressionGroups(HL.sce, da_results_status_anno, features=intersect(rownames(HL.sce), markers[1:10]),
                          #subset.nhoods = da_results$NhoodGroup %in% c(1, 2, 3), 
                          scale = TRUE,
                          grid.space = "fixed")














#DGE testing within a group

# In some cases you might want to test for differential expression between cells in different conditions within the same neighbourhood group. You can do that using testDiffExp:

#dge_6 <- testDiffExp(embryo_milo, da_results, design = ~ stage, meta.data = data.frame(colData(embryo_milo)),
#                    subset.row = rownames(embryo_milo)[1:5], subset.nhoods=da_results$NhoodGroup=="6")

#dge_6


```


### Plot frequency stats
```{r freq_stats}

stat.vals <- scCustomize::Cluster_Stats_All_Samples(Nils.seurat$CD45Pos, group_by_var = "Treatment")

stat.vals



```



## DEG between treatment and vehicle {.tabset}
# write to iterate over clusters

### DEG analysis CD45Pos dataset treatment vs vehicle
```{r DEG_CD45Pos_treatment}

# create output dir
output.dir <- paste0("results/CD45Pos_DEG_analysis_treatment_vs_vehicle/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}



Nils.seurat$CD45Pos <- PrepSCTFindMarkers(Nils.seurat$CD45Pos)

# set assay
DefaultAssay(Nils.seurat$CD45Pos) <- "SCT"


# set idents
Idents(Nils.seurat$CD45Pos) <- "seurat_clusters"

Nils.seurat$CD45Pos@meta.data$Treatment_cluster <- paste0(Nils.seurat$CD45Pos@meta.data$Treatment, "_", Nils.seurat$CD45Pos@meta.data$seurat_clusters)
Idents(Nils.seurat$CD45Pos) <- "Treatment_cluster"
unique(Idents(Nils.seurat$CD45Pos))



Nils.seurat$CD45Pos@meta.data$Treatment_anno <- paste0(Nils.seurat$CD45Pos@meta.data$Treatment, "_", Nils.seurat$CD45Pos@meta.data$predicted.celltype.l2)
Idents(Nils.seurat$CD45Pos) <- "Treatment_anno"
unique(Idents(Nils.seurat$CD45Pos))




clust.id <- "CD14 Mono"
deg.list <- FindMarkers(Nils.seurat$CD45Pos,
                        ident.1 = paste0("PDS_", clust.id),
                        ident.2 = paste0("Vehicle_", clust.id),
                        assay = "SCT",
                        only.pos = TRUE, 
                        verbose = TRUE)



deg.list %>%
  filter(p_val_adj < 0.05)








# save to file 
#write.csv(CD45Pos.de, paste0(output.dir, "CD45Pos_DEGs.csv"))


# aggregate expression for heatmap visualisation

pseudo.cd45pos <- AggregateExpression(Nils.seurat$CD45Pos,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("Treatment_anno"))



goi <- deg.list %>%
  filter(p_val_adj < 0.05) %>%
  top_n(20, avg_log2FC) %>%
  rownames()
Idents(pseudo.cd45pos)
pseudo.cd45pos <- subset(pseudo.cd45pos, idents = c("PDS-CD14 Mono", "Vehicle-CD14 Mono"))
pseudo.cd45pos <- ScaleData(pseudo.cd45pos)

DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          raster = FALSE) + NoLegend()

#dev.copy(pdf, paste0(output.dir, "Heatmaptop_10_DEGs_pseudobulk.pdf"))
#dev.off()

VlnPlot_scCustom(Nils.seurat$CD45Pos,
                 features = c("ISG20"), 
                 group.by = "Treatment_anno", 
                 idents = c("PDS_CD14 Mono", "Vehicle_CD14 Mono"))




```













### CELL CYCLE
```{r remedy001}


# cell cycle scoring 
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes


HL.seurat <- CellCycleScoring(HL.seurat, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)

head(HL.seurat@meta.data)
scCustomize::FeaturePlot_scCustom(HL.seurat, feature = "S.Score", reduction = "umap.cca.sct")
scCustomize::FeaturePlot_scCustom(HL.seurat, feature = "G2M.Score", reduction = "umap.cca.sct")




```


### below not written for this dataset


### Playground 
```{r remedy001}


# interrogate specific lineage genes 

# aggregate expression for heatmap visualisation

pseudo.cd45pos <- AggregateExpression(Nils.seurat$CD45Pos,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("seurat_clusters"))



# cytokine expression 

goi <- grep("^IL.*", rownames(Nils.seurat$CD45Pos), value = T)
goi <- grep("^IF.*", rownames(Nils.seurat$CD45Pos), value = T)
goi <- grep("^CX.*", rownames(Nils.seurat$CD45Pos), value = T)
goi <- grep("^GZ.*", rownames(Nils.seurat$CD45Pos), value = T)
goi


DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          raster = FALSE)

goi <- grep("^Serpinb.*", rownames(Nils.seurat$CD45Neg), value = T)
goi





# SERPINB14









rownames(Nils.seurat$CD45Neg)
rownames(Nils.seurat$CD45Pos)
# why is cd45pos dataset have caps genes and cd45neg has mouse gene annotation? 





```






