---
title: "Nils_dataset_preprocessing"
author: "Dillon Corvino"
date: "25/01/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}


```

### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)


# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- FALSE
do.plot <- TRUE
quick.load <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())

# script uses Seurat v5
#install.packages('remotes')
#remotes::install_github(repo = 'satijalab/seurat')


#setRepositories(ind = 1:3, addURLs = c('https://satijalab.r-universe.dev', 'https://bnprks.r-universe.dev/'))
#install.packages(c("BPCells", "presto", "glmGamPoi"))

#remotes::install_github("mojaveazure/seurat-object", quiet = TRUE)
#install.packages("Matrix")

#remotes::install_github("satijalab/seurat-data", quiet = TRUE)
#remotes::install_github("satijalab/azimuth", "seurat5", quiet = TRUE)
#remotes::install_github("satijalab/seurat-wrappers", "seurat5", quiet = TRUE)


library("Seurat")
library("SeuratObject")
library("SeuratData")
library("Azimuth")
library("SeuratWrappers")
library("BPCells")
library("presto")
library("dplyr")
library("scCustomize")
library("Seurat")

# if older version is needed
# remotes::install_version("Seurat", "4.4.0", repos = c("https://satijalab.r-universe.dev", getOption("repos")))

########################################
# Parallelization of seurat functions
########################################

parallelization <- TRUE


if(parallelization){
  library(future)
  # check the current active plan
  plan()
  
  # detect cores
  ncores <- parallel::detectCores()
  ncores <- ncores-5
  # set plan
  plan("multisession", workers = ncores)
  plan()
  
  # set max limit for global variables
  # note this will drastically increase RAM usage, be aware of machine limits
  # set value is in bites therefore 1000*1024^2 = 1GB Therefore set to 10GBs 
  options(future.globals.maxSize = 10000 * 1024^2, 
          future.rng.onMisuse="ignore") 
}



if(quick.load){
  
  Nils.seurat <- list(CD45Pos = LoadSeuratRds("saves/Nils_seurat_CD45Pos.rds"), CD45Neg = LoadSeuratRds("saves/Nils_seurat_CD45Neg.rds"))
  
}
```

### Load dataset
```{r load_dataset}

if(long.compute){
  
  samples <- list.dirs(path = "data", recursive = FALSE)
  
  # create list to store data 
  seurat.list <- list()
  
  for(i in seq_along(samples)){
    
    # define input path
    input.path <- paste0(samples[i])
    
    # read in data (GEX and ADT/hashtag)
    sequence.data <- Seurat::Read10X(data.dir = paste0(input.path))
    
    
    # extract sequence run info 
    sample.id <- gsub("data/", "", samples[i])
    sample.id <- gsub("_filtmatrix", "", sample.id)
    
    # create seurat object with gene expression 
    seurat.temp <- CreateSeuratObject(counts = sequence.data,
                                      project = paste0(sample.id))
    
    # store data in a seurat list
    seurat.list[[paste0(sample.id)]] <- seurat.temp
    
    # clean environment 
    rm(list = c("seurat.temp", "sequence.data"))
    gc()
    
  }
  
  # for some reason creating these objects sometimes has a glitch where nfeatures etc of metadata is not generated 
  # if this happens, restart R session and rerun from scratch
  #seurat.list$E4@meta.data
  
  seurat.list$E4@meta.data
  
  
  
  ########################################
  # Merge and annotate object
  ########################################
  
  # change names of objects to more user friendly names 
  
  # E4 = vehicle CD45Neg
  # F4 = vehicle CD45Pos
  # G4 = PDS CD45Neg
  # H4 = PDS CD45Pos
  
  names(seurat.list)
  names(seurat.list) <- c("Vehicle_CD45Neg", "Vehicle_CD45Pos", "PDS_CD45Neg", "PDS_CD45Pos")
  names(seurat.list)
  
  # create a vector of sequencing samples for later use in loops 
  sample.id <- names(seurat.list)
  sample.id
  
  
  # merge object for easier QC visualisation etc
  Nils.seurat <- scCustomize::Merge_Seurat_List(seurat.list, add.cell.ids = names(seurat.list))
  
  Nils.seurat # 32,285 genes across 35,302 cells
  unique(Nils.seurat@meta.data$orig.ident)
  
  Nils.seurat@meta.data$SampleID <- sub("^(.*?_.*?)(_.*)$", "\\1", rownames(Nils.seurat@meta.data))
  Nils.seurat@meta.data$Treatment <- gsub("_.*", "", Nils.seurat@meta.data$SampleID)
  Nils.seurat@meta.data$Compartment <- gsub(".*_", "", Nils.seurat@meta.data$SampleID)
  
  unique(Nils.seurat@meta.data$SampleID)
  unique(Nils.seurat@meta.data$Treatment)
  unique(Nils.seurat@meta.data$Compartment)
  
  # CD45Neg and CD45Pos datasets are very different and need to be QC'd differently and integrated differently 
  # split object to process compartments seperately 
  Nils.seurat <- SplitObject(Nils.seurat, split.by = "Compartment")
  
  # fully merge data by also combining layers for all assays, for clarity, thus downstream will explicitly split layers when required
  # layers can be combined for visualisation and subsetting 
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = JoinLayers, 
                        assay = "RNA")
  Nils.seurat
  
  rm(seurat.list)
  
}

```


# consider treating cd45 neg and pos fractions seperately for qc filtering, they have distinct patterns of umi per cell etc 


## QC and Filter {.tabset}

### QC & filter

```{r QC_and_filter}

if(long.compute){
  
  # Create output dir
  
  output.dir <- paste0("results/QC/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  for(i in seq_along(names(Nils.seurat))){
    
    # set idents to sampleID
    Idents(Nils.seurat[[i]]) <- Nils.seurat[[i]]$SampleID
    
    # set default assay 
    DefaultAssay(Nils.seurat[[i]]) <- "RNA"
    
  }
  
  
  # Add mitochondrial and ribosomal gene percentages
  
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = Add_Mito_Ribo_Seurat, 
                        species = "Mouse") 
  
  # Add cell complexity/novelty QC metric
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = Add_Cell_Complexity_Seurat) 
  
  
  
  
  # create cutoff dataframe for easy plotting and interation 
  
  cutoff.df <- as.data.frame(matrix(nrow = 2, ncol = 7))
  
  colnames(cutoff.df) <- c("feature_low", "feature_high", "umi_low", "umi_low", "mito_high", "ribo_high", "complexity_high")
  
  rownames(cutoff.df) <- c("CD45Neg", "CD45Pos")
  # Setting cutoff values 
  cutoff.df$feature_low <- c(200, 200)
  cutoff.df$feature_high <- c(5500, 5500)
  cutoff.df$umi_low <- c(1000, 1000)
  cutoff.df$umi_high <- c(45000, 45000)
  cutoff.df$mito_high <- c(10, 10)
  cutoff.df$ribo_high <- c(10, 6)
  cutoff.df$complexity_high <- c(0.75, 0.8)
  
  
  # note ribo high is maybe better called ribo low, its the lower bound 
  # same with complexity high, should be complexity low, its the lower bound of complexity 
  
  
  ########################
  # QC plots
  ########################
  if(do.plot){
    
    for(i in seq_along(names(Nils.seurat))){
      # nFeatures
      print(
        QC_Plots_Genes(seurat_object = Nils.seurat[[i]], 
                       low_cutoff = cutoff.df$feature_low[i], 
                       high_cutoff = cutoff.df$feature_high[i],
                       pt.size = 0.05) #how many genes per cell? See where to set the cutoffs later
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_genes.pdf"))
      dev.off()
      
      
      # UMI
      print(
        QC_Plots_UMIs(seurat_object = Nils.seurat[[i]], 
                      low_cutoff = cutoff.df$umi_low[i], 
                      high_cutoff = cutoff.df$umi_high[i], 
                      y_axis_log = T,
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_UMI.pdf"))
      dev.off()
      
      
      # Mito percentage
      print(
        QC_Plots_Mito(seurat_object = Nils.seurat[[i]], 
                      high_cutoff = cutoff.df$mito_high[i],
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_mito.pdf"))
      dev.off()
      
      
      # Ribo percentage
      print(
        QC_Plots_Feature(seurat_object = Nils.seurat[[i]],
                         feature = "percent_ribo",
                         high_cutoff = cutoff.df$ribo_high[i],
                         pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_Ribo.pdf"))
      dev.off()
      
      
      # Complexity
      print(
        QC_Plots_Complexity(seurat_object = Nils.seurat[[i]], 
                            high_cutoff = cutoff.df$complexity_high[i],
                            pt.size = 0.05)
      ) 
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_complexity.pdf"))
      dev.off()
      
    }
    
  }
  
  
  # note there is lots of ribo low cells in cd45pos PDS dataset this is where most cells will be lost
  
  # Filter dataset
  
  for(i in seq_along(names(Nils.seurat))){
    
    Nils.seurat[[i]] <- subset(Nils.seurat[[i]], 
                               subset = nFeature_RNA > cutoff.df$feature_low[i] & 
                                 nFeature_RNA < cutoff.df$feature_high[i] & 
                                 percent_mito < cutoff.df$mito_high[i] & 
                                 percent_ribo > cutoff.df$ribo_high[i] & 
                                 log10GenesPerUMI > cutoff.df$complexity_high[i])
    
  }
  
  
  
  if(do.plot){
    
    for(i in seq_along(names(Nils.seurat))){
      # nFeatures
      print(
        QC_Plots_Genes(seurat_object = Nils.seurat[[i]], 
                       low_cutoff = cutoff.df$feature_low[i], 
                       high_cutoff = cutoff.df$feature_high[i],
                       pt.size = 0.05) #how many genes per cell? See where to set the cutoffs later
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_genes_filtered.pdf"))
      dev.off()
      
      
      # UMI
      print(
        QC_Plots_UMIs(seurat_object = Nils.seurat[[i]], 
                      low_cutoff = cutoff.df$umi_low[i], 
                      high_cutoff = cutoff.df$umi_high[i], 
                      y_axis_log = T,
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_UMI_filtered.pdf"))
      dev.off()
      
      
      # Mito percentage
      print(
        QC_Plots_Mito(seurat_object = Nils.seurat[[i]], 
                      high_cutoff = cutoff.df$mito_high[i],
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_mito_filtered.pdf"))
      dev.off()
      
      
      # Ribo percentage
      print(
        QC_Plots_Feature(seurat_object = Nils.seurat[[i]],
                         feature = "percent_ribo",
                         high_cutoff = cutoff.df$ribo_high[i],
                         pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_Ribo_filtered.pdf"))
      dev.off()
      
      
      # Complexity
      print(
        QC_Plots_Complexity(seurat_object = Nils.seurat[[i]], 
                            high_cutoff = cutoff.df$complexity_high[i],
                            pt.size = 0.05)
      ) 
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_complexity_filtered.pdf"))
      dev.off()
      
    }
    
  }
  
  Nils.seurat$CD45Neg  # 32,285 features by 12,407 cells
  Nils.seurat$CD45Pos # 32,285 features by 13,511 cells
  
  table(Nils.seurat$CD45Neg@meta.data$Treatment) # 7,100 PDS and 5,307 Vehicle
  
  table(Nils.seurat$CD45Pos@meta.data$Treatment) # 4,869 PDS and 8,642 Vehicle
  
  
  
}


```


## Normalisation and integration {.tabset}

## Normalise & integrate data
```{r Normalise_and_integrate_data}

if(long.compute){
  
  # Note, 
  # I compared CCA integration and rpca integration, looks cca is better fit for this dataset 
  # also see documentation on cca vs rpca integration, cca more aligns with this dataset 
  
  # perform sct and default/standard normalisation before cca integration to compare both normalisation methods 
  
  
  #################################################################################
  # quick prediction of cell annotations, useful to assess integration
  #################################################################################
  
  # run azimuth only for CD45 compartment 
  
  
  # load library
  library("Azimuth")
  
  # set default assay
  DefaultAssay(Nils.seurat$CD45Pos) <- "RNA"
  
  # Run azimuth
  Nils.seurat$CD45Pos <- RunAzimuth(Nils.seurat$CD45Pos, reference = "pbmcref")
  
  
  Nils.seurat$CD45Pos@meta.data$predicted.celltype.l1
  
  # visualise results 
  DimPlot(Nils.seurat$CD45Pos, reduction = "umap.cca.sct", group.by = "predicted.celltype.l1")
  DimPlot(Nils.seurat$CD45Pos, reduction = "umap.cca.sct", group.by = "predicted.celltype.l2")
  
  
  
  # split object up based on sequencing run / in seurat V5 these are splitting the layers
  Nils.seurat$CD45Neg[["RNA"]] <- split(Nils.seurat$CD45Neg[["RNA"]],  f = Nils.seurat$CD45Neg$SampleID)
  Nils.seurat$CD45Pos[["RNA"]] <- split(Nils.seurat$CD45Pos[["RNA"]],  f = Nils.seurat$CD45Pos$SampleID)
  
  
  
  use.sct <- TRUE
  
  
  if(use.sct){
    
    #####################################################
    # Normalise with scTransform and integrate data
    #####################################################
    # note error with more recent version of matrixStats resulting in "useNames = NA is defunct" error. downgrade to matrixStats version 1.1.0 as a workaround
    # https://github.com/satijalab/seurat/issues/7501#issuecomment-1854571904
    #remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts  
    
    DefaultAssay(Nils.seurat[[1]]) <- "RNA"
    DefaultAssay(Nils.seurat[[2]]) <- "RNA"
    
    # note RNA assay is split into run level layers, below script will apply normalisation etc layer-wise
    
    # Normalise and scale with SCTransform v2
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = SCTransform, 
                          method = "glmGamPoi", 
                          vars.to.regress = "percent_mito", 
                          vst.flavor = "v2",
                          verbose = TRUE)
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = RunPCA, 
                          reduction.name = "pca_sct")
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = RunUMAP, 
                          reduction = "pca_sct", 
                          dims = 1:30, 
                          reduction.name = "umap.unintegrated.sct")
    
    
    DefaultAssay(Nils.seurat[[1]]) <- "SCT"
    DefaultAssay(Nils.seurat[[2]]) <- "SCT"
    
    
    # integrate layers using cca integration
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = IntegrateLayers, 
                          orig.reduction = "pca_sct",
                          new.reduction = "integrated.cca.sct",
                          method = CCAIntegration,
                          normalization.method = "SCT",
                          verbose = F)
    
    
    
    # find clusters and dim reduction using integrated data
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = FindNeighbors, 
                          dims = 1:30,
                          reduction = "integrated.cca.sct")
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = FindClusters, 
                          resolution = 1.5, 
                          cluster.name = "clusters_cca_sct")
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = RunUMAP, 
                          reduction = "integrated.cca.sct",
                          dims = 1:30, 
                          reduction.name = "umap.cca.sct")
    
    
    
  }else{
    
    
    ############################
    # standard normalisation 
    ############################
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = NormalizeData)
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = FindVariableFeatures)
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = ScaleData)
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = RunPCA, 
                          reduction.name = "pca")
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = RunUMAP, 
                          reduction = "pca", 
                          dims = 1:30, 
                          reduction.name = "umap.unintegrated")
    
    
    
    
    # # integrate layers using cca integration
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = IntegrateLayers, 
                          method = CCAIntegration,
                          orig.reduction = "pca",
                          new.reduction = "integrated.cca",
                          verbose = FALSE)
    
    
    # # find clusters and dim reduction using integrated data
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = FindNeighbors, 
                          reduction = "integrated.cca",
                          dims = 1:30)
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = FindClusters, 
                          resolution = 1.5,
                          cluster.name = "clusters_cca")
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = RunUMAP, 
                          reduction = "integrated.cca",
                          dims = 1:30, 
                          reduction.name = "umap.cca")
    
    
    
    Nils.seurat$CD45Neg[["RNA"]] <- JoinLayers(Nils.seurat$CD45Neg[["RNA"]],  assay = "RNA")
    Nils.seurat$CD45Pos[["RNA"]] <- JoinLayers(Nils.seurat$CD45Pos[["RNA"]],  assay = "RNA")
    
  }
  
  
  
  
}


```




## evaluate integration 

```{r eval_integration}

if(do.plot){
  
  # create output dir
  output.dir <- paste0("results/integration/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  if(use.sct){
    
    unintegrated.reduction <- "umap.unintegrated.sct"
    integrated.reduction <- "umap.cca.sct"
    title.var <- "SCT"
  }else{
    
    unintegrated.reduction <- "umap.unintegrated"
    integrated.reduction <- "umap.cca"
    title.var <- "Standard"
    
  }
  
  
  # evaluate integration and dim reduction
  
  # set default assay
  DefaultAssay(Nils.seurat[[1]]) <- "SCT"
  DefaultAssay(Nils.seurat[[2]]) <- "SCT"
  
  Nils.seurat$CD45Pos@meta.data
  
  
  
  
  for(j in seq_along(names(Nils.seurat))){
    
    if(names(Nils.seurat)[j] == "CD45Pos"){
      vars.plot <- c("Treatment", "SampleID", "seurat_clusters", "predicted.celltype.l1", "predicted.celltype.l2")
      
    }else{
      
      vars.plot <- c("Treatment", "SampleID", "seurat_clusters")
      
    }
    
    
    for(i in seq_along(vars.plot)){
      
      
      # remove "." from name
      name.var <- gsub("\\.", "_", vars.plot[i])
      
      before.int <- DimPlot(Nils.seurat[[j]],
                            reduction = unintegrated.reduction, 
                            label = T,
                            group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("Unintegrated ", title.var, " Normalisation"))
      
      after.int <- DimPlot(Nils.seurat[[j]],
                           reduction = integrated.reduction, 
                           label = T,
                           group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("CCA Integration using ", title.var, " Normalisation"))
      
      print(before.int | after.int)
      dev.copy(pdf, paste0(output.dir, "UMAP_before_after_integration_", name.var, "_", names(Nils.seurat)[j], "_dataset.pdf"))
      dev.off()
      
      
      
    }
  }
  
  
}

```





### Save seurat object
```{r save_seurat_object}

if(long.compute){
  
  library("SeuratDisk")
  
  
  # unsure why but error in trying to save as h5Seurat object
  
  # export and save seurat object
  # SaveH5Seurat(Nils.seurat$CD45Pos,
  #              filename = "saves/Nils_seurat_CD45Pos.h5Seurat",
  #              overwrite = TRUE, 
  #              verbose = TRUE)
  # 
  # SaveH5Seurat(Nils.seurat$CD45Neg,
  #              filename = "saves/Nils_seurat_CD45Neg.h5Seurat",
  #              overwrite = TRUE, 
  #              verbose = TRUE)
  # 
  
  
  SaveSeuratRds(Nils.seurat$CD45Pos, 
                file = "saves/Nils_seurat_CD45Pos.rds")
  
  SaveSeuratRds(Nils.seurat$CD45Neg, 
                file = "saves/Nils_seurat_CD45Neg.rds")
  
}

```




## Cluster identification and UMAP optimisation {.tabset}


### Cluster optimisation
```{r Cluster_optimisation}


if(long.compute){
  
  
  # create output dir
  output.dir <- paste0("results/Cluster_optimisation/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  Graphs(Nils.seurat$CD45Neg, slot = NULL)
  Neighbors(Nils.seurat$CD45Neg, slot = NULL)
  Reductions(Nils.seurat$CD45Neg, slot = NULL)
  
  # set assay 
  DefaultAssay(Nils.seurat[[1]]) <- "SCT"
  DefaultAssay(Nils.seurat[[2]]) <- "SCT"
  
  
  # Set seed for reproducibility
  set.seed(42)
  
  # Calculate neighbours graph using integrated data reduction 
  # this was previously calculated as part of normalisation and integration code chunk, but put here for clarity and to ensure neighbour graph is calculated on correct integrated dataset
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = FindNeighbors, 
                        dims = 1:30,
                        reduction = "integrated.cca.sct")
  
  
  # explicitly use the SCT_snn graph calculated for finding clusters 
  
  
  # Find clusters over an iteration of resolutions
  res.var <- 0
  
  while (res.var <= 1.5) {
    
    
    Nils.seurat <- lapply(Nils.seurat, 
                          FUN = FindClusters, 
                          resolution = res.var, 
                          graph.name = "SCT_snn")  
    res.var = res.var + 0.1
  }
  
  # Use clustree package to visualise impact of cluster resolution on clusters determined
  
  #install.packages("clustree")
  library(clustree) # clusttree needs to be loaded, error occurs if just call function from namespace
  
  head(Nils.seurat[[2]][[]])
  
  for(i in seq_along(names(Nils.seurat))){
    
    print(
      clustree::clustree(Nils.seurat[[i]], 
                         prefix = "SCT_snn_res.",
                         layout = "sugiyama")
    )
    dev.copy(pdf, paste0(output.dir, "dataset_", names(Nils.seurat)[i], "_clusttree.pdf"))
    dev.off()
    
    
  }
  
  
  # note resolution of 1.3 looks relatively ok for both datasets 
  # will use this until further annotation of clusters and refinement is done
  
  # set seurat clusters to this level 
  Nils.seurat[[1]]@meta.data$seurat_clusters <- Nils.seurat[[1]]@meta.data$SCT_snn_res.1.3
  Nils.seurat[[2]]@meta.data$seurat_clusters <- Nils.seurat[[2]]@meta.data$SCT_snn_res.1.3
  
  length(unique(Nils.seurat[[1]]@meta.data$seurat_clusters)) # 27 clusters
  length(unique(Nils.seurat[[2]]@meta.data$seurat_clusters)) # 31 clusters
  
  
  
  # Visualise clusters on UMAP projection 
  DimPlot(Nils.seurat$CD45Pos,
          reduction = "umap.cca.sct", 
          label = T,
          group.by = c("seurat_clusters")) + 
    NoLegend() + 
    ggtitle("SCT_snn_res.1.5")
  
  dev.copy(pdf, paste0(output.dir, "UMAP_CD45Pos_dataset_seurat_clusters.pdf"))
  dev.off()
  
  
  DimPlot(Nils.seurat$CD45Neg,
          reduction = "umap.cca.sct", 
          label = T,
          group.by = c("seurat_clusters")) + 
    NoLegend() + 
    ggtitle("SCT_snn_res.1.5")
  
  dev.copy(pdf, paste0(output.dir, "UMAP_CD45Neg_dataset_seurat_clusters.pdf"))
  dev.off()
  
}




```

```{r umap_optimisation}


if(long.compute){
  
  
  # create output dir
  output.dir <- paste0("results/UMAP_optimisation/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  
  
  # Optimise UMAP projection 
  
  # set init conditions and step sizes
  init.min.dist = 0.001
  init.neigh = 5
  min.dist.step = 2 
  neigh.step = 20
  
  
  
  for(i in seq_along(names(Nils.seurat))){
    
    if(names(Nils.seurat)[i] == "CD45Pos"){
      
      grouping.var <- "predicted.celltype.l2"
      
    }else{
      
      grouping.var <- "seurat_clusters"
      
    }
    min.dist.val <- init.min.dist
    
    # start loop 
    while (min.dist.val < 1) {
      
      n.neigh.val <- init.neigh
      
      while (n.neigh.val <= 50) {
        
        print(paste0("Calculating UMAP for min.dist = ", 
                     min.dist.val, " & n.neighbor = ", n.neigh.val))
        
        # caculate UMAP
        temp <- Seurat::RunUMAP(object = Nils.seurat[[i]],
                                reduction = "integrated.cca.sct", 
                                dims = 1:30,
                                n.neighbors = n.neigh.val, 
                                min.dist = min.dist.val,
                                reduction.name = "umap.temp",
                                seed.use = 42)
        
        # print UMAP
        p1 <- Seurat::DimPlot(object = temp, 
                              reduction = "umap.temp", 
                              label = TRUE, 
                              label.size = 4, 
                              group.by = grouping.var) + 
          ggtitle(paste0(grouping.var, " UMAP Min dist = ", 
                         min.dist.val, " n.val = ", n.neigh.val)) +
          NoLegend()
        
        
        
        print(p1)
        
        dev.copy(pdf, 
                 paste0(output.dir,  names(Nils.seurat)[i], 
                        "_UMAP_Min_dist_", min.dist.val, "_neighval_", n.neigh.val, ".pdf"))
        dev.off()
        
        n.neigh.val <- n.neigh.val + neigh.step
      }
      min.dist.val <- min.dist.val * min.dist.step
    }
    
    
  }
  
  
  
  # calculate UMAP using default params 
  
  Nils.seurat <- lapply(Nils.seurat, 
                        FUN = RunUMAP,
                        reduction = "integrated.cca.sct",
                        dims = 1:30)
  
}


```

### Save seurat object
```{r save_seurat_object_2}

if(long.compute){
  
  library("SeuratDisk")
  
  
  # unsure why but error in trying to save as h5Seurat object
  
  # export and save seurat object
  # SaveH5Seurat(Nils.seurat$CD45Pos,
  #              filename = "saves/Nils_seurat_CD45Pos.h5Seurat",
  #              overwrite = TRUE, 
  #              verbose = TRUE)
  # 
  # SaveH5Seurat(Nils.seurat$CD45Neg,
  #              filename = "saves/Nils_seurat_CD45Neg.h5Seurat",
  #              overwrite = TRUE, 
  #              verbose = TRUE)
  # 
  
  
  SaveSeuratRds(Nils.seurat$CD45Pos, 
                file = "saves/Nils_seurat_CD45Pos.rds")
  
  SaveSeuratRds(Nils.seurat$CD45Neg, 
                file = "saves/Nils_seurat_CD45Neg.rds")
  
}

```


### Broad overview of dataset
```{r broad_overview}

if(do.plot){
  
  
  # create output dir
  output.dir <- paste0("results/overview_visualisation/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  for(i in seq_along(names(Nils.seurat))){
    
    
    
    # set assay
    DefaultAssay(Nils.seurat[[i]]) <- "SCT"
    
    if(names(Nils.seurat)[i] == "CD45Pos"){
      vars.to.plot <- c("Treatment", "SampleID", "seurat_clusters", "predicted.celltype.l1", "predicted.celltype.l2", "predicted.celltype.l3")
    }else{
      
      vars.to.plot <- c("Treatment", "SampleID", "seurat_clusters")
      
    }
    for(j in seq_along(vars.to.plot)){
      name.var <- gsub("\\.", "_", vars.to.plot[j])
      
      print(
        scCustomize::DimPlot_scCustom(Nils.seurat[[i]],
                                      reduction = "umap", 
                                      label = TRUE,
                                      figure_plot = TRUE,
                                      group.by = vars.to.plot[j]) 
      )
      
      
      dev.copy(pdf, paste0(output.dir, names(Nils.seurat)[i], "_UMAP_", name.var, ".pdf"))
      dev.off()
      
    }
    
  }
  
  
  
  
  
  
  
}


```



## CD45Neg dataset analysis {.tabset}


## Cluster annotation CD45Neg dataset
```{r CD45Neg_data_cluster_annotation}


output.dir <- "results/CD45Neg_SingleR/"

# Create output directory
if(!dir.exists(paste0(output.dir))){
  dir.create(paste0(output.dir), 
             recursive = T)
}


# load required packages
library("SingleR")
library("celldex")

################################################################
# Use a couple databased to get broad overview of clusters 
################################################################

# error with collect() when trying to load celldex datasets

# i think its an error with biocmanager and has been fixed with a very recent update. > 3.16 
# but im running r 4.2.3 and this would require 4.3
# not keen to upgrade right now 
# so will try a work around of downgrading dbplyr to version 2.3.4 from 2.4 which im currently running 

# need to downgrade to this version 
#devtools::install_version("dbplyr", version = "2.3.4")
#devtools::install_version("dbplyr", version = "2.4")

# in theory if biocmanager is up to date, could just run below and should fix the problem

#library(BiocManager) # 3.16
#install("BiocFileCache")

##################################
# Use MouseRNAseqData
##################################
?MouseRNAseqData()
ref.mouse <-  celldex::MouseRNAseqData()

# need to ensure matrixStats is at version 1.1.0 
#remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts
#BiocManager::install(version = "3.18")

# Benayoun Lab that identified, downloaded and processed data sets on GEO that corresponded to sorted cell types (Benayoun et al., 2019).

# The dataset contains 358 mouse RNA-seq samples annotated to 18 main cell types
# further split into 28 subtypes under label.fine 


# First calcualte cell ID using all available labels
SingleR.pred <- SingleR(test = Nils.seurat$CD45Neg@assays$SCT$data, 
                        ref = ref.mouse, 
                        method = "cluster",
                        clusters = Nils.seurat$CD45Neg@meta.data$seurat_clusters,
                        labels = ref.mouse$label.fine)

# Plot 
plotScoreHeatmap(SingleR.pred, 
                 show_colnames = TRUE)

dev.copy(pdf, paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_MouseRNAseqData_db.pdf"))
dev.off()


SingleR::plotDeltaDistribution(SingleR.pred)

SingleR::plotScoreDistribution(SingleR.pred)



SingleR.pred <- SingleR(test = Nils.seurat$CD45Neg@assays$SCT$data, 
                        ref = ref.mouse, 
                        method = "cluster",
                        clusters = Nils.seurat$CD45Neg@meta.data$seurat_clusters,
                        labels = ref.mouse$label.main)

# Plot 
plotScoreHeatmap(SingleR.pred, 
                 show_colnames = TRUE)

dev.copy(pdf, paste0(output.dir, "Heatmap_cluster_annotations_main_labels_MouseRNAseqData_db.pdf"))
dev.off()



```


### DEG analysis CD45Neg dataset clusters
```{r DEG_CD45Neg_clusters}

# create output dir
output.dir <- paste0("results/CD45Neg_DEG_analysis/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}



Nils.seurat$CD45Neg <- PrepSCTFindMarkers(Nils.seurat$CD45Neg)

# set assay
DefaultAssay(Nils.seurat$CD45Neg) <- "SCT"


# set idents
Idents(Nils.seurat$CD45Neg) <- "seurat_clusters"

CD45Neg.de <- FindAllMarkers(Nils.seurat$CD45Neg,
                             assay = "SCT",
                             only.pos = TRUE, 
                             verbose = TRUE)


# save to file 
write.csv(CD45Neg.de, paste0(output.dir, "CD45Neg_DEGs.csv"))


# aggregate expression for heatmap visualisation

pseudo.cd45neg <- AggregateExpression(Nils.seurat$CD45Neg,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("seurat_clusters"))



goi <- CD45Neg.de %>%
  filter(p_val_adj < 0.05) %>%
  group_by(cluster) %>%
  top_n(10, avg_log2FC) %>%
  pull(gene)


DoHeatmap(pseudo.cd45neg, 
          features = goi, 
          draw.lines = FALSE, 
          raster = FALSE)

dev.copy(pdf, paste0(output.dir, "Heatmaptop_10_DEGs_pseudobulk.pdf"))
dev.off()

# note cluster 22 was predicted to be cardiomyocytes 
# clusters 9, 11, 16, 30 predicted to be endothelial cells 
# 14, 24, 29 adipocytes 
# cluster 20 granulocytes 
# cluster 27 DC/T/NK cells
# rest fibroblasts



# iterate over the top degs for each cluster 
cluster.id <- unique(CD45Neg.de$cluster)
for(i in seq_along(cluster.id)){
  
  
  print(cluster.id[i])
  
  goi <- CD45Neg.de %>%
    filter(p_val_adj < 0.05) %>%
    filter(cluster == cluster.id[i]) %>%
    top_n(50, avg_log2FC)
  
  goi <- unique(goi)
  
  print(
    DoHeatmap(pseudo.cd45neg, 
              features = goi$gene, 
              draw.lines = FALSE, 
              raster = FALSE)
  )
  
  dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
  dev.off()
  
}



```



## CD45Pos dataset analysis {.tabset}


## Cluster annotation CD45NPos dataset
```{r CD45Pos_data_cluster_annotation}


output.dir <- "results/CD45Pos_SingleR/"

# Create output directory
if(!dir.exists(paste0(output.dir))){
  dir.create(paste0(output.dir), 
             recursive = T)
}


# load required packages
library("SingleR")
library("celldex")

################################################################
# Use a couple databased to get broad overview of clusters 
################################################################

# error with collect() when trying to load celldex datasets

# i think its an error with biocmanager and has been fixed with a very recent update. > 3.16 
# but im running r 4.2.3 and this would require 4.3
# not keen to upgrade right now 
# so will try a work around of downgrading dbplyr to version 2.3.4 from 2.4 which im currently running 

# need to downgrade to this version 
#devtools::install_version("dbplyr", version = "2.3.4")
#devtools::install_version("dbplyr", version = "2.4")

# in theory if biocmanager is up to date, could just run below and should fix the problem

#library(BiocManager) # 3.16
#install("BiocFileCache")


# need to ensure matrixStats is at version 1.1.0 
#remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts
#BiocManager::install(version = "3.18")

#####################
# Use ImmGenData
#####################
# Info: Mouse, Microarray, 830 samples, 20 main labels, 253 fine lables, Hematopoietic & Immune focus

ImmGen.data <- celldex::ImmGenData()

SingleR.pred <- SingleR(test = Nils.seurat$CD45Pos@assays$SCT$data, 
                        ref = ImmGen.data, 
                        method = "cluster",
                        clusters = Nils.seurat$CD45Pos@meta.data$seurat_clusters,
                        labels = ImmGen.data$label.fine)


# Plot 
plotScoreHeatmap(SingleR.pred, 
                 show_colnames = TRUE)

dev.copy(pdf, paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_MouseRNAseqData_db.pdf"))
dev.off()




SingleR.pred <- SingleR(test = Nils.seurat$CD45Pos@assays$SCT$data, 
                        ref = ImmGen.data, 
                        method = "cluster",
                        clusters = Nils.seurat$CD45Pos@meta.data$seurat_clusters,
                        labels = ImmGen.data$label.main)


# Plot 
plotScoreHeatmap(SingleR.pred, 
                 show_colnames = TRUE)

dev.copy(pdf, paste0(output.dir, "Heatmap_cluster_annotations_main_labels_MouseRNAseqData_db.pdf"))
dev.off()


```


### DEG analysis CD45Pos dataset clusters
```{r DEG_CD45Pos_clusters}

# create output dir
output.dir <- paste0("results/CD45Pos_DEG_analysis/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}



Nils.seurat$CD45Pos <- PrepSCTFindMarkers(Nils.seurat$CD45Pos)

# set assay
DefaultAssay(Nils.seurat$CD45Pos) <- "SCT"


# set idents
Idents(Nils.seurat$CD45Pos) <- "seurat_clusters"

CD45Pos.de <- FindAllMarkers(Nils.seurat$CD45Pos,
                             assay = "SCT",
                             only.pos = TRUE, 
                             verbose = TRUE)


# save to file 
write.csv(CD45Pos.de, paste0(output.dir, "CD45Pos_DEGs.csv"))


# aggregate expression for heatmap visualisation

pseudo.cd45pos <- AggregateExpression(Nils.seurat$CD45Pos,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("seurat_clusters"))



goi <- CD45Pos.de %>%
  filter(p_val_adj < 0.05) %>%
  group_by(cluster) %>%
  top_n(10, avg_log2FC) %>%
  pull(gene)


DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          raster = FALSE)

dev.copy(pdf, paste0(output.dir, "Heatmaptop_10_DEGs_pseudobulk.pdf"))
dev.off()


# iterate over the top degs for each cluster 
cluster.id <- unique(CD45Pos.de$cluster)
for(i in seq_along(cluster.id)){
  
  
  print(cluster.id[i])
  
  goi <- CD45Pos.de %>%
    filter(p_val_adj < 0.05) %>%
    filter(cluster == cluster.id[i]) %>%
    top_n(50, avg_log2FC)
  
  goi <- unique(goi)
  
  print(
    DoHeatmap(pseudo.cd45pos, 
              features = goi$gene, 
              draw.lines = FALSE, 
              raster = FALSE)
  )
  
  dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
  dev.off()
  
}


# compare with azimuth annotation 
table(Nils.seurat$CD45Pos@meta.data$predicted.celltype.l1, Nils.seurat$CD45Pos@meta.data$seurat_clusters)

table(Nils.seurat$CD45Pos@meta.data$predicted.celltype.l2, Nils.seurat$CD45Pos@meta.data$seurat_clusters)
table(Nils.seurat$CD45Pos@meta.data$predicted.celltype.l3, Nils.seurat$CD45Pos@meta.data$seurat_clusters)

# how big are clusters
x <- table(Nils.seurat$CD45Pos@meta.data$seurat_clusters)
barplot(x, las = 2)
#write.csv(x, paste0())


# interrogate specific lineage genes 

# cytokine expression 

goi <- grep("^IL.*", rownames(Nils.seurat$CD45Pos), value = T)
goi <- grep("^IF.*", rownames(Nils.seurat$CD45Pos), value = T)
goi <- grep("^CX.*", rownames(Nils.seurat$CD45Pos), value = T)
goi <- grep("^GZ.*", rownames(Nils.seurat$CD45Pos), value = T)

DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          raster = FALSE)



```






```{r}






# quantify cluster distribution and associations


sample.sum <- table(HL.seurat@meta.data$SampleID)
temp.df <- table(HL.seurat@meta.data$predicted.celltype.l1, HL.seurat@meta.data$SampleID)

percent.sample.val <- sweep(temp.df, 2, sample.sum, FUN = "/")

percent.sample.val <- percent.sample.val*100
# sanity check 
colSums(percent.sample.val)


ggplot(percent.sample.val) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  scale_fill_hue(direction = 1) +
  labs(x = "SampleID", y = "Frequency", title = "Frequency of SampleID in each Celltype.l1") +
  theme_minimal() +
  theme(legend.position = "none") +
  facet_wrap(vars(Var1), ncol = 2L)



dev.copy(pdf, paste0(output.dir, "Barplot_frequency_of_sampleID_by_celltype_l1.pdf"))
dev.off()




disease.sum <- table(HL.seurat@meta.data$DiseaseStatus)
temp.df <- table(HL.seurat@meta.data$predicted.celltype.l1, HL.seurat@meta.data$DiseaseStatus)

percent.disease.val <- sweep(temp.df, 2, disease.sum, FUN = "/")

percent.disease.val <- percent.disease.val*100
# sanity check 
colSums(percent.disease.val)


ggplot(percent.disease.val) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  scale_fill_hue(direction = 1) +
  labs(x = "Disease Status", y = "Frequency", title = "Frequency of Disease status in each Celltype.l1") +
  theme_minimal() +
  theme(legend.position = "none") +
  facet_wrap(vars(Var1), ncol = 2L, scales = "free_y")



dev.copy(pdf, paste0(output.dir, "Barplot_frequency_of_disease_status_by_celltype_l1.pdf"))
dev.off()





cluster.sum <- table(HL.seurat@meta.data$seurat_clusters)
temp.df <- table(HL.seurat@meta.data$predicted.celltype.l1, HL.seurat@meta.data$seurat_clusters)

percent.cluster.val <- sweep(temp.df, 2, cluster.sum, FUN = "/")

percent.cluster.val <- percent.cluster.val*100
# sanity check 
colSums(percent.cluster.val)


ggplot(percent.cluster.val) +
  aes(x = factor(Var2, levels = colnames(percent.cluster.val)), y = Freq, fill = Var1) +
  geom_col() +
  scale_fill_brewer(palette = "Accent", 
                    direction = 1) +
  labs(x = "Seurat Clusters", y = "Frequency of cluster", title = "Frequency of cluster according to predicted cell type .l1", 
       fill = "Predicted cell type") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

dev.copy(pdf, paste0(output.dir, "Barplot_frequency_of_seurat_clusters_by_celltype_l1.pdf"))
dev.off()





cluster.sum <- table(HL.seurat@meta.data$seurat_clusters)
temp.df <- table(HL.seurat@meta.data$predicted.celltype.l2, HL.seurat@meta.data$seurat_clusters)

percent.cluster.val <- sweep(temp.df, 2, cluster.sum, FUN = "/")

percent.cluster.val <- percent.cluster.val*100
# sanity check 
colSums(percent.cluster.val)


nrow(percent.cluster.val) # 29 cell.type variables 


# Assuming each group of 4 rows should be displayed on a separate page
group_size <- 4
total_rows <- nrow(percent.cluster.val)
seq.var <- seq(1, total_rows, by = group_size)
seq.var <- seq.var[-length(seq.var)]

# Iterate over groups of 4 rows
for (start_row in seq.var) {
  end_row <- min(start_row + group_size - 1, total_rows)
  temp.df <- percent.cluster.val[start_row:end_row, ]
  
  print(
    ggplot(temp.df) +
      aes(x = factor(Var2, levels = colnames(temp.df)), y = Freq, fill = Var1) +
      geom_col() +
      labs(x = "Seurat Clusters", y = "Frequency of cluster", 
           title = "Frequency of cluster according to predicted cell type .l1", 
           fill = "Predicted cell type") +
      theme_minimal() +
      theme(legend.position = "none") +
      theme(plot.title = element_text(hjust = 0.5)) + 
      facet_wrap(vars(Var1), ncol = 2, scales = "free_y")
  )
  # Optionally, save the plots to different files
  ggsave(paste0(output.dir, "Barplot_", start_row, "_to_", end_row, ".png"), width = 8, height = 4)
}



```


### Differential gene expression
```{r DEG_analysis}





# create output dir
output.dir <- paste0("results/DEG_analysis/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}



HL.seurat <- PrepSCTFindMarkers(HL.seurat)

# set assay
DefaultAssay(HL.seurat) <- "RNA"

# create annotation for celltype and disease status 
HL.seurat@meta.data$celltype_l1_disease <- paste(HL.seurat@meta.data$predicted.celltype.l1, HL.seurat@meta.data$DiseaseStatus, sep = "_")


Idents(HL.seurat) <- "celltype_l1_disease"
DefaultAssay(HL.seurat) <- "RNA"

NK.de <- FindMarkers(HL.seurat, ident.1 = "CD8 T_Healthy", ident.2 = "CD8 T_Patient", assay = "RNA", verbose = FALSE)


NK.de %>%
  filter(p_val_adj < 0.05) %>%
  top_n(10, avg_log2FC)



DefaultAssay(HL.seurat) <- "SCT"

NK.de.sct <- FindMarkers(HL.seurat, ident.1 = "CD8 T_Healthy", ident.2 = "CD8 T_Patient", assay = "SCT", verbose = FALSE)


NK.de.sct %>%
  filter(p_val_adj < 0.05) %>%
  top_n(10, avg_log2FC)



norm.de <- NK.de %>%
  filter(p_val_adj < 0.05) %>%
  filter(avg_log2FC > 0)


sct.de <- NK.de.sct %>%
  filter(p_val_adj < 0.05) %>%
  filter(avg_log2FC > 0)





temp.logic <- rownames(sct.de) %in% rownames(norm.de)
temp.logic <- rownames(norm.de) %in% rownames(sct.de)

rownames(norm.de)[!temp.logic]



DefaultAssay(HL.seurat) <- "SCT"

Idents(HL.seurat) <- "celltype_l1_disease"

temp.seurat <- subset(HL.seurat, idents = c("CD8 T_Healthy", "CD8 T_Patient"))
temp.seurat <- subset(temp.seurat, downsample = 300)
temp.seurat <- ScaleData(temp.seurat, features =  unique(c(rownames(sct.de), rownames(norm.de))))

goi <- unique(c(rownames(sct.de), rownames(norm.de)))
DoHeatmap(temp.seurat, features = rownames(norm.de))




names(NK.de.sct) <- paste0(names(NK.de.sct), ".sct")
NK.de.sct$gene <- rownames(NK.de.sct)

names(NK.de) <- paste0(names(NK.de), ".norm")
NK.de$gene <- rownames(NK.de)

merge_dat <- merge(NK.de.sct, NK.de, by = "gene")
merge_dat <- merge_dat[order(merge_dat$p_val_adj.sct), ]

# Number of genes that are marginally significant in both; marginally significant only in bulk; and marginally significant only in single-cell
common <- merge_dat$gene[which(merge_dat$p_val_adj.sct < 0.05 & 
                                 merge_dat$p_val_adj.norm < 0.05)]
only_norm <- merge_dat$gene[which(merge_dat$p_val_adj.sct > 0.05 & 
                                    merge_dat$p_val_adj.norm < 0.05)]
only_sct <- merge_dat$gene[which(merge_dat$p_val_adj.sct < 0.05 & 
                                   merge_dat$p_val_adj.norm > 0.05)]
print(paste0('# Common: ',length(common)))
print(paste0('# Only in norm: ',length(only_norm)))
print(paste0('# Only in sct: ',length(only_sct)))

merge_dat %>%
  filter(gene %in% only_norm) %>%
  arrange(desc(avg_log2FC.norm))


DefaultAssay(HL.seurat) <- "RNA"

Idents(HL.seurat) <- "celltype_l1_disease"

temp.seurat <- subset(HL.seurat, idents = c("CD8 T_Healthy", "CD8 T_Patient"))

RidgePlot(temp.seurat, features = "NUCKS1")



VlnPlot(temp.seurat, features = "NUCKS1", group.by = "celltype_l1_disease") 



DefaultAssay(HL.seurat) <- "SCT"
Idents(HL.seurat) <- "predicted.celltype.l1"

conserved.sct <- FindConservedMarkers(HL.seurat, ident.1 = "CD8 T", grouping.var = "DiseaseStatus", assay = "SCT", verbose = FALSE)


sct.de <- conserved.sct %>%
  filter(p_val_adj < 0.05) %>%
  filter(avg_log2FC > 0)



# perform pseudobulk DEG analysis 

# pseudobulk the counts based on donor-condition-celltype
pseudo.HL <- AggregateExpression(HL.seurat, assays = "SCT", return.seurat = T, group.by = c("SampleID", "DiseaseStatus", "predicted.celltype.l1"))

# each 'cell' is a donor-condition-celltype pseudobulk profile
tail(Cells(pseudo.HL))

pseudo.HL@meta.data$celltype_l1_disease <- paste(pseudo.HL@meta.data$predicted.celltype.l1, pseudo.HL@meta.data$DiseaseStatus, sep = "_")

Idents(pseudo.HL) <- "celltype_l1_disease"


bulk.nk.de <- FindMarkers(object = pseudo.HL, 
                          ident.1 = "CD8 T_Patient", 
                          ident.2 = "CD8 T_Healthy",
                          test.use = "DESeq2")

head(bulk.nk.de, n = 15)
bulk.nk.de %>%
  filter(p_val_adj < 0.05) %>%
  top_n(10, avg_log2FC)


VlnPlot(HL.seurat, features = "SERPINA1", #idents = c("CD8 T_Patient", "CD8 T_Healthy"), 
        group.by = "SampleID") 


table(HL.seurat@meta.data$predicted.celltype.l1, HL.seurat@meta.data$DiseaseStatus)

DimPlot(HL.seurat,
        reduction = "umap.cca", 
        split.by = "RunID",
        group.by = c("RunID"))

DefaultAssay(HL.seurat) <- "SCT"
VlnPlot(HL.seurat, features = "OTULIN", group.by = "SampleID") 

scCustomize::FeaturePlot_scCustom(HL.seurat, "SERPINA1")


DefaultAssay(HL.seurat) <- "RNA"

VlnPlot(HL.seurat, 
        features = "CD3E", 
        group.by = "predicted.celltype.l1") + NoLegend()








############################################################
# Evaluate standard normalisation vs SCT normalisation !!!

############################################################




















if(do.plot){
  
}




```









### Differential Abundance
```{r}







scCustomize::FeaturePlot_scCustom(HL.seurat, feature = "percent_mito", reduction = "umap.cca.sct")
scCustomize::FeaturePlot_scCustom(HL.seurat, feature = "percent_ribo", reduction = "umap.cca.sct")

scCustomize::FeaturePlot_scCustom(HL.seurat, feature = "nCount_RNA", reduction = "umap.cca.sct")
scCustomize::FeaturePlot_scCustom(HL.seurat, feature = "nFeature_RNA", reduction = "umap.cca.sct")

scCustomize::FeaturePlot_scCustom(HL.seurat, feature = "log10GenesPerUMI", reduction = "umap.cca.sct")
scCustomize::FeaturePlot_scCustom(HL.seurat, feature = "mapping.score", reduction = "umap.cca.sct")


# cell cycle scoring 
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes


HL.seurat <- CellCycleScoring(HL.seurat, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)

head(HL.seurat@meta.data)
scCustomize::FeaturePlot_scCustom(HL.seurat, feature = "S.Score", reduction = "umap.cca.sct")
scCustomize::FeaturePlot_scCustom(HL.seurat, feature = "G2M.Score", reduction = "umap.cca.sct")


table(HL.seurat$hash.ID, HL.seurat$predicted.celltype.l3)

table(HL.seurat$RunID, HL.seurat$predicted.celltype.l3)





head(HL.seurat@meta.data)
unique(HL.seurat@meta.data$DiseaseStatus)
unique(HL.seurat@meta.data$SampleID)
unique(HL.seurat@meta.data$DiseaseOutcome)


DimPlot(HL.seurat,
        reduction = "umap.cca.sct", 
        label = T,
        group.by = c("DiseaseStatus")) + NoLegend() + ggtitle("clusters_cca_sct")


DimPlot(HL.seurat,
        reduction = "umap.cca.sct", 
        label = T,
        group.by = c("SampleID")) + NoLegend() + ggtitle("clusters_cca_sct")


DimPlot(HL.seurat,
        reduction = "umap.cca.sct", 
        label = T,
        group.by = c("DiseaseOutcome")) + NoLegend() + ggtitle("clusters_cca_sct")


DimPlot(HL.seurat,
        reduction = "umap.cca.sct", 
        split.by = c("DiseaseOutcome")) + NoLegend() + ggtitle("clusters_cca_sct")

table(HL.seurat$DiseaseOutcome)



# Why got NA values in disease outcome

#dev.copy(pdf, paste0(output.dir, "UMAP_before_after_integration_RunID_split.pdf"))
#dev.off()


}




```





## normalise ADT dataset
```{r}


if(long.compute){
  
  ############################################
  # Normalise and scale ADT data
  ############################################  
  
  
  # ensure ADT is the default assay and set variable features to all ADT markers in dataset
  
  DefaultAssay(HL.seurat) <- "ADT"
  
  VariableFeatures(HL.seurat[["ADT"]]) <- rownames(HL.seurat[["ADT"]])
  
  # split ADT assay according to sequencing run
  HL.seurat[["ADT"]] <- split(HL.seurat[["ADT"]],  f = HL.seurat$RunID)
  
  
  HL.seurat <- NormalizeData(HL.seurat,
                             normalization.method = "CLR", 
                             margin = "2", 
                             assay = "ADT") %>%
    ScaleData() %>%
    RunPCA(reduction.name = 'adtPCA_unintegrated') %>%
    FindNeighbors(reduction = "adtPCA_unintegrated") %>%
    FindClusters(resolution = 0.8, cluster.name = "adt_unintegrated_clusters") %>%
    RunUMAP(dims = 1:30, reduction = "adtPCA_unintegrated", reduction.name = "umap.adt.unintegrated")
  
  
  
  
}





```







```{r}












# Identify multimodal neighbors. These will be stored in the neighbors slot, 
# and can be accessed using bm[['weighted.nn']]
# The WNN graph can be accessed at bm[["wknn"]], 
# and the SNN graph used for clustering at bm[["wsnn"]]
# Cell-specific modality weights can be accessed at bm$RNA.weight
bm <- FindMultiModalNeighbors(
  bm, reduction.list = list("pca", "apca"), 
  dims.list = list(1:30, 1:18), modality.weight.name = "RNA.weight"
)


bm <- RunUMAP(bm, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_")
bm <- FindClusters(bm, graph.name = "wsnn", algorithm = 3, resolution = 2, verbose = FALSE)


p1 <- DimPlot(bm, reduction = 'wnn.umap', label = TRUE, repel = TRUE, label.size = 2.5) + NoLegend()
p2 <- DimPlot(bm, reduction = 'wnn.umap', group.by = 'celltype.l2', label = TRUE, repel = TRUE, label.size = 2.5) + NoLegend()
p1 + p2




Run.3.seurat <- Azimuth::RunAzimuth(Run.3.seurat, reference = "pbmcref")






```





