---
title: "CD45Pos_dataset_analysis"
author: "Dillon Corvino"
date: "25/01/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# Data 
# B16 - OVA 

# PDS = I.T - bind Gquadraplex
# day 4 post-treatment 


# Research questions 
# How does PDS induces an anti-tumor immune response?
# Known that after PDS treatment, certain DAMPs are induced, intratumoral chemokine levels go up drastically and subsequntly immune cells are recruited into the tumor.
# Main research questions are:
#1. Which cell subsets are producing the chemokines
#2. Are there differences between macrophages and DCs between PDS and Vehicle (because there are DAMPs that should activate them)
#3. How does the treatment effect the tumor cells (is it possible to discriminate them from other tissue cells, if not they express ovalbumin...this should make it at least possible to identify them.)




# key genes
# IL1b, IL6, IL12, IFNG, TNFA, 

# CCL2,3,4,5 
# CXCL1, CXCL9,10,13 





```


### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)


# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- TRUE
do.plot <- TRUE
quick.load <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())


library("Seurat")
library("SeuratObject")
library("SeuratData")
library("SeuratWrappers")
library("Azimuth")
library("dplyr")
library("scCustomize")
library("tibble")

########################################
# Parallelization of seurat functions
########################################

parallelization <- TRUE


if(parallelization){
  library(future)
  # check the current active plan
  plan()
  
  # detect cores
  ncores <- parallel::detectCores()
  ncores <- ncores-5
  # set plan
  plan("multisession", workers = ncores)
  plan()
  
  # set max limit for global variables
  # note this will drastically increase RAM usage, be aware of machine limits
  # set value is in bites therefore 1000*1024^2 = 1GB Therefore set to 10GBs 
  options(future.globals.maxSize = 10000 * 1024^2, 
          future.rng.onMisuse="ignore") 
}



if(quick.load){
  
  CD45pos.seurat <- LoadSeuratRds("saves/Nils_seurat_CD45Pos.rds")
  
}
```



## DEG calculation and visualisation {.tabset}

### DEG analysis CD45Pos dataset clusters
```{r DEG_CD45Pos_clusters}

# note this is a recalculation of what was done in pre-processing script and should not be different for cd45pos dataset, but as some clusters were removed in cd45neg dataset, i am redoing this deg calculation in these dataset specific scripts
# as a mechanism to explictly overwrite any old data and ensure deg calculations are done on most up-to-date seurat objects
# poor use of resources and compute time but done for clarity


# create output dir
output.dir <- paste0("results/CD45Pos_DEG_analysis/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  CD45pos.seurat <- PrepSCTFindMarkers(CD45pos.seurat)
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  # set idents
  Idents(CD45pos.seurat) <- "seurat_clusters"
  
  CD45Pos.de <- FindAllMarkers(CD45pos.seurat,
                               assay = "SCT",
                               only.pos = TRUE, 
                               verbose = TRUE)
  
  
  # save to file 
  write.csv(CD45Pos.de, paste0(output.dir, "CD45Pos_DEGs.csv"))
  
}else{
  
  CD45Pos.de <- read.csv(paste0(output.dir, "CD45Pos_DEGs.csv"))
  
  
}


# aggregate expression for heatmap visualisation

pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("seurat_clusters"))



goi <- CD45Pos.de %>%
  filter(p_val_adj < 0.05) %>%
  group_by(cluster) %>%
  top_n(10, avg_log2FC) %>%
  pull(gene)


DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          raster = FALSE)

dev.copy(pdf, paste0(output.dir, "Heatmaptop_10_DEGs_pseudobulk.pdf"))
dev.off()


if(do.plot){
  # iterate over the top degs for each cluster 
  cluster.id <- unique(CD45Pos.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- CD45Pos.de %>%
      filter(p_val_adj < 0.05) %>%
      filter(cluster == cluster.id[i]) %>%
      top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.cd45pos, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE)
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
}



```
 
 
 
 
## annotate clusters {.tabset}

```{r annotate_clusters}

########################################################
# provide annotation for clusters where possible 
########################################################


# create output dir
output.dir <- paste0("results/CD45Pos_cluster_vs_azimuth/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

# compare with azimuth annotation 

# calculate the frequency of each cluster that has an azimuth annotation, therefore the clusters which are primarily 1 annotation can easily be defined 
x <- table(CD45pos.seurat@meta.data$predicted.celltype.l1, CD45pos.seurat@meta.data$seurat_clusters)
cluster.totals <- colSums(x)
y <- sweep(x, 2, cluster.totals, FUN = "/")
y <- y*100
y


# set a threshold 
# clusters with greater than this value will become clusterX_Bcell or whatever annotation
# other clusters are left ambiguous to be decided later
threshold.limit <- 90

library(ggplot2)

ggplot(y) +
  aes(x = Var2, y = Freq, fill = Var1) +
  geom_col(show.legend = FALSE) +
  scale_fill_hue(direction = 1) +
  theme_minimal() +
  facet_wrap(vars(Var1), ncol = 2L) +
  geom_hline(yintercept = threshold.limit, color = "red", linetype = "solid")

dev.copy(pdf, paste0(output.dir, "CD45Pos_freq_of_cluster_by_azimuth_anno.pdf"))
dev.off()

# many clusters are quite exclusively annotated as a single cell subtype
# these can be defined and annotated as such for eaiser downstream interpretation 
# other clusters will be annotated biologically later

threshold.y <- y >= threshold.limit

colSums(threshold.y)


# use fine grain azimuth anno to try see if can define these populations more specifically 


# lower threshold for fine grain annotation  
threshold.limit <- 80

####################################
# inspect monocyte clusters 
####################################

temp.df <- CD45pos.seurat@meta.data %>%
  filter(seurat_clusters %in% mono.clusters)

temp.df$seurat_clusters <- droplevels(temp.df$seurat_clusters)

# calculate the frequency of each cluster using fine grain azimuth
x <- table(temp.df$predicted.celltype.l3, temp.df$seurat_clusters)
cluster.totals <- colSums(x)
y <- sweep(x, 2, cluster.totals, FUN = "/")
y <- y*100
y <- as.data.frame(y)

y_filtered <- y %>%
  filter(grepl("Mono", Var1))

ggplot(y_filtered) +
  aes(x = Var2, y = Freq, fill = Var1) +
  geom_col(show.legend = FALSE) +
  scale_fill_hue(direction = 1) +
  theme_minimal() +
  facet_wrap(vars(Var1), ncol = 2L) +
  geom_hline(yintercept = threshold.limit, color = "red", linetype = "solid")

dev.copy(pdf, paste0(output.dir, "CD45Pos_monocyte_clusters.pdf"))
dev.off()



####################################
# inspect DC clusters 
####################################

temp.df <- CD45pos.seurat@meta.data %>%
  filter(seurat_clusters %in% dc.clusters)

temp.df$seurat_clusters <- droplevels(temp.df$seurat_clusters)

# calculate the frequency of each cluster using fine grain azimuth
x <- table(temp.df$predicted.celltype.l3, temp.df$seurat_clusters)
cluster.totals <- colSums(x)
y <- sweep(x, 2, cluster.totals, FUN = "/")
y <- y*100
y <- as.data.frame(y)

y_filtered <- y %>%
  filter(grepl("DC", Var1))

ggplot(y_filtered) +
  aes(x = Var2, y = Freq, fill = Var1) +
  geom_col(show.legend = FALSE) +
  scale_fill_hue(direction = 1) +
  theme_minimal() +
  facet_wrap(vars(Var1), ncol = 2L) +
  geom_hline(yintercept = threshold.limit, color = "red", linetype = "solid")

dev.copy(pdf, paste0(output.dir, "CD45Pos_DC_clusters.pdf"))
dev.off()


# 26 is clearly a pDC looking cluster 
# but cluster 18 is unclear if cDC1 or cDC2, will need to inspect gene expression profile 




####################################
# inspect NK clusters 
####################################


temp.df <- CD45pos.seurat@meta.data %>%
  filter(seurat_clusters %in% nk.clusters)

temp.df$seurat_clusters <- droplevels(temp.df$seurat_clusters)

# calculate the frequency of each cluster using fine grain azimuth
x <- table(temp.df$predicted.celltype.l2, temp.df$seurat_clusters)
cluster.totals <- colSums(x)
y <- sweep(x, 2, cluster.totals, FUN = "/")
y <- y*100
y <- as.data.frame(y)

y_filtered <- y %>%
  filter(grepl("NK", Var1))

ggplot(y_filtered) +
  aes(x = Var2, y = Freq, fill = Var1) +
  geom_col(show.legend = FALSE) +
  scale_fill_hue(direction = 1) +
  theme_minimal() +
  facet_wrap(vars(Var1), ncol = 2L) +
  geom_hline(yintercept = threshold.limit, color = "red", linetype = "solid")

dev.copy(pdf, paste0(output.dir, "CD45Pos_NK_clusters.pdf"))
dev.off()

# cluster 8 seems to mostly be of NK_4 type annotation at level 3 and just NK at level 2


####################################
# inspect B cell clusters
####################################


temp.df <- CD45pos.seurat@meta.data %>%
  filter(seurat_clusters %in% bcell.clusters)

temp.df$seurat_clusters <- droplevels(temp.df$seurat_clusters)

# calculate the frequency of each cluster using fine grain azimuth
x <- table(temp.df$predicted.celltype.l2, temp.df$seurat_clusters)
cluster.totals <- colSums(x)
y <- sweep(x, 2, cluster.totals, FUN = "/")
y <- y*100
y <- as.data.frame(y)

y_filtered <- y %>%
  filter(grepl("B", Var1))

ggplot(y_filtered) +
  aes(x = Var2, y = Freq, fill = Var1) +
  geom_col(show.legend = FALSE) +
  scale_fill_hue(direction = 1) +
  theme_minimal() +
  facet_wrap(vars(Var1), ncol = 2L) +
  geom_hline(yintercept = threshold.limit, color = "red", linetype = "solid")

dev.copy(pdf, paste0(output.dir, "CD45Pos_B_clusters.pdf"))
dev.off()

# cluster 5 looks to be more like a B-naive cluster but 30% is representated as B-intermediate. 

# fine grain resolution of this will require gene expression data 


#
# different way to visualise data

#############################
# cluster by annotation 
#############################

# large grain annotation

# broad annotation 
## look at all clusters again across annotation. as some clusters maybe can be called X celltype where they are say 70% Xcelltype and rest is made up of small annotations 
# as opposed to clusters 50/50 split between annotations 


# calculate the frequency of each cluster that has an azimuth annotation, therefore the clusters which are primarily 1 annotation can easily be defined 
x <- table(CD45pos.seurat@meta.data$predicted.celltype.l1, CD45pos.seurat@meta.data$seurat_clusters)
cluster.totals <- colSums(x)
y <- sweep(x, 2, cluster.totals, FUN = "/")
y <- y*100
y <- as.data.frame(y)


cluster.id <- unique(CD45pos.seurat@meta.data$seurat_clusters)

for(i in seq_along(cluster.id)){
  
  y.filt <- y %>%
    filter(Var2 == cluster.id[i])
  
  
  library(ggplot2)
  
  p1 <- ggplot(y.filt) +
    aes(x = Var1, y = Freq, fill = Var1) +
    geom_col() +
    scale_fill_hue(direction = 1) +
    labs(title = paste0("Cluster ", cluster.id[i])) +
           coord_flip() +
           theme_minimal() +
           theme(legend.position = "none") +
           ylim(0, 100)
         
         print(p1)
         
         dev.copy(pdf, paste0(output.dir, "Cluster_", cluster.id[i], "_broad_azimuth_annotation_freq.pdf"))
         dev.off()
         
}


#############################
# medium grain annotation 
#############################

## look at all clusters again across annotation. as some clusters maybe can be called X celltype where they are say 70% Xcelltype and rest is made up of small annotations 
# as opposed to clusters 50/50 split between annotations 


# calculate the frequency of each cluster that has an azimuth annotation, therefore the clusters which are primarily 1 annotation can easily be defined 
x <- table(CD45pos.seurat@meta.data$predicted.celltype.l2, CD45pos.seurat@meta.data$seurat_clusters)
cluster.totals <- colSums(x)
y <- sweep(x, 2, cluster.totals, FUN = "/")
y <- y*100
y <- as.data.frame(y)


cluster.id <- unique(CD45pos.seurat@meta.data$seurat_clusters)

for(i in seq_along(cluster.id)){
  
  y.filt <- y %>%
    filter(Var2 == cluster.id[i])
  
  
  library(ggplot2)
  
  p1 <- ggplot(y.filt) +
    aes(x = Var1, y = Freq, fill = Var1) +
    geom_col() +
    scale_fill_hue(direction = 1) +
    labs(title = paste0("Cluster ", cluster.id[i])) +
           coord_flip() +
           theme_minimal() +
           theme(legend.position = "none") +
           ylim(0, 100)
         
         print(p1)
         
         dev.copy(pdf, paste0(output.dir, "Cluster_", cluster.id[i], "_medium_grain_azimuth_annotation_freq.pdf"))
         dev.off()
         
}

#############################
# fine grain annotation 
#############################

## look at all clusters again across annotation. as some clusters maybe can be called X celltype where they are say 70% Xcelltype and rest is made up of small annotations 
# as opposed to clusters 50/50 split between annotations 


# calculate the frequency of each cluster that has an azimuth annotation, therefore the clusters which are primarily 1 annotation can easily be defined 
x <- table(CD45pos.seurat@meta.data$predicted.celltype.l3, CD45pos.seurat@meta.data$seurat_clusters)
cluster.totals <- colSums(x)
y <- sweep(x, 2, cluster.totals, FUN = "/")
y <- y*100
y <- as.data.frame(y)


cluster.id <- unique(CD45pos.seurat@meta.data$seurat_clusters)

for(i in seq_along(cluster.id)){
  
  y.filt <- y %>%
    filter(Var2 == cluster.id[i])
  
  
  library(ggplot2)
  
  p1 <- ggplot(y.filt) +
    aes(x = Var1, y = Freq, fill = Var1) +
    geom_col() +
    scale_fill_hue(direction = 1) +
    labs(title = paste0("Cluster ", cluster.id[i])) +
           coord_flip() +
           theme_minimal() +
           theme(legend.position = "none") +
           ylim(0, 100)
         
         print(p1)
         
         dev.copy(pdf, paste0(output.dir, "Cluster_", cluster.id[i], "_fine_grain_azimuth_annotation_freq.pdf"))
         dev.off()
         
}


######################################################
# using all above plots annotation definitions are 
######################################################

# broad annotations
# cluster annotations 
mono.clusters <- c(2, 4, 6, 7, 9, 12, 13, 14, 17, 19, 20, 21, 22)
nk.clusters <- c(8, 16)
bcell.clusters <- c(5)
dc.clusters <- c(11, 18, 26)
cd4.clusters <- c(23)
cd8.clusters <- c(3)
erythrocyte.clusters <- c(27)


# fine grain annotations
pDC.clusters <- c(26)
cd14.mono.clusters <- c(2, 4, 6, 9, 14, 20, 22)
cd16.mono.clusters <- c(13, 17)
cDC1.clusters <- c(18)
cDC2.clusters <- c(11)
cd8.EM <- c(3)
cd4.cm <- c(23)



# ambig clusters 
# cluster 25 = nk/cd8
# cluster 1 cd4/cd8
# cluster 10 nk/cd8
# cluster 0 mono/dc
# cluster 24 and 15 = mono/other
# cluster 27 = other maybe erythrocytes


# for defined clusters confirm identity with connonical markers


CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
  dplyr::mutate(Anno_l1 = case_when(seurat_clusters %in% mono.clusters ~ paste0("Mono_cluster_", seurat_clusters), 
                                    seurat_clusters %in% nk.clusters ~ paste0("NK_cluster_", seurat_clusters),
                                    seurat_clusters %in% bcell.clusters ~ paste0("Bcell_cluster_", seurat_clusters),
                                    seurat_clusters %in% dc.clusters ~ paste0("DC_cluster_", seurat_clusters),
                                    seurat_clusters %in% cd4.clusters ~ paste0("CD4_cluster_", seurat_clusters),
                                    seurat_clusters %in% cd8.clusters ~ paste0("CD8_cluster_", seurat_clusters),
                                    seurat_clusters %in% erythrocyte.clusters ~ paste0("Erythrocyte_cluster_", seurat_clusters),
                                    TRUE ~ seurat_clusters), 
                Anno_l2 = case_when(seurat_clusters %in% pDC.clusters ~ paste0("pDC_cluster_", seurat_clusters),
                                    seurat_clusters %in% cd14.mono.clusters ~ paste0("CD14_Mono_cluster_", seurat_clusters), 
                                    seurat_clusters %in% cd16.mono.clusters ~ paste0("CD16_Mono_cluster_", seurat_clusters), 
                                    seurat_clusters %in% cDC1.clusters ~ paste0("cDC1_cluster_", seurat_clusters), 
                                    seurat_clusters %in% cDC2.clusters ~ paste0("cDC2_cluster_", seurat_clusters), 
                                    seurat_clusters %in% cd8.EM ~ paste0("CD8_EM_cluster_", seurat_clusters), 
                                    seurat_clusters %in% cd4.cm ~ paste0("CD4_CM_cluster_", seurat_clusters), 
                                    TRUE ~ Anno_l1))








DimPlot_scCustom(CD45pos.seurat, 
                 label = T,
                 group.by = "Anno_l1") + NoLegend()

DimPlot_scCustom(CD45pos.seurat, 
                 label = T,
                 group.by = "Anno_l2") + NoLegend()








# cluster 5 looks to be clearly B cell cluster 


# how big are clusters
x <- table(CD45pos.seurat@meta.data$seurat_clusters)
barplot(x, las = 2)
#write.csv(x, paste0())


```








# NOT WORKING - check all code 
# also not written again for this dataset specific analysis - check all code 


## Which cell subsets are producing the chemokines {.tabset}

## WORK IN PROGRESS
### Vis chemokine/cytokine expression
```{r chemokine_cytokine_expression}

# create output dir
output.dir <- paste0("results/chemo_cyto_expression/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}




# aggregate expression for heatmap visualisation
pseudo.cd45pos <- AggregateExpression(Nils.seurat$CD45Pos,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("seurat_clusters"))


pseudo.cd45neg <- AggregateExpression(Nils.seurat$CD45Neg,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("seurat_clusters"))



# identify chemokines/cytokines of interest
goi <- grep("^IL.*", rownames(pseudo.cd45pos), value = T)
goi


goi <- grep("^IF.*", rownames(Nils.seurat$CD45Pos), value = T)
goi <- grep("^CX.*", rownames(Nils.seurat$CD45Pos), value = T)
goi <- grep("^GZ.*", rownames(Nils.seurat$CD45Pos), value = T)
goi



goi <- c("IL1B", "IL6", "IFNG", "TNF", "CCL2", "CCL3", "CCL4", "CCL5", "CXCL1", "CXCL9", "CXCL10", "CXCL13")

DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          raster = FALSE)


head(Nils.seurat$CD45Pos@meta.data)
scCustomize::DimPlot_scCustom(Nils.seurat$CD45Pos,
                              group.by = "predicted.celltype.l2", 
                              reduction = "umap",
                              label = T)

scCustomize::FeaturePlot_scCustom(Nils.seurat$CD45Pos,
                                  feature = "IL7R", 
                                  reduction = "umap")









goi <- grep("^Il1b.*", rownames(pseudo.cd45neg), value = T)
goi


goi <- c("Il1b", "Il6", "Ifng", "Tnf")
goi <- c("Ccl2", "Ccl3", "Ccl4", "Ccl5")
goi <- c("Cxcl1", "Cxcl9", "Cxcl10", "Cxcl13")
DoHeatmap(pseudo.cd45neg, 
          features = goi, 
          draw.lines = FALSE, 
          raster = FALSE)

#dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
#dev.off()


# key genes
# IL1b, IL6, IL12, IFNG, TNFA, 

# CCL2,3,4,5 
# CXCL1, CXCL9,10,13 




```











# NOT WORKING - check all code 
## Difference between treatment and vehicle {.tabset}
# Cant really do DA without replicate samples 

### Differential Abundance testing
```{r Differential_Abundance_testing}

#BiocManager::install("miloR")

library("miloR")
library("SingleCellExperiment")
library("scater")
library("scran")


#############################################
# Perform analysis for CD45Pos dataset
#############################################


# set assay
DefaultAssay(Nils.seurat$CD45Pos) <- "SCT"



# create output dir
output.dir <- paste0("results/DA_analysis_CD45Pos/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}





# Note: 
# milo object capatilises lots of variables, ie umap = UMAP
# below is method was used before i realised this, keeping for reference 
# some error in buildGraph() call prevents specifiying to use integrated.cca as reduced.dim, so it recalculates PCA, therefore rewrite PCA to be integrated.cca
#HL.seurat[["pca"]] <- CreateDimReducObject(embeddings = HL.seurat[["integrated.cca"]]@cell.embeddings, key = "PC_ ", assay = DefaultAssay(HL.seurat))
#names of dim reductions get capatilized for some reason, therefore reduced.dim = "INTEGRATED.CCA"


#######################
# Create milo object
#######################

# create single cell experiment object
CD45Pos.sce <- Seurat::as.SingleCellExperiment(Nils.seurat[["CD45Pos"]],
                                               assay = "SCT")

# convert object into a milo object
CD45Pos.sce <- Milo(CD45Pos.sce)
CD45Pos.sce



#######################
# Create KNN graph
#######################

# use nn graph generated as part of seurat pipeline
Graphs(CD45Pos.sce, slot = NULL)
Neighbors(CD45Pos.sce, slot = NULL)
Reductions(CD45Pos.sce, slot = NULL)


CD45Pos.sce@graph$SCT_nn <- Nils.seurat[["CD45Pos"]]@graphs$SCT_nn


# build from the adjacency matrix
milo.graph <- miloR::buildFromAdjacency(CD45Pos.sce@graph$SCT_nn)

# append graph to milo object
CD45Pos.sce@graph <- milo.graph@graph


# set some global-ish variables 
k.parm.use <- 20
d.use <- 50


# conversly, recaculate knn graph 
#HL.sce <- buildGraph(HL.sce,
#                 k = k.parm.use,
#                  d = 30, 
#            reduced.dim = "INTEGRATED.CCA")


########################
# make Neighbourhoods
########################

CD45Pos.sce <- makeNhoods(CD45Pos.sce,
                          prop = 0.1,
                          k = k.parm.use, # default for seurat findneighbours is k.parm = 20
                          d = d.use, # dim reduction integrated.cca is 50 dims
                          reduced_dims = "INTEGRATED.CCA.SCT",
                          refined = TRUE, 
                          refinement_scheme = "graph") # make nhoods using graph-only as this is faster

# Rule of thumb, aim for average neighbourhood size over 5*n_samples 
# therefore 5*10 = 50
#Empirically, we found it’s best to have a distribution peaking between 50 and 100


plotNhoodSizeHist(CD45Pos.sce, bins = 100)

dev.copy(pdf, paste0(output.dir, "Neighbourhood_size_histogram.pdf"))
dev.off()


########################
# count cells 
########################

CD45Pos.sce <- countCells(CD45Pos.sce, 
                          meta.data = data.frame(colData(CD45Pos.sce)),
                          samples = "Treatment") # make the nhood X sample counts matrix

head(nhoodCounts(CD45Pos.sce))

##############################
# calculate nhood distance
##############################

# longest compute time of this entire pipeline 
CD45Pos.sce <- calcNhoodDistance(CD45Pos.sce, 
                                 d = d.use, 
                                 reduced.dim = "INTEGRATED.CCA.SCT")


####################################
# Differential Abundance testing
####################################

# design model matrix for DA testing

DA.design <- data.frame(colData(CD45Pos.sce))[,c("SampleID", "Treatment")]
DA.design <- distinct(DA.design)
rownames(DA.design) <- DA.design$SampleID

## Reorder rownames to match columns of nhoodCounts(milo)
DA.design <- DA.design[colnames(nhoodCounts(CD45Pos.sce)), , drop=FALSE]
table(DA.design$SampleID)

table(DA.design$Treatment)

# create contrasts
#contrast.1 <- c("Age1wk - Age4wk") # the syntax is <VariableName><ConditionLevel> - <VariableName><ControlLevel>
contrast.status <- c("SampleIDPDS_CD45Pos - SampleIDVehicle_CD45Pos")


# this is the edgeR code called by `testNhoods`
# check it looks correct
model <- model.matrix(~ 0 + SampleID, data=DA.design)
mod.constrast <- makeContrasts(contrasts=contrast.status, levels=model)

mod.constrast


# perform DA test
da_results_status <- testNhoods(CD45Pos.sce, 
                                design = ~ SampleID,
                                design.df = DA.design, 
                                model.contrasts = contrast.status,
                                fdr.weighting= "graph-overlap", 
                                reduced.dim = "INTEGRATED.CCA.SCT")






# old single contrast code 

# 
matrix_design <- data.frame(colData(CD45Pos.sce))[,c("Treatment", "SampleID")]
matrix_design
# 
# ## factorise Treatment
matrix_design$Treatment <- as.factor(matrix_design$Treatment) 
matrix_design <- distinct(matrix_design)
rownames(matrix_design) <- matrix_design$Treatment

matrix_design
# 
# 
# # test nhoods for DA
da_results <- testNhoods(CD45Pos.sce,
                         design = ~ Treatment,
                         design.df = matrix_design, 
                         reduced.dim = "INTEGRATED.CCA.SCT")



#######################
# overview of results 
#######################


# how many nhoods are DA
da_results_status %>%
  filter(SpatialFDR < 0.05) %>%
  arrange(SpatialFDR) %>%
  head()

table(da_results_status$SpatialFDR < 0.05) # 513 DA nhoods
table(da_results_outcome$SpatialFDR < 0.05) # 3 DA nhoods

length(da_results_status$Nhood) # 2,833 nhoods
length(da_results_outcome$Nhood) # 2,833 nhoods

# distribution of pvalue

ggplot(da_results_status, aes(PValue)) + geom_histogram(bins=50)

dev.copy(pdf, paste0(output.dir, "DA_result_pvalue_histogram_status.pdf"))
dev.off()


ggplot(da_results_outcome, aes(PValue)) + geom_histogram(bins=50)

dev.copy(pdf, paste0(output.dir, "DA_result_pvalue_histogram_outcome.pdf"))
dev.off()


# volcano plots 


ggplot(da_results_status, aes(logFC, -log10(SpatialFDR))) + 
  geom_point() +
  geom_hline(yintercept = 1) ## Mark significance threshold (10% FDR

dev.copy(pdf, paste0(output.dir, "DA_result_volcano_status.pdf"))
dev.off()


ggplot(da_results_outcome, aes(logFC, -log10(SpatialFDR))) + 
  geom_point() +
  geom_hline(yintercept = 1) ## Mark significance threshold (10% FDR

dev.copy(pdf, paste0(output.dir, "DA_result_volcano_outcome.pdf"))
dev.off()




head(da_results)

######################################################
# visualise nhoods on umap projection
######################################################

# Build nhood graph
HL.sce <- buildNhoodGraph(HL.sce)



# Plot single-cell UMAP
umap_pl <- plotReducedDim(HL.sce, 
                          dimred = "UMAP",
                          colour_by="predicted.celltype.l1",
                          text_by = "predicted.celltype.l1", 
                          text_size = 3, 
                          point_size = 0.5) +
  guides(fill = "none")

## Plot neighbourhood graph
nh_graph_status <- plotNhoodGraphDA(HL.sce,
                                    da_results_status, 
                                    layout="UMAP",
                                    alpha=0.1) 


nh_graph_outcome <- plotNhoodGraphDA(HL.sce,
                                     da_results_outcome, 
                                     layout="UMAP",
                                     alpha=0.1) 

# print plots
umap_pl + nh_graph_status + 
  patchwork::plot_layout(guides = "collect")

dev.copy(pdf, paste0(output.dir, "UMAP_Nhood_DA_graph_status.pdf"))
dev.off()

umap_pl + nh_graph_outcome + 
  patchwork::plot_layout(guides = "collect")

dev.copy(pdf, paste0(output.dir, "UMAP_Nhood_DA_graph_outcome.pdf"))
dev.off()



# for context show umap with disease status highlighted
plotReducedDim(HL.sce, 
               dimred = "UMAP",
               colour_by="DiseaseStatus",
               text_by = "DiseaseStatus", 
               text_size = 3, 
               point_size = 0.5) +
  guides(fill = "none")

dev.copy(pdf, paste0(output.dir, "UMAP_DiseaseStatus.pdf"))
dev.off()


plotReducedDim(HL.sce, 
               dimred = "UMAP",
               colour_by="DiseaseOutcome",
               text_by = "DiseaseOutcome", 
               text_size = 3, 
               point_size = 0.5) +
  guides(fill = "none")

dev.copy(pdf, paste0(output.dir, "UMAP_DiseaseOutcome.pdf"))
dev.off()




# proceed with analysis for DiseaseStatus only, currently dataset seems insufficient for calling DiseaseOutcome 


table(HL.seurat@meta.data$predicted.celltype.l3)

#############################
# annotate neighbourhoods 
#############################

da_results_status_anno <- annotateNhoods(HL.sce, da_results_status, coldata_col = "predicted.celltype.l1")
da_results_status_anno <- annotateNhoods(HL.sce, da_results_status_anno, coldata_col = "predicted.celltype.l2")
da_results_status_anno <- annotateNhoods(HL.sce, da_results_status_anno, coldata_col = "predicted.celltype.l3")
da_results_status_anno <- annotateNhoods(HL.sce, da_results_status_anno, coldata_col = "seurat_clusters")

head(da_results_status_anno)


# Neighbourhoods tend to be homogeneous, we can define a threshold for celltype_fraction to exclude neighbourhoods that are a mix of cell types.

# visualise at each level of annotation 
anno.var <- c("predicted.celltype.l1", "predicted.celltype.l2", "predicted.celltype.l3", "seurat_clusters")


for(i in seq_along(anno.var)){
  
  plot.var <- paste0(anno.var[i], "_fraction")
  name.var <- gsub("\\.", "_", anno.var[i])
  print(ggplot(da_results_status_anno, aes(!!sym(plot.var))) + geom_histogram(bins=100))
  
  dev.copy(pdf, paste0(output.dir, "DA_results_DiseaseOutcome_histogram_nhood_", name.var, "fraction.pdf"))
  dev.off()
}



# use above histograms to define thresholds for annotating nhoods as mixed at each resolution of annotation 

threshold <- 0.9


da_results_status_anno$celltype.l1 <- ifelse(da_results_status_anno$predicted.celltype.l1_fraction < threshold, "Mixed", da_results_status_anno$predicted.celltype.l1)
da_results_status_anno$celltype.l2 <- ifelse(da_results_status_anno$predicted.celltype.l2_fraction < threshold, "Mixed", da_results_status_anno$predicted.celltype.l2)
da_results_status_anno$celltype.l3 <- ifelse(da_results_status_anno$predicted.celltype.l3_fraction < threshold, "Mixed", da_results_status_anno$predicted.celltype.l3)
da_results_status_anno$seurat_clusters <- ifelse(da_results_status_anno$seurat_clusters_fraction < threshold, "Mixed", da_results_status_anno$seurat_clusters)



length(da_results_status_anno$Nhood) # 2,833 nhoods
table(da_results_status_anno$celltype.l1 == "Mixed") # 700 nhoods became mixed
table(da_results_status_anno$celltype.l2 == "Mixed") # 1335 nhoods became mixed
table(da_results_status_anno$celltype.l3 == "Mixed") # 1891 nhoods became mixed
table(da_results_status_anno$seurat_clusters == "Mixed") # 1045 nhoods became mixed






# visualise at each level of annotation 
anno.var <- c("predicted.celltype.l1", "predicted.celltype.l2", "predicted.celltype.l3", "seurat_clusters")


for(i in seq_along(anno.var)){
  
  plot.var <- paste0(anno.var[i], "_fraction")
  name.var <- gsub("\\.", "_", anno.var[i])
  
  
  # Create a ggplot with histogram and line
  print(
    ggplot(da_results_status_anno, aes(x = !!sym(plot.var))) +
      geom_histogram(aes(fill = ifelse(!!sym(plot.var) > threshold, "Above", "Below")), bins = 100, color = "black") +
      geom_vline(xintercept = threshold, color = "green", linetype = "dashed") +
      scale_fill_manual(values = c("red", "blue"), name = "Category") +
      theme_minimal()  # Add additional themes as needed
  )
  
  dev.copy(pdf, paste0(output.dir, "DA_results_DiseaseOutcome_histogram_nhood_", name.var, "fraction_withthreshold.pdf"))
  dev.off()
}




# distribution of DA nhoods across annotation 
table(da_results_status_anno$SpatialFDR < 0.05, da_results_status_anno$celltype.l1)
table(da_results_status_anno$SpatialFDR < 0.05, da_results_status_anno$celltype.l2)
table(da_results_status_anno$SpatialFDR < 0.05, da_results_status_anno$celltype.l3)
table(da_results_status_anno$SpatialFDR < 0.05, da_results_status_anno$seurat_clusters)



# Now we can visualize the distribution of DA Fold Changes in different cell types

# iterate over annotation levels 

anno.var <- c("celltype.l1", "celltype.l2", "celltype.l3", "seurat_clusters")

# factorise for plotting
da_results_status_anno$seurat_clusters <- factor(da_results_status_anno$seurat_clusters, levels = c(levels(HL.seurat$seurat_clusters), "Mixed"))



for(i in seq_along(anno.var)){
  
  name.var <- gsub("\\.", "_", anno.var[i])
  
  print(
    plotDAbeeswarm(da_results_status_anno, 
                   group.by = anno.var[i])
  )
  dev.copy(pdf, paste0(output.dir, "Beeswarm_DA_DiseaseStatus_results_", name.var, ".pdf"))
  dev.off()
  
}

#####################################################################################
# Group nhoods into larger groups for later deg identification
#####################################################################################

# once neighbourhoods are identified that are sig between conditions
# we want to identify markers of these DA neighbourhoods

#The function findNhoodGroupMarkers runs a one-VS-all differential gene expression test to identify marker genes for a group of neighbourhoods of interest. 
# Before running this function you will need to define your neighbourhood groups depending on your biological question, that need to be stored as a NhoodGroup column in the da_results data.frame.

# In many cases, such as this example, DA neighbourhoods are found in different areas of the KNN graph, and grouping together all significant DA populations might not be ideal, as they might include cells of very different celltypes. 
# For this kind of scenario, we have implemented a neighbourhood function that uses community detection to partition neighbourhoods into groups on the basis of
# (1) the number of shared cells between 2 neighbourhoods; (2) the direction of fold-change for DA neighbourhoods; (3) the difference in fold change.

## Run buildNhoodGraph to store nhood adjacency matrix
HL.sce <- buildNhoodGraph(HL.sce)

# group DA neighbourhoods together using above metrics of shared cells, direction FC and difference FC. 


# test different metrics to get nice grouping of nhoods 
overlap.var <- 5
max.fc.delta.var <- 10

da_results_status_anno <- groupNhoods(HL.sce,
                                      da_results_status_anno, 
                                      da.fdr = 0.05,
                                      overlap = overlap.var,
                                      max.lfc.delta = max.fc.delta.var)


head(da_results_status_anno) # found 513 DA neighbourhoods at FDR 5%
length(unique(da_results_status_anno$NhoodGroup)) # number of groups




# plot beeswarm
plotDAbeeswarm(da_results_status_anno, group.by = "NhoodGroup") + 
  ggtitle(paste0("overlap = ", overlap.var, " max LFC delta = ", max.fc.delta.var))

dev.copy(pdf, paste0(output.dir, "Beeswarm_nhood_groups_DiseaseStatus.pdf"))
dev.off()

# unsure what is happening but looks like some nhood groups contain nhoods which are DA in different directions. this should not be occuring 


# plot on nhood groups on umap projection 
p1 <- plotNhoodGroups(HL.sce, da_results_status_anno, layout = "UMAP") + NoLegend()

nh_graph_status + p1 + patchwork::plot_layout(guides = "collect")

dev.copy(pdf, paste0(output.dir, "UMAP_nhood_groups_DiseaseStatus.pdf"))
dev.off()


# seurat clusters already offers a decent grouping of nhoods, perhaps use this or more refined clustering from seurat package to define nhood groups
plotDAbeeswarm(da_results_status_anno, "seurat_clusters")


# note, need at least 2 neighbourhoods assigned to each group for further processing 


##################################################
# finding gene signatures for neighbourhoods
##################################################

# restrict to only highly variable genes 

## Exclude zero counts genes
keep.rows <- rowSums(logcounts(HL.sce)) != 0
HL.sce <- HL.sce[keep.rows, ]

## Find HVGs
set.seed(101)
dec <- scran::modelGeneVar(HL.sce)
hvgs <- scran::getTopHVGs(dec, n = 2000)
head(hvgs)



# We run findNhoodGroupMarkers to test for one-vs-all differential gene expression for each neighbourhood group

# use seurat groups 
#da_results_status_anno$NhoodGroup <- da_results_status_anno$seurat_clusters

#da_results_status_anno <- da_results_status_anno %>%
# filter(NhoodGroup != "Mixed")

#da_results_status_anno$NhoodGroup <- droplevels(da_results_status_anno$NhoodGroup)

set.seed(42)

nhood_markers <- findNhoodGroupMarkers(HL.sce, 
                                       da_results_status_anno, 
                                       subset.row = hvgs, 
                                       aggregate.samples = TRUE, 
                                       sample_col = "SampleID")

head(nhood_markers)
table(da_results_status_anno$NhoodGroup, da_results_status_anno$celltype.l1)

# inspect degs of nhood group1
# which had many downregulated nhoods



## ANALYSIS WRITTEN UP TO HERE 









nhood_markers %>%
  filter(adj.P.Val_13 < 0.05) %>%
  arrange(desc(logFC_13)) %>%
  select(c(GeneID, logFC_13, adj.P.Val_13))

# visualise detexted markers 
# select markers for group 1 at FDR 1% and logFC > 1
table(nhood_markers$adj.P.Val_13 < 0.01) # 35 genes

ggplot(nhood_markers, aes(logFC_13, -log10(adj.P.Val_13))) + 
  geom_point(alpha=0.5, size=0.5) +
  geom_hline(yintercept = 3)


markers <- nhood_markers$GeneID[nhood_markers$adj.P.Val_13 < 0.01 & nhood_markers$logFC_13 > 0]


set.seed(42)
plotNhoodExpressionGroups(HL.sce, da_results_status_anno, features=intersect(rownames(HL.sce), markers[1:10]),
                          #subset.nhoods = da_results$NhoodGroup %in% c(1, 2, 3), 
                          scale = TRUE,
                          grid.space = "fixed")














#DGE testing within a group

# In some cases you might want to test for differential expression between cells in different conditions within the same neighbourhood group. You can do that using testDiffExp:

#dge_6 <- testDiffExp(embryo_milo, da_results, design = ~ stage, meta.data = data.frame(colData(embryo_milo)),
#                    subset.row = rownames(embryo_milo)[1:5], subset.nhoods=da_results$NhoodGroup=="6")

#dge_6


```


### Plot frequency stats
```{r freq_stats}

stat.vals <- scCustomize::Cluster_Stats_All_Samples(Nils.seurat$CD45Pos, group_by_var = "Treatment")

stat.vals



```



## DEG between treatment and vehicle {.tabset}
# write to iterate over clusters

### DEG analysis CD45Pos dataset treatment vs vehicle
```{r DEG_CD45Pos_treatment}

# create output dir
output.dir <- paste0("results/CD45Pos_DEG_analysis_treatment_vs_vehicle/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}



Nils.seurat$CD45Pos <- PrepSCTFindMarkers(Nils.seurat$CD45Pos)

# set assay
DefaultAssay(Nils.seurat$CD45Pos) <- "SCT"


# set idents
Idents(Nils.seurat$CD45Pos) <- "seurat_clusters"

Nils.seurat$CD45Pos@meta.data$Treatment_cluster <- paste0(Nils.seurat$CD45Pos@meta.data$Treatment, "_", Nils.seurat$CD45Pos@meta.data$seurat_clusters)
Idents(Nils.seurat$CD45Pos) <- "Treatment_cluster"
unique(Idents(Nils.seurat$CD45Pos))



Nils.seurat$CD45Pos@meta.data$Treatment_anno <- paste0(Nils.seurat$CD45Pos@meta.data$Treatment, "_", Nils.seurat$CD45Pos@meta.data$predicted.celltype.l2)
Idents(Nils.seurat$CD45Pos) <- "Treatment_anno"
unique(Idents(Nils.seurat$CD45Pos))




clust.id <- "CD14 Mono"
deg.list <- FindMarkers(Nils.seurat$CD45Pos,
                        ident.1 = paste0("PDS_", clust.id),
                        ident.2 = paste0("Vehicle_", clust.id),
                        assay = "SCT",
                        only.pos = TRUE, 
                        verbose = TRUE)



deg.list %>%
  filter(p_val_adj < 0.05)








# save to file 
#write.csv(CD45Pos.de, paste0(output.dir, "CD45Pos_DEGs.csv"))


# aggregate expression for heatmap visualisation

pseudo.cd45pos <- AggregateExpression(Nils.seurat$CD45Pos,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("Treatment_anno"))



goi <- deg.list %>%
  filter(p_val_adj < 0.05) %>%
  top_n(20, avg_log2FC) %>%
  rownames()
Idents(pseudo.cd45pos)
pseudo.cd45pos <- subset(pseudo.cd45pos, idents = c("PDS-CD14 Mono", "Vehicle-CD14 Mono"))
pseudo.cd45pos <- ScaleData(pseudo.cd45pos)

DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          raster = FALSE) + NoLegend()

#dev.copy(pdf, paste0(output.dir, "Heatmaptop_10_DEGs_pseudobulk.pdf"))
#dev.off()

VlnPlot_scCustom(Nils.seurat$CD45Pos,
                 features = c("ISG20"), 
                 group.by = "Treatment_anno", 
                 idents = c("PDS_CD14 Mono", "Vehicle_CD14 Mono"))




```













### CELL CYCLE
```{r remedy001}


# cell cycle scoring 
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes


HL.seurat <- CellCycleScoring(HL.seurat, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)

head(HL.seurat@meta.data)
scCustomize::FeaturePlot_scCustom(HL.seurat, feature = "S.Score", reduction = "umap.cca.sct")
scCustomize::FeaturePlot_scCustom(HL.seurat, feature = "G2M.Score", reduction = "umap.cca.sct")




```





