---
title: "CD45Pos_dataset_analysis"
author: "Dillon Corvino"
date: "25/01/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# Data 
# B16 - OVA 

# PDS = I.T - bind Gquadraplex
# day 4 post-treatment 


# Research questions 
# How does PDS induces an anti-tumor immune response?
# Known that after PDS treatment, certain DAMPs are induced, intratumoral chemokine levels go up drastically and subsequntly immune cells are recruited into the tumor.
# Main research questions are:
#1. Which cell subsets are producing the chemokines
#2. Are there differences between macrophages and DCs between PDS and Vehicle (because there are DAMPs that should activate them)
#3. How does the treatment effect the tumor cells (is it possible to discriminate them from other tissue cells, if not they express ovalbumin...this should make it at least possible to identify them.)




# key genes
# IL1b, IL6, IL12, IFNG, TNFA, 

# CCL2,3,4,5 
# CXCL1, CXCL9,10,13 





```


### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)


# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- TRUE
do.plot <- TRUE
quick.load <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())


library("Seurat")
library("SeuratObject")
library("SeuratData")
library("SeuratWrappers")
library("Azimuth")
library("dplyr")
library("scCustomize")
library("tibble")

########################################
# Parallelization of seurat functions
########################################

parallelization <- TRUE


if(parallelization){
  library(future)
  # check the current active plan
  plan()
  
  # detect cores
  ncores <- parallel::detectCores()
  ncores <- ncores-5
  # set plan
  plan("multisession", workers = ncores)
  plan()
  
  # set max limit for global variables
  # note this will drastically increase RAM usage, be aware of machine limits
  # set value is in bites therefore 1000*1024^2 = 1GB Therefore set to 10GBs 
  options(future.globals.maxSize = 10000 * 1024^2, 
          future.rng.onMisuse="ignore") 
}



if(quick.load){
  
  CD45pos.seurat <- LoadSeuratRds("saves/Nils_seurat_CD45Pos.rds")
  
  # set levels for easier downstream plotting 
  CD45pos.seurat@meta.data$Treatment <- factor(CD45pos.seurat@meta.data$Treatment, levels = c("Vehicle", "PDS"))
  
}
```




## Overview of dataset {.tabset}

### Broad overview of dataset
```{r broad_overview}

if(do.plot){
  
  
  # create output dir
  output.dir <- paste0("results/CD45Pos_Overview/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  vars.to.plot <- c("Treatment", "SampleID", "seurat_clusters")
  
  
  for(j in seq_along(vars.to.plot)){
    name.var <- gsub("\\.", "_", vars.to.plot[j])
    
    print(
      scCustomize::DimPlot_scCustom(CD45pos.seurat,
                                    reduction = "umap", 
                                    label = TRUE,
                                    figure_plot = TRUE,
                                    group.by = vars.to.plot[j]) 
    )
    
    
    dev.copy(pdf, paste0(output.dir, "CD45Pos_UMAP_", name.var, ".pdf"))
    dev.off()
    
  }
  
  
  # split umap plots
  print(
    scCustomize::DimPlot_scCustom(CD45pos.seurat,
                                  reduction = "umap", 
                                  label = TRUE,
                                  figure_plot = TRUE,
                                  group.by = "seurat_clusters", 
                                  split.by = "Treatment") 
  )
  
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_UMAP_seurat_clusters_split_by_treatment.pdf"))
  dev.off()
  
}


```

### Cluster stats
```{r cluster_stats}

# create output dir
output.dir <- paste0("results/CD45Pos_Overview/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

#############################################
# Number of cells per cluster
#############################################

cluster.n <- table(CD45pos.seurat@meta.data$seurat_clusters)

cluster.n <- as.data.frame(cluster.n)
cluster.n$Var1 <- factor(cluster.n$Var1, levels = cluster.n$Var1)


ggplot(cluster.n) +
  aes(x = Var1, y = Freq, fill = Var1) +
  geom_col() +
  scale_fill_viridis_d(option = "viridis", 
                       direction = -1) +
  labs(x = "Cluster ID", y = "Number of cells", title = "Number of cells per cluster") +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

ggsave(paste0(output.dir, "Barplot_number_of_cells_per_cluster.pdf"))


#############################################
# Freq of clusters across treatment 
#############################################

dist.table <- table(CD45pos.seurat@meta.data$seurat_clusters, CD45pos.seurat@meta.data$Treatment)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Vehicle", "PDS"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Treatment", y = "% of treatment", title = "Frequency of clusters per treatment") +
  scale_fill_manual(values = c(Vehicle = "grey", PDS = "darkred")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_clusters_per_treatment.pdf"))


```

### Cell Cycle scoring
```{r cell_cycle}

# create output dir
output.dir <- paste0("results/CD45Pos_cell_cycle/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

################################
# Get cell cycle genes 
################################
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

# pseudo-map to mouse gene (captures most except Mlf1ip, Fam64a and Hn1)
s.genes <- stringr::str_to_title(s.genes)
g2m.genes <- stringr::str_to_title(g2m.genes)

#####################
# score cell cycle
#####################

CD45pos.seurat <- CellCycleScoring(CD45pos.seurat,
                                   s.features = s.genes,
                                   g2m.features = g2m.genes,
                                   set.ident = TRUE)


#####################
# Visualise score
#####################

head(CD45pos.seurat@meta.data)

# Feature plots 

# S-score 
scCustomize::FeaturePlot_scCustom(CD45pos.seurat, 
                                  feature = "S.Score", 
                                  reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Pos_S_score.pdf"))
dev.off()

# G2M-score 
scCustomize::FeaturePlot_scCustom(CD45pos.seurat, 
                                  feature = "G2M.Score", 
                                  reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Pos_G2M_score.pdf"))
dev.off()

# Dim plot split by phase
scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                              group.by = "seurat_clusters",
                              split.by = "Phase",
                              label = TRUE,
                              reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_CD45Pos_cluster_by_phase.pdf"))
dev.off()


# Stacked VlnPlot
scCustomize::Stacked_VlnPlot(CD45pos.seurat,
                             features = c("S.Score", "G2M.Score"), 
                             group.by = "seurat_clusters")

dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Pos_cell_cycle.pdf"))
dev.off()




###################
# Quantification 
###################


# Frequency of phase per cluster
dist.table <- table(CD45pos.seurat@meta.data$seurat_clusters, CD45pos.seurat@meta.data$Phase)

cluster.total <- rowSums(dist.table)

freq.table <- sweep(dist.table, 1, cluster.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin
rowSums(freq.table)

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("G1", "G2M", "S"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Predicted phase", y = "% of cluster", title = "Frequency of phase per cluster") +
  scale_fill_manual(values = c(G1 = "lightblue", G2M = "orange", S = "red")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_phase_per_cluster.pdf"))


# Frequency of phase per treatment

dist.table <- table(CD45pos.seurat@meta.data$Treatment, CD45pos.seurat@meta.data$Phase)

treatment.total <- rowSums(dist.table)

freq.table <- sweep(dist.table, 1, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin
rowSums(freq.table)

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("G1", "G2M", "S"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Predicted phase", y = "% of treatment", title = "Frequency of phase per treatment") +
  scale_fill_manual(values = c(G1 = "lightblue", G2M = "orange", S = "red")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_phase_per_treatment.pdf"))



```



## DEG analysis {.tabset}

### Cluster-wise DEGs - calculate
```{r Clusterwise_DEGs_calculation}


# create output dir
output.dir <- paste0("results/CD45Pos_cluster_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  
  CD45pos.seurat <- PrepSCTFindMarkers(CD45pos.seurat)
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  # set idents
  Idents(CD45pos.seurat) <- "seurat_clusters"
  
  CD45Pos.de <- FindAllMarkers(CD45pos.seurat,
                               assay = "SCT",
                               only.pos = TRUE, 
                               verbose = TRUE)
  
  
  # save to file 
  write.csv(CD45Pos.de, paste0(output.dir, "CD45Pos_DEGs.csv"))
  
  
  
}else{
  
  # load DEG table
  CD45Pos.de <- read.csv(paste0(output.dir, "CD45Pos_DEGs.csv"))
  
}



```

### Cluster-wise DEGs - heatmaps
```{r clusterwise_DEGs_heatmaps}

if(do.plot){
  # aggregate expression for heatmap visualisation
  
  pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("seurat_clusters"))
  
  
  
  goi <- CD45Pos.de %>%
    filter(p_val_adj < 0.05) %>%
    group_by(cluster) %>%
    top_n(10, avg_log2FC) %>%
    pull(gene)
  
  
  DoHeatmap(pseudo.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_top_10_DEGs_pseudobulk.pdf"))
  dev.off()
  
  
  
  
  # iterate over the top degs for each cluster 
  cluster.id <- unique(CD45Pos.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- CD45Pos.de %>%
      filter(p_val_adj < 0.05) %>%
      filter(cluster == cluster.id[i]) %>%
      top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.cd45pos, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE) + 
        scico::scale_fill_scico(palette = "batlow", 
                                direction = 1, 
                                na.value = "white")
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
}

```


### Treatment-wise DEGs - calculate
```{r Treatmentwise_DEGs_calculation}

# create output dir
output.dir <- paste0("results/CD45Pos_treatment_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  
  CD45pos.seurat <- PrepSCTFindMarkers(CD45pos.seurat)
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  
  # add annotation for treatment 
  CD45pos.seurat@meta.data$Treatment_cluster <- paste0(CD45pos.seurat@meta.data$Treatment, "_", CD45pos.seurat@meta.data$seurat_clusters)
  
  # set Idents
  Idents(CD45pos.seurat) <- "Treatment_cluster"
  
  
  cluster.ids <- unique(CD45pos.seurat@meta.data$seurat_clusters)
  
  # create list to store output
  DEG.list <- list()
  
  # run through DEG specifically between vehicle and treatment for each cluster 
  
  for(i in seq_along(cluster.ids)){
    temp.de <- FindMarkers(CD45pos.seurat,
                           assay = "SCT",
                           ident.1 = paste0("PDS_", cluster.ids[i]),
                           ident.2 = paste0("Vehicle_", cluster.ids[i]),
                           only.pos = FALSE, 
                           recorrect_umi = FALSE,
                           verbose = TRUE)
    
    DEG.list[[i]] <- temp.de
    
    # save to file 
    write.csv(temp.de, paste0(output.dir, "DEGs_PDS_", cluster.ids[i], "_vs_Vehicle_", cluster.ids[i], ".csv"))
    
    
  }
  # add names to DEG.list
  names(DEG.list) <- cluster.ids
  
  # save DEG list to file 
  saveRDS(DEG.list, paste0(output.dir, "DEG_list_object.rds"))
  
  
}else{
  
  DEG.list <- readRDS(paste0(output.dir, "DEG_list_object.rds"))
  
}


```





### Treatment-wise DEGs - Volcano
```{r treatmentwise_DEGs_volcano}

# visualise treatment-wise DEGs as volcano plots


if(do.plot){
  
  for(i in 1:length(DEG.list)){
    
    FDR.theshold <- 0.01
    logFC.threshold <- 1
    
    # filter data
    input.data <- DEG.list[[i]] %>%
      dplyr::rename(logFC = avg_log2FC) %>%
      dplyr::rename(FDR = p_val_adj) %>%
      filter(FDR < FDR.theshold)
    
    # color points
    volcano.col <- usefulfunctions::colour.points.volcano(input.data,
                                                          increase.col = "Red",
                                                          decreased.col = "Blue",
                                                          FDR.cutoff = FDR.theshold, 
                                                          logFC.cutoff = logFC.threshold)
    # plot heatmap
    print(
      usefulfunctions::custom.enhanced.volcano(input.data, 
                                               pCutoff = FDR.theshold,
                                               FCcutoff = logFC.threshold,
                                               title = paste0(names(DEG.list)[i], " PDS vs Vehicle"),
                                               subtitle = paste0("adjPval threshold = ", FDR.theshold, " & log2FC threshold = ", logFC.threshold),
                                               colCustom = volcano.col)
    )
    dev.copy(pdf, paste0(output.dir, "VolcanoPlot_PDS_vs_Vehicle_cluster_", names(DEG.list)[i], ".pdf"))
    dev.off()
    
    
  }
  
}

```


## Cluster annotation {.tabset}

### SingleR cluster annotation
```{r SingleR_cluster_annotation}

# Notes:
# there is an error with collect() when trying to load celldex datasets
# i think its an error with biocmanager and has been fixed with a very recent update. > 3.16 
# but im running r 4.2.3 and this would require 4.3
# not keen to upgrade right now 
# so will try a work around of downgrading dbplyr to version 2.3.4 from 2.4 which im currently running 
# need to downgrade to this version 
#devtools::install_version("dbplyr", version = "2.3.4")
#devtools::install_version("dbplyr", version = "2.4")
# in theory if biocmanager is up to date, could just run below and should fix the problem
#library(BiocManager) # 3.16
#install("BiocFileCache")
# also:
# need to ensure matrixStats is at version 1.1.0 
#remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts
#BiocManager::install(version = "3.18")



if(long.compute){
  
  output.dir <- "results/CD45Pos_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # load required packages
  library("SingleR")
  library("celldex")
  
  
  ##################################
  # Using MouseRNAseqData
  ##################################
  
  ref.mouse <-  celldex::MouseRNAseqData()
  
  
  # Benayoun Lab that identified, downloaded and processed data sets on GEO that corresponded to sorted cell types (Benayoun et al., 2019).
  
  # The dataset contains 358 mouse RNA-seq samples annotated to 18 main cell types
  # further split into 28 subtypes under label.fine 
  
  
  ######################
  # Fine labels 
  ######################
  SingleR.pred <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45pos.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.fine)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
  #SingleR::plotDeltaDistribution(SingleR.pred)
  #SingleR::plotScoreDistribution(SingleR.pred)
  
  ######################
  # Main labels 
  ######################
  
  SingleR.pred <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45pos.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.main)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_main_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
  
  #####################
  # Use ImmGenData
  #####################
  # Info: Mouse, Microarray, 830 samples, 20 main labels, 253 fine lables, Hematopoietic & Immune focus
  
  ImmGen.data <- celldex::ImmGenData()
  
  
  # Fine labels
  SingleR.pred.fine <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                               ref = ImmGen.data, 
                               method = "cluster",
                               clusters = CD45pos.seurat@meta.data$seurat_clusters,
                               labels = ImmGen.data$label.fine)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred.fine, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_ImmGen_db.pdf"), plot = p1)
  
  
  # Main labels
  SingleR.pred.main <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                               ref = ImmGen.data, 
                               method = "cluster",
                               clusters = CD45pos.seurat@meta.data$seurat_clusters,
                               labels = ImmGen.data$label.main)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred.main, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_main_labels_ImmGen_db.pdf"), plot = p1)
  
  
}


```
 
### Transfer SingleR annotations to seurat object 
```{r transfer_singleR_annotations}


if(long.compute){
  
  output.dir <- "results/CD45Pos_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  

########################################
# transfer labels to seurat object 
########################################

table(predicted=SingleR.pred.main$labels, cluster=rownames(SingleR.pred.main))
table(predicted=SingleR.pred.fine$labels, cluster=rownames(SingleR.pred.fine))

# Main labels
main.labels <- SingleR.pred.main$labels
names(main.labels) <- rownames(SingleR.pred.main)

# Fine labels
fine.labels <- SingleR.pred.fine$labels
names(fine.labels) <- rownames(SingleR.pred.fine)

####################
# rename clusters 
####################

# Main labels
Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$seurat_clusters
CD45pos.seurat <- Rename_Clusters(CD45pos.seurat, new_idents = main.labels)
CD45pos.seurat@meta.data$SingleR_main <- Idents(CD45pos.seurat)

# Fine labels
Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$seurat_clusters
CD45pos.seurat <- Rename_Clusters(CD45pos.seurat, new_idents = fine.labels)
CD45pos.seurat@meta.data$SingleR_fine <- Idents(CD45pos.seurat)


############
# Visualise
############

# Main label
p.main <- scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                                        group.by = "SingleR_main", 
                                        label = TRUE,
                                        reduction = "umap")

print(p.main)

dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_main.pdf"))
dev.off()


# Fine label
p.fine <- scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                                        group.by = "SingleR_fine", 
                                        label = TRUE,
                                        reduction = "umap")

print(p.fine)

dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_fine.pdf"))
dev.off()


# side-by-side with cluster numbers
p.clusters <- scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                                            group.by = "seurat_clusters", 
                                            label = TRUE,
                                            reduction = "umap")

# clusters plus main
print(p.clusters | p.main)

dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_main_plus_clusters.pdf"))
dev.off()

# clusters plus fine
print(p.clusters | p.fine)

dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_fine_plus_clusters.pdf"))
dev.off()
  
}

```


# Neutrophil gene  heatmap
```{r remedy001}



```

## RQ1: Which CD45Neg clusters express chemokines or cytokines {.tabset}

### Chemokine and Cytokine expression visualisation 
```{r chemokine_cytokine_expression}

# create output dir
output.dir <- paste0("results/CD45Pos_chemokine_cytokines/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# aggregate expression for heatmap visualisation
pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("seurat_clusters"))


########################################
# List gene sets of interest to plot
########################################

cxc.genes <- grep("^Cxc.*", rownames(pseudo.cd45pos), value = T)
ccr.genes <- grep("^Ccr.*", rownames(pseudo.cd45pos), value = T)
il.genes <- grep("^Il.*", rownames(pseudo.cd45pos), value = T)
gzm.genes <- grep("^Gzm.*", rownames(pseudo.cd45pos), value = T)
ifn.genes <- grep("^Ifn.*", rownames(pseudo.cd45pos), value = T)
custom.genes <- c("Il1b", "Il6", "Il12a", "Il12b", "Ifng", "Tnf", "Ccl2", "Ccl3", "Ccl4", "Ccl5", "Cxcl1", "Cxcl9", "Cxcl10") # Cxcl13 not present in CD45Pos dataset

goi <- list(Cxcr_genes = cxc.genes, 
            Ccr_genes = ccr.genes,
            Interleukin_genes = il.genes,
            Gzm_genes = gzm.genes,
            Ifn_genes = ifn.genes,
            Custom_genes = custom.genes)


###################
# Plot heatmaps
###################

for(i in seq_along(names(goi))){
  
  print(
    DoHeatmap(pseudo.cd45pos, 
              features = goi[[i]], 
              draw.lines = FALSE, 
              angle = 90,
              raster = FALSE) + 
      scico::scale_fill_scico(palette = "batlow", 
                              direction = 1, 
                              na.value = "white")
  )
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_", names(goi)[i], ".pdf"))
  dev.off()
  
}


##############################
# Stacked vln plot
##############################

for(i in seq_along(names(goi))){
  
  if(length(goi[[i]]) <= 15){
    
    
    print(
      scCustomize::Stacked_VlnPlot(CD45pos.seurat, 
                                   features = goi[[i]],
                                   group.by = "seurat_clusters",
                                   pt.size = 0)
    )
    
    dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Pos_", names(goi)[i], ".pdf"))
    dev.off()
    
  }else{
    
    temp.goi <- goi[[i]]
    
    # Split the vector into multiple vectors
    split_indices <- seq(1, length(temp.goi), by = 15)
    split_list <- lapply(split_indices, function(i) temp.goi[i:min(i + 14, length(temp.goi))])
    
    
    for(j in seq_along(length(split_list))){
      
      print(
        scCustomize::Stacked_VlnPlot(CD45pos.seurat, 
                                     features = split_list[[j]],
                                     group.by = "seurat_clusters",
                                     pt.size = 0)
      )
      
      dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Pos_", names(goi)[i], "_", j, ".pdf"))
      dev.off()
      
    }
    
  }
  
}


```







# re-write below based on analysis layout in cd45neg dataset 




## Cluster annotation CD45Pos dataset
```{r CD45Pos_data_cluster_annotation}

if(long.compute){
  output.dir <- "results/CD45Pos_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # load required packages
  library("SingleR")
  library("celldex")
  
  ################################################################
  # Use a couple databased to get broad overview of clusters 
  ################################################################
  
  # error with collect() when trying to load celldex datasets
  
  # i think its an error with biocmanager and has been fixed with a very recent update. > 3.16 
  # but im running r 4.2.3 and this would require 4.3
  # not keen to upgrade right now 
  # so will try a work around of downgrading dbplyr to version 2.3.4 from 2.4 which im currently running 
  
  # need to downgrade to this version 
  #devtools::install_version("dbplyr", version = "2.3.4")
  #devtools::install_version("dbplyr", version = "2.4")
  
  # in theory if biocmanager is up to date, could just run below and should fix the problem
  
  #library(BiocManager) # 3.16
  #install("BiocFileCache")
  
  
  # need to ensure matrixStats is at version 1.1.0 
  #remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts
  #BiocManager::install(version = "3.18")
  
  #####################
  # Use ImmGenData
  #####################
  # Info: Mouse, Microarray, 830 samples, 20 main labels, 253 fine lables, Hematopoietic & Immune focus
  
  ImmGen.data <- celldex::ImmGenData()
  
  SingleR.pred <- SingleR(test = Nils.seurat$CD45Pos@assays$SCT$data, 
                          ref = ImmGen.data, 
                          method = "cluster",
                          clusters = Nils.seurat$CD45Pos@meta.data$seurat_clusters,
                          labels = ImmGen.data$label.fine)
  
  
  # Plot 
  plotScoreHeatmap(SingleR.pred, 
                   show_colnames = TRUE)
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_MouseRNAseqData_db.pdf"))
  dev.off()
  
  
  
  
  SingleR.pred <- SingleR(test = Nils.seurat$CD45Pos@assays$SCT$data, 
                          ref = ImmGen.data, 
                          method = "cluster",
                          clusters = Nils.seurat$CD45Pos@meta.data$seurat_clusters,
                          labels = ImmGen.data$label.main)
  
  
  # Plot 
  plotScoreHeatmap(SingleR.pred, 
                   show_colnames = TRUE)
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_cluster_annotations_main_labels_MouseRNAseqData_db.pdf"))
  dev.off()
  
}

```














## RQ1: Which cell subsets are producing the chemokines {.tabset}

### Vis chemokine/cytokine expression
```{r chemokine_cytokine_expression}

# create output dir
output.dir <- paste0("results/RQ1_which_celltypes_express_chemokines/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# aggregate expression for heatmap visualisation
pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("seurat_clusters"))





# CXCR chemokines
goi <- grep("^Cxc.*", rownames(pseudo.cd45pos), value = T)
goi


DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          angle = 90,
          raster = FALSE) + 
  scico::scale_fill_scico(palette = "batlow", 
                          direction = 1, 
                          na.value = "white")

dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_Cxc_chemokines.pdf"))
dev.off()


# CCR genes
goi <- grep("^Ccr.*", rownames(pseudo.cd45pos), value = T)
goi

DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          angle = 90,
          raster = FALSE) + 
  scico::scale_fill_scico(palette = "batlow", 
                          direction = 1, 
                          na.value = "white")



dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_Ccrs.pdf"))
dev.off()


# Interleukins
goi <- grep("^Il.*", rownames(pseudo.cd45pos), value = T)
goi

DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          angle = 90,
          raster = FALSE) + 
  scico::scale_fill_scico(palette = "batlow", 
                          direction = 1, 
                          na.value = "white")



dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_interleukins.pdf"))
dev.off()



# Granzymes
goi <- grep("^Gzm.*", rownames(pseudo.cd45pos), value = T)
goi

DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          angle = 90,
          raster = FALSE) + 
  scico::scale_fill_scico(palette = "batlow", 
                          direction = 1, 
                          na.value = "white")

dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_granzymes.pdf"))
dev.off()



# Interferons
goi <- grep("^Ifn.*", rownames(pseudo.cd45pos), value = T)
goi

DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          angle = 90,
          raster = FALSE) + 
  scico::scale_fill_scico(palette = "batlow", 
                          direction = 1, 
                          na.value = "white")


dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_interferons.pdf"))
dev.off()


#######################
# custom set of genes
#######################
# Nils provided this set of genes he is particularly interested in 

# key genes
# IL1b, IL6, IL12, IFNG, TNFA, 
# CCL2,3,4,5 
# CXCL1, CXCL9,10,13 

goi <- c("Il1b", "Il6", "Il12a", "Il12b", "Ifng", "Tnf", "Ccl2", "Ccl3", "Ccl4", "Ccl5", "Cxcl1", "Cxcl9", "Cxcl10", "Cxcl13")


DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          angle = 90,
          raster = FALSE) + 
  scico::scale_fill_scico(palette = "batlow", 
                          direction = 1, 
                          na.value = "white")


dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_custom_genes.pdf"))
dev.off()


```




## RQ2: Are there differences between macrophages and DCs between PDS and Vehicle (because there are DAMPs that should activate them) {.tabset}


```{r treatment_differences_in_DCs_Macrophages}


# create output dir
output.dir <- paste0("results/RQ2_Macrophage_DC_treatment_differences/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# Firstly are there abundance differences or polarisations torwards particular subsets of marophages or DCs between untreated and treated 

dist.table <- table(CD45pos.seurat@meta.data$seurat_clusters, CD45pos.seurat@meta.data$Treatment)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Vehicle", "PDS"))


library(ggplot2)

ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  scale_fill_hue(direction = 1) +
  theme_minimal() +
  facet_wrap(vars(Var1))

dev.copy(pdf, paste0(output.dir, "Barplot_freq_of_clusters_as_percentage_of_treatment.pdf"))
dev.off()      


# no replicates in dataset so cant perform any statistical analysis on this type of oberservation. 
# but some clusters do seem to be largely polarised/captured more in treatment or ctrl in comparison to the other. 
# notably, CD14_Mono_14, CD14_Mono_22, CD14_Mono_4 are enriched in Treatment dataset 

# where as cDC1_18, cDC2_11, Monocyte_7, and pDC_26 are slightly enriched in vehicle dataset

# largest polarisations seem to be Bcell_5 and Tcell_1 enrichment in vehicle dataset


# lets dig into the monocyte and DC populations, are there any genes different between vehicle and treatment 
# not sure this will be useful, expect not much differences. really should be comparing CD14_Mono_14 to other mono populations, as these are the enriched in treatment ones. 
# but will perform the analysis regardless to see what occurs 


# plot umaps split by treatment 
scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                              group.by = "seurat_clusters", 
                              split.by = "Treatment", 
                              label = TRUE,
                              reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_split_by_treatment.pdf"))
dev.off()


```






## cluster highlight plots
```{r remedy001}

scCustomize::Cluster_Highlight_Plot(CD45pos.seurat, cluster_name = 25, reduction = "umap")


```



# NOT WORKING - check all code 
# also not written again for this dataset specific analysis - check all code 



## NO LONGER USED
## annotate clusters {.tabset}
```{r annotate_clusters}
# 
# if(long.compute){
#   
# ########################################################
# # provide annotation for clusters where possible 
# ########################################################
# 
# 
# # create output dir
# output.dir <- paste0("results/CD45Pos_cluster_vs_azimuth/")
# 
# if(!dir.exists(output.dir)){
#   dir.create(output.dir, 
#              recursive = T)
#   
# }
# 
# 
# #############################
# # cluster by annotation 
# #############################
# 
# # large grain annotation
# 
# # broad annotation 
# ## look at all clusters again across annotation. as some clusters maybe can be called X celltype where they are say 70% Xcelltype and rest is made up of small annotations 
# # as opposed to clusters 50/50 split between annotations 
# 
# 
# # calculate the frequency of each cluster that has an azimuth annotation, therefore the clusters which are primarily 1 annotation can easily be defined 
# x <- table(CD45pos.seurat@meta.data$predicted.celltype.l1, CD45pos.seurat@meta.data$seurat_clusters)
# cluster.totals <- colSums(x)
# y <- sweep(x, 2, cluster.totals, FUN = "/")
# y <- y*100
# y <- as.data.frame(y)
# 
# 
# cluster.id <- unique(CD45pos.seurat@meta.data$seurat_clusters)
# 
# for(i in seq_along(cluster.id)){
#   
#   y.filt <- y %>%
#     filter(Var2 == cluster.id[i])
#   
#   
#   library(ggplot2)
#   
#   p1 <- ggplot(y.filt) +
#     aes(x = Var1, y = Freq, fill = Var1) +
#     geom_col() +
#     scale_fill_hue(direction = 1) +
#     labs(title = paste0("Cluster ", cluster.id[i])) +
#            coord_flip() +
#            theme_minimal() +
#            theme(legend.position = "none") +
#            ylim(0, 100)
#          
#          print(p1)
#          
#          dev.copy(pdf, paste0(output.dir, "Cluster_", cluster.id[i], "_broad_azimuth_annotation_freq.pdf"))
#          dev.off()
#          
# }
# 
# 
# #############################
# # medium grain annotation 
# #############################
# 
# ## look at all clusters again across annotation. as some clusters maybe can be called X celltype where they are say 70% Xcelltype and rest is made up of small annotations 
# # as opposed to clusters 50/50 split between annotations 
# 
# 
# # calculate the frequency of each cluster that has an azimuth annotation, therefore the clusters which are primarily 1 annotation can easily be defined 
# x <- table(CD45pos.seurat@meta.data$predicted.celltype.l2, CD45pos.seurat@meta.data$seurat_clusters)
# cluster.totals <- colSums(x)
# y <- sweep(x, 2, cluster.totals, FUN = "/")
# y <- y*100
# y <- as.data.frame(y)
# 
# 
# cluster.id <- unique(CD45pos.seurat@meta.data$seurat_clusters)
# 
# for(i in seq_along(cluster.id)){
#   
#   y.filt <- y %>%
#     filter(Var2 == cluster.id[i])
#   
#   
#   library(ggplot2)
#   
#   p1 <- ggplot(y.filt) +
#     aes(x = Var1, y = Freq, fill = Var1) +
#     geom_col() +
#     scale_fill_hue(direction = 1) +
#     labs(title = paste0("Cluster ", cluster.id[i])) +
#            coord_flip() +
#            theme_minimal() +
#            theme(legend.position = "none") +
#            ylim(0, 100)
#          
#          print(p1)
#          
#          dev.copy(pdf, paste0(output.dir, "Cluster_", cluster.id[i], "_medium_grain_azimuth_annotation_freq.pdf"))
#          dev.off()
#          
# }
# 
# #############################
# # fine grain annotation 
# #############################
# 
# ## look at all clusters again across annotation. as some clusters maybe can be called X celltype where they are say 70% Xcelltype and rest is made up of small annotations 
# # as opposed to clusters 50/50 split between annotations 
# 
# 
# # calculate the frequency of each cluster that has an azimuth annotation, therefore the clusters which are primarily 1 annotation can easily be defined 
# x <- table(CD45pos.seurat@meta.data$predicted.celltype.l3, CD45pos.seurat@meta.data$seurat_clusters)
# cluster.totals <- colSums(x)
# y <- sweep(x, 2, cluster.totals, FUN = "/")
# y <- y*100
# y <- as.data.frame(y)
# 
# 
# cluster.id <- unique(CD45pos.seurat@meta.data$seurat_clusters)
# 
# for(i in seq_along(cluster.id)){
#   
#   y.filt <- y %>%
#     filter(Var2 == cluster.id[i])
#   
#   
#   library(ggplot2)
#   
#   p1 <- ggplot(y.filt) +
#     aes(x = Var1, y = Freq, fill = Var1) +
#     geom_col() +
#     scale_fill_hue(direction = 1) +
#     labs(title = paste0("Cluster ", cluster.id[i])) +
#            coord_flip() +
#            theme_minimal() +
#            theme(legend.position = "none") +
#            ylim(0, 100)
#          
#          print(p1)
#          
#          dev.copy(pdf, paste0(output.dir, "Cluster_", cluster.id[i], "_fine_grain_azimuth_annotation_freq.pdf"))
#          dev.off()
#          
# }
# 
# 
# 
# 
# 
# 
# ########################################################
# # Use gene expression profiles to investigate clusters
# ########################################################
# 
# 
# # aggregate expression for heatmap visualisation
# 
# pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
#                                       assays = "SCT",
#                                       return.seurat = TRUE, 
#                                       group.by = c("seurat_clusters"))
# 
# 
# 
# # load deg data
# CD45Pos.de <- read.csv("results/CD45Pos_DEG_analysis/CD45Pos_DEGs.csv")
# 
# 
# 
# # some clusters were easily annotated based on overlap and call by azimuth reference 
# # some are ambiguous 
# # inspect gene expression profiles for these ambiguous clusters
# 
# # this section was done interactively comparing previous plots with visualising certain genes and DEG for each cluster
# 
# grep("FCGR", rownames(pseudo.cd45pos), value = T)
# 
# 
# goi <- CD45Pos.de %>%
#   filter(p_val_adj < 0.05) %>%
#   group_by(cluster) %>%
#   filter(cluster == 10) %>%
#   top_n(50, avg_log2FC) %>%
#   pull(gene)
# 
# goi
#   
#   DoHeatmap(pseudo.cd45pos, 
#           features = goi, 
#           draw.lines = FALSE, 
#           raster = FALSE)
# 
# 
# scCustomize::Cluster_Highlight_Plot(CD45pos.seurat, cluster_name = 25, reduction = "umap")
# 
# VlnPlot_scCustom(CD45pos.seurat, features = "IL2RA", pt.size = 0.1)
# 
# 
# 
# 
# ######################################################
# # using all above plots annotation definitions are 
# ######################################################
# 
# # broad annotations
# # cluster annotations 
# myeloid.clusters <- c(0)
# mono.clusters <- c(2, 4, 6, 7, 9, 12, 13, 14, 15, 17, 19, 20, 21, 22, 24)
# dc.clusters <- c(11, 18, 26)
# 
# bcell.clusters <- c(5)
# erythrocyte.clusters <- c(27)
# 
# t.cell.clusters <- c(1, 3, 23)
# nk.clusters <- c(8, 16)
# nk.or.t.cell.clusters <- c(10, 25) # as in ambig if nk or t or nkt
# 
# 
# # fine grain annotations
# cd14.mono.clusters <- c(2, 4, 6, 9, 14, 20, 22)
# cd16.mono.clusters <- c(13, 17)
# 
# pDC.clusters <- c(26)
# cDC1.clusters <- c(18)
# cDC2.clusters <- c(11)
# 
# cd8.clusters <- c(3)
# cd4.clusters <- c(23)
# 
# 
# 
# 
# #
# # Cluster 23 might be treg cd4
# 
# 
# ############################################################
# # Define clusters based on above data and interpretation
# ############################################################
# 
# 
# CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
#   dplyr::mutate(Anno_l1 = case_when(seurat_clusters %in% myeloid.clusters ~ paste0("Myeloid_", seurat_clusters), 
#                                     seurat_clusters %in% mono.clusters ~ paste0("Monocyte_", seurat_clusters), 
#                                     seurat_clusters %in% dc.clusters ~ paste0("DC_", seurat_clusters), 
#                                     seurat_clusters %in% bcell.clusters ~ paste0("Bcell_", seurat_clusters), 
#                                     seurat_clusters %in% erythrocyte.clusters ~ paste0("Erythrocyte_", seurat_clusters), 
#                                     seurat_clusters %in% t.cell.clusters ~ paste0("Tcell_", seurat_clusters), 
#                                     seurat_clusters %in% nk.clusters ~ paste0("NK_", seurat_clusters), 
#                                     seurat_clusters %in% nk.or.t.cell.clusters ~ paste0("NKorTcell_", seurat_clusters), 
#                                     TRUE ~ seurat_clusters), 
#                 Anno_l2 = case_when(seurat_clusters %in% cd14.mono.clusters ~ paste0("CD14_Mono_", seurat_clusters),
#                                     seurat_clusters %in% cd16.mono.clusters ~ paste0("CD16_Mono_", seurat_clusters),
#                                     seurat_clusters %in% pDC.clusters ~ paste0("pDC_", seurat_clusters),
#                                     seurat_clusters %in% cDC1.clusters ~ paste0("cDC1_", seurat_clusters),
#                                     seurat_clusters %in% cDC2.clusters ~ paste0("cDC2_", seurat_clusters),
#                                     seurat_clusters %in% cd8.clusters ~ paste0("CD8T_", seurat_clusters),
#                                     seurat_clusters %in% cd4.clusters ~ paste0("CD4T_", seurat_clusters),
#                                     TRUE ~ Anno_l1))
# 
# 
# ##############################
# # visualise annotations 
# ##############################
# 
# # create output dir
# output.dir <- paste0("results/CD45Pos_cluster_annotations/")
# 
# if(!dir.exists(output.dir)){
#   dir.create(output.dir, 
#              recursive = T)
#   
# }
# 
# 
# DimPlot_scCustom(CD45pos.seurat, 
#                  label = TRUE,
#                  group.by = "Anno_l1", 
#                  reduction = "umap") + NoLegend()
# 
# dev.copy(pdf, paste0(output.dir, "UMAP_clusters_annotated_l1.pdf"))
# dev.off()
# 
# 
# DimPlot_scCustom(CD45pos.seurat, 
#                  label = TRUE,
#                  group.by = "Anno_l2", 
#                  reduction = "umap") + NoLegend()
# 
# dev.copy(pdf, paste0(output.dir, "UMAP_clusters_annotated_l2.pdf"))
# dev.off()
# 
# 
# 
# # save seurat object after annotation 
# SaveSeuratRds(CD45pos.seurat, 
#               file = "saves/Nils_seurat_CD45Pos.rds")
# }




# for insight into markers can use the below resource 
# load cell marker info from CellMarker2.0 (http://bio-bigdata.hrbmu.edu.cn/CellMarker/index.html)
# cellmarkers <- readxl::read_xlsx("data/Misc/Cell_marker_All.xlsx")
# 
# unique(cellmarkers$tissue_class)
# 
# # filter
# mouse.normal <- cellmarkers %>%
#   filter(species == "Mouse") %>%
#   filter(cell_type == "Normal cell")
# 
# unique(mouse.normal$cell_name)
# 
# 
# Tcell.markers <- mouse.normal %>%
#   filter(grepl("Natural killer", cell_name)) %>%
#   distinct(Symbol)
# 
# print(Tcell.markers, n = 100)


```






