---
title: "CD45Pos_dataset_analysis"
author: "Dillon Corvino"
date: "25/01/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# Data 
# B16 - OVA 

# PDS = I.T - bind Gquadraplex
# day 4 post-treatment 
# day 9 post-treatment

# Research questions 
# How does PDS induces an anti-tumor immune response?
# Known that after PDS treatment, certain DAMPs are induced, intratumoral chemokine levels go up drastically and subsequntly immune cells are recruited into the tumor.
# Main research questions are:
#1. Which cell subsets are producing the chemokines
#2. Are there differences between macrophages and DCs between PDS and Vehicle (because there are DAMPs that should activate them)
#3. How does the treatment effect the tumor cells (is it possible to discriminate them from other tissue cells, if not they express ovalbumin...this should make it at least possible to identify them.)




# key genes
# IL1b, IL6, IL12, IFNG, TNFA, 

# CCL2,3,4,5 
# CXCL1, CXCL9,10,13 





```


### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)


# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- FALSE
do.plot <- TRUE
quick.load <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())


library("Seurat")
library("SeuratObject")
library("SeuratData")
library("SeuratWrappers")
library("dplyr")
library("scCustomize")
library("tibble")


# load output from preprocessing script
if(long.compute){
  
  CD45pos.seurat <- LoadSeuratRds("saves/Preprocessed_CD45Pos.rds")
  
  # set levels for easier downstream plotting 
  CD45pos.seurat@meta.data$Treatment <- factor(CD45pos.seurat@meta.data$Treatment, levels = c("Vehicle", "PDS"))
  CD45pos.seurat@meta.data$Time <- factor(CD45pos.seurat@meta.data$Time, levels = c("Day4", "Day9"))
  CD45pos.seurat@meta.data$SampleID <- factor(CD45pos.seurat@meta.data$SampleID, levels = c("Day4_Vehicle_CD45Pos", "Day4_PDS_CD45Pos", "Day9_Vehicle_CD45Pos", "Day9_PDS_CD45Pos"))
  
}else if(quick.load){
  
  # load cleaned dataset 
  CD45pos.seurat <- LoadSeuratRds("saves/CD45pos_seurat.rds")
  
  
}


```



## Cluster annotation {.tabset}

### SingleR cluster annotation
```{r SingleR_cluster_annotation}

# Notes:
# there is an error with collect() when trying to load celldex datasets
# i think its an error with biocmanager and has been fixed with a very recent update. > 3.16 
# but im running r 4.2.3 and this would require 4.3
# not keen to upgrade right now 
# so will try a work around of downgrading dbplyr to version 2.3.4 from 2.4 which im currently running 
# need to downgrade to this version 
#devtools::install_version("dbplyr", version = "2.3.4")
#devtools::install_version("dbplyr", version = "2.4")
# in theory if biocmanager is up to date, could just run below and should fix the problem
#library(BiocManager) # 3.16
#install("BiocFileCache")
# also:
# need to ensure matrixStats is at version 1.1.0 
#remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts
#BiocManager::install(version = "3.18")



if(long.compute){
  
  output.dir <- "results/CD45Pos_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # load required packages
  library("SingleR")
  library("celldex")
  
  
  ##################################
  # Using MouseRNAseqData
  ##################################
  
  ref.mouse <-  celldex::MouseRNAseqData()
  
  
  # Benayoun Lab that identified, downloaded and processed data sets on GEO that corresponded to sorted cell types (Benayoun et al., 2019).
  
  # The dataset contains 358 mouse RNA-seq samples annotated to 18 main cell types
  # further split into 28 subtypes under label.fine 
  
  
  ######################
  # Fine labels 
  ######################
  SingleR.pred <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45pos.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.fine)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
  #SingleR::plotDeltaDistribution(SingleR.pred)
  #SingleR::plotScoreDistribution(SingleR.pred)
  
  ######################
  # Main labels 
  ######################
  
  SingleR.pred <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45pos.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.main)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_main_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
  
  #####################
  # Use ImmGenData
  #####################
  # Info: Mouse, Microarray, 830 samples, 20 main labels, 253 fine lables, Hematopoietic & Immune focus
  
  ImmGen.data <- celldex::ImmGenData()
  
  
  # Fine labels
  SingleR.pred.fine <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                               ref = ImmGen.data, 
                               method = "cluster",
                               clusters = CD45pos.seurat@meta.data$seurat_clusters,
                               labels = ImmGen.data$label.fine)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred.fine, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_ImmGen_db.pdf"), plot = p1)
  
  
  # Main labels
  SingleR.pred.main <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                               ref = ImmGen.data, 
                               method = "cluster",
                               clusters = CD45pos.seurat@meta.data$seurat_clusters,
                               labels = ImmGen.data$label.main)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred.main, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_main_labels_ImmGen_db.pdf"), plot = p1)
  
  
}


```


### Transfer SingleR annotations to seurat object 
```{r transfer_singleR_annotations}


if(long.compute){
  
  output.dir <- "results/CD45Pos_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  
  ########################################
  # transfer labels to seurat object 
  ########################################
  
  table(predicted=SingleR.pred.main$labels, cluster=rownames(SingleR.pred.main))
  table(predicted=SingleR.pred.fine$labels, cluster=rownames(SingleR.pred.fine))
  
  # Main labels
  main.labels <- SingleR.pred.main$labels
  names(main.labels) <- rownames(SingleR.pred.main)
  
  # Fine labels
  fine.labels <- SingleR.pred.fine$labels
  names(fine.labels) <- rownames(SingleR.pred.fine)
  
  ####################
  # rename clusters 
  ####################
  
  # Main labels
  Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$seurat_clusters
  CD45pos.seurat <- Rename_Clusters(CD45pos.seurat, new_idents = main.labels)
  CD45pos.seurat@meta.data$SingleR_main <- Idents(CD45pos.seurat)
  
  # Fine labels
  Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$seurat_clusters
  CD45pos.seurat <- Rename_Clusters(CD45pos.seurat, new_idents = fine.labels)
  CD45pos.seurat@meta.data$SingleR_fine <- Idents(CD45pos.seurat)
  
  
  ############
  # Visualise
  ############
  
  # Main label
  p.main <- scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                                          group.by = "SingleR_main", 
                                          label = TRUE,
                                          reduction = "umap")
  
  print(p.main)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_main.pdf"))
  dev.off()
  
  
  # Fine label
  p.fine <- scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                                          group.by = "SingleR_fine", 
                                          label = TRUE,
                                          reduction = "umap")
  
  print(p.fine)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_fine.pdf"))
  dev.off()
  
  
  # side-by-side with cluster numbers
  p.clusters <- scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                                              group.by = "seurat_clusters", 
                                              label = TRUE,
                                              reduction = "umap")
  
  # clusters plus main
  print(p.clusters | p.main)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_main_plus_clusters.pdf"))
  dev.off()
  
  # clusters plus fine
  print(p.clusters | p.fine)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_fine_plus_clusters.pdf"))
  dev.off()
  
  
}










```


### Verify cluster annotation
```{r verify_and_finalise_cluster_annotation}


if(long.compute){
  
  output.dir <- "results/CD45Pos_SingleR_annotation_validation/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # overview of data 
  table(CD45pos.seurat@meta.data$seurat_clusters, CD45pos.seurat@meta.data$SingleR_fine)
  
  table(CD45pos.seurat@meta.data$seurat_clusters, CD45pos.seurat@meta.data$SingleR_main)
  table(CD45pos.seurat@meta.data$SingleR_main, CD45pos.seurat@meta.data$SingleR_fine)
  
  # inconsistency between main and fine annotation with respect to T-cell populations
  # Tcells in main are annotated as NKT
  # NKT in main are annotated as treg 
  
  
  # also seurat clusters nicely line up with singleR fine annotations, i.e one cluster per annotation 
  # only monocytes is with higher resolution using seurat clusters, therefore consider seperating into monocytes cluster X and cluster Y
  
  
  # look at annotations and simplify 
  
  
  #################################################
  # calculate DEGs using fine level resolution 
  #################################################
  
  CD45pos.seurat <- PrepSCTFindMarkers(CD45pos.seurat)
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  # set idents
  Idents(CD45pos.seurat) <- "SingleR_fine"
  
  CD45Pos.de <- FindAllMarkers(CD45pos.seurat,
                               assay = "SCT",
                               only.pos = TRUE, 
                               verbose = TRUE)
  
  
  # save to file 
  write.csv(CD45Pos.de, paste0(output.dir, "CD45Pos_DEGs_singleR_fine.csv"))
  
  
  ##############
  # plot DEGs
  ##############
  
  pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("SingleR_fine"))
  
  
  
  goi <- CD45Pos.de %>%
    filter(p_val_adj < 0.05) %>%
    group_by(cluster) %>%
    top_n(10, avg_log2FC) %>%
    pull(gene)
  
  
  DoHeatmap(pseudo.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_top_10_singleR_fine_DEGs_pseudobulk.pdf"))
  dev.off()
  
  # populations are very cleanly resolved 
  
  
  
  
  # iterate over the top degs for each cluster 
  cluster.id <- unique(CD45Pos.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- CD45Pos.de %>%
      filter(p_val_adj < 0.05) %>%
      filter(cluster == cluster.id[i]) %>%
      top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.cd45pos, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE) + 
        scico::scale_fill_scico(palette = "batlow", 
                                direction = 1, 
                                na.value = "white")
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
  
  # happy with using fine level annotation 
  # except, defining NKT from T-cells is a challenge. will just call them T-cells as at main level annotation these are T-cells
  # NKTs at fine level annotation are split into Treg and Teff - and these look to make sense
  # will therefore use fine level annotation fo all, but change NKT to just T-cell 
  # will also simplify the annotation for all clusters but keep their distinct'ness 
  
  
}






```


### Clean up seurat object metadata and export
```{r clean_seurat_object_metadata}

if(long.compute){
  # reannotate singleR_fine cluster annotations based on above analysis 
  # namely, simplify the annotations 
  # also, convert NKT to just T-cells
  
  CD45pos.seurat <- backup.object.for.now
  
  # note the clusters are essentially one cell type one cluster. except for macrophages which is present in two clusters of macrophages and monocytes present in 3 clusters
  table(CD45pos.seurat@meta.data$SingleR_fine, CD45pos.seurat@meta.data$seurat_clusters)
  
  library(stringr) # For using the `str_extract` function
  
  
  # simplify the annotations, to just the cell_type, but keep cluster_seperation 
  CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
    # Annotate clusters based on pattern matching
    mutate(cluster_anno = case_when(
      # Use `str_extract` to extract the cell type from SingleR_fine
      TRUE ~ paste0(
        str_extract(SingleR_fine, "^[^(]+"), # Extract the cell type name
        "_", seurat_clusters
      )
    )) %>%
    # Remove any spaces in the cluster_anno name
    mutate(cluster_anno = str_replace_all(cluster_anno, " ", ""))
  
  unique(CD45pos.seurat@meta.data$cluster_anno)
  table(CD45pos.seurat@meta.data$SingleR_fine, CD45pos.seurat@meta.data$cluster_anno)
  
  
  
  # simplify the annotations, so there are numbered within cell_type
  
  
  
  # little code to count up how many distinct cell types there are 
CD45pos.seurat@meta.data %>%
    dplyr::select(c(cluster_anno, SingleR_fine)) %>%
         # Extract cell_type from cluster_anno
     mutate(cell_type = str_extract(cluster_anno, "^[^_]+")) %>%
     # Group by cell_type
     group_by(cell_type) %>%
     # Calculate the number of unique annotations for each cell_type and rank them
    mutate(
         # Calculate the count of unique cluster_anno
         unique_count = n_distinct(cluster_anno) ) %>%
    ungroup() %>%
    distinct(cluster_anno, .keep_all = TRUE)



CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
  # Extract cell_type from cluster_anno
  mutate(cell_type = str_extract(cluster_anno, "^[^_]+")) %>%
  mutate(cluster_anno_new = case_when(
    cluster_anno == "Macrophages_1" ~ "Macrophages_1", 
    cluster_anno == "Macrophages_9" ~ "Macrophages_2", 
    
    cluster_anno == "Bcells_8" ~ "B_cells", 
    
    cluster_anno == "DC_15" ~ "pDC", 
    cluster_anno == "DC_14" ~ "DC_1", 
    cluster_anno == "DC_12" ~ "DC_2",
    
    cluster_anno == "Monocytes_4" ~ "Monocytes_1", 
    cluster_anno == "Monocytes_5" ~ "Monocytes_2", 
    cluster_anno == "Monocytes_7" ~ "Monocytes_3", 
    
    cluster_anno == "Macrophages_1" ~ "Macrophages_1", 
    cluster_anno == "Macrophages_9" ~ "Macrophages_2", 
    
    cluster_anno == "Neutrophils_3" ~ "Neutrophils_1", 
    cluster_anno == "Neutrophils_6" ~ "Neutrophils_2", 
    
    cluster_anno == "Stromalcells_16" ~ "Stromal_cells", 
    
    cluster_anno == "NKcells_2" ~ "NK_cells", 
    
    cluster_anno == "Basophils_17" ~ "Basophils", 
    
    cluster_anno == "Tcells_0" ~ "T_cells_1", 
    cluster_anno == "Tcells_10" ~ "T_cells_2", 
    cluster_anno == "Tcells_13" ~ "T_cells_3", 
    cluster_anno == "Tcells_11" ~ "Tregs", 
    TRUE ~ "ERROR")) %>%
  mutate(cluster_anno = cluster_anno_new) %>%
  select(-cluster_anno_new)
   
  
  unique(CD45pos.seurat@meta.data$cluster_anno)
    unique(CD45pos.seurat@meta.data$SingleR_fine)

  
  # clean up object, remove unnecessary columns keep only what is needed
  CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
    select(c("orig.ident", "nCount_RNA", "nFeature_RNA", "nCount_SCT", "nFeature_SCT", "percent_mito", "percent_ribo", "percent_mito_ribo", "log10GenesPerUMI",  "SampleID", "Time", "Compartment", "Treatment",  "cluster_anno", "seurat_clusters", "SCT_snn_res.0.4", "SingleR_main", "SingleR_fine", "cell_type"))
  
  
  # Display the first few rows of the updated data frame to verify the results
  head(as.data.frame(CD45pos.seurat@meta.data[1:10, ]))
  
  
  
  table(CD45pos.seurat@meta.data$SingleR_fine, CD45pos.seurat@meta.data$cell_type)
  table(CD45pos.seurat@meta.data$SingleR_fine, CD45pos.seurat@meta.data$cluster_anno)
  table(CD45pos.seurat@meta.data$cell_type, CD45pos.seurat@meta.data$cluster_anno)
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  # clean up sample ID annotation - no need to say CD45Pos still 
  
  CD45pos.seurat@meta.data$SampleID <- gsub("_CD45Pos", "", CD45pos.seurat@meta.data$SampleID)
  
  unique(CD45pos.seurat@meta.data$cluster_anno)
  head(CD45pos.seurat@meta.data)
  
  # set factorisation 
  CD45pos.seurat@meta.data$Treatment <- factor(CD45pos.seurat@meta.data$Treatment, 
                                               levels = c("Vehicle", "PDS"))
  
  CD45pos.seurat@meta.data$Time <- factor(CD45pos.seurat@meta.data$Time, 
                                          levels = c("Day4", "Day9"))
  
  CD45pos.seurat@meta.data$SampleID <- factor(CD45pos.seurat@meta.data$SampleID, 
                                              levels = c("Day4_Vehicle", "Day9_Vehicle", "Day4_PDS", "Day9_PDS"))
  
  
  CD45pos.seurat@meta.data$cluster_anno <- factor(CD45pos.seurat@meta.data$cluster_anno)
  
  # save output
  SaveSeuratRds(CD45pos.seurat, file = "saves/CD45pos_seurat.rds")
}


```



## Overview of dataset {.tabset}

### Broad overview of dataset
```{r broad_overview}

if(do.plot){
  
  
  # create output dir
  output.dir <- paste0("results/CD45Pos_Overview/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  vars.to.plot <- c("Treatment", "SampleID", "cluster_anno", "Time", "cell_type")
  
  
  for(j in seq_along(vars.to.plot)){
    name.var <- gsub("\\.", "_", vars.to.plot[j])
    
    print(
      scCustomize::DimPlot_scCustom(CD45pos.seurat,
                                    reduction = "umap", 
                                    label = TRUE,
                                    figure_plot = TRUE,
                                    group.by = vars.to.plot[j]) 
    )
    
    
    dev.copy(pdf, paste0(output.dir, "CD45Pos_UMAP_", name.var, ".pdf"))
    dev.off()
    
  }
  
  
  # for split comparison plots, downsample to make visual comparisons more fair 
  table(CD45pos.seurat@meta.data$Treatment)
  Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$Treatment
  temp.seurat <- subset(CD45pos.seurat, downsample = 15000)
  # split umap plots
  print(
    scCustomize::DimPlot_scCustom(temp.seurat,
                                  reduction = "umap", 
                                  label = TRUE,
                                  figure_plot = TRUE,
                                  group.by = "cluster_anno", 
                                  split.by = "Treatment") 
  )
  
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_UMAP_cluster_anno_split_by_treatment.pdf"))
  dev.off()
  
  
  table(CD45pos.seurat@meta.data$SampleID)
  Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$SampleID
  temp.seurat <- subset(CD45pos.seurat, downsample = 5000)
  # split umap plots
  print(
    scCustomize::DimPlot_scCustom(temp.seurat,
                                  reduction = "umap", 
                                  label = FALSE,
                                  figure_plot = TRUE,
                                  group.by = "cluster_anno", 
                                  split.by = "SampleID") 
  )
  
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_UMAP_cluster_anno_split_by_SampleID.pdf"))
  dev.off()
  
  rm(temp.seurat)
  
}



```


### Cluster stats
```{r cluster_stats}

# create output dir
output.dir <- paste0("results/CD45Pos_Overview/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

#############################################
# Number of cells per cluster
#############################################

cluster.n <- table(CD45pos.seurat@meta.data$cluster_anno)

cluster.n <- as.data.frame(cluster.n)

cluster.n$Var1 <- factor(cluster.n$Var1, levels = cluster.n[order(cluster.n$Freq, decreasing = T), ]$Var1)


ggplot(cluster.n) +
  aes(x = Var1, y = Freq, fill = Var1) +
  geom_col() +
  scale_fill_viridis_d(option = "viridis", 
                       direction = -1) +
  labs(x = "Cluster", y = "Number of cells", title = "Number of cells per cluster") +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

ggsave(paste0(output.dir, "Barplot_number_of_cells_per_cluster.pdf"))



#############################################
# Freq of cluster across treatment 
#############################################

dist.table <- table(CD45pos.seurat@meta.data$cluster_anno, CD45pos.seurat@meta.data$Treatment)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Vehicle", "PDS"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Treatment", y = "% of treatment", title = "Frequency of cluster per treatment") +
  scale_fill_manual(values = c(Vehicle = "grey", PDS = "darkred")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_cluster_per_treatment.pdf"))


#############################################
# Freq of clusters per treatment across time
#############################################


#######################################
#  Ordered by treatment
#######################################

dist.table <- table(CD45pos.seurat@meta.data$cluster_anno, CD45pos.seurat@meta.data$SampleID)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Day4_Vehicle", "Day9_Vehicle", "Day4_PDS", "Day9_PDS"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Treatment", y = "% of treatment", title = "Frequency of cluster per treatment") +
  scale_fill_manual(values = c(Day4_Vehicle = "lightgreen", Day9_Vehicle = "darkgreen", Day4_PDS = "red", Day9_PDS = "darkred")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_cluster_per_treatment_across_time.pdf"))



#######################################
#  Ordered by time
#######################################

dist.table <- table(CD45pos.seurat@meta.data$cluster_anno, CD45pos.seurat@meta.data$SampleID)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Day4_Vehicle", "Day4_PDS", "Day9_Vehicle", "Day9_PDS"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Treatment", y = "% of treatment", title = "Frequency of cluster per treatment") +
  scale_fill_manual(values = c(Day4_Vehicle = "lightgreen", Day9_Vehicle = "darkgreen", Day4_PDS = "red", Day9_PDS = "darkred")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_cluster_per_treatment_across_time_ordered_time.pdf"))


```

### Cell Cycle scoring
```{r cell_cycle}

# create output dir
output.dir <- paste0("results/CD45Pos_cell_cycle/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

################################
# Get cell cycle genes 
################################
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

# pseudo-map to mouse gene (captures most except Mlf1ip, Fam64a and Hn1)
s.genes <- stringr::str_to_title(s.genes)
g2m.genes <- stringr::str_to_title(g2m.genes)

#####################
# score cell cycle
#####################

CD45pos.seurat <- CellCycleScoring(CD45pos.seurat,
                                   s.features = s.genes,
                                   g2m.features = g2m.genes,
                                   set.ident = TRUE)


#####################
# Visualise score
#####################

head(CD45pos.seurat@meta.data)

# Feature plots 

# S-score 
scCustomize::FeaturePlot_scCustom(CD45pos.seurat, 
                                  feature = "S.Score", 
                                  reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Pos_S_score.pdf"))
dev.off()

# G2M-score 
scCustomize::FeaturePlot_scCustom(CD45pos.seurat, 
                                  feature = "G2M.Score", 
                                  reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Pos_G2M_score.pdf"))
dev.off()

# Dim plot split by phase
scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                              group.by = "cluster_anno",
                              split.by = "Phase",
                              label = TRUE,
                              reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_CD45Pos_cluster_by_phase.pdf"))
dev.off()


# Stacked VlnPlot
scCustomize::Stacked_VlnPlot(CD45pos.seurat,
                             features = c("S.Score", "G2M.Score"), 
                             group.by = "cluster_anno")

dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Pos_cell_cycle.pdf"))
dev.off()




###################
# Quantification 
###################


# Frequency of phase per cluster
dist.table <- table(CD45pos.seurat@meta.data$cluster_anno, CD45pos.seurat@meta.data$Phase)

cluster.total <- rowSums(dist.table)

freq.table <- sweep(dist.table, 1, cluster.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin
rowSums(freq.table)

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("G1", "G2M", "S"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Predicted phase", y = "% of cluster", title = "Frequency of phase per cluster") +
  scale_fill_manual(values = c(G1 = "lightblue", G2M = "orange", S = "red")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_phase_per_cluster.pdf"))


# Frequency of phase per sampleID

dist.table <- table(CD45pos.seurat@meta.data$SampleID, CD45pos.seurat@meta.data$Phase)

treatment.total <- rowSums(dist.table)

freq.table <- sweep(dist.table, 1, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin
rowSums(freq.table)

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("G1", "G2M", "S"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Predicted phase", y = "% of treatment", title = "Frequency of phase per sampleID") +
  scale_fill_manual(values = c(G1 = "lightblue", G2M = "orange", S = "red")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_phase_per_sampleID.pdf"))

```



## DEG analysis {.tabset}

### Cluster-wise DEGs - calculate
```{r Clusterwise_DEGs_calculation}


# create output dir
output.dir <- paste0("results/CD45Pos_cluster_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  
  CD45pos.seurat <- PrepSCTFindMarkers(CD45pos.seurat)
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  # set idents
  Idents(CD45pos.seurat) <- "cluster_anno"
  
  CD45Pos.de <- FindAllMarkers(CD45pos.seurat,
                               assay = "SCT",
                               only.pos = TRUE, 
                               verbose = TRUE)
  
  
  # save to file 
  write.csv(CD45Pos.de, paste0(output.dir, "CD45Pos_DEGs.csv"))
  
  
  
}else{
  
  # load DEG table
  CD45Pos.de <- read.csv(paste0(output.dir, "CD45Pos_DEGs.csv"))
  
}



```

### Cluster-wise DEGs - heatmaps
```{r clusterwise_DEGs_heatmaps}

# create output dir
output.dir <- paste0("results/CD45Pos_cluster_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

if(do.plot){
  # aggregate expression for heatmap visualisation
  
  pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("cluster_anno"))
  
  
  
  goi <- CD45Pos.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::group_by(cluster) %>%
    dplyr::top_n(10, avg_log2FC) %>%
    dplyr::pull(gene)
  
  
  DoHeatmap(pseudo.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_top_10_DEGs_pseudobulk.pdf"))
  dev.off()
  
  
  
  
  # iterate over the top degs for each cluster 
  cluster.id <- unique(CD45Pos.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- CD45Pos.de %>%
      dplyr::filter(p_val_adj < 0.05) %>%
      dplyr::filter(cluster == cluster.id[i]) %>%
      dplyr::top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.cd45pos, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE) + 
        scico::scale_fill_scico(palette = "batlow", 
                                direction = 1, 
                                na.value = "white")
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
}

```






### Treatment-wise DEGs - calculate
```{r Treatmentwise_DEGs_calculation}

# create output dir
output.dir <- paste0("results/CD45Pos_treatment_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  
  CD45pos.seurat <- PrepSCTFindMarkers(CD45pos.seurat)
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  
  # add annotation for treatment 
  CD45pos.seurat@meta.data$Treatment_cluster <- paste0(CD45pos.seurat@meta.data$Treatment, "_", CD45pos.seurat@meta.data$cluster_anno)
  
  # set Idents
  Idents(CD45pos.seurat) <- "Treatment_cluster"
  
  
  cluster.ids <- unique(CD45pos.seurat@meta.data$cluster_anno)
  
  # create list to store output
  DEG.list <- list()
  
  # run through DEG specifically between vehicle and treatment for each cluster 
  
  for(i in seq_along(cluster.ids)){
    temp.de <- FindMarkers(CD45pos.seurat,
                           assay = "SCT",
                           ident.1 = paste0("PDS_", cluster.ids[i]),
                           ident.2 = paste0("Vehicle_", cluster.ids[i]),
                           only.pos = FALSE, 
                           recorrect_umi = FALSE,
                           verbose = TRUE)
    
    DEG.list[[i]] <- temp.de
    
    # save to file 
    write.csv(temp.de, paste0(output.dir, "DEGs_PDS_", cluster.ids[i], "_vs_Vehicle_", cluster.ids[i], ".csv"))
    
    
  }
  # add names to DEG.list
  names(DEG.list) <- cluster.ids
  
  # save DEG list to file 
  saveRDS(DEG.list, paste0(output.dir, "DEG_list_object.rds"))
  
  
}else{
  
  DEG.list <- readRDS(paste0(output.dir, "DEG_list_object.rds"))
  
}


```


### Treatment-wise DEGs - Volcano
```{r treatmentwise_DEGs_volcano}


# create output dir
output.dir <- paste0("results/CD45Pos_treatment_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# visualise treatment-wise DEGs as volcano plots


if(do.plot){
  
  for(i in 1:length(DEG.list)){
    
    FDR.theshold <- 0.01
    logFC.threshold <- 1
    
    # filter data
    input.data <- DEG.list[[i]] %>%
      dplyr::rename(logFC = avg_log2FC) %>%
      dplyr::rename(FDR = p_val_adj) %>%
      filter(FDR < FDR.theshold)
    
    # color points
    volcano.col <- usefulfunctions::colour.points.volcano(input.data,
                                                          increase.col = "Red",
                                                          decreased.col = "Blue",
                                                          FDR.cutoff = FDR.theshold, 
                                                          logFC.cutoff = logFC.threshold)
    # plot heatmap
    print(
      usefulfunctions::custom.enhanced.volcano(input.data, 
                                               pCutoff = FDR.theshold,
                                               FCcutoff = logFC.threshold,
                                               title = paste0(names(DEG.list)[i], " PDS vs Vehicle"),
                                               subtitle = paste0("adjPval threshold = ", FDR.theshold, " & log2FC threshold = ", logFC.threshold),
                                               colCustom = volcano.col)
    )
    dev.copy(pdf, paste0(output.dir, "VolcanoPlot_PDS_vs_Vehicle_cluster_", names(DEG.list)[i], ".pdf"))
    dev.off()
    
    
  }
  
}

```


### Export aggregated counts table
```{r export_aggregated_counts}

# create output dir
output.dir <- paste0("results/tables/CD45Pos/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  DefaultAssay(CD45pos.seurat) <- "SCT"

  # Aggregate expression by cluster 
  pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("cluster_anno"))
  
  # Write to file
  temp.out <- GetAssayData(pseudo.cd45pos,
                           layer = "data",
                           assay = "SCT")
  
  write.csv(temp.out, paste0(output.dir, "CD45pos_Aggregate_SCT_data_by_cluster.csv"))
  
  # Aggregate expression by cluster and SampleID
    DefaultAssay(CD45pos.seurat) <- "SCT"

  pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("cluster_anno", "SampleID"))
  # write to file
  temp.out <- GetAssayData(pseudo.cd45pos,
                           layer = "data",
                           assay = "SCT")
  
  write.csv(temp.out, paste0(output.dir, "CD45Pos_Aggregate_SCT_data_by_cluster_and_SampleID.csv"))
  
  # clean environment
  rm(pseudo.cd45pos)
  rm(temp.out)
  
}



```


## RQ1: Which CD45Pos clusters express chemokines or cytokines {.tabset}

### Chemokine and Cytokine expression visualisation 
```{r chemokine_cytokine_expression}

# create output dir
output.dir <- paste0("results/CD45Pos_chemokine_cytokines/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# aggregate expression for heatmap visualisation
DefaultAssay(CD45pos.seurat) <- "SCT"
pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("cluster_anno"))


########################################
# List gene sets of interest to plot
########################################

cxc.genes <- grep("^Cxc.*", rownames(pseudo.cd45pos), value = T)
ccr.genes <- grep("^Ccr.*", rownames(pseudo.cd45pos), value = T)
il.genes <- grep("^Il.*", rownames(pseudo.cd45pos), value = T)
gzm.genes <- grep("^Gzm.*", rownames(pseudo.cd45pos), value = T)
ifn.genes <- grep("^Ifn.*", rownames(pseudo.cd45pos), value = T)
custom.genes <- c("Il1b", "Il6", "Il12a", "Il12b", "Ifng", "Tnf", "Ccl2", "Ccl3", "Ccl4", "Ccl5", "Cxcl1", "Cxcl9", "Cxcl10") # Cxcl13 not present in CD45Pos dataset

goi <- list(Cxcr_genes = cxc.genes, 
            Ccr_genes = ccr.genes,
            Interleukin_genes = il.genes,
            Gzm_genes = gzm.genes,
            Ifn_genes = ifn.genes,
            Custom_genes = custom.genes)


###################
# Plot heatmaps
###################

for(i in seq_along(names(goi))){
  
  print(
    DoHeatmap(pseudo.cd45pos, 
              features = goi[[i]], 
              draw.lines = FALSE, 
              angle = 90,
              raster = FALSE) + 
      scico::scale_fill_scico(palette = "batlow", 
                              direction = 1, 
                              na.value = "white")
  )
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_", names(goi)[i], ".pdf"))
  dev.off()
  
}


##############################
# Stacked vln plot
##############################

for(i in seq_along(names(goi))){
  
  if(length(goi[[i]]) <= 15){
    
    
    print(
      scCustomize::Stacked_VlnPlot(CD45pos.seurat, 
                                   features = goi[[i]],
                                   group.by = "cluster_anno",
                                   pt.size = 0)
    )
    
    dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Pos_", names(goi)[i], ".pdf"))
    dev.off()
    
  }else{
    
    temp.goi <- goi[[i]]
    
    # Split the vector into multiple vectors
    split_indices <- seq(1, length(temp.goi), by = 15)
    split_list <- lapply(split_indices, function(i) temp.goi[i:min(i + 14, length(temp.goi))])
    
    
    for(j in seq_along(length(split_list))){
      
      print(
        scCustomize::Stacked_VlnPlot(CD45pos.seurat, 
                                     features = split_list[[j]],
                                     group.by = "cluster_anno",
                                     pt.size = 0)
      )
      
      dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Pos_", names(goi)[i], "_", j, ".pdf"))
      dev.off()
      
    }
    
  }
  
}


```




# i am up to here 


# re-write below based on analysis layout in cd45neg dataset 


# TO do for Nils dataset
# CD45Neg dataset 
# cytokine chemokine across treatment from tumor cells
# apoptosis across treatment in tumor cells - PDS should induce apoptosis 

# CD45Pos dataset
# Time 1 - activation, myeloid, neutrophil compartments 
# time 1 - myeloid M1 vs M2, neutrophil, antigen presentation 
# time 2 - tcell, activation, exhasution, 


# i am up to this chunk 

## RQ2: Are there differences between macrophages and DCs between PDS and Vehicle (because there are DAMPs that should activate them) {.tabset}


```{r treatment_differences_in_DCs_Macrophages}


# create output dir
output.dir <- paste0("results/RQ2_Macrophage_DC_treatment_differences/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# i am up to here
# to do this bit










# Firstly are there abundance differences or polarisations torwards particular subsets of marophages or DCs between untreated and treated 

dist.table <- table(CD45pos.seurat@meta.data$cluster_anno, CD45pos.seurat@meta.data$Treatment)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Vehicle", "PDS"))


library(ggplot2)

ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  scale_fill_hue(direction = 1) +
  theme_minimal() +
  facet_wrap(vars(Var1))

dev.copy(pdf, paste0(output.dir, "Barplot_freq_of_clusters_as_percentage_of_treatment.pdf"))
dev.off()      



# plot umaps split by treatment 
scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                              group.by = "cluster_anno", 
                              split.by = "Treatment", 
                              label = TRUE,
                              reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_split_by_treatment.pdf"))
dev.off()








```






## cluster highlight plots
```{r remedy001}

scCustomize::Cluster_Highlight_Plot(CD45pos.seurat, cluster_name = 25, reduction = "umap")


```



# NOT WORKING - check all code 
# also not written again for this dataset specific analysis - check all code 



## NO LONGER USED
## annotate clusters {.tabset}
```{r annotate_clusters}
# 
# if(long.compute){
#   
# ########################################################
# # provide annotation for clusters where possible 
# ########################################################
# 
# 
# # create output dir
# output.dir <- paste0("results/CD45Pos_cluster_vs_azimuth/")
# 
# if(!dir.exists(output.dir)){
#   dir.create(output.dir, 
#              recursive = T)
#   
# }
# 
# 
# #############################
# # cluster by annotation 
# #############################
# 
# # large grain annotation
# 
# # broad annotation 
# ## look at all clusters again across annotation. as some clusters maybe can be called X celltype where they are say 70% Xcelltype and rest is made up of small annotations 
# # as opposed to clusters 50/50 split between annotations 
# 
# 
# # calculate the frequency of each cluster that has an azimuth annotation, therefore the clusters which are primarily 1 annotation can easily be defined 
# x <- table(CD45pos.seurat@meta.data$predicted.celltype.l1, CD45pos.seurat@meta.data$seurat_clusters)
# cluster.totals <- colSums(x)
# y <- sweep(x, 2, cluster.totals, FUN = "/")
# y <- y*100
# y <- as.data.frame(y)
# 
# 
# cluster.id <- unique(CD45pos.seurat@meta.data$seurat_clusters)
# 
# for(i in seq_along(cluster.id)){
#   
#   y.filt <- y %>%
#     filter(Var2 == cluster.id[i])
#   
#   
#   library(ggplot2)
#   
#   p1 <- ggplot(y.filt) +
#     aes(x = Var1, y = Freq, fill = Var1) +
#     geom_col() +
#     scale_fill_hue(direction = 1) +
#     labs(title = paste0("Cluster ", cluster.id[i])) +
#            coord_flip() +
#            theme_minimal() +
#            theme(legend.position = "none") +
#            ylim(0, 100)
#          
#          print(p1)
#          
#          dev.copy(pdf, paste0(output.dir, "Cluster_", cluster.id[i], "_broad_azimuth_annotation_freq.pdf"))
#          dev.off()
#          
# }
# 
# 
# #############################
# # medium grain annotation 
# #############################
# 
# ## look at all clusters again across annotation. as some clusters maybe can be called X celltype where they are say 70% Xcelltype and rest is made up of small annotations 
# # as opposed to clusters 50/50 split between annotations 
# 
# 
# # calculate the frequency of each cluster that has an azimuth annotation, therefore the clusters which are primarily 1 annotation can easily be defined 
# x <- table(CD45pos.seurat@meta.data$predicted.celltype.l2, CD45pos.seurat@meta.data$seurat_clusters)
# cluster.totals <- colSums(x)
# y <- sweep(x, 2, cluster.totals, FUN = "/")
# y <- y*100
# y <- as.data.frame(y)
# 
# 
# cluster.id <- unique(CD45pos.seurat@meta.data$seurat_clusters)
# 
# for(i in seq_along(cluster.id)){
#   
#   y.filt <- y %>%
#     filter(Var2 == cluster.id[i])
#   
#   
#   library(ggplot2)
#   
#   p1 <- ggplot(y.filt) +
#     aes(x = Var1, y = Freq, fill = Var1) +
#     geom_col() +
#     scale_fill_hue(direction = 1) +
#     labs(title = paste0("Cluster ", cluster.id[i])) +
#            coord_flip() +
#            theme_minimal() +
#            theme(legend.position = "none") +
#            ylim(0, 100)
#          
#          print(p1)
#          
#          dev.copy(pdf, paste0(output.dir, "Cluster_", cluster.id[i], "_medium_grain_azimuth_annotation_freq.pdf"))
#          dev.off()
#          
# }
# 
# #############################
# # fine grain annotation 
# #############################
# 
# ## look at all clusters again across annotation. as some clusters maybe can be called X celltype where they are say 70% Xcelltype and rest is made up of small annotations 
# # as opposed to clusters 50/50 split between annotations 
# 
# 
# # calculate the frequency of each cluster that has an azimuth annotation, therefore the clusters which are primarily 1 annotation can easily be defined 
# x <- table(CD45pos.seurat@meta.data$predicted.celltype.l3, CD45pos.seurat@meta.data$seurat_clusters)
# cluster.totals <- colSums(x)
# y <- sweep(x, 2, cluster.totals, FUN = "/")
# y <- y*100
# y <- as.data.frame(y)
# 
# 
# cluster.id <- unique(CD45pos.seurat@meta.data$seurat_clusters)
# 
# for(i in seq_along(cluster.id)){
#   
#   y.filt <- y %>%
#     filter(Var2 == cluster.id[i])
#   
#   
#   library(ggplot2)
#   
#   p1 <- ggplot(y.filt) +
#     aes(x = Var1, y = Freq, fill = Var1) +
#     geom_col() +
#     scale_fill_hue(direction = 1) +
#     labs(title = paste0("Cluster ", cluster.id[i])) +
#            coord_flip() +
#            theme_minimal() +
#            theme(legend.position = "none") +
#            ylim(0, 100)
#          
#          print(p1)
#          
#          dev.copy(pdf, paste0(output.dir, "Cluster_", cluster.id[i], "_fine_grain_azimuth_annotation_freq.pdf"))
#          dev.off()
#          
# }
# 
# 
# 
# 
# 
# 
# ########################################################
# # Use gene expression profiles to investigate clusters
# ########################################################
# 
# 
# # aggregate expression for heatmap visualisation
# 
# pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
#                                       assays = "SCT",
#                                       return.seurat = TRUE, 
#                                       group.by = c("seurat_clusters"))
# 
# 
# 
# # load deg data
# CD45Pos.de <- read.csv("results/CD45Pos_DEG_analysis/CD45Pos_DEGs.csv")
# 
# 
# 
# # some clusters were easily annotated based on overlap and call by azimuth reference 
# # some are ambiguous 
# # inspect gene expression profiles for these ambiguous clusters
# 
# # this section was done interactively comparing previous plots with visualising certain genes and DEG for each cluster
# 
# grep("FCGR", rownames(pseudo.cd45pos), value = T)
# 
# 
# goi <- CD45Pos.de %>%
#   filter(p_val_adj < 0.05) %>%
#   group_by(cluster) %>%
#   filter(cluster == 10) %>%
#   top_n(50, avg_log2FC) %>%
#   pull(gene)
# 
# goi
#   
#   DoHeatmap(pseudo.cd45pos, 
#           features = goi, 
#           draw.lines = FALSE, 
#           raster = FALSE)
# 
# 
# scCustomize::Cluster_Highlight_Plot(CD45pos.seurat, cluster_name = 25, reduction = "umap")
# 
# VlnPlot_scCustom(CD45pos.seurat, features = "IL2RA", pt.size = 0.1)
# 
# 
# 
# 
# ######################################################
# # using all above plots annotation definitions are 
# ######################################################
# 
# # broad annotations
# # cluster annotations 
# myeloid.clusters <- c(0)
# mono.clusters <- c(2, 4, 6, 7, 9, 12, 13, 14, 15, 17, 19, 20, 21, 22, 24)
# dc.clusters <- c(11, 18, 26)
# 
# bcell.clusters <- c(5)
# erythrocyte.clusters <- c(27)
# 
# t.cell.clusters <- c(1, 3, 23)
# nk.clusters <- c(8, 16)
# nk.or.t.cell.clusters <- c(10, 25) # as in ambig if nk or t or nkt
# 
# 
# # fine grain annotations
# cd14.mono.clusters <- c(2, 4, 6, 9, 14, 20, 22)
# cd16.mono.clusters <- c(13, 17)
# 
# pDC.clusters <- c(26)
# cDC1.clusters <- c(18)
# cDC2.clusters <- c(11)
# 
# cd8.clusters <- c(3)
# cd4.clusters <- c(23)
# 
# 
# 
# 
# #
# # Cluster 23 might be treg cd4
# 
# 
# ############################################################
# # Define clusters based on above data and interpretation
# ############################################################
# 
# 
# CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
#   dplyr::mutate(Anno_l1 = case_when(seurat_clusters %in% myeloid.clusters ~ paste0("Myeloid_", seurat_clusters), 
#                                     seurat_clusters %in% mono.clusters ~ paste0("Monocyte_", seurat_clusters), 
#                                     seurat_clusters %in% dc.clusters ~ paste0("DC_", seurat_clusters), 
#                                     seurat_clusters %in% bcell.clusters ~ paste0("Bcell_", seurat_clusters), 
#                                     seurat_clusters %in% erythrocyte.clusters ~ paste0("Erythrocyte_", seurat_clusters), 
#                                     seurat_clusters %in% t.cell.clusters ~ paste0("Tcell_", seurat_clusters), 
#                                     seurat_clusters %in% nk.clusters ~ paste0("NK_", seurat_clusters), 
#                                     seurat_clusters %in% nk.or.t.cell.clusters ~ paste0("NKorTcell_", seurat_clusters), 
#                                     TRUE ~ seurat_clusters), 
#                 Anno_l2 = case_when(seurat_clusters %in% cd14.mono.clusters ~ paste0("CD14_Mono_", seurat_clusters),
#                                     seurat_clusters %in% cd16.mono.clusters ~ paste0("CD16_Mono_", seurat_clusters),
#                                     seurat_clusters %in% pDC.clusters ~ paste0("pDC_", seurat_clusters),
#                                     seurat_clusters %in% cDC1.clusters ~ paste0("cDC1_", seurat_clusters),
#                                     seurat_clusters %in% cDC2.clusters ~ paste0("cDC2_", seurat_clusters),
#                                     seurat_clusters %in% cd8.clusters ~ paste0("CD8T_", seurat_clusters),
#                                     seurat_clusters %in% cd4.clusters ~ paste0("CD4T_", seurat_clusters),
#                                     TRUE ~ Anno_l1))
# 
# 
# ##############################
# # visualise annotations 
# ##############################
# 
# # create output dir
# output.dir <- paste0("results/CD45Pos_cluster_annotations/")
# 
# if(!dir.exists(output.dir)){
#   dir.create(output.dir, 
#              recursive = T)
#   
# }
# 
# 
# DimPlot_scCustom(CD45pos.seurat, 
#                  label = TRUE,
#                  group.by = "Anno_l1", 
#                  reduction = "umap") + NoLegend()
# 
# dev.copy(pdf, paste0(output.dir, "UMAP_clusters_annotated_l1.pdf"))
# dev.off()
# 
# 
# DimPlot_scCustom(CD45pos.seurat, 
#                  label = TRUE,
#                  group.by = "Anno_l2", 
#                  reduction = "umap") + NoLegend()
# 
# dev.copy(pdf, paste0(output.dir, "UMAP_clusters_annotated_l2.pdf"))
# dev.off()
# 
# 
# 
# # save seurat object after annotation 
# SaveSeuratRds(CD45pos.seurat, 
#               file = "saves/Nils_seurat_CD45Pos.rds")
# }




# for insight into markers can use the below resource 
# load cell marker info from CellMarker2.0 (http://bio-bigdata.hrbmu.edu.cn/CellMarker/index.html)
# cellmarkers <- readxl::read_xlsx("data/Misc/Cell_marker_All.xlsx")
# 
# unique(cellmarkers$tissue_class)
# 
# # filter
# mouse.normal <- cellmarkers %>%
#   filter(species == "Mouse") %>%
#   filter(cell_type == "Normal cell")
# 
# unique(mouse.normal$cell_name)
# 
# 
# Tcell.markers <- mouse.normal %>%
#   filter(grepl("Natural killer", cell_name)) %>%
#   distinct(Symbol)
# 
# print(Tcell.markers, n = 100)


```





## Lineage analysis {.tabset}

### Lineage analysis using GEMLI

```{r GEMLI_analysis}

#devtools::install_github("UPSUTER/GEMLI", subdir="GEMLI_package_v0")
library("GEMLI")


load("~/Downloads/GEMLI_example_data_matrix.RData")

#GEMLI_items[['gene_expression']] = data_matrix

DefaultAssay(CD45pos.seurat) <- "SCT"
Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$SampleID
table(Idents(CD45pos.seurat))
temp.object <- subset(CD45pos.seurat, downsample = 3000)

count.matrix <- LayerData(temp.object, assay = "SCT", layer = "data")

GEMLI_items = list()
GEMLI_items[['gene_expression']] = as.matrix(count.matrix)

GEMLI_items <- predict_lineages(GEMLI_items)

library(igraph)

# style = "fr", "kk", "grid"
visualize_as_network(GEMLI_items, cutoff=90, layout_style = "kk") # top image
visualize_as_network(GEMLI_items, cutoff=50, layout_style = "kk") # lower image

GEMLI_items = prediction_to_lineage_information(GEMLI_items, cutoff=50)

GEMLI_items$predicted_lineage_table[1:50,]

# trim lineages that are too big
suggest_network_trimming_to_size(GEMLI_items, max_size=2, cutoff=50)

GEMLI_items_post_processed = trim_network_to_size(GEMLI_items, max_size=2, cutoff=50)
visualize_as_network(GEMLI_items_post_processed, cutoff=50) 



#load('~/Downloads/GEMLI_crypts_example_cell_type_annotation.RData')
clust.anno <- temp.object@meta.data %>%
  dplyr::select("seurat_clusters") %>%
  dplyr::rename(cell.type = "seurat_clusters")

GEMLI_items_post_processed[['cell_type']] = clust.anno


visualize_as_network(GEMLI_items_post_processed, cutoff=70, max_edge_width=5, display_orphan=F, include_labels=F, ground_truth=F, highlight_FPs=T, layout_style="kk", cell_type_colors=T)

GEMLI_items_post_processed = prediction_to_lineage_information(GEMLI_items_post_processed, cutoff=50)
cell_type_composition_plot(GEMLI_items_post_processed, cell_type_colors=T, type=c("bubble"))

unique(GEMLI_items_post_processed[['cell_type']]$cell.type)



```




