---
title: "CD45Pos_dataset_analysis"
author: "Dillon Corvino"
date: "25/01/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# Data 
# B16 - OVA 

# PDS = I.T - bind Gquadraplex
# day 4 post-treatment 
# day 9 post-treatment

# Research questions 
# How does PDS induces an anti-tumor immune response?
# Known that after PDS treatment, certain DAMPs are induced, intratumoral chemokine levels go up drastically and subsequntly immune cells are recruited into the tumor.
# Main research questions are:
#1. Which cell subsets are producing the chemokines
#2. Are there differences between macrophages and DCs between PDS and Vehicle (because there are DAMPs that should activate them)
#3. How does the treatment effect the tumor cells (is it possible to discriminate them from other tissue cells, if not they express ovalbumin...this should make it at least possible to identify them.)




# key genes
# IL1b, IL6, IL12, IFNG, TNFA, 

# CCL2,3,4,5 
# CXCL1, CXCL9,10,13 





```


### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)


# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- FALSE
do.plot <- TRUE
quick.load <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())


library("Seurat")
library("SeuratObject")
library("SeuratData")
library("SeuratWrappers")
library("dplyr")
library("scCustomize")
library("tibble")


# load output from preprocessing script
if(long.compute){
  
  CD45pos.seurat <- LoadSeuratRds("saves/Preprocessed_CD45Pos.rds")
  
  # set levels for easier downstream plotting 
  CD45pos.seurat@meta.data$Treatment <- factor(CD45pos.seurat@meta.data$Treatment, levels = c("Vehicle", "PDS"))
  CD45pos.seurat@meta.data$Time <- factor(CD45pos.seurat@meta.data$Time, levels = c("Day4", "Day9"))
  CD45pos.seurat@meta.data$SampleID <- factor(CD45pos.seurat@meta.data$SampleID, levels = c("Day4_Vehicle_CD45Pos", "Day4_PDS_CD45Pos", "Day9_Vehicle_CD45Pos", "Day9_PDS_CD45Pos"))
  
}else if(quick.load){
  
  # load cleaned dataset 
  CD45pos.seurat <- LoadSeuratRds("saves/CD45pos_seurat.rds")
  
  
}


```



## Cluster annotation {.tabset}

### SingleR cluster annotation
```{r SingleR_cluster_annotation}

# Notes:
# there is an error with collect() when trying to load celldex datasets
# i think its an error with biocmanager and has been fixed with a very recent update. > 3.16 
# but im running r 4.2.3 and this would require 4.3
# not keen to upgrade right now 
# so will try a work around of downgrading dbplyr to version 2.3.4 from 2.4 which im currently running 
# need to downgrade to this version 
#devtools::install_version("dbplyr", version = "2.3.4")
#devtools::install_version("dbplyr", version = "2.4")
# in theory if biocmanager is up to date, could just run below and should fix the problem
#library(BiocManager) # 3.16
#install("BiocFileCache")
# also:
# need to ensure matrixStats is at version 1.1.0 
#remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts
#BiocManager::install(version = "3.18")



if(long.compute){
  
  output.dir <- "results/CD45Pos_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # load required packages
  library("SingleR")
  library("celldex")
  
  
  ##################################
  # Using MouseRNAseqData
  ##################################
  
  ref.mouse <-  celldex::MouseRNAseqData()
  
  
  # Benayoun Lab that identified, downloaded and processed data sets on GEO that corresponded to sorted cell types (Benayoun et al., 2019).
  
  # The dataset contains 358 mouse RNA-seq samples annotated to 18 main cell types
  # further split into 28 subtypes under label.fine 
  
  
  ######################
  # Fine labels 
  ######################
  SingleR.pred <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45pos.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.fine)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
  #SingleR::plotDeltaDistribution(SingleR.pred)
  #SingleR::plotScoreDistribution(SingleR.pred)
  
  ######################
  # Main labels 
  ######################
  
  SingleR.pred <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45pos.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.main)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_main_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
  
  #####################
  # Use ImmGenData
  #####################
  # Info: Mouse, Microarray, 830 samples, 20 main labels, 253 fine lables, Hematopoietic & Immune focus
  
  ImmGen.data <- celldex::ImmGenData()
  
  
  # Fine labels
  SingleR.pred.fine <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                               ref = ImmGen.data, 
                               method = "cluster",
                               clusters = CD45pos.seurat@meta.data$seurat_clusters,
                               labels = ImmGen.data$label.fine)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred.fine, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_ImmGen_db.pdf"), plot = p1)
  
  
  # Main labels
  SingleR.pred.main <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                               ref = ImmGen.data, 
                               method = "cluster",
                               clusters = CD45pos.seurat@meta.data$seurat_clusters,
                               labels = ImmGen.data$label.main)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred.main, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_main_labels_ImmGen_db.pdf"), plot = p1)
  
  
}


```


### Transfer SingleR annotations to seurat object 
```{r transfer_singleR_annotations}


if(long.compute){
  
  output.dir <- "results/CD45Pos_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  
  ########################################
  # transfer labels to seurat object 
  ########################################
  
  table(predicted=SingleR.pred.main$labels, cluster=rownames(SingleR.pred.main))
  table(predicted=SingleR.pred.fine$labels, cluster=rownames(SingleR.pred.fine))
  
  # Main labels
  main.labels <- SingleR.pred.main$labels
  names(main.labels) <- rownames(SingleR.pred.main)
  
  # Fine labels
  fine.labels <- SingleR.pred.fine$labels
  names(fine.labels) <- rownames(SingleR.pred.fine)
  
  ####################
  # rename clusters 
  ####################
  
  # Main labels
  Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$seurat_clusters
  CD45pos.seurat <- Rename_Clusters(CD45pos.seurat, new_idents = main.labels)
  CD45pos.seurat@meta.data$SingleR_main <- Idents(CD45pos.seurat)
  
  # Fine labels
  Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$seurat_clusters
  CD45pos.seurat <- Rename_Clusters(CD45pos.seurat, new_idents = fine.labels)
  CD45pos.seurat@meta.data$SingleR_fine <- Idents(CD45pos.seurat)
  
  
  ############
  # Visualise
  ############
  
  # Main label
  p.main <- scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                                          group.by = "SingleR_main", 
                                          label = TRUE,
                                          reduction = "umap")
  
  print(p.main)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_main.pdf"))
  dev.off()
  
  
  # Fine label
  p.fine <- scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                                          group.by = "SingleR_fine", 
                                          label = TRUE,
                                          reduction = "umap")
  
  print(p.fine)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_fine.pdf"))
  dev.off()
  
  
  # side-by-side with cluster numbers
  p.clusters <- scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                                              group.by = "seurat_clusters", 
                                              label = TRUE,
                                              reduction = "umap")
  
  # clusters plus main
  print(p.clusters | p.main)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_main_plus_clusters.pdf"))
  dev.off()
  
  # clusters plus fine
  print(p.clusters | p.fine)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_fine_plus_clusters.pdf"))
  dev.off()
  
  
}










```


### Verify cluster annotation
```{r verify_and_finalise_cluster_annotation}


if(long.compute){
  
  output.dir <- "results/CD45Pos_SingleR_annotation_validation/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # overview of data 
  table(CD45pos.seurat@meta.data$seurat_clusters, CD45pos.seurat@meta.data$SingleR_fine)
  
  table(CD45pos.seurat@meta.data$seurat_clusters, CD45pos.seurat@meta.data$SingleR_main)
  table(CD45pos.seurat@meta.data$SingleR_main, CD45pos.seurat@meta.data$SingleR_fine)
  
  # inconsistency between main and fine annotation with respect to T-cell populations
  # Tcells in main are annotated as NKT
  # NKT in main are annotated as treg 
  
  
  # also seurat clusters nicely line up with singleR fine annotations, i.e one cluster per annotation 
  # only monocytes is with higher resolution using seurat clusters, therefore consider seperating into monocytes cluster X and cluster Y
  
  
  # look at annotations and simplify 
  
  
  #################################################
  # calculate DEGs using fine level resolution 
  #################################################
  
  CD45pos.seurat <- PrepSCTFindMarkers(CD45pos.seurat)
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  # set idents
  Idents(CD45pos.seurat) <- "SingleR_fine"
  
  CD45Pos.de <- FindAllMarkers(CD45pos.seurat,
                               assay = "SCT",
                               only.pos = TRUE, 
                               verbose = TRUE)
  
  
  # save to file 
  write.csv(CD45Pos.de, paste0(output.dir, "CD45Pos_DEGs_singleR_fine.csv"))
  
  
  ##############
  # plot DEGs
  ##############
  
  pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("SingleR_fine"))
  
  
  
  goi <- CD45Pos.de %>%
    filter(p_val_adj < 0.05) %>%
    group_by(cluster) %>%
    top_n(10, avg_log2FC) %>%
    pull(gene)
  
  
  DoHeatmap(pseudo.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_top_10_singleR_fine_DEGs_pseudobulk.pdf"))
  dev.off()
  
  # populations are very cleanly resolved 
  
  
  
  
  # iterate over the top degs for each cluster 
  cluster.id <- unique(CD45Pos.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- CD45Pos.de %>%
      filter(p_val_adj < 0.05) %>%
      filter(cluster == cluster.id[i]) %>%
      top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.cd45pos, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE) + 
        scico::scale_fill_scico(palette = "batlow", 
                                direction = 1, 
                                na.value = "white")
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
  
  # happy with using fine level annotation 
  # except, defining NKT from T-cells is a challenge. will just call them T-cells as at main level annotation these are T-cells
  # NKTs at fine level annotation are split into Treg and Teff - and these look to make sense
  # will therefore use fine level annotation fo all, but change NKT to just T-cell 
  # will also simplify the annotation for all clusters but keep their distinct'ness 
  
  
}






```


### Clean up seurat object metadata and export
```{r clean_seurat_object_metadata}

if(long.compute){
  # reannotate singleR_fine cluster annotations based on above analysis 
  # namely, simplify the annotations 
  # also, convert NKT to just T-cells
  
  CD45pos.seurat <- backup.object.for.now
  
  # note the clusters are essentially one cell type one cluster. except for macrophages which is present in two clusters of macrophages and monocytes present in 3 clusters
  table(CD45pos.seurat@meta.data$SingleR_fine, CD45pos.seurat@meta.data$seurat_clusters)
  
  library(stringr) # For using the `str_extract` function
  
  
  # simplify the annotations, to just the cell_type, but keep cluster_seperation 
  CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
    # Annotate clusters based on pattern matching
    mutate(cluster_anno = case_when(
      # Use `str_extract` to extract the cell type from SingleR_fine
      TRUE ~ paste0(
        str_extract(SingleR_fine, "^[^(]+"), # Extract the cell type name
        "_", seurat_clusters
      )
    )) %>%
    # Remove any spaces in the cluster_anno name
    mutate(cluster_anno = str_replace_all(cluster_anno, " ", ""))
  
  unique(CD45pos.seurat@meta.data$cluster_anno)
  table(CD45pos.seurat@meta.data$SingleR_fine, CD45pos.seurat@meta.data$cluster_anno)
  
  
  
  # simplify the annotations, so there are numbered within cell_type
  
  
  
  # little code to count up how many distinct cell types there are 
  CD45pos.seurat@meta.data %>%
    dplyr::select(c(cluster_anno, SingleR_fine)) %>%
    # Extract cell_type from cluster_anno
    mutate(cell_type = str_extract(cluster_anno, "^[^_]+")) %>%
    # Group by cell_type
    group_by(cell_type) %>%
    # Calculate the number of unique annotations for each cell_type and rank them
    mutate(
      # Calculate the count of unique cluster_anno
      unique_count = n_distinct(cluster_anno) ) %>%
    ungroup() %>%
    distinct(cluster_anno, .keep_all = TRUE)
  
  
  
  CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
    # Extract cell_type from cluster_anno
    mutate(cell_type = str_extract(cluster_anno, "^[^_]+")) %>%
    mutate(cluster_anno_new = case_when(
      cluster_anno == "Macrophages_1" ~ "Macrophages_1", 
      cluster_anno == "Macrophages_9" ~ "Macrophages_2", 
      
      cluster_anno == "Bcells_8" ~ "B_cells", 
      
      cluster_anno == "DC_15" ~ "pDC", 
      cluster_anno == "DC_14" ~ "DC_1", 
      cluster_anno == "DC_12" ~ "DC_2",
      
      cluster_anno == "Monocytes_4" ~ "Monocytes_1", 
      cluster_anno == "Monocytes_5" ~ "Monocytes_2", 
      cluster_anno == "Monocytes_7" ~ "Monocytes_3", 
      
      cluster_anno == "Macrophages_1" ~ "Macrophages_1", 
      cluster_anno == "Macrophages_9" ~ "Macrophages_2", 
      
      cluster_anno == "Neutrophils_3" ~ "Neutrophils_1", 
      cluster_anno == "Neutrophils_6" ~ "Neutrophils_2", 
      
      cluster_anno == "Stromalcells_16" ~ "Stromal_cells", 
      
      cluster_anno == "NKcells_2" ~ "NK_cells", 
      
      cluster_anno == "Basophils_17" ~ "Basophils", 
      
      cluster_anno == "Tcells_0" ~ "T_cells_1", 
      cluster_anno == "Tcells_10" ~ "T_cells_2", 
      cluster_anno == "Tcells_13" ~ "T_cells_3", 
      cluster_anno == "Tcells_11" ~ "Tregs", 
      TRUE ~ "ERROR")) %>%
    mutate(cluster_anno = cluster_anno_new) %>%
    select(-cluster_anno_new)
  
  
  unique(CD45pos.seurat@meta.data$cluster_anno)
  unique(CD45pos.seurat@meta.data$SingleR_fine)
  
  
  # clean up object, remove unnecessary columns keep only what is needed
  CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
    select(c("orig.ident", "nCount_RNA", "nFeature_RNA", "nCount_SCT", "nFeature_SCT", "percent_mito", "percent_ribo", "percent_mito_ribo", "log10GenesPerUMI",  "SampleID", "Time", "Compartment", "Treatment",  "cluster_anno", "seurat_clusters", "SCT_snn_res.0.4", "SingleR_main", "SingleR_fine", "cell_type"))
  
  
  # Display the first few rows of the updated data frame to verify the results
  head(as.data.frame(CD45pos.seurat@meta.data[1:10, ]))
  
  
  
  table(CD45pos.seurat@meta.data$SingleR_fine, CD45pos.seurat@meta.data$cell_type)
  table(CD45pos.seurat@meta.data$SingleR_fine, CD45pos.seurat@meta.data$cluster_anno)
  table(CD45pos.seurat@meta.data$cell_type, CD45pos.seurat@meta.data$cluster_anno)
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  # clean up sample ID annotation - no need to say CD45Pos still 
  
  CD45pos.seurat@meta.data$SampleID <- gsub("_CD45Pos", "", CD45pos.seurat@meta.data$SampleID)
  
  unique(CD45pos.seurat@meta.data$cluster_anno)
  head(CD45pos.seurat@meta.data)
  
  # set factorisation 
  CD45pos.seurat@meta.data$Treatment <- factor(CD45pos.seurat@meta.data$Treatment, 
                                               levels = c("Vehicle", "PDS"))
  
  CD45pos.seurat@meta.data$Time <- factor(CD45pos.seurat@meta.data$Time, 
                                          levels = c("Day4", "Day9"))
  
  CD45pos.seurat@meta.data$SampleID <- factor(CD45pos.seurat@meta.data$SampleID, 
                                              levels = c("Day4_Vehicle", "Day9_Vehicle", "Day4_PDS", "Day9_PDS"))
  
  
  CD45pos.seurat@meta.data$cluster_anno <- factor(CD45pos.seurat@meta.data$cluster_anno)
  
  # save output
  SaveSeuratRds(CD45pos.seurat, file = "saves/CD45pos_seurat.rds")
}


```



## Overview of dataset {.tabset}

### Broad overview of dataset
```{r broad_overview}

if(do.plot){
  
  
  # create output dir
  output.dir <- paste0("results/CD45Pos_Overview/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  vars.to.plot <- c("Treatment", "SampleID", "cluster_anno", "Time", "cell_type")
  
  
  for(j in seq_along(vars.to.plot)){
    name.var <- gsub("\\.", "_", vars.to.plot[j])
    
    print(
      scCustomize::DimPlot_scCustom(CD45pos.seurat,
                                    reduction = "umap", 
                                    label = TRUE,
                                    figure_plot = TRUE,
                                    group.by = vars.to.plot[j]) 
    )
    
    
    dev.copy(pdf, paste0(output.dir, "CD45Pos_UMAP_", name.var, ".pdf"))
    dev.off()
    
  }
  
  
  # for split comparison plots, downsample to make visual comparisons more fair 
  table(CD45pos.seurat@meta.data$Treatment)
  Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$Treatment
  temp.seurat <- subset(CD45pos.seurat, downsample = 15000)
  # split umap plots
  print(
    scCustomize::DimPlot_scCustom(temp.seurat,
                                  reduction = "umap", 
                                  label = TRUE,
                                  figure_plot = TRUE,
                                  group.by = "cluster_anno", 
                                  split.by = "Treatment") 
  )
  
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_UMAP_cluster_anno_split_by_treatment.pdf"))
  dev.off()
  
  
  table(CD45pos.seurat@meta.data$SampleID)
  Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$SampleID
  temp.seurat <- subset(CD45pos.seurat, downsample = 5000)
  # split umap plots
  print(
    scCustomize::DimPlot_scCustom(temp.seurat,
                                  reduction = "umap", 
                                  label = FALSE,
                                  figure_plot = TRUE,
                                  group.by = "cluster_anno", 
                                  split.by = "SampleID") 
  )
  
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_UMAP_cluster_anno_split_by_SampleID.pdf"))
  dev.off()
  
  rm(temp.seurat)
  
}



```


### Cluster stats
```{r cluster_stats}

# create output dir
output.dir <- paste0("results/CD45Pos_Overview/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

#############################################
# Number of cells per cluster
#############################################

cluster.n <- table(CD45pos.seurat@meta.data$cluster_anno)

cluster.n <- as.data.frame(cluster.n)

cluster.n$Var1 <- factor(cluster.n$Var1, levels = cluster.n[order(cluster.n$Freq, decreasing = T), ]$Var1)


ggplot(cluster.n) +
  aes(x = Var1, y = Freq, fill = Var1) +
  geom_col() +
  scale_fill_viridis_d(option = "viridis", 
                       direction = -1) +
  labs(x = "Cluster", y = "Number of cells", title = "Number of cells per cluster") +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

ggsave(paste0(output.dir, "Barplot_number_of_cells_per_cluster.pdf"))



#############################################
# Freq of cluster across treatment 
#############################################

dist.table <- table(CD45pos.seurat@meta.data$cluster_anno, CD45pos.seurat@meta.data$Treatment)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Vehicle", "PDS"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Treatment", y = "% of treatment", title = "Frequency of cluster per treatment") +
  scale_fill_manual(values = c(Vehicle = "grey", PDS = "darkred")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_cluster_per_treatment.pdf"))


#############################################
# Freq of clusters per treatment across time
#############################################


#######################################
#  Ordered by treatment
#######################################

dist.table <- table(CD45pos.seurat@meta.data$cluster_anno, CD45pos.seurat@meta.data$SampleID)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Day4_Vehicle", "Day9_Vehicle", "Day4_PDS", "Day9_PDS"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Treatment", y = "% of treatment", title = "Frequency of cluster per treatment") +
  scale_fill_manual(values = c(Day4_Vehicle = "lightgreen", Day9_Vehicle = "darkgreen", Day4_PDS = "red", Day9_PDS = "darkred")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_cluster_per_treatment_across_time.pdf"))



#######################################
#  Ordered by time
#######################################

dist.table <- table(CD45pos.seurat@meta.data$cluster_anno, CD45pos.seurat@meta.data$SampleID)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Day4_Vehicle", "Day4_PDS", "Day9_Vehicle", "Day9_PDS"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Treatment", y = "% of treatment", title = "Frequency of cluster per treatment") +
  scale_fill_manual(values = c(Day4_Vehicle = "lightgreen", Day9_Vehicle = "darkgreen", Day4_PDS = "red", Day9_PDS = "darkred")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_cluster_per_treatment_across_time_ordered_time.pdf"))


```

### Cell Cycle scoring
```{r cell_cycle}

# create output dir
output.dir <- paste0("results/CD45Pos_cell_cycle/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

################################
# Get cell cycle genes 
################################
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

# pseudo-map to mouse gene (captures most except Mlf1ip, Fam64a and Hn1)
s.genes <- stringr::str_to_title(s.genes)
g2m.genes <- stringr::str_to_title(g2m.genes)

#####################
# score cell cycle
#####################

CD45pos.seurat <- CellCycleScoring(CD45pos.seurat,
                                   s.features = s.genes,
                                   g2m.features = g2m.genes,
                                   set.ident = TRUE)


#####################
# Visualise score
#####################

head(CD45pos.seurat@meta.data)

# Feature plots 

# S-score 
scCustomize::FeaturePlot_scCustom(CD45pos.seurat, 
                                  feature = "S.Score", 
                                  reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Pos_S_score.pdf"))
dev.off()

# G2M-score 
scCustomize::FeaturePlot_scCustom(CD45pos.seurat, 
                                  feature = "G2M.Score", 
                                  reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Pos_G2M_score.pdf"))
dev.off()

# Dim plot split by phase
scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                              group.by = "cluster_anno",
                              split.by = "Phase",
                              label = TRUE,
                              reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_CD45Pos_cluster_by_phase.pdf"))
dev.off()


# Stacked VlnPlot
scCustomize::Stacked_VlnPlot(CD45pos.seurat,
                             features = c("S.Score", "G2M.Score"), 
                             group.by = "cluster_anno")

dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Pos_cell_cycle.pdf"))
dev.off()




###################
# Quantification 
###################


# Frequency of phase per cluster
dist.table <- table(CD45pos.seurat@meta.data$cluster_anno, CD45pos.seurat@meta.data$Phase)

cluster.total <- rowSums(dist.table)

freq.table <- sweep(dist.table, 1, cluster.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin
rowSums(freq.table)

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("G1", "G2M", "S"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Predicted phase", y = "% of cluster", title = "Frequency of phase per cluster") +
  scale_fill_manual(values = c(G1 = "lightblue", G2M = "orange", S = "red")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_phase_per_cluster.pdf"))


# Frequency of phase per sampleID

dist.table <- table(CD45pos.seurat@meta.data$SampleID, CD45pos.seurat@meta.data$Phase)

treatment.total <- rowSums(dist.table)

freq.table <- sweep(dist.table, 1, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin
rowSums(freq.table)

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("G1", "G2M", "S"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Predicted phase", y = "% of treatment", title = "Frequency of phase per sampleID") +
  scale_fill_manual(values = c(G1 = "lightblue", G2M = "orange", S = "red")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_phase_per_sampleID.pdf"))

```


## DEG analysis {.tabset}

### Cluster-wise DEGs - calculate
```{r Clusterwise_DEGs_calculation}


# create output dir
output.dir <- paste0("results/CD45Pos_cluster_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  
  CD45pos.seurat <- PrepSCTFindMarkers(CD45pos.seurat)
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  # set idents
  Idents(CD45pos.seurat) <- "cluster_anno"
  
  CD45Pos.de <- FindAllMarkers(CD45pos.seurat,
                               assay = "SCT",
                               only.pos = TRUE, 
                               verbose = TRUE)
  
  
  # save to file 
  write.csv(CD45Pos.de, paste0(output.dir, "CD45Pos_DEGs.csv"))
  
  
  
}else{
  
  # load DEG table
  CD45Pos.de <- read.csv(paste0(output.dir, "CD45Pos_DEGs.csv"))
  
}



```

### Cluster-wise DEGs - heatmaps
```{r clusterwise_DEGs_heatmaps}

# create output dir
output.dir <- paste0("results/CD45Pos_cluster_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

if(do.plot){
  # aggregate expression for heatmap visualisation
  
  pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("cluster_anno"))
  
  
  
  goi <- CD45Pos.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::group_by(cluster) %>%
    dplyr::top_n(10, avg_log2FC) %>%
    dplyr::pull(gene)
  
  
  DoHeatmap(pseudo.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_top_10_DEGs_pseudobulk.pdf"))
  dev.off()
  
  
  
  
  # iterate over the top degs for each cluster 
  cluster.id <- unique(CD45Pos.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- CD45Pos.de %>%
      dplyr::filter(p_val_adj < 0.05) %>%
      dplyr::filter(cluster == cluster.id[i]) %>%
      dplyr::top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.cd45pos, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE) + 
        scico::scale_fill_scico(palette = "batlow", 
                                direction = 1, 
                                na.value = "white")
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
}

```


### Treatment-wise DEGs - calculate
```{r Treatmentwise_DEGs_calculation}

# create output dir
output.dir <- paste0("results/CD45Pos_treatment_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  
  CD45pos.seurat <- PrepSCTFindMarkers(CD45pos.seurat)
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  
  # add annotation for treatment 
  CD45pos.seurat@meta.data$Treatment_cluster <- paste0(CD45pos.seurat@meta.data$Treatment, "_", CD45pos.seurat@meta.data$cluster_anno)
  
  # set Idents
  Idents(CD45pos.seurat) <- "Treatment_cluster"
  
  
  cluster.ids <- unique(CD45pos.seurat@meta.data$cluster_anno)
  
  # create list to store output
  DEG.list <- list()
  
  # run through DEG specifically between vehicle and treatment for each cluster 
  
  for(i in seq_along(cluster.ids)){
    temp.de <- FindMarkers(CD45pos.seurat,
                           assay = "SCT",
                           ident.1 = paste0("PDS_", cluster.ids[i]),
                           ident.2 = paste0("Vehicle_", cluster.ids[i]),
                           only.pos = FALSE, 
                           recorrect_umi = FALSE,
                           verbose = TRUE)
    
    DEG.list[[i]] <- temp.de
    
    # save to file 
    write.csv(temp.de, paste0(output.dir, "DEGs_PDS_", cluster.ids[i], "_vs_Vehicle_", cluster.ids[i], ".csv"))
    
    
  }
  # add names to DEG.list
  names(DEG.list) <- cluster.ids
  
  # save DEG list to file 
  saveRDS(DEG.list, paste0(output.dir, "DEG_list_object.rds"))
  
  
}else{
  
  DEG.list <- readRDS(paste0(output.dir, "DEG_list_object.rds"))
  
}


```


### Treatment-wise DEGs - Volcano
```{r treatmentwise_DEGs_volcano}


# create output dir
output.dir <- paste0("results/CD45Pos_treatment_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# visualise treatment-wise DEGs as volcano plots


if(do.plot){
  
  for(i in 1:length(DEG.list)){
    
    FDR.theshold <- 0.01
    logFC.threshold <- 1
    
    # filter data
    input.data <- DEG.list[[i]] %>%
      dplyr::rename(logFC = avg_log2FC) %>%
      dplyr::rename(FDR = p_val_adj) %>%
      filter(FDR < FDR.theshold)
    
    # color points
    volcano.col <- usefulfunctions::colour.points.volcano(input.data,
                                                          increase.col = "Red",
                                                          decreased.col = "Blue",
                                                          FDR.cutoff = FDR.theshold, 
                                                          logFC.cutoff = logFC.threshold)
    # plot heatmap
    print(
      usefulfunctions::custom.enhanced.volcano(input.data, 
                                               pCutoff = FDR.theshold,
                                               FCcutoff = logFC.threshold,
                                               title = paste0(names(DEG.list)[i], " PDS vs Vehicle"),
                                               subtitle = paste0("adjPval threshold = ", FDR.theshold, " & log2FC threshold = ", logFC.threshold),
                                               colCustom = volcano.col)
    )
    dev.copy(pdf, paste0(output.dir, "VolcanoPlot_PDS_vs_Vehicle_cluster_", names(DEG.list)[i], ".pdf"))
    dev.off()
    
    
  }
  
}

```


### Export aggregated counts table
```{r export_aggregated_counts}

# create output dir
output.dir <- paste0("results/tables/CD45Pos/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  # Aggregate expression by cluster 
  pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("cluster_anno"))
  
  # Write to file
  temp.out <- GetAssayData(pseudo.cd45pos,
                           layer = "data",
                           assay = "SCT")
  
  write.csv(temp.out, paste0(output.dir, "CD45pos_Aggregate_SCT_data_by_cluster.csv"))
  
  # Aggregate expression by cluster and SampleID
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("cluster_anno", "SampleID"))
  # write to file
  temp.out <- GetAssayData(pseudo.cd45pos,
                           layer = "data",
                           assay = "SCT")
  
  write.csv(temp.out, paste0(output.dir, "CD45Pos_Aggregate_SCT_data_by_cluster_and_SampleID.csv"))
  
  # clean environment
  rm(pseudo.cd45pos)
  rm(temp.out)
  
}



```




# TO DO
# for CD45Neg dataset
# group Tumor together, DEG between vehicle and treatment 

# for CD45Pos dataset
# split by time 
# day 4 - group monocytes, macrophage, neutrophils etc, DEG between vehicle and treatment 
# day 9 group t-cells, group neutrophils, degs between vehcile treatment, 
# neutrophil treatment day 9 vs day 4 neutrophils 
# zoom in on t-cells, can find cytotoxic cd4s?


## RQ1: Which CD45Pos clusters express chemokines or cytokines {.tabset}

### Chemokine and Cytokine expression visualisation 
```{r chemokine_cytokine_expression}

# create output dir
output.dir <- paste0("results/CD45Pos_chemokine_cytokines/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# aggregate expression for heatmap visualisation
DefaultAssay(CD45pos.seurat) <- "SCT"
pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("cluster_anno"))


########################################
# List gene sets of interest to plot
########################################

cxc.genes <- grep("^Cxc.*", rownames(pseudo.cd45pos), value = T)
ccr.genes <- grep("^Ccr.*", rownames(pseudo.cd45pos), value = T)
il.genes <- grep("^Il.*", rownames(pseudo.cd45pos), value = T)
gzm.genes <- grep("^Gzm.*", rownames(pseudo.cd45pos), value = T)
ifn.genes <- grep("^Ifn.*", rownames(pseudo.cd45pos), value = T)
custom.genes <- c("Il1b", "Il6", "Il12a", "Il12b", "Ifng", "Tnf", "Ccl2", "Ccl3", "Ccl4", "Ccl5", "Cxcl1", "Cxcl9", "Cxcl10") # Cxcl13 not present in CD45Pos dataset

goi <- list(Cxcr_genes = cxc.genes, 
            Ccr_genes = ccr.genes,
            Interleukin_genes = il.genes,
            Gzm_genes = gzm.genes,
            Ifn_genes = ifn.genes,
            Custom_genes = custom.genes)


###################
# Plot heatmaps
###################

for(i in seq_along(names(goi))){
  
  print(
    DoHeatmap(pseudo.cd45pos, 
              features = goi[[i]], 
              draw.lines = FALSE, 
              angle = 90,
              raster = FALSE) + 
      scico::scale_fill_scico(palette = "batlow", 
                              direction = 1, 
                              na.value = "white")
  )
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_", names(goi)[i], ".pdf"))
  dev.off()
  
}


##############################
# Stacked vln plot
##############################

for(i in seq_along(names(goi))){
  
  if(length(goi[[i]]) <= 15){
    
    
    print(
      scCustomize::Stacked_VlnPlot(CD45pos.seurat, 
                                   features = goi[[i]],
                                   group.by = "cluster_anno",
                                   pt.size = 0)
    )
    
    dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Pos_", names(goi)[i], ".pdf"))
    dev.off()
    
  }else{
    
    temp.goi <- goi[[i]]
    
    # Split the vector into multiple vectors
    split_indices <- seq(1, length(temp.goi), by = 15)
    split_list <- lapply(split_indices, function(i) temp.goi[i:min(i + 14, length(temp.goi))])
    
    
    for(j in seq_along(length(split_list))){
      
      print(
        scCustomize::Stacked_VlnPlot(CD45pos.seurat, 
                                     features = split_list[[j]],
                                     group.by = "cluster_anno",
                                     pt.size = 0)
      )
      
      dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Pos_", names(goi)[i], "_", j, ".pdf"))
      dev.off()
      
    }
    
  }
  
}


```


## RQ2: What is the difference between macrophage clusters and between neutrophil clusters {.tabset}
# PDS should induce DAMPs that activate macrophage and neutrophil populations

```{r Macrophage_neutrophil_cluster_differences}


# create output dir
output.dir <- paste0("results/CD45Pos_Day_4_Neutrophil_Macrophage_cluster_differences/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}



# Myeloid populations at early timepoint are expected to play a large role
# investigate myeloid populations at early timepoint between PDS and Vehicle 


# reduce dataset to just day 4 timepoint 
Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$Time
day4.seurat <- subset(CD45pos.seurat, idents = "Day4")






# Firstly are there abundance differences or polarisations torwards particular subsets of marophages or DCs between untreated and treated 

dist.table <- table(day4.seurat@meta.data$cluster_anno, day4.seurat@meta.data$Treatment)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Vehicle", "PDS"))


library(ggplot2)

ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  scale_fill_hue(direction = 1) +
  theme_minimal() +
  facet_wrap(vars(Var1))

dev.copy(pdf, paste0(output.dir, "Barplot_freq_of_clusters_as_percentage_of_treatment_day4_only.pdf"))
dev.off()      


# note: 
# more B cells, DC_1, DC_2, and T_cells_1, and T_cells_2 in vehicle 
# major increase in Neutrophils_1 and Neutrophils_2 in PDS
# increase in Macrophages_2 but not Macrophages_1 in PDS

# downsample for accurate representation of clusters

table(day4.seurat@meta.data$SampleID) # 9,950 vehicle and 8,082 PDS
# reduce to 6,000 cells each 
Idents(day4.seurat) <- day4.seurat@meta.data$SampleID
temp.seurat <- subset(day4.seurat, downsample = 6000)




# plot umaps split by treatment 
scCustomize::DimPlot_scCustom(temp.seurat, 
                              group.by = "cluster_anno", 
                              split.by = "Treatment", 
                              label = TRUE,
                              pt.size = 1,
                              reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_split_by_treatment_downsampled.pdf"))
dev.off()


rm(temp.seurat)

# DEG between myeloid populations 

# how are Neutrophil_1 and Neutrophil_2 populations different 



if(long.compute){
  
  
  day4.seurat <- PrepSCTFindMarkers(day4.seurat)
  
  # set assay
  DefaultAssay(day4.seurat) <- "SCT"
  
  
  
  # set Idents
  Idents(day4.seurat) <- "cluster_anno"
  
  
  
  # Macrophage 1 vs 2
  D4.macrophage.de <- FindMarkers(day4.seurat,
                                  assay = "SCT",
                                  ident.1 = "Macrophages_1",
                                  ident.2 = "Macrophages_2",
                                  only.pos = FALSE, 
                                  recorrect_umi = FALSE,
                                  verbose = TRUE)
  
  
  # save to file 
  write.csv(D4.macrophage.de, paste0(output.dir, "Day_4_CD45Pos_DEGs_Macrophages_1_vs_Macrophages_2.csv"))
  
  
  # Neutrophils_1 vs 2
  D4.neutrophil.de <- FindMarkers(day4.seurat,
                                  assay = "SCT",
                                  ident.1 = "Neutrophils_1",
                                  ident.2 = "Neutrophils_2",
                                  only.pos = FALSE, 
                                  recorrect_umi = FALSE,
                                  verbose = TRUE)
  
  
  # save to file 
  write.csv(D4.neutrophil.de, paste0(output.dir, "Day_4_CD45Pos_DEGs_Neutrophils_1_vs_Neutrohils_2.csv"))
  
}


if(do.plot){
  
  #####################
  # heatmaps of top DEGs
  #####################
  
  
  # Neutrophils 1 vs 2
  Idents(day4.seurat) <- day4.seurat@meta.data$cluster_anno
  
  
  temp.seurat <- subset(day4.seurat, idents = c("Neutrophils_1", "Neutrophils_2"))
  
  pseudo.D4.cd45pos <- AggregateExpression(temp.seurat,
                                           assays = "SCT",
                                           return.seurat = TRUE, 
                                           group.by = c("cluster_anno"))
  
  
  # Extract top fold change DEGs 
  
  goi <- D4.neutrophil.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(abs(avg_log2FC) > 2) %>%
    dplyr::arrange(desc(avg_log2FC)) %>%
    rownames()
  
  
  # plot heatmap
  DoHeatmap(pseudo.D4.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_Day4_Heatmap_top_DEGs_logfold_2_neutrophils_1_vs_2.pdf"))
  dev.off()
  
  
  # extract just top 10 up and down degs
  goi.up <- D4.neutrophil.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::top_n(10, avg_log2FC) %>%
    rownames()
  
  goi.dn <- D4.neutrophil.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::top_n(-10, avg_log2FC) %>%
    rownames()
  
  goi <- unique(c(goi.up, goi.dn))
  
  DoHeatmap(pseudo.D4.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_Day4_Heatmap_top_DEGs_10_up_down_neutrophils_1_vs_2.pdf"))
  dev.off()
  
  
  
  
  ##################
  # Macrophages 1 vs 2
  ##################
  Idents(day4.seurat) <- day4.seurat@meta.data$cluster_anno
  
  
  temp.seurat <- subset(day4.seurat, idents = c("Macrophages_1", "Macrophages_2"))
  
  pseudo.D4.cd45pos <- AggregateExpression(temp.seurat,
                                           assays = "SCT",
                                           return.seurat = TRUE, 
                                           group.by = c("cluster_anno"))
  
  
  # Extract top fold change DEGs 
  
  goi <- D4.macrophage.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(abs(avg_log2FC) > 2) %>%
    dplyr::arrange(desc(avg_log2FC)) %>%
    rownames()
  
  
  # plot heatmap
  DoHeatmap(pseudo.D4.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_Day4_Heatmap_top_DEGs_logfold_2_macrophage_1_vs_2.pdf"))
  dev.off()
  
  
  # extract just top 10 up and down degs
  goi.up <- D4.macrophage.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::top_n(10, avg_log2FC) %>%
    rownames()
  
  goi.dn <- D4.macrophage.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::top_n(-10, avg_log2FC) %>%
    rownames()
  
  goi <- unique(c(goi.up, goi.dn))
  
  DoHeatmap(pseudo.D4.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_Day4_Heatmap_top_DEGs_10_up_down_macrophage_1_vs_2.pdf"))
  dev.off()
  
  
}

####################################
# perform gene ontology analysis 
####################################

if(do.plot){
  
  # load packages
  #install.packages("enrichR")
  library("enrichR")
  #install.packages("SCpubr")
  library("SCpubr")
  
  # set serach to human genes
  enrichR::setEnrichrSite(site = "Enrichr")
  
  dbs <- enrichR::listEnrichrDbs()
  dbs <- sort(dbs$libraryName)
  
  # Choose the dataset to query against.
  dbs_use <- c("GO_Biological_Process_2023", "GO_Cellular_Component_2023", "GO_Molecular_Function_2023", "KEGG_2021_Human", "WikiPathway_2023_Human")
  
  
  
  ##################################################
  # Perform GO for neutrophil_1 vs neutrophil_2 
  ##################################################
  
  title.var <- "Neutrophils"
  
  # filter DEGs for sig up genes 
  Neutrophils_1_sig <- D4.neutrophil.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(avg_log2FC > 1)
  
  # filter DEGs for sig down genes 
  Neutrophils_2_sig <- D4.neutrophil.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(avg_log2FC < -1)
  
  
  ########################################################################
  # Perform enrichment for genes upregulated in Neutrophils_1
  ########################################################################
  
  # List of genes to use as input.
  neutrophil.1.genes <- unique(rownames(Neutrophils_1_sig))
  
  # Retrieve the enriched terms.
  up.enriched.terms <- enrichR::enrichr(neutrophil.1.genes, dbs_use)
  
  # Plot enriched terms 
  bp.1 <- plotEnrich(up.enriched.terms$GO_Biological_Process_2023, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched BPs"))
  cc.1 <- plotEnrich(up.enriched.terms$GO_Cellular_Component_2023, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched CCs"))
  mf.1 <- plotEnrich(up.enriched.terms$GO_Molecular_Function_2023, showTerms = 20, numChar = 50, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched MFs"))
  kegg.1 <- plotEnrich(up.enriched.terms$KEGG_2021_Human, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched KEGG"))
  wiki.1 <- plotEnrich(up.enriched.terms$WikiPathway_2023_Human, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched Wiki paths"))
  
  
  ############################################################
  # Perform enrichment analysis for Downregulated genes 
  ############################################################
  
  
  # List of genes to use as input.
  neutrophil.2.genes <- unique(rownames(Neutrophils_2_sig))
  
  # Retrieve the enriched terms.
  dn.enriched.terms <- enrichR::enrichr(neutrophil.2.genes, dbs_use)
  
  # Plot enriched terms 
  
  bp.2 <- plotEnrich(dn.enriched.terms$GO_Biological_Process_2023, showTerms = 20, numChar = 50, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched BPs"))
  cc.2 <- plotEnrich(dn.enriched.terms$GO_Cellular_Component_2023, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched CCs"))
  mf.2 <- plotEnrich(dn.enriched.terms$GO_Molecular_Function_2023, showTerms = 20, numChar = 50, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched MFs"))
  kegg.2 <- plotEnrich(dn.enriched.terms$KEGG_2021_Human, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched KEGG"))
  wiki.2 <- plotEnrich(dn.enriched.terms$WikiPathway_2023_Human, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched Wiki paths"))
  
  
  
  print(bp.1|bp.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_GO_BP.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  print(cc.1|cc.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_GO_CC.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  
  print(mf.1|mf.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_GO_MF.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  
  print(kegg.1|kegg.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_KEGG.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  
  print(wiki.1|wiki.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_WIKI.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  
  
  
  ##################################################
  # Perform GO for Macrophages_1 vs Macrophages_2 
  ##################################################
  
  title.var <- "Macrophages"
  
  # filter DEGs for sig up genes 
  Macrophages_1_sig <- D4.macrophage.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(avg_log2FC > 1)
  
  # filter DEGs for sig down genes 
  Macrophages_2_sig <- D4.macrophage.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(avg_log2FC < -1)
  
  
  ########################################################################
  # Perform enrichment for genes upregulated in Neutrophils_1
  ########################################################################
  
  # List of genes to use as input.
  macrophages.1.genes <- unique(rownames(Macrophages_1_sig))
  
  # Retrieve the enriched terms.
  up.enriched.terms <- enrichR::enrichr(macrophages.1.genes, dbs_use)
  
  # Plot enriched terms 
  bp.1 <- plotEnrich(up.enriched.terms$GO_Biological_Process_2023, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched BPs"))
  cc.1 <- plotEnrich(up.enriched.terms$GO_Cellular_Component_2023, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched CCs"))
  mf.1 <- plotEnrich(up.enriched.terms$GO_Molecular_Function_2023, showTerms = 20, numChar = 50, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched MFs"))
  kegg.1 <- plotEnrich(up.enriched.terms$KEGG_2021_Human, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched KEGG"))
  wiki.1 <- plotEnrich(up.enriched.terms$WikiPathway_2023_Human, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched Wiki paths"))
  
  
  ############################################################
  # Perform enrichment analysis for Downregulated genes 
  ############################################################
  
  
  # List of genes to use as input.
  macrophages.2.genes <- unique(rownames(Macrophages_2_sig))
  
  # Retrieve the enriched terms.
  dn.enriched.terms <- enrichR::enrichr(macrophages.2.genes, dbs_use)
  
  # Plot enriched terms 
  
  bp.2 <- plotEnrich(dn.enriched.terms$GO_Biological_Process_2023, showTerms = 20, numChar = 50, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched BPs"))
  cc.2 <- plotEnrich(dn.enriched.terms$GO_Cellular_Component_2023, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched CCs"))
  mf.2 <- plotEnrich(dn.enriched.terms$GO_Molecular_Function_2023, showTerms = 20, numChar = 50, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched MFs"))
  kegg.2 <- plotEnrich(dn.enriched.terms$KEGG_2021_Human, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched KEGG"))
  wiki.2 <- plotEnrich(dn.enriched.terms$WikiPathway_2023_Human, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched Wiki paths"))
  
  
  
  print(bp.1|bp.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_GO_BP.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  print(cc.1|cc.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_GO_CC.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  
  print(mf.1|mf.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_GO_MF.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  
  print(kegg.1|kegg.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_KEGG.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  
  print(wiki.1|wiki.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_WIKI.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  
}

```




## RQ3: Day 4 - what is the treatment difference within neutrophils, macrophages, monocytes {.tabset}

```{r RQ3_day_4_treatment_differences}


# create output dir
output.dir <- paste0("results/CD45Pos_Day_4_RQ3_treatment_difference_within_populations/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}



# reduce dataset to just day 4 timepoint 
Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$Time
day4.seurat <- subset(CD45pos.seurat, idents = "Day4")



if(long.compute){
  
  
  day4.seurat <- PrepSCTFindMarkers(day4.seurat)
  
  # set assay
  DefaultAssay(day4.seurat) <- "SCT"
  
  
  # create new cell_type plus treatment 
  day4.seurat@meta.data$cell_type_treat <- paste0(day4.seurat@meta.data$cell_type, "_", day4.seurat@meta.data$Treatment)
  
  # set Idents
  Idents(day4.seurat) <- "cell_type_treat"
  
  
  # cell types to perform analysis for 
  celltype.vec <- c("Macrophages", "DC", "Monocytes", "Neutrophils")
  
  # create list to store output
  DEG.list <- list()
  
  # run through DEG specifically between vehicle and treatment for each cluster 
  
  for(i in seq_along(celltype.vec)){
    
    temp.de <- FindMarkers(day4.seurat,
                           assay = "SCT",
                           ident.1 = paste0(celltype.vec[i], "_PDS"),
                           ident.2 = paste0(celltype.vec[i], "_Vehicle"),
                           only.pos = FALSE, 
                           recorrect_umi = FALSE,
                           verbose = TRUE)
    
    DEG.list[[i]] <- temp.de
    
    # save to file 
    write.csv(temp.de, paste0(output.dir, "DEGs_Day_4_PDS_", celltype.vec[i], "_vs_Vehicle_", celltype.vec[i], ".csv"))
    
    
  }
  # add names to DEG.list
  names(DEG.list) <- celltype.vec
  
  # save DEG list to file 
  saveRDS(DEG.list, paste0(output.dir, "DEG_Day_4_PDS_vs_Vehicle_celltypes_list_object.rds"))
  
    rm(day4.seurat)

}else{
  
  DEG.list <- readRDS(paste0(output.dir, "DEG_Day_4_PDS_vs_Vehicle_celltypes_list_object.rds"))
  
}


# plot DEGs 


if(do.plot){
  
  for(i in 1:length(DEG.list)){
    
    FDR.theshold <- 0.01
    logFC.threshold <- 1
    
    # filter data
    input.data <- DEG.list[[i]] %>%
      dplyr::rename(logFC = avg_log2FC) %>%
      dplyr::rename(FDR = p_val_adj) %>%
      filter(FDR < FDR.theshold)
    
    # color points
    volcano.col <- usefulfunctions::colour.points.volcano(input.data,
                                                          increase.col = "Red",
                                                          decreased.col = "Blue",
                                                          FDR.cutoff = FDR.theshold, 
                                                          logFC.cutoff = logFC.threshold)
    # plot heatmap
    print(
      usefulfunctions::custom.enhanced.volcano(input.data, 
                                               pCutoff = FDR.theshold,
                                               FCcutoff = logFC.threshold,
                                               title = paste0(names(DEG.list)[i], " PDS vs Vehicle"),
                                               subtitle = paste0("adjPval threshold = ", FDR.theshold, " & log2FC threshold = ", logFC.threshold),
                                               colCustom = volcano.col)
    )
    dev.copy(pdf, paste0(output.dir, "VolcanoPlot_Day_4_PDS_vs_Vehicle_cluster_", names(DEG.list)[i], ".pdf"))
    dev.off()
    
    
  }
  
}



```



## RQ4: Day 9 - what is the treatment difference within neutrophils, macrophages, monocytes, T-cells {.tabset}

```{r RQ4_treatment_difference_Day9_celltypes}


# create output dir
output.dir <- paste0("results/CD45Pos_Day_9_RQ4_treatment_difference_within_populations/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}



# reduce dataset to just day 4 timepoint 
Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$Time
day9.seurat <- subset(CD45pos.seurat, idents = "Day9")



if(long.compute){
  
  
  day9.seurat <- PrepSCTFindMarkers(day9.seurat)
  
  # set assay
  DefaultAssay(day9.seurat) <- "SCT"
  
  
  # create new cell_type plus treatment 
  
  # currently tregs is grouped with t-cells, split this back out 
  
  day9.seurat@meta.data <- day9.seurat@meta.data %>%
    dplyr::mutate(cell_type = case_when(cluster_anno == "Tregs" ~ "Tregs", 
                                        TRUE ~ cell_type))
  
  table(day9.seurat@meta.data$cell_type)
  
  day9.seurat@meta.data$cell_type_treat <- paste0(day9.seurat@meta.data$cell_type, "_", day9.seurat@meta.data$Treatment)
  
  # set Idents
  Idents(day9.seurat) <- "cell_type_treat"
  
  
  # cell types to perform analysis for 
  celltype.vec <- c("Macrophages", "DC", "Monocytes", "Neutrophils", "Tcells", "NKcells")
  
  # create list to store output
  DEG.list <- list()
  
  # run through DEG specifically between vehicle and treatment for each cluster 
  
  for(i in seq_along(celltype.vec)){
    
    temp.de <- FindMarkers(day9.seurat,
                           assay = "SCT",
                           ident.1 = paste0(celltype.vec[i], "_PDS"),
                           ident.2 = paste0(celltype.vec[i], "_Vehicle"),
                           only.pos = FALSE, 
                           recorrect_umi = FALSE,
                           verbose = TRUE)
    
    DEG.list[[i]] <- temp.de
    
    # save to file 
    write.csv(temp.de, paste0(output.dir, "DEGs_Day_9_PDS_", celltype.vec[i], "_vs_Vehicle_", celltype.vec[i], ".csv"))
    
    
  }
  # add names to DEG.list
  names(DEG.list) <- celltype.vec
  
  # save DEG list to file 
  saveRDS(DEG.list, paste0(output.dir, "DEG_Day_9_PDS_vs_Vehicle_celltypes_list_object.rds"))
  
  rm(day9.seurat)
  
}else{
  
  DEG.list <- readRDS(paste0(output.dir, "DEG_Day_9_PDS_vs_Vehicle_celltypes_list_object.rds"))
  
}


# plot DEGs 


if(do.plot){
  
  for(i in 1:length(DEG.list)){
    
    FDR.theshold <- 0.01
    logFC.threshold <- 1
    
    # filter data
    input.data <- DEG.list[[i]] %>%
      dplyr::rename(logFC = avg_log2FC) %>%
      dplyr::rename(FDR = p_val_adj) %>%
      filter(FDR < FDR.theshold)
    
    # color points
    volcano.col <- usefulfunctions::colour.points.volcano(input.data,
                                                          increase.col = "Red",
                                                          decreased.col = "Blue",
                                                          FDR.cutoff = FDR.theshold, 
                                                          logFC.cutoff = logFC.threshold)
    # plot heatmap
    print(
      usefulfunctions::custom.enhanced.volcano(input.data, 
                                               pCutoff = FDR.theshold,
                                               FCcutoff = logFC.threshold,
                                               title = paste0(names(DEG.list)[i], " PDS vs Vehicle"),
                                               subtitle = paste0("adjPval threshold = ", FDR.theshold, " & log2FC threshold = ", logFC.threshold),
                                               colCustom = volcano.col)
    )
    dev.copy(pdf, paste0(output.dir, "VolcanoPlot_Day_9_PDS_vs_Vehicle_cluster_", names(DEG.list)[i], ".pdf"))
    dev.off()
    
    
  }
  
}


```


## RQ5: DEG Neutrophils PDS Day 9 vs Neutrophils PDS Day 4 {.tabset}

```{r RQ5_D9_Neutrophils_PDS_vs_D4_Neutrophils_PDS}


# create output dir
output.dir <- paste0("results/CD45Pos_RQ5_D9_Neutrophils_PDS_vs_D4_Neutrophils_PDS/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}





# reduce dataset to just day 4 timepoint 
Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$cell_type
neutrophil.seurat <- subset(CD45pos.seurat, idents = "Neutrophils")



if(long.compute){
  
  
  neutrophil.seurat <- PrepSCTFindMarkers(neutrophil.seurat)
  
  # set assay
  DefaultAssay(neutrophil.seurat) <- "SCT"
  
  
  # create new time plus treatment 
 
  neutrophil.seurat@meta.data$Time_treat <- paste0(neutrophil.seurat@meta.data$Time, "_", neutrophil.seurat@meta.data$Treatment)
  
  # set Idents
  Idents(neutrophil.seurat) <- "Time_treat"
  
  
  
    temp.de <- FindMarkers(neutrophil.seurat,
                           assay = "SCT",
                           ident.1 = "Day9_PDS",
                           ident.2 = "Day4_PDS",
                           only.pos = FALSE, 
                           recorrect_umi = FALSE,
                           verbose = TRUE)
    
    # save to file 
    write.csv(temp.de, paste0(output.dir, "DEGs_Day_9_PDS_neutrophils_vs_Day_4_PDS_neutrophils.csv"))
    
    
  
}


# plot DEGs 


if(do.plot){
  
    FDR.theshold <- 0.01
    logFC.threshold <- 1
    
    # filter data
    input.data <- temp.de %>%
      dplyr::rename(logFC = avg_log2FC) %>%
      dplyr::rename(FDR = p_val_adj) %>%
      filter(FDR < FDR.theshold)
    
    # color points
    volcano.col <- usefulfunctions::colour.points.volcano(input.data,
                                                          increase.col = "Red",
                                                          decreased.col = "Blue",
                                                          FDR.cutoff = FDR.theshold, 
                                                          logFC.cutoff = logFC.threshold)
    # plot heatmap
    print(
      usefulfunctions::custom.enhanced.volcano(input.data, 
                                               pCutoff = FDR.theshold,
                                               FCcutoff = logFC.threshold,
                                               title = paste0("Neutrophils Day 9 PDS vs Day 4 PDS"),
                                               subtitle = paste0("adjPval threshold = ", FDR.theshold, " & log2FC threshold = ", logFC.threshold),
                                               colCustom = volcano.col)
    )
    dev.copy(pdf, paste0(output.dir, "VolcanoPlot_Day_9_neutrophils_PDS_vs_Day_4_neutrophils_PDS.pdf"))
    dev.off()
    
    
  
    
  #####################
  # heatmaps of top DEGs
  #####################
  
  
  
  Idents(neutrophil.seurat) <- neutrophil.seurat@meta.data$Time_treat
  
  

  pseudo.neutrophils <- AggregateExpression(neutrophil.seurat,
                                           assays = "SCT",
                                           return.seurat = TRUE, 
                                           group.by = c("Time_treat"))
  
  
  # Extract top fold change DEGs 
  
  goi <- temp.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(abs(avg_log2FC) > 2) %>%
    dplyr::arrange(desc(avg_log2FC)) %>%
    rownames()
  
  
  # plot heatmap
  DoHeatmap(pseudo.neutrophils, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_Heatmap_Day_9_PDS_neutrophils_vs_Day_4_PDS_neutrophils_logfold_2.pdf"))
  dev.off()
  
  
  # extract just top 10 up and down degs
  goi.up <- temp.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::top_n(10, avg_log2FC) %>%
    rownames()
  
  goi.dn <- temp.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::top_n(-10, avg_log2FC) %>%
    rownames()
  
  goi <- unique(c(goi.up, goi.dn))
  
  DoHeatmap(pseudo.neutrophils, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_Heatmap_Day_9_PDS_neutrophils_vs_Day_4_PDS_neutrophils_top_10_DEGs_up_down.pdf"))
  dev.off()
  
}
  


# clean environment 
  rm(neutrophil.seurat)
  gc()

```





## RQ6: Zoom in on T-cells, can identify Cytotoxic CD4 T-cells? {.tabset}

```{r Cytotoxic_CD4_Tcells}


# create output dir
output.dir <- paste0("results/CD45Pos_Cytotoxic_CD4_Tcells/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# currently tregs is grouped with t-cells, split this back out 
CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
  dplyr::mutate(cell_type = case_when(cluster_anno == "Tregs" ~ "Tregs", 
                                      TRUE ~ cell_type))


# reduce dataset to just day 4 timepoint 
Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$cell_type
Tcell.seurat <- subset(CD45pos.seurat, idents = c("Tcells", "Tregs"))

DefaultAssay(Tcell.seurat) <- "RNA"

Tcell.seurat <- FindVariableFeatures(Tcell.seurat, assay = "RNA")

imputation.features <- VariableFeatures(Tcell.seurat)

# ensure cd8 and cd4 key markers are included 
imputation.features <- c(paste0(imputation.features), "Cd4", "Cd8a", "Cd8b1", "Foxp3")
imputation.features <- unique(imputation.features)

DefaultAssay(Tcell.seurat) <- "SCT"



######################
# Impute values 
######################
if(long.compute){
  Tcell.seurat <- RunALRA(Tcell.seurat, 
                             genes.use = imputation.features)
}


DefaultAssay(Tcell.seurat) <- "SCT"


# recalculate UMAP on just T-cell population 
Tcell.seurat <- RunUMAP(Tcell.seurat, 
                        reduction = "integrated.cca.sct",
                        dims = 1:30)
  
# find clusters
Tcell.seurat <- FindClusters(Tcell.seurat, 
                          resolution = 1, 
                          graph.name = "SCT_snn")  

table(Tcell.seurat@meta.data$seurat_clusters)

  

Seurat::DimPlot(object = Tcell.seurat, 
                  reduction = "umap", 
                  label = TRUE, 
                  label.size = 4, 
                  group.by = "cell_type") +
    NoLegend()
  
Seurat::DimPlot(object = Tcell.seurat, 
                  reduction = "umap", 
                  label = TRUE, 
                  label.size = 4, 
                  group.by = "seurat_clusters") +
    NoLegend()
  


  grep("Cd8.*", rownames(Tcell.seurat), value = T)

  # visualise gene expression - use imputed values 
  DefaultAssay(Tcell.seurat) <- "alra"

  Stacked_VlnPlot(Tcell.seurat, 
                features = c("Cd8a", "Cd8b1", "Cd4", "Gzmb", "Gzma", "Nkg7", "Foxp3"))
  
  
  
  scCustomize::FeaturePlot_scCustom(Tcell.seurat, reduction = "umap", feature = "Cd4")
    scCustomize::FeaturePlot_scCustom(Tcell.seurat, reduction = "umap", feature = "Gzmb")

  
    
  
  
  Seurat::DimPlot(object = Tcell.seurat, 
                  reduction = "umap", 
                  label = TRUE, 
                  label.size = 4, 
                  group.by = "SampleID") +
    NoLegend()




```




## Imputation {.tabset}

### Imputation calculation

```{r seurat_imputation, eval = FALSE}

output.dir <- "results/scRNAseq/figures/QC/"

# Create output directory
if(!dir.exists(paste0(output.dir))){
  dir.create(paste0(output.dir), 
             recursive = T)
}

######################
# Impute values 
######################
if(long.compute){
  seurat.combined <- RunALRA(seurat.combined, 
                             genes.use = rownames(seurat.combined))
}

# 8.37% of the values became negative in the scaling process and were set to zero
# The matrix went from 29.63% nonzero to 38.95% nonzero


####################
# Calculate k value
####################
if(long.compute){
  
  Idents(seurat.combined) <- seurat.combined@meta.data$seurat_clusters
  seurat.combined.small <- subset(seurat.combined, downsample = 100)
  
  ALRA.out <- RunALRA(seurat.combined.small,
                      k.only = TRUE)
  
  ggouts <- ALRAChooseKPlot(ALRA.out)
  
  ggouts
  dev.copy(pdf, paste0(output.dir, "Imputation_K_plot.pdf"))
  dev.off()
  
}


# set default assay back to RNA
DefaultAssay(seurat.combined) <- "RNA"
Idents(seurat.combined) <- seurat.combined@meta.data$seurat_clusters

```
