---
title: "CD45Pos_dataset_analysis"
author: "Dillon Corvino"
date: "25/01/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# Data 
# B16 - OVA 

# PDS = I.T - bind Gquadraplex
# day 4 post-treatment 
# day 9 post-treatment

# Research questions 
# How does PDS induces an anti-tumor immune response?
# Known that after PDS treatment, certain DAMPs are induced, intratumoral chemokine levels go up drastically and subsequntly immune cells are recruited into the tumor.
# Main research questions are:
#1. Which cell subsets are producing the chemokines
#2. Are there differences between macrophages and DCs between PDS and Vehicle (because there are DAMPs that should activate them)
#3. How does the treatment effect the tumor cells (is it possible to discriminate them from other tissue cells, if not they express ovalbumin...this should make it at least possible to identify them.)




# key genes
# IL1b, IL6, IL12, IFNG, TNFA, 

# CCL2,3,4,5 
# CXCL1, CXCL9,10,13 





```


### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)


# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- FALSE
do.plot <- TRUE
quick.load <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())


library("Seurat")
library("SeuratObject")
library("SeuratData")
library("SeuratWrappers")
library("dplyr")
library("scCustomize")
library("tibble")


# load output from preprocessing script
if(long.compute){
  
  CD45pos.seurat <- LoadSeuratRds("saves/Preprocessed_CD45Pos.rds")
  
  # set levels for easier downstream plotting 
  CD45pos.seurat@meta.data$Treatment <- factor(CD45pos.seurat@meta.data$Treatment, levels = c("Vehicle", "PDS"))
  CD45pos.seurat@meta.data$Time <- factor(CD45pos.seurat@meta.data$Time, levels = c("Day4", "Day9"))
  CD45pos.seurat@meta.data$SampleID <- factor(CD45pos.seurat@meta.data$SampleID, levels = c("Day4_Vehicle_CD45Pos", "Day4_PDS_CD45Pos", "Day9_Vehicle_CD45Pos", "Day9_PDS_CD45Pos"))
  
}else if(quick.load){
  
  # load cleaned dataset 
  CD45pos.seurat <- LoadSeuratRds("saves/CD45pos_seurat.rds")
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
}


```



## Cluster annotation {.tabset}

### SingleR cluster annotation
```{r SingleR_cluster_annotation}

# Notes:
# there is an error with collect() when trying to load celldex datasets
# i think its an error with biocmanager and has been fixed with a very recent update. > 3.16 
# but im running r 4.2.3 and this would require 4.3
# not keen to upgrade right now 
# so will try a work around of downgrading dbplyr to version 2.3.4 from 2.4 which im currently running 
# need to downgrade to this version 
#devtools::install_version("dbplyr", version = "2.3.4")
#devtools::install_version("dbplyr", version = "2.4")
# in theory if biocmanager is up to date, could just run below and should fix the problem
#library(BiocManager) # 3.16
#install("BiocFileCache")
# also:
# need to ensure matrixStats is at version 1.1.0 
#remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts
#BiocManager::install(version = "3.18")



if(long.compute){
  
  output.dir <- "results/CD45Pos_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # load required packages
  library("SingleR")
  library("celldex")
  
  
  ##################################
  # Using MouseRNAseqData
  ##################################
  
  ref.mouse <-  celldex::MouseRNAseqData()
  
  
  # Benayoun Lab that identified, downloaded and processed data sets on GEO that corresponded to sorted cell types (Benayoun et al., 2019).
  
  # The dataset contains 358 mouse RNA-seq samples annotated to 18 main cell types
  # further split into 28 subtypes under label.fine 
  
  
  ######################
  # Fine labels 
  ######################
  SingleR.pred <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45pos.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.fine)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
  #SingleR::plotDeltaDistribution(SingleR.pred)
  #SingleR::plotScoreDistribution(SingleR.pred)
  
  ######################
  # Main labels 
  ######################
  
  SingleR.pred <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45pos.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.main)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_main_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
  
  #####################
  # Use ImmGenData
  #####################
  # Info: Mouse, Microarray, 830 samples, 20 main labels, 253 fine lables, Hematopoietic & Immune focus
  
  ImmGen.data <- celldex::ImmGenData()
  
  
  # Fine labels
  SingleR.pred.fine <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                               ref = ImmGen.data, 
                               method = "cluster",
                               clusters = CD45pos.seurat@meta.data$seurat_clusters,
                               labels = ImmGen.data$label.fine)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred.fine, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_ImmGen_db.pdf"), plot = p1)
  
  
  # Main labels
  SingleR.pred.main <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                               ref = ImmGen.data, 
                               method = "cluster",
                               clusters = CD45pos.seurat@meta.data$seurat_clusters,
                               labels = ImmGen.data$label.main)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred.main, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_main_labels_ImmGen_db.pdf"), plot = p1)
  
  
}


```


### Transfer SingleR annotations to seurat object 
```{r transfer_singleR_annotations}


if(long.compute){
  
  output.dir <- "results/CD45Pos_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  
  ########################################
  # transfer labels to seurat object 
  ########################################
  
  table(predicted=SingleR.pred.main$labels, cluster=rownames(SingleR.pred.main))
  table(predicted=SingleR.pred.fine$labels, cluster=rownames(SingleR.pred.fine))
  
  # Main labels
  main.labels <- SingleR.pred.main$labels
  names(main.labels) <- rownames(SingleR.pred.main)
  
  # Fine labels
  fine.labels <- SingleR.pred.fine$labels
  names(fine.labels) <- rownames(SingleR.pred.fine)
  
  ####################
  # rename clusters 
  ####################
  
  # Main labels
  Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$seurat_clusters
  CD45pos.seurat <- Rename_Clusters(CD45pos.seurat, new_idents = main.labels)
  CD45pos.seurat@meta.data$SingleR_main <- Idents(CD45pos.seurat)
  
  # Fine labels
  Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$seurat_clusters
  CD45pos.seurat <- Rename_Clusters(CD45pos.seurat, new_idents = fine.labels)
  CD45pos.seurat@meta.data$SingleR_fine <- Idents(CD45pos.seurat)
  
  
  ############
  # Visualise
  ############
  
  # Main label
  p.main <- scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                                          group.by = "SingleR_main", 
                                          label = TRUE,
                                          reduction = "umap")
  
  print(p.main)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_main.pdf"))
  dev.off()
  
  
  # Fine label
  p.fine <- scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                                          group.by = "SingleR_fine", 
                                          label = TRUE,
                                          reduction = "umap")
  
  print(p.fine)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_fine.pdf"))
  dev.off()
  
  
  # side-by-side with cluster numbers
  p.clusters <- scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                                              group.by = "seurat_clusters", 
                                              label = TRUE,
                                              reduction = "umap")
  
  # clusters plus main
  print(p.clusters | p.main)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_main_plus_clusters.pdf"))
  dev.off()
  
  # clusters plus fine
  print(p.clusters | p.fine)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_fine_plus_clusters.pdf"))
  dev.off()
  
  
}










```


### Verify cluster annotation
```{r verify_and_finalise_cluster_annotation}


if(long.compute){
  
  output.dir <- "results/CD45Pos_SingleR_annotation_validation/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # overview of data 
  table(CD45pos.seurat@meta.data$seurat_clusters, CD45pos.seurat@meta.data$SingleR_fine)
  
  table(CD45pos.seurat@meta.data$seurat_clusters, CD45pos.seurat@meta.data$SingleR_main)
  table(CD45pos.seurat@meta.data$SingleR_main, CD45pos.seurat@meta.data$SingleR_fine)
  
  # inconsistency between main and fine annotation with respect to T-cell populations
  # Tcells in main are annotated as NKT
  # NKT in main are annotated as treg 
  
  
  # also seurat clusters nicely line up with singleR fine annotations, i.e one cluster per annotation 
  # only monocytes is with higher resolution using seurat clusters, therefore consider seperating into monocytes cluster X and cluster Y
  
  
  # look at annotations and simplify 
  
  
  #################################################
  # calculate DEGs using fine level resolution 
  #################################################
  
  CD45pos.seurat <- PrepSCTFindMarkers(CD45pos.seurat)
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  # set idents
  Idents(CD45pos.seurat) <- "SingleR_fine"
  
  CD45Pos.de <- FindAllMarkers(CD45pos.seurat,
                               assay = "SCT",
                               only.pos = TRUE, 
                               verbose = TRUE)
  
  
  # save to file 
  write.csv(CD45Pos.de, paste0(output.dir, "CD45Pos_DEGs_singleR_fine.csv"))
  
  
  ##############
  # plot DEGs
  ##############
  
  pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("SingleR_fine"))
  
  
  
  goi <- CD45Pos.de %>%
    filter(p_val_adj < 0.05) %>%
    group_by(cluster) %>%
    top_n(10, avg_log2FC) %>%
    pull(gene)
  
  
  DoHeatmap(pseudo.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_top_10_singleR_fine_DEGs_pseudobulk.pdf"))
  dev.off()
  
  # populations are very cleanly resolved 
  
  
  
  
  # iterate over the top degs for each cluster 
  cluster.id <- unique(CD45Pos.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- CD45Pos.de %>%
      filter(p_val_adj < 0.05) %>%
      filter(cluster == cluster.id[i]) %>%
      top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.cd45pos, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE) + 
        scico::scale_fill_scico(palette = "batlow", 
                                direction = 1, 
                                na.value = "white")
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
  
  # happy with using fine level annotation 
  # except, defining NKT from T-cells is a challenge. will just call them T-cells as at main level annotation these are T-cells
  # NKTs at fine level annotation are split into Treg and Teff - and these look to make sense
  # will therefore use fine level annotation fo all, but change NKT to just T-cell 
  # will also simplify the annotation for all clusters but keep their distinct'ness 
  
  
}






```


### Clean up seurat object metadata and export
```{r clean_seurat_object_metadata}

if(long.compute){
  # reannotate singleR_fine cluster annotations based on above analysis 
  # namely, simplify the annotations 
  # also, convert NKT to just T-cells
  
  CD45pos.seurat <- backup.object.for.now
  
  # note the clusters are essentially one cell type one cluster. except for macrophages which is present in two clusters of macrophages and monocytes present in 3 clusters
  table(CD45pos.seurat@meta.data$SingleR_fine, CD45pos.seurat@meta.data$seurat_clusters)
  
  library(stringr) # For using the `str_extract` function
  
  
  # simplify the annotations, to just the cell_type, but keep cluster_seperation 
  CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
    # Annotate clusters based on pattern matching
    mutate(cluster_anno = case_when(
      # Use `str_extract` to extract the cell type from SingleR_fine
      TRUE ~ paste0(
        str_extract(SingleR_fine, "^[^(]+"), # Extract the cell type name
        "_", seurat_clusters
      )
    )) %>%
    # Remove any spaces in the cluster_anno name
    mutate(cluster_anno = str_replace_all(cluster_anno, " ", ""))
  
  unique(CD45pos.seurat@meta.data$cluster_anno)
  table(CD45pos.seurat@meta.data$SingleR_fine, CD45pos.seurat@meta.data$cluster_anno)
  
  
  
  # simplify the annotations, so there are numbered within cell_type
  
  
  
  # little code to count up how many distinct cell types there are 
  CD45pos.seurat@meta.data %>%
    dplyr::select(c(cluster_anno, SingleR_fine)) %>%
    # Extract cell_type from cluster_anno
    mutate(cell_type = str_extract(cluster_anno, "^[^_]+")) %>%
    # Group by cell_type
    group_by(cell_type) %>%
    # Calculate the number of unique annotations for each cell_type and rank them
    mutate(
      # Calculate the count of unique cluster_anno
      unique_count = n_distinct(cluster_anno) ) %>%
    ungroup() %>%
    distinct(cluster_anno, .keep_all = TRUE)
  
  
  
  CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
    # Extract cell_type from cluster_anno
    mutate(cell_type = str_extract(cluster_anno, "^[^_]+")) %>%
    mutate(cluster_anno_new = case_when(
      cluster_anno == "Macrophages_1" ~ "Macrophages_1", 
      cluster_anno == "Macrophages_9" ~ "Macrophages_2", 
      
      cluster_anno == "Bcells_8" ~ "B_cells", 
      
      cluster_anno == "DC_15" ~ "pDC", 
      cluster_anno == "DC_14" ~ "DC_1", 
      cluster_anno == "DC_12" ~ "DC_2",
      
      cluster_anno == "Monocytes_4" ~ "Monocytes_1", 
      cluster_anno == "Monocytes_5" ~ "Monocytes_2", 
      cluster_anno == "Monocytes_7" ~ "Monocytes_3", 
      
      cluster_anno == "Macrophages_1" ~ "Macrophages_1", 
      cluster_anno == "Macrophages_9" ~ "Macrophages_2", 
      
      cluster_anno == "Neutrophils_3" ~ "Neutrophils_1", 
      cluster_anno == "Neutrophils_6" ~ "Neutrophils_2", 
      
      cluster_anno == "Stromalcells_16" ~ "Stromal_cells", 
      
      cluster_anno == "NKcells_2" ~ "NK_cells", 
      
      cluster_anno == "Basophils_17" ~ "Basophils", 
      
      cluster_anno == "Tcells_0" ~ "T_cells_1", 
      cluster_anno == "Tcells_10" ~ "T_cells_2", 
      cluster_anno == "Tcells_13" ~ "T_cells_3", 
      cluster_anno == "Tcells_11" ~ "Tregs", 
      TRUE ~ "ERROR")) %>%
    mutate(cluster_anno = cluster_anno_new) %>%
    select(-cluster_anno_new)
  
  
  unique(CD45pos.seurat@meta.data$cluster_anno)
  unique(CD45pos.seurat@meta.data$SingleR_fine)
  
  
  # clean up object, remove unnecessary columns keep only what is needed
  CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
    select(c("orig.ident", "nCount_RNA", "nFeature_RNA", "nCount_SCT", "nFeature_SCT", "percent_mito", "percent_ribo", "percent_mito_ribo", "log10GenesPerUMI",  "SampleID", "Time", "Compartment", "Treatment",  "cluster_anno", "seurat_clusters", "SCT_snn_res.0.4", "SingleR_main", "SingleR_fine", "cell_type"))
  
  
  # Display the first few rows of the updated data frame to verify the results
  head(as.data.frame(CD45pos.seurat@meta.data[1:10, ]))
  
  
  
  table(CD45pos.seurat@meta.data$SingleR_fine, CD45pos.seurat@meta.data$cell_type)
  table(CD45pos.seurat@meta.data$SingleR_fine, CD45pos.seurat@meta.data$cluster_anno)
  table(CD45pos.seurat@meta.data$cell_type, CD45pos.seurat@meta.data$cluster_anno)
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  # clean up sample ID annotation - no need to say CD45Pos still 
  
  CD45pos.seurat@meta.data$SampleID <- gsub("_CD45Pos", "", CD45pos.seurat@meta.data$SampleID)
  
  unique(CD45pos.seurat@meta.data$cluster_anno)
  head(CD45pos.seurat@meta.data)
  
  # set factorisation 
  CD45pos.seurat@meta.data$Treatment <- factor(CD45pos.seurat@meta.data$Treatment, 
                                               levels = c("Vehicle", "PDS"))
  
  CD45pos.seurat@meta.data$Time <- factor(CD45pos.seurat@meta.data$Time, 
                                          levels = c("Day4", "Day9"))
  
  CD45pos.seurat@meta.data$SampleID <- factor(CD45pos.seurat@meta.data$SampleID, 
                                              levels = c("Day4_Vehicle", "Day9_Vehicle", "Day4_PDS", "Day9_PDS"))
  
  
  CD45pos.seurat@meta.data$cluster_anno <- factor(CD45pos.seurat@meta.data$cluster_anno)
  
  # save output
  SaveSeuratRds(CD45pos.seurat, file = "saves/CD45pos_seurat.rds")
}


```



## Overview of dataset {.tabset}

### Broad overview of dataset
```{r broad_overview}

if(do.plot){
  
  
  # create output dir
  output.dir <- paste0("results/CD45Pos_Overview/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  vars.to.plot <- c("Treatment", "SampleID", "cluster_anno", "Time", "cell_type")
  
  
  for(j in seq_along(vars.to.plot)){
    name.var <- gsub("\\.", "_", vars.to.plot[j])
    
    print(
      scCustomize::DimPlot_scCustom(CD45pos.seurat,
                                    reduction = "umap", 
                                    label = TRUE,
                                    figure_plot = TRUE,
                                    group.by = vars.to.plot[j]) 
    )
    
    
    dev.copy(pdf, paste0(output.dir, "CD45Pos_UMAP_", name.var, ".pdf"))
    dev.off()
    
  }
  
  
  # for split comparison plots, downsample to make visual comparisons more fair 
  table(CD45pos.seurat@meta.data$Treatment)
  Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$Treatment
  temp.seurat <- subset(CD45pos.seurat, downsample = 15000)
  # split umap plots
  print(
    scCustomize::DimPlot_scCustom(temp.seurat,
                                  reduction = "umap", 
                                  label = TRUE,
                                  figure_plot = TRUE,
                                  group.by = "cluster_anno", 
                                  split.by = "Treatment") 
  )
  
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_UMAP_cluster_anno_split_by_treatment.pdf"))
  dev.off()
  
  
  table(CD45pos.seurat@meta.data$SampleID)
  Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$SampleID
  temp.seurat <- subset(CD45pos.seurat, downsample = 5000)
  # split umap plots
  print(
    scCustomize::DimPlot_scCustom(temp.seurat,
                                  reduction = "umap", 
                                  label = FALSE,
                                  figure_plot = TRUE,
                                  group.by = "cluster_anno", 
                                  split.by = "SampleID") 
  )
  
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_UMAP_cluster_anno_split_by_SampleID.pdf"))
  dev.off()
  
  rm(temp.seurat)
  
}



```


### Cluster stats
```{r cluster_stats}

# create output dir
output.dir <- paste0("results/CD45Pos_Overview/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

#############################################
# Number of cells per cluster
#############################################

cluster.n <- table(CD45pos.seurat@meta.data$cluster_anno)

cluster.n <- as.data.frame(cluster.n)

cluster.n$Var1 <- factor(cluster.n$Var1, levels = cluster.n[order(cluster.n$Freq, decreasing = T), ]$Var1)


ggplot(cluster.n) +
  aes(x = Var1, y = Freq, fill = Var1) +
  geom_col() +
  scale_fill_viridis_d(option = "viridis", 
                       direction = -1) +
  labs(x = "Cluster", y = "Number of cells", title = "Number of cells per cluster") +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

ggsave(paste0(output.dir, "Barplot_number_of_cells_per_cluster.pdf"))



#############################################
# Freq of cluster across treatment 
#############################################

dist.table <- table(CD45pos.seurat@meta.data$cluster_anno, CD45pos.seurat@meta.data$Treatment)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Vehicle", "PDS"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Treatment", y = "% of treatment", title = "Frequency of cluster per treatment") +
  scale_fill_manual(values = c(Vehicle = "grey", PDS = "darkred")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_cluster_per_treatment.pdf"))


#############################################
# Freq of clusters per treatment across time
#############################################


#######################################
#  Ordered by treatment
#######################################

dist.table <- table(CD45pos.seurat@meta.data$cluster_anno, CD45pos.seurat@meta.data$SampleID)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Day4_Vehicle", "Day9_Vehicle", "Day4_PDS", "Day9_PDS"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Treatment", y = "% of treatment", title = "Frequency of cluster per treatment") +
  scale_fill_manual(values = c(Day4_Vehicle = "lightgreen", Day9_Vehicle = "darkgreen", Day4_PDS = "red", Day9_PDS = "darkred")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_cluster_per_treatment_across_time.pdf"))



#######################################
#  Ordered by time
#######################################

dist.table <- table(CD45pos.seurat@meta.data$cluster_anno, CD45pos.seurat@meta.data$SampleID)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Day4_Vehicle", "Day4_PDS", "Day9_Vehicle", "Day9_PDS"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Treatment", y = "% of treatment", title = "Frequency of cluster per treatment") +
  scale_fill_manual(values = c(Day4_Vehicle = "lightgreen", Day9_Vehicle = "darkgreen", Day4_PDS = "red", Day9_PDS = "darkred")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_cluster_per_treatment_across_time_ordered_time.pdf"))


```

### Cell Cycle scoring
```{r cell_cycle}

# create output dir
output.dir <- paste0("results/CD45Pos_cell_cycle/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

################################
# Get cell cycle genes 
################################
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

# pseudo-map to mouse gene (captures most except Mlf1ip, Fam64a and Hn1)
s.genes <- stringr::str_to_title(s.genes)
g2m.genes <- stringr::str_to_title(g2m.genes)

#####################
# score cell cycle
#####################

CD45pos.seurat <- CellCycleScoring(CD45pos.seurat,
                                   s.features = s.genes,
                                   g2m.features = g2m.genes,
                                   set.ident = TRUE)


#####################
# Visualise score
#####################

head(CD45pos.seurat@meta.data)

# Feature plots 

# S-score 
scCustomize::FeaturePlot_scCustom(CD45pos.seurat, 
                                  feature = "S.Score", 
                                  reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Pos_S_score.pdf"))
dev.off()

# G2M-score 
scCustomize::FeaturePlot_scCustom(CD45pos.seurat, 
                                  feature = "G2M.Score", 
                                  reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Pos_G2M_score.pdf"))
dev.off()

# Dim plot split by phase
scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                              group.by = "cluster_anno",
                              split.by = "Phase",
                              label = TRUE,
                              reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_CD45Pos_cluster_by_phase.pdf"))
dev.off()


# Stacked VlnPlot
scCustomize::Stacked_VlnPlot(CD45pos.seurat,
                             features = c("S.Score", "G2M.Score"), 
                             group.by = "cluster_anno")

dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Pos_cell_cycle.pdf"))
dev.off()




###################
# Quantification 
###################


# Frequency of phase per cluster
dist.table <- table(CD45pos.seurat@meta.data$cluster_anno, CD45pos.seurat@meta.data$Phase)

cluster.total <- rowSums(dist.table)

freq.table <- sweep(dist.table, 1, cluster.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin
rowSums(freq.table)

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("G1", "G2M", "S"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Predicted phase", y = "% of cluster", title = "Frequency of phase per cluster") +
  scale_fill_manual(values = c(G1 = "lightblue", G2M = "orange", S = "red")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_phase_per_cluster.pdf"))


# Frequency of phase per sampleID

dist.table <- table(CD45pos.seurat@meta.data$SampleID, CD45pos.seurat@meta.data$Phase)

treatment.total <- rowSums(dist.table)

freq.table <- sweep(dist.table, 1, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin
rowSums(freq.table)

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("G1", "G2M", "S"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Predicted phase", y = "% of treatment", title = "Frequency of phase per sampleID") +
  scale_fill_manual(values = c(G1 = "lightblue", G2M = "orange", S = "red")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_phase_per_sampleID.pdf"))

```


## DEG analysis {.tabset}

### Cluster-wise DEGs - calculate
```{r Clusterwise_DEGs_calculation}


# create output dir
output.dir <- paste0("results/CD45Pos_cluster_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  
  CD45pos.seurat <- PrepSCTFindMarkers(CD45pos.seurat)
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  # set idents
  Idents(CD45pos.seurat) <- "cluster_anno"
  
  CD45Pos.de <- FindAllMarkers(CD45pos.seurat,
                               assay = "SCT",
                               only.pos = TRUE, 
                               verbose = TRUE)
  
  
  # save to file 
  write.csv(CD45Pos.de, paste0(output.dir, "CD45Pos_DEGs.csv"))
  
  
  
}else{
  
  # load DEG table
  CD45Pos.de <- read.csv(paste0(output.dir, "CD45Pos_DEGs.csv"))
  
}



```

### Cluster-wise DEGs - heatmaps
```{r clusterwise_DEGs_heatmaps}

# create output dir
output.dir <- paste0("results/CD45Pos_cluster_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

if(do.plot){
  # aggregate expression for heatmap visualisation
  
  pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("cluster_anno"))
  
  
  
  goi <- CD45Pos.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::group_by(cluster) %>%
    dplyr::top_n(10, avg_log2FC) %>%
    dplyr::pull(gene)
  
  
  DoHeatmap(pseudo.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_top_10_DEGs_pseudobulk.pdf"))
  dev.off()
  
  
  
  
  # iterate over the top degs for each cluster 
  cluster.id <- unique(CD45Pos.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- CD45Pos.de %>%
      dplyr::filter(p_val_adj < 0.05) %>%
      dplyr::filter(cluster == cluster.id[i]) %>%
      dplyr::top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.cd45pos, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE) + 
        scico::scale_fill_scico(palette = "batlow", 
                                direction = 1, 
                                na.value = "white")
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
}

```


### Treatment-wise DEGs - calculate
```{r Treatmentwise_DEGs_calculation}

# create output dir
output.dir <- paste0("results/CD45Pos_treatment_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  
  CD45pos.seurat <- PrepSCTFindMarkers(CD45pos.seurat)
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  
  # add annotation for treatment 
  CD45pos.seurat@meta.data$Treatment_cluster <- paste0(CD45pos.seurat@meta.data$Treatment, "_", CD45pos.seurat@meta.data$cluster_anno)
  
  # set Idents
  Idents(CD45pos.seurat) <- "Treatment_cluster"
  
  
  cluster.ids <- unique(CD45pos.seurat@meta.data$cluster_anno)
  
  # create list to store output
  DEG.list <- list()
  
  # run through DEG specifically between vehicle and treatment for each cluster 
  
  for(i in seq_along(cluster.ids)){
    temp.de <- FindMarkers(CD45pos.seurat,
                           assay = "SCT",
                           ident.1 = paste0("PDS_", cluster.ids[i]),
                           ident.2 = paste0("Vehicle_", cluster.ids[i]),
                           only.pos = FALSE, 
                           recorrect_umi = FALSE,
                           verbose = TRUE)
    
    DEG.list[[i]] <- temp.de
    
    # save to file 
    write.csv(temp.de, paste0(output.dir, "DEGs_PDS_", cluster.ids[i], "_vs_Vehicle_", cluster.ids[i], ".csv"))
    
    
  }
  # add names to DEG.list
  names(DEG.list) <- cluster.ids
  
  # save DEG list to file 
  saveRDS(DEG.list, paste0(output.dir, "DEG_list_object.rds"))
  
  
}else{
  
  DEG.list <- readRDS(paste0(output.dir, "DEG_list_object.rds"))
  
}


```


### Treatment-wise DEGs - Volcano
```{r treatmentwise_DEGs_volcano}


# create output dir
output.dir <- paste0("results/CD45Pos_treatment_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# visualise treatment-wise DEGs as volcano plots


if(do.plot){
  
  for(i in 1:length(DEG.list)){
    
    FDR.theshold <- 0.01
    logFC.threshold <- 1
    
    # filter data
    input.data <- DEG.list[[i]] %>%
      dplyr::rename(logFC = avg_log2FC) %>%
      dplyr::rename(FDR = p_val_adj) %>%
      filter(FDR < FDR.theshold)
    
    # color points
    volcano.col <- usefulfunctions::colour.points.volcano(input.data,
                                                          increase.col = "Red",
                                                          decreased.col = "Blue",
                                                          FDR.cutoff = FDR.theshold, 
                                                          logFC.cutoff = logFC.threshold)
    # plot heatmap
    print(
      usefulfunctions::custom.enhanced.volcano(input.data, 
                                               pCutoff = FDR.theshold,
                                               FCcutoff = logFC.threshold,
                                               title = paste0(names(DEG.list)[i], " PDS vs Vehicle"),
                                               subtitle = paste0("adjPval threshold = ", FDR.theshold, " & log2FC threshold = ", logFC.threshold),
                                               colCustom = volcano.col)
    )
    dev.copy(pdf, paste0(output.dir, "VolcanoPlot_PDS_vs_Vehicle_cluster_", names(DEG.list)[i], ".pdf"))
    dev.off()
    
    
  }
  
}

```


### Export aggregated counts table
```{r export_aggregated_counts}

# create output dir
output.dir <- paste0("results/tables/CD45Pos/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  # Aggregate expression by cluster 
  pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("cluster_anno"))
  
  # Write to file
  temp.out <- GetAssayData(pseudo.cd45pos,
                           layer = "data",
                           assay = "SCT")
  
  write.csv(temp.out, paste0(output.dir, "CD45pos_Aggregate_SCT_data_by_cluster.csv"))
  
  # Aggregate expression by cluster and SampleID
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("cluster_anno", "SampleID"))
  # write to file
  temp.out <- GetAssayData(pseudo.cd45pos,
                           layer = "data",
                           assay = "SCT")
  
  write.csv(temp.out, paste0(output.dir, "CD45Pos_Aggregate_SCT_data_by_cluster_and_SampleID.csv"))
  
  # clean environment
  rm(pseudo.cd45pos)
  rm(temp.out)
  
}



```




# TO DO
# for CD45Neg dataset
# group Tumor together, DEG between vehicle and treatment 

# for CD45Pos dataset
# split by time 
# day 4 - group monocytes, macrophage, neutrophils etc, DEG between vehicle and treatment 
# day 9 group t-cells, group neutrophils, degs between vehcile treatment, 
# neutrophil treatment day 9 vs day 4 neutrophils 
# zoom in on t-cells, can find cytotoxic cd4s?


## RQ1: Which CD45Pos clusters express chemokines or cytokines {.tabset}

### Chemokine and Cytokine expression visualisation 
```{r chemokine_cytokine_expression}

# create output dir
output.dir <- paste0("results/CD45Pos_chemokine_cytokines/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# aggregate expression for heatmap visualisation
DefaultAssay(CD45pos.seurat) <- "SCT"
pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("cluster_anno"))


########################################
# List gene sets of interest to plot
########################################

cxc.genes <- grep("^Cxc.*", rownames(pseudo.cd45pos), value = T)
ccr.genes <- grep("^Ccr.*", rownames(pseudo.cd45pos), value = T)
il.genes <- grep("^Il.*", rownames(pseudo.cd45pos), value = T)
gzm.genes <- grep("^Gzm.*", rownames(pseudo.cd45pos), value = T)
ifn.genes <- grep("^Ifn.*", rownames(pseudo.cd45pos), value = T)
custom.genes <- c("Il1b", "Il6", "Il12a", "Il12b", "Ifng", "Tnf", "Ccl2", "Ccl3", "Ccl4", "Ccl5", "Cxcl1", "Cxcl9", "Cxcl10") # Cxcl13 not present in CD45Pos dataset

goi <- list(Cxcr_genes = cxc.genes, 
            Ccr_genes = ccr.genes,
            Interleukin_genes = il.genes,
            Gzm_genes = gzm.genes,
            Ifn_genes = ifn.genes,
            Custom_genes = custom.genes)


###################
# Plot heatmaps
###################

for(i in seq_along(names(goi))){
  
  print(
    DoHeatmap(pseudo.cd45pos, 
              features = goi[[i]], 
              draw.lines = FALSE, 
              angle = 90,
              raster = FALSE) + 
      scico::scale_fill_scico(palette = "batlow", 
                              direction = 1, 
                              na.value = "white")
  )
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_", names(goi)[i], ".pdf"))
  dev.off()
  
}


##############################
# Stacked vln plot
##############################

for(i in seq_along(names(goi))){
  
  if(length(goi[[i]]) <= 15){
    
    
    print(
      scCustomize::Stacked_VlnPlot(CD45pos.seurat, 
                                   features = goi[[i]],
                                   group.by = "cluster_anno",
                                   pt.size = 0)
    )
    
    dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Pos_", names(goi)[i], ".pdf"))
    dev.off()
    
  }else{
    
    temp.goi <- goi[[i]]
    
    # Split the vector into multiple vectors
    split_indices <- seq(1, length(temp.goi), by = 15)
    split_list <- lapply(split_indices, function(i) temp.goi[i:min(i + 14, length(temp.goi))])
    
    
    for(j in seq_along(length(split_list))){
      
      print(
        scCustomize::Stacked_VlnPlot(CD45pos.seurat, 
                                     features = split_list[[j]],
                                     group.by = "cluster_anno",
                                     pt.size = 0)
      )
      
      dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Pos_", names(goi)[i], "_", j, ".pdf"))
      dev.off()
      
    }
    
  }
  
}


```


## RQ2: What is the difference between macrophage clusters and between neutrophil clusters {.tabset}
# PDS should induce DAMPs that activate macrophage and neutrophil populations

```{r Macrophage_neutrophil_cluster_differences}


# create output dir
output.dir <- paste0("results/CD45Pos_Day_4_Neutrophil_Macrophage_cluster_differences/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}



# Myeloid populations at early timepoint are expected to play a large role
# investigate myeloid populations at early timepoint between PDS and Vehicle 


# reduce dataset to just day 4 timepoint 
Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$Time
day4.seurat <- subset(CD45pos.seurat, idents = "Day4")






# Firstly are there abundance differences or polarisations torwards particular subsets of marophages or DCs between untreated and treated 

dist.table <- table(day4.seurat@meta.data$cluster_anno, day4.seurat@meta.data$Treatment)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Vehicle", "PDS"))


library(ggplot2)

ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  scale_fill_hue(direction = 1) +
  theme_minimal() +
  facet_wrap(vars(Var1))

dev.copy(pdf, paste0(output.dir, "Barplot_freq_of_clusters_as_percentage_of_treatment_day4_only.pdf"))
dev.off()      


# note: 
# more B cells, DC_1, DC_2, and T_cells_1, and T_cells_2 in vehicle 
# major increase in Neutrophils_1 and Neutrophils_2 in PDS
# increase in Macrophages_2 but not Macrophages_1 in PDS

# downsample for accurate representation of clusters

table(day4.seurat@meta.data$SampleID) # 9,950 vehicle and 8,082 PDS
# reduce to 6,000 cells each 
Idents(day4.seurat) <- day4.seurat@meta.data$SampleID
temp.seurat <- subset(day4.seurat, downsample = 6000)




# plot umaps split by treatment 
scCustomize::DimPlot_scCustom(temp.seurat, 
                              group.by = "cluster_anno", 
                              split.by = "Treatment", 
                              label = TRUE,
                              pt.size = 1,
                              reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_split_by_treatment_downsampled.pdf"))
dev.off()


rm(temp.seurat)

# DEG between myeloid populations 

# how are Neutrophil_1 and Neutrophil_2 populations different 



if(long.compute){
  
  
  day4.seurat <- PrepSCTFindMarkers(day4.seurat)
  
  # set assay
  DefaultAssay(day4.seurat) <- "SCT"
  
  
  
  # set Idents
  Idents(day4.seurat) <- "cluster_anno"
  
  
  
  # Macrophage 1 vs 2
  D4.macrophage.de <- FindMarkers(day4.seurat,
                                  assay = "SCT",
                                  ident.1 = "Macrophages_1",
                                  ident.2 = "Macrophages_2",
                                  only.pos = FALSE, 
                                  recorrect_umi = FALSE,
                                  verbose = TRUE)
  
  
  # save to file 
  write.csv(D4.macrophage.de, paste0(output.dir, "Day_4_CD45Pos_DEGs_Macrophages_1_vs_Macrophages_2.csv"))
  
  
  # Neutrophils_1 vs 2
  D4.neutrophil.de <- FindMarkers(day4.seurat,
                                  assay = "SCT",
                                  ident.1 = "Neutrophils_1",
                                  ident.2 = "Neutrophils_2",
                                  only.pos = FALSE, 
                                  recorrect_umi = FALSE,
                                  verbose = TRUE)
  
  
  # save to file 
  write.csv(D4.neutrophil.de, paste0(output.dir, "Day_4_CD45Pos_DEGs_Neutrophils_1_vs_Neutrohils_2.csv"))
  
}


if(do.plot){
  
  #####################
  # heatmaps of top DEGs
  #####################
  
  
  # Neutrophils 1 vs 2
  Idents(day4.seurat) <- day4.seurat@meta.data$cluster_anno
  
  
  temp.seurat <- subset(day4.seurat, idents = c("Neutrophils_1", "Neutrophils_2"))
  
  pseudo.D4.cd45pos <- AggregateExpression(temp.seurat,
                                           assays = "SCT",
                                           return.seurat = TRUE, 
                                           group.by = c("cluster_anno"))
  
  
  # Extract top fold change DEGs 
  
  goi <- D4.neutrophil.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(abs(avg_log2FC) > 2) %>%
    dplyr::arrange(desc(avg_log2FC)) %>%
    rownames()
  
  
  # plot heatmap
  DoHeatmap(pseudo.D4.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_Day4_Heatmap_top_DEGs_logfold_2_neutrophils_1_vs_2.pdf"))
  dev.off()
  
  
  # extract just top 10 up and down degs
  goi.up <- D4.neutrophil.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::top_n(10, avg_log2FC) %>%
    rownames()
  
  goi.dn <- D4.neutrophil.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::top_n(-10, avg_log2FC) %>%
    rownames()
  
  goi <- unique(c(goi.up, goi.dn))
  
  DoHeatmap(pseudo.D4.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_Day4_Heatmap_top_DEGs_10_up_down_neutrophils_1_vs_2.pdf"))
  dev.off()
  
  
  
  
  ##################
  # Macrophages 1 vs 2
  ##################
  Idents(day4.seurat) <- day4.seurat@meta.data$cluster_anno
  
  
  temp.seurat <- subset(day4.seurat, idents = c("Macrophages_1", "Macrophages_2"))
  
  pseudo.D4.cd45pos <- AggregateExpression(temp.seurat,
                                           assays = "SCT",
                                           return.seurat = TRUE, 
                                           group.by = c("cluster_anno"))
  
  
  # Extract top fold change DEGs 
  
  goi <- D4.macrophage.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(abs(avg_log2FC) > 2) %>%
    dplyr::arrange(desc(avg_log2FC)) %>%
    rownames()
  
  
  # plot heatmap
  DoHeatmap(pseudo.D4.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_Day4_Heatmap_top_DEGs_logfold_2_macrophage_1_vs_2.pdf"))
  dev.off()
  
  
  # extract just top 10 up and down degs
  goi.up <- D4.macrophage.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::top_n(10, avg_log2FC) %>%
    rownames()
  
  goi.dn <- D4.macrophage.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::top_n(-10, avg_log2FC) %>%
    rownames()
  
  goi <- unique(c(goi.up, goi.dn))
  
  DoHeatmap(pseudo.D4.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_Day4_Heatmap_top_DEGs_10_up_down_macrophage_1_vs_2.pdf"))
  dev.off()
  
  
}

####################################
# perform gene ontology analysis 
####################################

if(do.plot){
  
  # load packages
  #install.packages("enrichR")
  library("enrichR")
  #install.packages("SCpubr")
  library("SCpubr")
  
  # set serach to human genes
  enrichR::setEnrichrSite(site = "Enrichr")
  
  dbs <- enrichR::listEnrichrDbs()
  dbs <- sort(dbs$libraryName)
  
  # Choose the dataset to query against.
  dbs_use <- c("GO_Biological_Process_2023", "GO_Cellular_Component_2023", "GO_Molecular_Function_2023", "KEGG_2021_Human", "WikiPathway_2023_Human")
  
  
  
  ##################################################
  # Perform GO for neutrophil_1 vs neutrophil_2 
  ##################################################
  
  title.var <- "Neutrophils"
  
  # filter DEGs for sig up genes 
  Neutrophils_1_sig <- D4.neutrophil.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(avg_log2FC > 1)
  
  # filter DEGs for sig down genes 
  Neutrophils_2_sig <- D4.neutrophil.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(avg_log2FC < -1)
  
  
  ########################################################################
  # Perform enrichment for genes upregulated in Neutrophils_1
  ########################################################################
  
  # List of genes to use as input.
  neutrophil.1.genes <- unique(rownames(Neutrophils_1_sig))
  
  # Retrieve the enriched terms.
  up.enriched.terms <- enrichR::enrichr(neutrophil.1.genes, dbs_use)
  
  # Plot enriched terms 
  bp.1 <- plotEnrich(up.enriched.terms$GO_Biological_Process_2023, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched BPs"))
  cc.1 <- plotEnrich(up.enriched.terms$GO_Cellular_Component_2023, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched CCs"))
  mf.1 <- plotEnrich(up.enriched.terms$GO_Molecular_Function_2023, showTerms = 20, numChar = 50, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched MFs"))
  kegg.1 <- plotEnrich(up.enriched.terms$KEGG_2021_Human, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched KEGG"))
  wiki.1 <- plotEnrich(up.enriched.terms$WikiPathway_2023_Human, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched Wiki paths"))
  
  
  ############################################################
  # Perform enrichment analysis for Downregulated genes 
  ############################################################
  
  
  # List of genes to use as input.
  neutrophil.2.genes <- unique(rownames(Neutrophils_2_sig))
  
  # Retrieve the enriched terms.
  dn.enriched.terms <- enrichR::enrichr(neutrophil.2.genes, dbs_use)
  
  # Plot enriched terms 
  
  bp.2 <- plotEnrich(dn.enriched.terms$GO_Biological_Process_2023, showTerms = 20, numChar = 50, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched BPs"))
  cc.2 <- plotEnrich(dn.enriched.terms$GO_Cellular_Component_2023, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched CCs"))
  mf.2 <- plotEnrich(dn.enriched.terms$GO_Molecular_Function_2023, showTerms = 20, numChar = 50, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched MFs"))
  kegg.2 <- plotEnrich(dn.enriched.terms$KEGG_2021_Human, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched KEGG"))
  wiki.2 <- plotEnrich(dn.enriched.terms$WikiPathway_2023_Human, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched Wiki paths"))
  
  
  
  print(bp.1|bp.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_GO_BP.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  print(cc.1|cc.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_GO_CC.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  
  print(mf.1|mf.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_GO_MF.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  
  print(kegg.1|kegg.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_KEGG.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  
  print(wiki.1|wiki.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_WIKI.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  
  
  
  ##################################################
  # Perform GO for Macrophages_1 vs Macrophages_2 
  ##################################################
  
  title.var <- "Macrophages"
  
  # filter DEGs for sig up genes 
  Macrophages_1_sig <- D4.macrophage.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(avg_log2FC > 1)
  
  # filter DEGs for sig down genes 
  Macrophages_2_sig <- D4.macrophage.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(avg_log2FC < -1)
  
  
  ########################################################################
  # Perform enrichment for genes upregulated in Neutrophils_1
  ########################################################################
  
  # List of genes to use as input.
  macrophages.1.genes <- unique(rownames(Macrophages_1_sig))
  
  # Retrieve the enriched terms.
  up.enriched.terms <- enrichR::enrichr(macrophages.1.genes, dbs_use)
  
  # Plot enriched terms 
  bp.1 <- plotEnrich(up.enriched.terms$GO_Biological_Process_2023, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched BPs"))
  cc.1 <- plotEnrich(up.enriched.terms$GO_Cellular_Component_2023, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched CCs"))
  mf.1 <- plotEnrich(up.enriched.terms$GO_Molecular_Function_2023, showTerms = 20, numChar = 50, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched MFs"))
  kegg.1 <- plotEnrich(up.enriched.terms$KEGG_2021_Human, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched KEGG"))
  wiki.1 <- plotEnrich(up.enriched.terms$WikiPathway_2023_Human, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_1 enriched Wiki paths"))
  
  
  ############################################################
  # Perform enrichment analysis for Downregulated genes 
  ############################################################
  
  
  # List of genes to use as input.
  macrophages.2.genes <- unique(rownames(Macrophages_2_sig))
  
  # Retrieve the enriched terms.
  dn.enriched.terms <- enrichR::enrichr(macrophages.2.genes, dbs_use)
  
  # Plot enriched terms 
  
  bp.2 <- plotEnrich(dn.enriched.terms$GO_Biological_Process_2023, showTerms = 20, numChar = 50, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched BPs"))
  cc.2 <- plotEnrich(dn.enriched.terms$GO_Cellular_Component_2023, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched CCs"))
  mf.2 <- plotEnrich(dn.enriched.terms$GO_Molecular_Function_2023, showTerms = 20, numChar = 50, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched MFs"))
  kegg.2 <- plotEnrich(dn.enriched.terms$KEGG_2021_Human, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched KEGG"))
  wiki.2 <- plotEnrich(dn.enriched.terms$WikiPathway_2023_Human, showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = paste0(title.var, "_2 enriched Wiki paths"))
  
  
  
  print(bp.1|bp.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_GO_BP.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  print(cc.1|cc.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_GO_CC.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  
  print(mf.1|mf.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_GO_MF.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  
  print(kegg.1|kegg.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_KEGG.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  
  print(wiki.1|wiki.2)
  ggsave(filename = paste0(output.dir, "Day_4_", title.var, "_WIKI.pdf"),
         width = 20, 
         height = 20, 
         units = "in",
         device = "pdf")
  
  
}

```


## RQ3: Day 4 - what is the treatment difference within neutrophils, macrophages, monocytes {.tabset}

```{r RQ3_day_4_treatment_differences}


# create output dir
output.dir <- paste0("results/CD45Pos_Day_4_RQ3_treatment_difference_within_populations/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}



# reduce dataset to just day 4 timepoint 
Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$Time
day4.seurat <- subset(CD45pos.seurat, idents = "Day4")



if(long.compute){
  
  
  day4.seurat <- PrepSCTFindMarkers(day4.seurat)
  
  # set assay
  DefaultAssay(day4.seurat) <- "SCT"
  
  
  # create new cell_type plus treatment 
  day4.seurat@meta.data$cell_type_treat <- paste0(day4.seurat@meta.data$cell_type, "_", day4.seurat@meta.data$Treatment)
  
  # set Idents
  Idents(day4.seurat) <- "cell_type_treat"
  
  
  # cell types to perform analysis for 
  celltype.vec <- c("Macrophages", "DC", "Monocytes", "Neutrophils")
  
  # create list to store output
  DEG.list <- list()
  
  # run through DEG specifically between vehicle and treatment for each cluster 
  
  for(i in seq_along(celltype.vec)){
    
    temp.de <- FindMarkers(day4.seurat,
                           assay = "SCT",
                           ident.1 = paste0(celltype.vec[i], "_PDS"),
                           ident.2 = paste0(celltype.vec[i], "_Vehicle"),
                           only.pos = FALSE, 
                           recorrect_umi = FALSE,
                           verbose = TRUE)
    
    DEG.list[[i]] <- temp.de
    
    # save to file 
    write.csv(temp.de, paste0(output.dir, "DEGs_Day_4_PDS_", celltype.vec[i], "_vs_Vehicle_", celltype.vec[i], ".csv"))
    
    
  }
  # add names to DEG.list
  names(DEG.list) <- celltype.vec
  
  # save DEG list to file 
  saveRDS(DEG.list, paste0(output.dir, "DEG_Day_4_PDS_vs_Vehicle_celltypes_list_object.rds"))
  
  rm(day4.seurat)
  
}else{
  
  DEG.list <- readRDS(paste0(output.dir, "DEG_Day_4_PDS_vs_Vehicle_celltypes_list_object.rds"))
  
}


# plot DEGs 


if(do.plot){
  
  for(i in 1:length(DEG.list)){
    
    FDR.theshold <- 0.01
    logFC.threshold <- 1
    
    # filter data
    input.data <- DEG.list[[i]] %>%
      dplyr::rename(logFC = avg_log2FC) %>%
      dplyr::rename(FDR = p_val_adj) %>%
      filter(FDR < FDR.theshold)
    
    # color points
    volcano.col <- usefulfunctions::colour.points.volcano(input.data,
                                                          increase.col = "Red",
                                                          decreased.col = "Blue",
                                                          FDR.cutoff = FDR.theshold, 
                                                          logFC.cutoff = logFC.threshold)
    # plot heatmap
    print(
      usefulfunctions::custom.enhanced.volcano(input.data, 
                                               pCutoff = FDR.theshold,
                                               FCcutoff = logFC.threshold,
                                               title = paste0(names(DEG.list)[i], " PDS vs Vehicle"),
                                               subtitle = paste0("adjPval threshold = ", FDR.theshold, " & log2FC threshold = ", logFC.threshold),
                                               colCustom = volcano.col)
    )
    dev.copy(pdf, paste0(output.dir, "VolcanoPlot_Day_4_PDS_vs_Vehicle_cluster_", names(DEG.list)[i], ".pdf"))
    dev.off()
    
    
  }
  
}



```



### Gene set enrichment
```{r geneset_enrichment_Day4_myeloid}

########################################################################################################################
# gene set enrichment analysis of Day 4 myeloid populations PDS vs vehicle 
########################################################################################################################

# load packages
library("enrichR")
#install.packages("SCpubr")
library("SCpubr")


# create output dir
output.dir <- paste0("results/CD45Pos_Day_4_Myeloid_population_GeneSet_enrichment_PDS_vs_Vehicle/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# set variables
enrichR::setEnrichrSite(site = "Enrichr")

dbs <- enrichR::listEnrichrDbs()
dbs <- sort(dbs$libraryName)

#grep("Mouse", dbs, value = T)

# Choose the dataset to query against.
dbs_use <- c("GO_Biological_Process_2021", "WikiPathways_2019_Mouse", "MSigDB_Hallmark_2020", "Mouse_Gene_Atlas", "KEGG_2019_Mouse", "HDSigDB_Mouse_2021", "GO_Molecular_Function_2023", "KOMP2_Mouse_Phenotypes_2022")



############################################################
# Filter DEGs for up and down-regulated
############################################################

for(i in seq_along(DEG.list)){
  
  cell.type <- names(DEG.list)[i]
  
  
  # filter DEGs for sig up genes 
  sig.up <- DEG.list[[i]] %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(avg_log2FC > 0) 
  
  # filter DEGs for sig down genes 
  sig.dn <- DEG.list[[i]]  %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(avg_log2FC < 0) 
  
  
  
  
  ############################################################
  # Perform enrichment analysis for Upregulated genes 
  ############################################################
  
  # List of genes to use as input.
  up.genes <- unique(rownames(sig.up))
  
  # Retrieve the enriched terms.
  up.enriched.terms <- enrichR::enrichr(up.genes, dbs_use)
  
  
  for(j in seq_along(up.enriched.terms)){
    database.var <- names(up.enriched.terms)[j]
    
    # plot enrichment
    up.plot <- plotEnrich(up.enriched.terms[[j]], 
                          showTerms = 20, 
                          numChar = 40,
                          y = "Count", 
                          orderBy = "P.value", 
                          title = paste0("Upregulated ", database.var))
    
    ggsave(filename = paste0(output.dir, "Upregulated_", cell.type, "_PDS_vs_Vehicle_",  database.var, ".pdf"),
           plot = up.plot, 
           width = 20, 
           height = 20, 
           units = "in",
           device = "pdf")
    
  }
  
  
  
  ############################################################
  # Perform enrichment analysis for Downregulated genes 
  ############################################################
  
  # List of genes to use as input.
  dn.genes <- unique(rownames(sig.dn))
  
  # Retrieve the enriched terms.
  dn.enriched.terms <- enrichR::enrichr(dn.genes, dbs_use)
  
  
  for(j in seq_along(dn.enriched.terms)){
    database.var <- names(dn.enriched.terms)[j]
    
    # plot enrichment
    down.plot <- plotEnrich(dn.enriched.terms[[j]], 
                            showTerms = 20, 
                            numChar = 40,
                            y = "Count", 
                            orderBy = "P.value", 
                            title = paste0("Downregulated ", database.var))
    
    
    ggsave(filename = paste0(output.dir, "Downregulated_", cell.type, "_PDS_vs_Vehicle_",  database.var, ".pdf"),
           plot = down.plot, 
           width = 20, 
           height = 20, 
           units = "in",
           device = "pdf")
    
  }
  
  
}
```



## RQ4: Day 9 - what is the treatment difference within neutrophils, macrophages, monocytes, T-cells {.tabset}

```{r RQ4_treatment_difference_Day9_celltypes}


# create output dir
output.dir <- paste0("results/CD45Pos_Day_9_RQ4_treatment_difference_within_populations/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}



# reduce dataset to just day 4 timepoint 
Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$Time
day9.seurat <- subset(CD45pos.seurat, idents = "Day9")



if(long.compute){
  
  
  day9.seurat <- PrepSCTFindMarkers(day9.seurat)
  
  # set assay
  DefaultAssay(day9.seurat) <- "SCT"
  
  
  # create new cell_type plus treatment 
  
  # currently tregs is grouped with t-cells, split this back out 
  
  day9.seurat@meta.data <- day9.seurat@meta.data %>%
    dplyr::mutate(cell_type = case_when(cluster_anno == "Tregs" ~ "Tregs", 
                                        TRUE ~ cell_type))
  
  table(day9.seurat@meta.data$cell_type)
  
  day9.seurat@meta.data$cell_type_treat <- paste0(day9.seurat@meta.data$cell_type, "_", day9.seurat@meta.data$Treatment)
  
  # set Idents
  Idents(day9.seurat) <- "cell_type_treat"
  
  
  # cell types to perform analysis for 
  celltype.vec <- c("Macrophages", "DC", "Monocytes", "Neutrophils", "Tcells", "NKcells")
  
  # create list to store output
  DEG.list <- list()
  
  # run through DEG specifically between vehicle and treatment for each cluster 
  
  for(i in seq_along(celltype.vec)){
    
    temp.de <- FindMarkers(day9.seurat,
                           assay = "SCT",
                           ident.1 = paste0(celltype.vec[i], "_PDS"),
                           ident.2 = paste0(celltype.vec[i], "_Vehicle"),
                           only.pos = FALSE, 
                           recorrect_umi = FALSE,
                           verbose = TRUE)
    
    DEG.list[[i]] <- temp.de
    
    # save to file 
    write.csv(temp.de, paste0(output.dir, "DEGs_Day_9_PDS_", celltype.vec[i], "_vs_Vehicle_", celltype.vec[i], ".csv"))
    
    
  }
  # add names to DEG.list
  names(DEG.list) <- celltype.vec
  
  # save DEG list to file 
  saveRDS(DEG.list, paste0(output.dir, "DEG_Day_9_PDS_vs_Vehicle_celltypes_list_object.rds"))
  
  rm(day9.seurat)
  
}else{
  
  DEG.list <- readRDS(paste0(output.dir, "DEG_Day_9_PDS_vs_Vehicle_celltypes_list_object.rds"))
  
}


# plot DEGs 


if(do.plot){
  
  for(i in 1:length(DEG.list)){
    
    FDR.theshold <- 0.01
    logFC.threshold <- 1
    
    # filter data
    input.data <- DEG.list[[i]] %>%
      dplyr::rename(logFC = avg_log2FC) %>%
      dplyr::rename(FDR = p_val_adj) %>%
      filter(FDR < FDR.theshold)
    
    # color points
    volcano.col <- usefulfunctions::colour.points.volcano(input.data,
                                                          increase.col = "Red",
                                                          decreased.col = "Blue",
                                                          FDR.cutoff = FDR.theshold, 
                                                          logFC.cutoff = logFC.threshold)
    # plot heatmap
    print(
      usefulfunctions::custom.enhanced.volcano(input.data, 
                                               pCutoff = FDR.theshold,
                                               FCcutoff = logFC.threshold,
                                               title = paste0(names(DEG.list)[i], " PDS vs Vehicle"),
                                               subtitle = paste0("adjPval threshold = ", FDR.theshold, " & log2FC threshold = ", logFC.threshold),
                                               colCustom = volcano.col)
    )
    dev.copy(pdf, paste0(output.dir, "VolcanoPlot_Day_9_PDS_vs_Vehicle_cluster_", names(DEG.list)[i], ".pdf"))
    dev.off()
    
    
  }
  
}


```


## RQ5: DEG Neutrophils PDS Day 9 vs Neutrophils PDS Day 4 {.tabset}

```{r RQ5_D9_Neutrophils_PDS_vs_D4_Neutrophils_PDS}


# create output dir
output.dir <- paste0("results/CD45Pos_RQ5_D9_Neutrophils_PDS_vs_D4_Neutrophils_PDS/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}





# reduce dataset to just day 4 timepoint 
Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$cell_type
neutrophil.seurat <- subset(CD45pos.seurat, idents = "Neutrophils")



if(long.compute){
  
  
  neutrophil.seurat <- PrepSCTFindMarkers(neutrophil.seurat)
  
  # set assay
  DefaultAssay(neutrophil.seurat) <- "SCT"
  
  
  # create new time plus treatment 
  
  neutrophil.seurat@meta.data$Time_treat <- paste0(neutrophil.seurat@meta.data$Time, "_", neutrophil.seurat@meta.data$Treatment)
  
  # set Idents
  Idents(neutrophil.seurat) <- "Time_treat"
  
  
  
  temp.de <- FindMarkers(neutrophil.seurat,
                         assay = "SCT",
                         ident.1 = "Day9_PDS",
                         ident.2 = "Day4_PDS",
                         only.pos = FALSE, 
                         recorrect_umi = FALSE,
                         verbose = TRUE)
  
  # save to file 
  write.csv(temp.de, paste0(output.dir, "DEGs_Day_9_PDS_neutrophils_vs_Day_4_PDS_neutrophils.csv"))
  
  
  
}


# plot DEGs 


if(do.plot){
  
  FDR.theshold <- 0.01
  logFC.threshold <- 1
  
  # filter data
  input.data <- temp.de %>%
    dplyr::rename(logFC = avg_log2FC) %>%
    dplyr::rename(FDR = p_val_adj) %>%
    filter(FDR < FDR.theshold)
  
  # color points
  volcano.col <- usefulfunctions::colour.points.volcano(input.data,
                                                        increase.col = "Red",
                                                        decreased.col = "Blue",
                                                        FDR.cutoff = FDR.theshold, 
                                                        logFC.cutoff = logFC.threshold)
  # plot heatmap
  print(
    usefulfunctions::custom.enhanced.volcano(input.data, 
                                             pCutoff = FDR.theshold,
                                             FCcutoff = logFC.threshold,
                                             title = paste0("Neutrophils Day 9 PDS vs Day 4 PDS"),
                                             subtitle = paste0("adjPval threshold = ", FDR.theshold, " & log2FC threshold = ", logFC.threshold),
                                             colCustom = volcano.col)
  )
  dev.copy(pdf, paste0(output.dir, "VolcanoPlot_Day_9_neutrophils_PDS_vs_Day_4_neutrophils_PDS.pdf"))
  dev.off()
  
  
  
  
  #####################
  # heatmaps of top DEGs
  #####################
  
  
  
  Idents(neutrophil.seurat) <- neutrophil.seurat@meta.data$Time_treat
  
  
  
  pseudo.neutrophils <- AggregateExpression(neutrophil.seurat,
                                            assays = "SCT",
                                            return.seurat = TRUE, 
                                            group.by = c("Time_treat"))
  
  
  # Extract top fold change DEGs 
  
  goi <- temp.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(abs(avg_log2FC) > 2) %>%
    dplyr::arrange(desc(avg_log2FC)) %>%
    rownames()
  
  
  # plot heatmap
  DoHeatmap(pseudo.neutrophils, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_Heatmap_Day_9_PDS_neutrophils_vs_Day_4_PDS_neutrophils_logfold_2.pdf"))
  dev.off()
  
  
  # extract just top 10 up and down degs
  goi.up <- temp.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::top_n(10, avg_log2FC) %>%
    rownames()
  
  goi.dn <- temp.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::top_n(-10, avg_log2FC) %>%
    rownames()
  
  goi <- unique(c(goi.up, goi.dn))
  
  DoHeatmap(pseudo.neutrophils, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "CD45Pos_Heatmap_Day_9_PDS_neutrophils_vs_Day_4_PDS_neutrophils_top_10_DEGs_up_down.pdf"))
  dev.off()
  
}



# clean environment 
rm(neutrophil.seurat)
gc()

```





## RQ6: Zoom in on T-cells, can identify Cytotoxic CD4 T-cells? {.tabset}

```{r Cytotoxic_CD4_Tcells}


# create output dir
output.dir <- paste0("results/CD45Pos_Cytotoxic_CD4_Tcells/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# currently tregs is grouped with t-cells, split this back out 
CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
  dplyr::mutate(cell_type = case_when(cluster_anno == "Tregs" ~ "Tregs", 
                                      TRUE ~ cell_type))


# reduce dataset to just day 4 timepoint 
Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$cell_type
Tcell.seurat <- subset(CD45pos.seurat, idents = c("Tcells", "Tregs"))

DefaultAssay(Tcell.seurat) <- "RNA"

Tcell.seurat <- FindVariableFeatures(Tcell.seurat, assay = "RNA")

imputation.features <- VariableFeatures(Tcell.seurat)

# ensure cd8 and cd4 key markers are included 
imputation.features <- c(paste0(imputation.features), "Cd4", "Cd8a", "Cd8b1", "Foxp3")
imputation.features <- unique(imputation.features)

DefaultAssay(Tcell.seurat) <- "SCT"



######################
# Impute values 
######################
if(long.compute){
  Tcell.seurat <- RunALRA(Tcell.seurat, 
                          genes.use = imputation.features)
}


DefaultAssay(Tcell.seurat) <- "SCT"


# recalculate UMAP on just T-cell population 
Tcell.seurat <- RunUMAP(Tcell.seurat, 
                        reduction = "integrated.cca.sct",
                        dims = 1:30)

# find clusters
Tcell.seurat <- FindClusters(Tcell.seurat, 
                             resolution = 1, 
                             graph.name = "SCT_snn")  

table(Tcell.seurat@meta.data$seurat_clusters)



Seurat::DimPlot(object = Tcell.seurat, 
                reduction = "umap", 
                label = TRUE, 
                label.size = 4, 
                group.by = "cell_type") +
  NoLegend()

Seurat::DimPlot(object = Tcell.seurat, 
                reduction = "umap", 
                label = TRUE, 
                label.size = 4, 
                group.by = "seurat_clusters") +
  NoLegend()



grep("Cd8.*", rownames(Tcell.seurat), value = T)

# visualise gene expression - use imputed values 
DefaultAssay(Tcell.seurat) <- "alra"

Stacked_VlnPlot(Tcell.seurat, 
                features = c("Cd8a", "Cd8b1", "Cd4", "Gzmb", "Gzma", "Nkg7", "Foxp3"))



scCustomize::FeaturePlot_scCustom(Tcell.seurat, reduction = "umap", feature = "Cd4")
scCustomize::FeaturePlot_scCustom(Tcell.seurat, reduction = "umap", feature = "Gzmb")





Seurat::DimPlot(object = Tcell.seurat, 
                reduction = "umap", 
                label = TRUE, 
                label.size = 4, 
                group.by = "SampleID") +
  NoLegend()




```



## Plots for manuscript {.tabset}

### Set up object
```{r set_up_object}

# create output dir
output.dir <- paste0("results/CD45Pos_manuscript_plots/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# Day 4 dataset, PDS vs Vehicle overview
CD45pos.seurat@meta.data

Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$Time
day4.seurat <- subset(CD45pos.seurat, idents = "Day4")

table(day4.seurat@meta.data$cell_type)
# note that day 4 only has 2 basophils | if only day 4 is plotted, this is weird and also no analysi can be done on 2 cells anyway
# for clearer plot, remove these 2 basophils 
Idents(day4.seurat) <- day4.seurat@meta.data$cell_type
day4.seurat <- subset(day4.seurat, idents = "Basophils", invert = T)

```


### UMAP of Vehicle vs PDS Day 4
```{r UMAP_PDS_Vehicle_D4}


###################################################################################################################################################
#1. 2 UMAPs (PDS and Vehicle day 4) annotated with the broad immune cell supsets (like neutrophil/monocyte/CD4 T cell/CD8 Tcell/NK cell etc.)
###################################################################################################################################################


# for split comparison plots, downsample to make visual comparisons more fair 
# i dont like how squished the plots are when split.by is used. therefore plot each one individually for more prettyness 

table(day4.seurat@meta.data$Treatment) # 9,950 and 8,082 cells
Idents(day4.seurat) <- day4.seurat@meta.data$Treatment
small.seurat <- subset(day4.seurat, downsample = 7000)


Idents(small.seurat) <- small.seurat@meta.data$Treatment
temp.seurat <- subset(small.seurat, idents = "Vehicle")


# Vehicle
scCustomize::DimPlot_scCustom(temp.seurat,
                              reduction = "umap", 
                              label = TRUE,
                              figure_plot = TRUE,
                              group.by = "cell_type", 
                              pt.size = 1)

dev.copy(pdf, paste0(output.dir, "CD45Pos_UMAP_Day_4_no_basophils_downsampled_vehicle.pdf"))
dev.off()


Idents(small.seurat) <- small.seurat@meta.data$Treatment
temp.seurat <- subset(small.seurat, idents = "PDS")

# PDS
scCustomize::DimPlot_scCustom(temp.seurat,
                              reduction = "umap", 
                              label = TRUE,
                              figure_plot = TRUE,
                              group.by = "cell_type", 
                              pt.size = 1)

dev.copy(pdf, paste0(output.dir, "CD45Pos_UMAP_Day_4_no_basophils_downsampled_PDS.pdf"))
dev.off()


```

### Cluster distribution
```{r cluster_distribution}

##################################################################################################################
#2. A simple graph with the distribution (not total numbers but percentage) of these subsets with Vehicle vs PDS to highlight that especially neutrophils are enriched in PDS
##################################################################################################################

dist.table <- table(day4.seurat@meta.data$cell_type, day4.seurat@meta.data$Treatment)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Vehicle", "PDS"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Treatment", y = "% of treatment", title = "Frequency of cluster per treatment") +
  scale_fill_manual(values = c(Vehicle = "grey", PDS = "darkred")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_cluster_per_treatment_day4_only.pdf"))

# rather make this plot in prism as it looks nicer 
# export data for prism plotting 
write.csv(freq.table, paste0(output.dir, "Barplot_freq_of_cluster_per_treatment_day_4_only.csv"))


```


### Heatmap with custom genes
```{r custom_gene_heatmap}
##################################################################################################################
#3. The Violin plot with the custom genes either as violin or as heatmap with the same clusters as above
##################################################################################################################

# Note, I am unclear if Nils wants heatmap sep by treatment and cluster or just by cluster. 
# will generate both plots 

# firstly, treatment and cluster sep 

# aggregate expression for heatmap visualisation
DefaultAssay(day4.seurat) <- "SCT"
pseudo.cd45pos <- AggregateExpression(day4.seurat,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("cell_type", "Treatment"))

# genes to plot provided by Nils
custom.genes <- c("Il1b", "Il6", "Il12a", "Il12b", "Ifng", "Tnf", "Ccl2", "Ccl3", "Ccl4", "Ccl5", "Cxcl1", "Cxcl9", "Cxcl10") # Cxcl13 not present in CD45Pos dataset


###################
# Plot heatmap
###################

DoHeatmap(pseudo.cd45pos, 
          features = custom.genes, 
          draw.lines = FALSE, 
          angle = 90,
          raster = FALSE) + 
  scico::scale_fill_scico(palette = "batlow", 
                          direction = 1, 
                          na.value = "white")

dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_Day_4_only_treatment_cluster_custom_genes.pdf"))
dev.off()




# secondly, cluster sep only

# aggregate expression for heatmap visualisation
DefaultAssay(day4.seurat) <- "SCT"
pseudo.cd45pos <- AggregateExpression(day4.seurat,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("cell_type"))


###################
# Plot heatmap
###################

DoHeatmap(pseudo.cd45pos, 
          features = custom.genes, 
          draw.lines = FALSE, 
          angle = 90,
          raster = FALSE) + 
  scico::scale_fill_scico(palette = "batlow", 
                          direction = 1, 
                          na.value = "white")

dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_Day_4_only_cluster_custom_genes.pdf"))
dev.off()


```


### Dot plots of select GO terms
```{r Dotplots_GO_terms}

###################################################################################################################################################
#4. Dotplots for GO terms Neutrophils, monocytes, macrophages and DCs 
###################################################################################################################################################

# firstly calculate DEGs treatment-wise - this is redone as this object is just Day 4 and no basophils

if(long.compute){
  
  
  day4.seurat <- PrepSCTFindMarkers(day4.seurat)
  
  # set assay
  DefaultAssay(day4.seurat) <- "SCT"
  
  
  
  # add annotation for treatment 
  day4.seurat@meta.data$Treatment_cluster <- paste0(day4.seurat@meta.data$Treatment, "_", day4.seurat@meta.data$cell_type)
  
  # set Idents
  Idents(day4.seurat) <- "Treatment_cluster"
  
  
  cluster.ids <- unique(day4.seurat@meta.data$cell_type)
  
  # create list to store output
  DEG.list <- list()
  
  # run through DEG specifically between vehicle and treatment for each cluster 
  
  for(i in seq_along(cluster.ids)){
    temp.de <- FindMarkers(day4.seurat,
                           assay = "SCT",
                           ident.1 = paste0("PDS_", cluster.ids[i]),
                           ident.2 = paste0("Vehicle_", cluster.ids[i]),
                           only.pos = FALSE, 
                           recorrect_umi = FALSE,
                           verbose = TRUE)
    
    DEG.list[[i]] <- temp.de
    
    # save to file 
    write.csv(temp.de, paste0(output.dir, "DEGs_PDS_", cluster.ids[i], "_vs_Vehicle_", cluster.ids[i], ".csv"))
    
    
  }
  # add names to DEG.list
  names(DEG.list) <- cluster.ids
  
  # save DEG list to file 
  saveRDS(DEG.list, paste0(output.dir, "DEG_list_object_day4_no_Baso.rds"))
  
  
}else{
  
  DEG.list <- readRDS(paste0(output.dir, "DEG_list_object_day4_no_Baso.rds"))
  
}


########################
# Set up enrichr params
########################

# load library
library("enrichR")
library("clusterProfiler")
library("dplyr")
library("ggplot2")
library("scales")  # For the log transformation and labels

# set variables
enrichR::setEnrichrSite(site = "Enrichr")

dbs <- enrichR::listEnrichrDbs()
dbs <- sort(dbs$libraryName)


# Choose the dataset to query against.
dbs_use <- c("GO_Biological_Process_2021", "KEGG_2019_Mouse")


##############
# Functions 
##############
# Define the function
filter_terms <- function(data, go_terms, group_name, prefix = "") {
  # Perform the filtering and mutation
  result <- data %>%
    dplyr::select(c(Term, Overlap, Adjusted.P.value, Odds.Ratio, Combined.Score)) %>%
    filter(grepl(paste(go_terms, collapse = "|"), Term, ignore.case = TRUE)) %>%
    mutate(Term = paste0(prefix, Term), 
           group = group_name)
  
  print(result)
  return(result)
}


process_data <- function(input.1, input.2, input.3, input.4, pval_threshold = 0.05){
  
  # join up and down term tables together
  combined.term.table <- bind_rows(input.1, input.2, input.3, input.4)
  
  # additional formatting of dataframe 
  combined.term.table <- combined.term.table %>%
    mutate(Overlap_Ratio = as.numeric(sub("/.*", "", Overlap)) / as.numeric(sub(".*/", "", Overlap))) %>% # Create overlap ratio
    mutate(color_value = ifelse(Adjusted.P.value > pval_threshold, NA, Combined.Score)) %>% # create new column so that in.sig values are greyed out
    mutate(color_value_log = log1p(color_value)) # Using log1p to handle zero values
  
  # Reorder the 'group' factor levels
  combined.term.table$group <- factor(combined.term.table$group, levels = c("PDS", "Vehicle"))
  
  # Order dataframe by group and combined score descending
  combined.term.table <- combined.term.table %>%
    arrange(group, desc(Combined.Score)) %>% 
    mutate(Term_order = factor(Term, levels = unique(Term)))
  
  return(combined.term.table)
  
}

plot_all <- function(term.table, title.var = " Ontology Terms", save.path){
  
  p1 <- ggplot(term.table, aes(x = group, y = Term_order, size = Overlap_Ratio)) +
    geom_point(aes(color = color_value_log), na.rm = TRUE) +
    scale_color_gradient(low = "blue", high = "red", na.value = "gray",
                         name = "Combined Score",
                         labels = scales::label_log()) +  # Apply log scale labels
    scale_size_continuous(range = c(0, 10)) +  # Adjust range to enhance size difference
    scale_y_discrete(limits = rev) +  # Reverse the order of the y-axis
    theme_minimal() +
    labs(title = title.var, x = "Treatment group (Day 4)", y = NULL)
  print(p1)
  ggsave(save.path, p1)
  
}

plot_pds_sig <- function(term.table, title.var = "", pval_threshold = 0.05, save.path){
  
  
  # Filter significant pathways in the PDS group
  significant_pds_terms <- term.table %>%
    filter(group == "PDS" & Adjusted.P.value < pval_threshold) %>%
    pull(Term)  # Extract terms as a vector
  
  # Filter the combined table to keep significant terms in PDS and all terms in Vehicle group
  filtered_combined_table <- term.table %>%
    filter(Term %in% significant_pds_terms)
  
  
  p1 <- ggplot(filtered_combined_table, aes(x = group, y = Term_order, size = Overlap_Ratio)) +
    geom_point(aes(color = color_value_log), na.rm = TRUE) +
    scale_color_gradient(low = "blue", high = "red", na.value = "gray",
                         name = "Combined Score",
                         labels = scales::label_log()) +  # Apply log scale labels
    scale_size_continuous(range = c(0, 10)) +  # Adjust range to enhance size difference
    scale_y_discrete(limits = rev) +  # Reverse the order of the y-axis
    theme_minimal() +
    labs(title = title.var, x = "Treatment group (Day 4)", y = NULL)
  
  print(p1)
  ggsave(save.path, p1)
  
}
##################
#a. Neutrophils: 
##################

# subset seurat object for cell_type in question
DefaultAssay(day4.seurat) <- "SCT"

Idents(day4.seurat) <- day4.seurat@meta.data$cell_type
Neutrophil.seurat <- subset(day4.seurat, idents = "Neutrophils")
table(Neutrophil.seurat@meta.data$Treatment)
Idents(Neutrophil.seurat) <- Neutrophil.seurat@meta.data$Treatment

# calculate enrichment

# filter for upregulated genes
up.genes <- DEG.list$Neutrophils %>%
  dplyr::filter(p_val_adj < 0.05) %>%
  dplyr::filter(avg_log2FC > 0) %>% 
  rownames()

dn.genes <- DEG.list$Neutrophils %>%
  dplyr::filter(p_val_adj < 0.05) %>%
  dplyr::filter(avg_log2FC < 0) %>% 
  rownames()


# Retrieve the enriched terms.
up.enriched.terms <- enrichR::enrichr(up.genes, dbs_use)
dn.enriched.terms <- enrichR::enrichr(dn.genes, dbs_use)

# Filter GO output for plotting and based on terms Nils would like to highlight

# Neutrophil GO-BP terms 
GO.terms <- c("cytokine-mediated signaling pathway", "neutrophil degranulation", "neutrophil mediated immunity", "cellular response to type I interferon", 
                         "phagosome acidification", "phagosome maturation", "translation", "gene expression", "ribosome biogenesis")
GO.terms <- paste0("^", GO.terms, " ")

length(GO.terms)

PDS.go <- filter_terms(up.enriched.terms$GO_Biological_Process_2021,
                       go_terms = GO.terms,
                       group_name = "PDS")

Vehcile.go <- filter_terms(dn.enriched.terms$GO_Biological_Process_2021,
                           go_terms = GO.terms,
                           group_name = "Vehicle")

# Neutrophil KEGG terms 
KEGG.terms <- c("Lysosome", "Phagosome", "NOD-like receptor signaling pathway", "Apoptosis", "NF-kappa B signaling pathway", "TNF signaling pathway",
                "Antigen processing and presentation", "RIG-I-like receptor signaling pathway", "Ribosome", "Oxidative phosphorylation")
KEGG.terms <- paste0("^", KEGG.terms, "$")
length(KEGG.terms)

PDS.kegg <- filter_terms(up.enriched.terms$KEGG_2019_Mouse,
                         go_terms = KEGG.terms,
                         group_name = "PDS", 
                         prefix = "KEGG: ")

Vehcile.kegg <- filter_terms(dn.enriched.terms$KEGG_2019_Mouse,
                             go_terms = KEGG.terms,
                             group_name = "Vehicle", 
                             prefix = "KEGG: ")

term.table.out <- process_data(PDS.go, PDS.kegg, Vehcile.go, Vehcile.kegg)

plot_all(term.table = term.table.out, save.path = paste0(output.dir, "test.pdf"))

plot_pds_sig(term.table = term.table.out, save.path = paste0(output.dir, "test_sig.pdf"))



# combine all data

# join up and down term tables together
combined.term.table <- bind_rows(PDS.go, PDS.kegg, Vehcile.go, Vehcile.kegg)

# Set the p-value threshold
pval_threshold <- 0.05


# additional formatting of dataframe 
combined.term.table <- combined.term.table %>%
  mutate(Overlap_Ratio = as.numeric(sub("/.*", "", Overlap)) / as.numeric(sub(".*/", "", Overlap))) %>% # Create overlap ratio
  mutate(color_value = ifelse(Adjusted.P.value > pval_threshold, NA, Combined.Score)) %>% # create new column so that in.sig values are greyed out
  mutate(color_value_log = log1p(color_value)) # Using log1p to handle zero values
  
# Reorder the 'group' factor levels
combined.term.table$group <- factor(combined.term.table$group, levels = c("PDS", "Vehicle"))

# Order dataframe by group and combined score descending
combined.term.table <- combined.term.table %>%
  arrange(group, desc(Combined.Score)) %>% 
  mutate(Term_order = factor(Term, levels = unique(Term)))


# Plot with the custom color scale on a log scale
ggplot(combined.term.table, aes(x = group, y = Term_order, size = Overlap_Ratio)) +
  geom_point(aes(color = color_value_log), na.rm = TRUE) +
  scale_color_gradient(low = "blue", high = "red", na.value = "gray",
                       name = "Combined Score",
                       labels = scales::label_log()) +  # Apply log scale labels
    scale_size_continuous(range = c(0, 10)) +  # Adjust range to enhance size difference
    scale_y_discrete(limits = rev) +  # Reverse the order of the y-axis
  theme_minimal() +
  labs(title = "Neutrophil Ontology terms", x = "Treatment group (Day 4)", y = NULL)










# Calculate the min and max adjusted p-values below the threshold
#pval_min <- min(combined.term.table$color_value, na.rm = TRUE)
#pval_max <- max(combined.term.table$color_value, na.rm = TRUE)

```



### Heatmaps highlighting genes in specific genesets
```{r Heatmaps_plots_highlighting_specific_genesets}

###################################################################################################################################################
#4. Volcano plots for Neutrophils, monocytes, macrophages and DCs highlighting and labeling top regulated genes with the following gene sets:
###################################################################################################################################################

# firstly calculate DEGs treatment-wise - this is redone as this object is just Day 4 and no basophils

if(long.compute){
  
  
  day4.seurat <- PrepSCTFindMarkers(day4.seurat)
  
  # set assay
  DefaultAssay(day4.seurat) <- "SCT"
  
  
  
  # add annotation for treatment 
  day4.seurat@meta.data$Treatment_cluster <- paste0(day4.seurat@meta.data$Treatment, "_", day4.seurat@meta.data$cell_type)
  
  # set Idents
  Idents(day4.seurat) <- "Treatment_cluster"
  
  
  cluster.ids <- unique(day4.seurat@meta.data$cell_type)
  
  # create list to store output
  DEG.list <- list()
  
  # run through DEG specifically between vehicle and treatment for each cluster 
  
  for(i in seq_along(cluster.ids)){
    temp.de <- FindMarkers(day4.seurat,
                           assay = "SCT",
                           ident.1 = paste0("PDS_", cluster.ids[i]),
                           ident.2 = paste0("Vehicle_", cluster.ids[i]),
                           only.pos = FALSE, 
                           recorrect_umi = FALSE,
                           verbose = TRUE)
    
    DEG.list[[i]] <- temp.de
    
    # save to file 
    write.csv(temp.de, paste0(output.dir, "DEGs_PDS_", cluster.ids[i], "_vs_Vehicle_", cluster.ids[i], ".csv"))
    
    
  }
  # add names to DEG.list
  names(DEG.list) <- cluster.ids
  
  # save DEG list to file 
  saveRDS(DEG.list, paste0(output.dir, "DEG_list_object_day4_no_Baso.rds"))
  
  
}else{
  
  DEG.list <- readRDS(paste0(output.dir, "DEG_list_object_day4_no_Baso.rds"))
  
}


########################
# Set up enrichr params
########################

# load library
library(enrichR)

# set variables
enrichR::setEnrichrSite(site = "Enrichr")

dbs <- enrichR::listEnrichrDbs()
dbs <- sort(dbs$libraryName)


# Choose the dataset to query against.
dbs_use <- c("GO_Biological_Process_2021", "KEGG_2019_Mouse", "GO_Molecular_Function_2023")

############################################################
# Define function for downstream filtering and plotting
############################################################
go.heatmap <- function(input.data = NULL, GO.genes = NULL, celltype.name = "TEST", GO.name = "TEST"){
  
  # remove low expressed genes
  goi <- input.data@assays$SCT@data %>%
    as.data.frame() %>%
    tibble::rownames_to_column() %>%
    dplyr::filter(rowname %in% GO.genes) %>%
    dplyr::rowwise() %>%
    dplyr::mutate(counts = sum(c_across(-rowname))) %>%
    dplyr::filter(counts > 100) %>%
    dplyr::pull(rowname)
  
  # scale data
  input.data <- ScaleData(input.data, features = goi)
  
  # plot heatmap
  p1 <- DoHeatmap(input.data, 
                  features = goi, 
                  draw.lines = T, 
                  angle = 90,
                  group.by = "Treatment",
                  raster = FALSE) + 
    ggtitle(GO.name) +
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white") 
  ggsave(paste0(output.dir, "Heatmap_Day4_", celltype.name, "_only_group_by_treatment_", GO.name, ".pdf"), plot = p1)
  
}

##################
#a. Neutrophils: 
##################
#neutrophil degranulation (GO:0043312), neutrophil mediated immunity (GO:0002446...), Phagosome+Lysosome (KEGG)

# subset seurat object for cell_type in question
DefaultAssay(day4.seurat) <- "SCT"

Idents(day4.seurat) <- day4.seurat@meta.data$cell_type
Neutrophil.seurat <- subset(day4.seurat, idents = "Neutrophils")
table(Neutrophil.seurat@meta.data$Treatment)
Idents(Neutrophil.seurat) <- Neutrophil.seurat@meta.data$Treatment

#Neutrophil.seurat <- subset(Neutrophil.seurat, downsample = 500)


########################
# calculate enrichment
########################

# filter for upregulated genes
sig.up <- DEG.list$Neutrophils %>%
  dplyr::filter(p_val_adj < 0.05) %>%
  dplyr::filter(avg_log2FC > 0) 

# List of genes to use as input.
up.genes <- unique(rownames(sig.up))

# Retrieve the enriched terms.
up.enriched.terms <- enrichR::enrichr(up.genes, dbs_use)

term.table <- up.enriched.terms$GO_Biological_Process_2021 %>%
  dplyr::select(c(Term, Overlap, Adjusted.P.value, Odds.Ratio))



################################
# GO Biological process terms 
################################

# Note that the genes in degranulation and neutrophil mediated immunity gene sets are identical. this is not an error just a reflection of the overlap in these genesets

# show upregulated GO Biological process terms
up.enriched.terms$GO_Biological_Process_2021 %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  pull(Term)

# Degranulation - GO:0043312
GO.Degranulation <- up.enriched.terms$GO_Biological_Process_2021 %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  dplyr::filter(grepl("GO:0043312", Term)) %>%
  pull(Genes) %>%
  strsplit(";") %>%
  unlist()

GO.Degranulation <- stringr::str_to_title(GO.Degranulation)



library("clusterProfiler")


dotplot(ego, showCategory=30) + ggtitle("dotplot for ORA")
# Simulated data example
df <- data.frame(
  Description = c("T cell activation", "JAK-STAT cascade", "Cytokine production", 
                  "Regulation of cell killing", "STAT cascade", "Cellular response to lipid", 
                  "Protein folding", "Negative regulation of multi-organism process"),
  GeneRatio = c(0.12, 0.08, 0.15, 0.10, 0.07, 0.09, 0.11, 0.06),
  p.adjust = c(0.02, 0.05, 0.01, 0.03, 0.04, 0.05, 0.02, 0.01),
  Count = c(25, 40, 30, 20, 10, 15, 18, 12),
  group = rep(c("Cbl", "WT"), each = 4)
)

# Plotting
ggplot(df, aes(x = group, y = reorder(Description, -GeneRatio), size = GeneRatio, color = p.adjust)) +
  geom_point() +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "Biological Process", x = "Group", y = NULL)





head(ego)
head(ggo)
# Entrez gene ID
head(gene)
edo2 <- gseGO(up.genes, ont = "BP", OrgDb = org.Mm.eg.db)
dotplot(edo, showCategory=30) + ggtitle("dotplot for ORA")
dotplot(edo2, showCategory=30) + ggtitle("dotplot for GSEA")


?SCpubr::do_DotPlot()


DotPlot_scCustom(Neutrophil.seurat, features = GO.Degranulation, flip_axes = T)



# Filter and plot 
go.heatmap(input.data = Neutrophil.seurat, 
           GO.genes = GO.Degranulation, 
           celltype.name = "Neutrophils", 
           GO.name = "Degranulation_0043312")


# Neutrophil mediated immunity - GO:0002446
GO.Neut.immunity <- up.enriched.terms$GO_Biological_Process_2021 %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  dplyr::filter(grepl("GO:0002446", Term)) %>%
  pull(Genes) %>%
  strsplit(";") %>%
  unlist()

GO.Neut.immunity <- stringr::str_to_title(GO.Neut.immunity)

# Filter and plot 
go.heatmap(input.data = Neutrophil.seurat, 
           GO.genes = GO.Neut.immunity, 
           celltype.name = "Neutrophils", 
           GO.name = "Neut_Immunity_0002446")



################################
# KEGG pathways
################################

# show upregulated Kegg terms
up.enriched.terms$KEGG_2019_Mouse %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  pull(Term)

# Lysosome and Phagosome (KEGG)
KEGG.Phago.lyso <- up.enriched.terms$KEGG_2019_Mouse %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  dplyr::filter(grepl(paste0(c("Lysosome", "Phagosome"), collapse = "|"), Term)) %>%
  pull(Genes) %>%
  strsplit(";") %>%
  unlist()

KEGG.Phago.lyso <- stringr::str_to_title(KEGG.Phago.lyso)

# Filter and plot 
go.heatmap(input.data = Neutrophil.seurat, 
           GO.genes = KEGG.Phago.lyso, 
           celltype.name = "Neutrophils", 
           GO.name = "KEGG_Lyso_Phagosome")

rm(Neutrophil.seurat)


##################
#b. Monocytes: 
##################
#positive regulation of response to exter.. (GO), positive regulation of inflammatory response (GO), 
#Phagosome (KEGG), Toll-like receptor signaling pathway (KEGG)

# subset seurat object for cell_type in question
DefaultAssay(day4.seurat) <- "SCT"

Idents(day4.seurat) <- day4.seurat@meta.data$cell_type
Monocytes.seurat <- subset(day4.seurat, idents = "Monocytes")
table(Monocytes.seurat@meta.data$Treatment)
Idents(Monocytes.seurat) <- Monocytes.seurat@meta.data$Treatment

Monocytes.seurat <- subset(Monocytes.seurat, downsample = 500)


########################
# calculate enrichment
########################

# filter for upregulated genes
sig.up <- DEG.list$Monocytes %>%
  dplyr::filter(p_val_adj < 0.05) %>%
  dplyr::filter(avg_log2FC > 0) 

# List of genes to use as input.
up.genes <- unique(rownames(sig.up))

# Retrieve the enriched terms.
up.enriched.terms <- enrichR::enrichr(up.genes, dbs_use)


################################
# GO Biological process terms 
################################

# show upregulated GO Biological process terms
up.enriched.terms$GO_Biological_Process_2021 %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  pull(Term)

# positive regulation of response to external stimulus (GO:0032103)
GO.external.stim <- up.enriched.terms$GO_Biological_Process_2021 %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  dplyr::filter(grepl("GO:0032103", Term)) %>%
  pull(Genes) %>%
  strsplit(";") %>%
  unlist()

GO.external.stim <- stringr::str_to_title(GO.external.stim)

# Filter and plot 
go.heatmap(input.data = Monocytes.seurat, 
           GO.genes = GO.external.stim, 
           celltype.name = "Monocytes", 
           GO.name = "Pos_reg_external_stim_0032103")

################################
# KEGG pathways
################################

# show upregulated Kegg terms
up.enriched.terms$KEGG_2019_Mouse %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  pull(Term)

# Phagosome (KEGG)
KEGG.Phago <- up.enriched.terms$KEGG_2019_Mouse %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  dplyr::filter(grepl("Phagosome", Term)) %>%
  pull(Genes) %>%
  strsplit(";") %>%
  unlist()

KEGG.Phago <- stringr::str_to_title(KEGG.Phago)

# Filter and plot 
go.heatmap(input.data = Monocytes.seurat, 
           GO.genes = KEGG.Phago, 
           celltype.name = "Monocytes", 
           GO.name = "KEGG_Phagosome")


# Toll-like receptor signaling pathway (KEGG)
KEGG.TLR <- up.enriched.terms$KEGG_2019_Mouse %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  dplyr::filter(grepl("Toll-like receptor signaling pathway", Term)) %>%
  pull(Genes) %>%
  strsplit(";") %>%
  unlist()

KEGG.TLR <- stringr::str_to_title(KEGG.TLR)

# Filter and plot 
go.heatmap(input.data = Monocytes.seurat, 
           GO.genes = KEGG.TLR, 
           celltype.name = "Monocytes", 
           GO.name = "KEGG_TLR")

rm(Monocytes.seurat)

####################
#c. Macrophages: 
####################
#pattern-recognition receptor signaling (GO),
#Toll-like receptor signaling pathway (KEGG), Antigen processing and presentation (KEGG), Phagosome (KEGG)


# subset seurat object for cell_type in question
DefaultAssay(day4.seurat) <- "SCT"

Idents(day4.seurat) <- day4.seurat@meta.data$cell_type
Macrophages.seurat <- subset(day4.seurat, idents = "Macrophages")
table(Macrophages.seurat@meta.data$Treatment)
Idents(Macrophages.seurat) <- Macrophages.seurat@meta.data$Treatment

Macrophages.seurat <- subset(Macrophages.seurat, downsample = 500)


########################
# calculate enrichment
########################

# filter for upregulated genes
sig.up <- DEG.list$Macrophages %>%
  dplyr::filter(p_val_adj < 0.05) %>%
  dplyr::filter(avg_log2FC > 0) 

# List of genes to use as input.
up.genes <- unique(rownames(sig.up))

# Retrieve the enriched terms.
up.enriched.terms <- enrichR::enrichr(up.genes, dbs_use)


################################
# GO Biological process terms 
################################

# show upregulated GO Biological process terms
up.enriched.terms$GO_Biological_Process_2021 %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  pull(Term)

# pattern recognition receptor signaling pathway (GO:0002221)
GO.pattern <- up.enriched.terms$GO_Biological_Process_2021 %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  dplyr::filter(grepl("GO:0002221", Term)) %>%
  pull(Genes) %>%
  strsplit(";") %>%
  unlist()

GO.pattern <- stringr::str_to_title(GO.pattern)

# Filter and plot 
go.heatmap(input.data = Macrophages.seurat, 
           GO.genes = GO.pattern, 
           celltype.name = "Macrophages", 
           GO.name = "PRR_signaling_pathway_0002221")

################################
# KEGG pathways
################################

# show upregulated Kegg terms
up.enriched.terms$KEGG_2019_Mouse %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  pull(Term)

# Toll-like receptor signaling pathway (KEGG)
KEGG.TLR <- up.enriched.terms$KEGG_2019_Mouse %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  dplyr::filter(grepl("Toll-like receptor signaling pathway", Term)) %>%
  pull(Genes) %>%
  strsplit(";") %>%
  unlist()

KEGG.TLR <- stringr::str_to_title(KEGG.TLR)

# Filter and plot 
go.heatmap(input.data = Macrophages.seurat, 
           GO.genes = KEGG.TLR, 
           celltype.name = "Macrophages", 
           GO.name = "KEGG_TLR")

# Phagosome (KEGG)
KEGG.Phago <- up.enriched.terms$KEGG_2019_Mouse %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  dplyr::filter(grepl("Phagosome", Term)) %>%
  pull(Genes) %>%
  strsplit(";") %>%
  unlist()

KEGG.Phago <- stringr::str_to_title(KEGG.Phago)

# Filter and plot 
go.heatmap(input.data = Macrophages.seurat, 
           GO.genes = KEGG.Phago, 
           celltype.name = "Macrophages", 
           GO.name = "KEGG_Phagosome")



# Antigen processing and presentation (KEGG)
KEGG.APP <- up.enriched.terms$KEGG_2019_Mouse %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  dplyr::filter(grepl("Antigen processing and presentation", Term)) %>%
  pull(Genes) %>%
  strsplit(";") %>%
  unlist()

KEGG.APP <- stringr::str_to_title(KEGG.APP)

# Filter and plot 
go.heatmap(input.data = Macrophages.seurat, 
           GO.genes = KEGG.APP, 
           celltype.name = "Macrophages", 
           GO.name = "KEGG_Antigen_Processing")

###########
#d. DCs: 
###########
# inflammatory response (GO:0006954)
#Toll-like receptor signaling pathway (KEGG), 

# subset seurat object for cell_type in question
DefaultAssay(day4.seurat) <- "SCT"

Idents(day4.seurat) <- day4.seurat@meta.data$cell_type
DC.seurat <- subset(day4.seurat, idents = "DC")
table(DC.seurat@meta.data$Treatment)
Idents(DC.seurat) <- DC.seurat@meta.data$Treatment

DC.seurat <- subset(DC.seurat, downsample = 300)


########################
# calculate enrichment
########################

# filter for upregulated genes
sig.up <- DEG.list$DC %>%
  dplyr::filter(p_val_adj < 0.05) %>%
  dplyr::filter(avg_log2FC > 0) 

# List of genes to use as input.
up.genes <- unique(rownames(sig.up))

# Retrieve the enriched terms.
up.enriched.terms <- enrichR::enrichr(up.genes, dbs_use)


################################
# GO Biological process terms 
################################

# show upregulated GO Biological process terms
up.enriched.terms$GO_Biological_Process_2021 %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  pull(Term)

# inflammatory response (GO:0006954)
GO.inflam <- up.enriched.terms$GO_Biological_Process_2021 %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  dplyr::filter(grepl("GO:0006954", Term)) %>%
  pull(Genes) %>%
  strsplit(";") %>%
  unlist()

GO.inflam <- stringr::str_to_title(GO.inflam)

# Filter and plot 
go.heatmap(input.data = DC.seurat, 
           GO.genes = GO.inflam, 
           celltype.name = "DC", 
           GO.name = "Inflam_0006954")

################################
# KEGG pathways
################################

# show upregulated Kegg terms
up.enriched.terms$KEGG_2019_Mouse %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  pull(Term)

# Toll-like receptor signaling pathway (KEGG)
KEGG.TLR <- up.enriched.terms$KEGG_2019_Mouse %>%
  dplyr::filter(Adjusted.P.value < 0.05) %>%
  dplyr::filter(grepl("Toll-like receptor signaling pathway", Term)) %>%
  pull(Genes) %>%
  strsplit(";") %>%
  unlist()

KEGG.TLR <- stringr::str_to_title(KEGG.TLR)

# Filter and plot 
go.heatmap(input.data = DC.seurat, 
           GO.genes = KEGG.TLR, 
           celltype.name = "DC", 
           GO.name = "KEGG_TLR")











FDR.theshold <- 0.01
logFC.threshold <- 1

# filter data
input.data <- DEG.list$Macrophages %>%
  dplyr::rename(logFC = avg_log2FC) %>%
  dplyr::rename(FDR = p_val_adj) %>%
  filter(FDR < FDR.theshold)

# color points
volcano.col <- usefulfunctions::colour.points.volcano(input.data,
                                                      increase.col = "Red",
                                                      decreased.col = "Blue",
                                                      FDR.cutoff = FDR.theshold, 
                                                      logFC.cutoff = logFC.threshold)
# plot heatmap
   
      usefulfunctions::custom.enhanced.volcano(input.data, 
                                               selectLab = KEGG.APP,
                                               pCutoff = FDR.theshold,
                                               FCcutoff = logFC.threshold,
                                               title = paste0(names(DEG.list)[i], " PDS vs Vehicle"),
                                               subtitle = paste0("adjPval threshold = ", FDR.theshold, " & log2FC threshold = ", logFC.threshold),
                                               colCustom = volcano.col)

















# firstly, extract vectors of genes in the above requested pathways 
custom.genes <- c("Il1b", "Il6", "Il12a", "Il12b", "Ifng", "Tnf", "Ccl2", "Ccl3", "Ccl4", "Ccl5", "Cxcl1", "Cxcl9", "Cxcl10") # Cxcl13 not present in CD45Pos dataset






















```



#The Volcanos should show nicely that the immune cells are activated and mediating immunity after PDS treatment. If the "custom genes" (e.g. CCL3, CCL4, Il-12, IL-6, Il-1b,etc.) are not already in the GOs would be cool if you could highlight them as well to show that PDS treatment induces the production of those.


#Additionally it would be cool if you could dive into the deeper analysis of the T cells on day 9 and see if there are any differences in smaller subsets in either numbers or expression between PDS and Vehicle. Will probably not make the cut into this paper, but in case reviewers are asking about the T cells or ortherwise we're planning another paper on PDS and T cells for which that would be great.















## Imputation {.tabset}

### Imputation calculation

```{r seurat_imputation, eval = FALSE}

output.dir <- "results/imputation/"

# Create output directory
if(!dir.exists(paste0(output.dir))){
  dir.create(paste0(output.dir), 
             recursive = T)
}

######################
# Impute values 
######################
if(long.compute){
  day4.seurat <- RunALRA(day4.seurat, 
                             genes.use = rownames(day4.seurat))

  
  # save object 
  SaveSeuratRds(day4.seurat, "saves/CD45Pos_Day4_only_imputed.rds")
  
  }

# 0.00% of the values became negative in the scaling process and were set to zero
# The matrix went from 5.72% nonzero to 27.74% nonzero


####################
# Calculate k value
####################
if(long.compute){
  
  Idents(day4.seurat) <- day4.seurat@meta.data$seurat_clusters
  day4.seurat.small <- subset(day4.seurat, downsample = 100)
  
  ALRA.out <- RunALRA(day4.seurat.small,
                      k.only = TRUE)
  
  ggouts <- ALRAChooseKPlot(ALRA.out)
  
  ggouts
  dev.copy(pdf, paste0(output.dir, "Imputation_K_plot.pdf"))
  dev.off()
  
}


# set default assay back to RNA
DefaultAssay(day4.seurat) <- "SCT"

```
