---
title: "CD45Pos_dataset_annotation"
author: "Dillon Corvino"
date: "25/01/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

Annotation stage for the CD45+ compartment: SingleR cluster annotation,
manual review/simplification of labels, and export of the final annotated
object to `saves/CD45pos_seurat.rds`. This is preprocessing/methods
provenance -- downstream scripts (`manuscript_figures.R`,
`scripts/archive/02_CD45Pos_exploratory.Rmd`) load that saved object rather
than re-deriving it. Split out of the original combined
`02_CD45Pos_data_analysis.Rmd` on 2026-08-14; the exploratory analyses that
used to follow this content now live in
`scripts/archive/02_CD45Pos_exploratory.Rmd`.

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# Data 
# B16 - OVA 

# PDS = I.T - bind Gquadraplex
# day 4 post-treatment 
# day 9 post-treatment

# Research questions 
# How does PDS induces an anti-tumor immune response?
# Known that after PDS treatment, certain DAMPs are induced, intratumoral chemokine levels go up drastically and subsequntly immune cells are recruited into the tumor.
# Main research questions are:
#1. Which cell subsets are producing the chemokines
#2. Are there differences between macrophages and DCs between PDS and Vehicle (because there are DAMPs that should activate them)
#3. How does the treatment effect the tumor cells (is it possible to discriminate them from other tissue cells, if not they express ovalbumin...this should make it at least possible to identify them.)




# key genes
# IL1b, IL6, IL12, IFNG, TNFA, 

# CCL2,3,4,5 
# CXCL1, CXCL9,10,13 





```


### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)


# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- FALSE
do.plot <- TRUE
quick.load <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())


library("Seurat")
library("SeuratObject")
library("SeuratData")
library("SeuratWrappers")
library("dplyr")
library("scCustomize")
library("tibble")


# load output from preprocessing script
if(long.compute){
  
  CD45pos.seurat <- LoadSeuratRds("saves/Preprocessed_CD45Pos.rds")
  
  # set levels for easier downstream plotting 
  CD45pos.seurat@meta.data$Treatment <- factor(CD45pos.seurat@meta.data$Treatment, levels = c("Vehicle", "PDS"))
  CD45pos.seurat@meta.data$Time <- factor(CD45pos.seurat@meta.data$Time, levels = c("Day4", "Day9"))
  CD45pos.seurat@meta.data$SampleID <- factor(CD45pos.seurat@meta.data$SampleID, levels = c("Day4_Vehicle_CD45Pos", "Day4_PDS_CD45Pos", "Day9_Vehicle_CD45Pos", "Day9_PDS_CD45Pos"))
  
}else if(quick.load){
  
  # load cleaned dataset 
  CD45pos.seurat <- LoadSeuratRds("saves/CD45pos_seurat.rds")
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
}


```



## Cluster annotation {.tabset}

### SingleR cluster annotation
```{r SingleR_cluster_annotation}

# Notes:
# there is an error with collect() when trying to load celldex datasets
# i think its an error with biocmanager and has been fixed with a very recent update. > 3.16 
# but im running r 4.2.3 and this would require 4.3
# not keen to upgrade right now 
# so will try a work around of downgrading dbplyr to version 2.3.4 from 2.4 which im currently running 
# need to downgrade to this version 
#devtools::install_version("dbplyr", version = "2.3.4")
#devtools::install_version("dbplyr", version = "2.4")
# in theory if biocmanager is up to date, could just run below and should fix the problem
#library(BiocManager) # 3.16
#install("BiocFileCache")
# also:
# need to ensure matrixStats is at version 1.1.0 
#remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts
#BiocManager::install(version = "3.18")



if(long.compute){
  
  output.dir <- "results/CD45Pos_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # load required packages
  library("SingleR")
  library("celldex")
  
  
  ##################################
  # Using MouseRNAseqData
  ##################################
  
  ref.mouse <-  celldex::MouseRNAseqData()
  
  
  # Benayoun Lab that identified, downloaded and processed data sets on GEO that corresponded to sorted cell types (Benayoun et al., 2019).
  
  # The dataset contains 358 mouse RNA-seq samples annotated to 18 main cell types
  # further split into 28 subtypes under label.fine 
  
  
  ######################
  # Fine labels 
  ######################
  SingleR.pred <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45pos.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.fine)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
  #SingleR::plotDeltaDistribution(SingleR.pred)
  #SingleR::plotScoreDistribution(SingleR.pred)
  
  ######################
  # Main labels 
  ######################
  
  SingleR.pred <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45pos.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.main)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_main_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
  
  #####################
  # Use ImmGenData
  #####################
  # Info: Mouse, Microarray, 830 samples, 20 main labels, 253 fine lables, Hematopoietic & Immune focus
  
  ImmGen.data <- celldex::ImmGenData()
  
  
  # Fine labels
  SingleR.pred.fine <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                               ref = ImmGen.data, 
                               method = "cluster",
                               clusters = CD45pos.seurat@meta.data$seurat_clusters,
                               labels = ImmGen.data$label.fine)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred.fine, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_ImmGen_db.pdf"), plot = p1)
  
  
  # Main labels
  SingleR.pred.main <- SingleR(test = CD45pos.seurat@assays$SCT$data, 
                               ref = ImmGen.data, 
                               method = "cluster",
                               clusters = CD45pos.seurat@meta.data$seurat_clusters,
                               labels = ImmGen.data$label.main)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred.main, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_main_labels_ImmGen_db.pdf"), plot = p1)
  
  
}


```


### Transfer SingleR annotations to seurat object 
```{r transfer_singleR_annotations}


if(long.compute){
  
  output.dir <- "results/CD45Pos_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  
  ########################################
  # transfer labels to seurat object 
  ########################################
  
  table(predicted=SingleR.pred.main$labels, cluster=rownames(SingleR.pred.main))
  table(predicted=SingleR.pred.fine$labels, cluster=rownames(SingleR.pred.fine))
  
  # Main labels
  main.labels <- SingleR.pred.main$labels
  names(main.labels) <- rownames(SingleR.pred.main)
  
  # Fine labels
  fine.labels <- SingleR.pred.fine$labels
  names(fine.labels) <- rownames(SingleR.pred.fine)
  
  ####################
  # rename clusters 
  ####################
  
  # Main labels
  Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$seurat_clusters
  CD45pos.seurat <- Rename_Clusters(CD45pos.seurat, new_idents = main.labels)
  CD45pos.seurat@meta.data$SingleR_main <- Idents(CD45pos.seurat)
  
  # Fine labels
  Idents(CD45pos.seurat) <- CD45pos.seurat@meta.data$seurat_clusters
  CD45pos.seurat <- Rename_Clusters(CD45pos.seurat, new_idents = fine.labels)
  CD45pos.seurat@meta.data$SingleR_fine <- Idents(CD45pos.seurat)
  
  
  ############
  # Visualise
  ############
  
  # Main label
  p.main <- scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                                          group.by = "SingleR_main", 
                                          label = TRUE,
                                          reduction = "umap")
  
  print(p.main)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_main.pdf"))
  dev.off()
  
  
  # Fine label
  p.fine <- scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                                          group.by = "SingleR_fine", 
                                          label = TRUE,
                                          reduction = "umap")
  
  print(p.fine)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_fine.pdf"))
  dev.off()
  
  
  # side-by-side with cluster numbers
  p.clusters <- scCustomize::DimPlot_scCustom(CD45pos.seurat, 
                                              group.by = "seurat_clusters", 
                                              label = TRUE,
                                              reduction = "umap")
  
  # clusters plus main
  print(p.clusters | p.main)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_main_plus_clusters.pdf"))
  dev.off()
  
  # clusters plus fine
  print(p.clusters | p.fine)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_fine_plus_clusters.pdf"))
  dev.off()
  
  
}










```


### Verify cluster annotation
```{r verify_and_finalise_cluster_annotation}


if(long.compute){
  
  output.dir <- "results/CD45Pos_SingleR_annotation_validation/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # overview of data 
  table(CD45pos.seurat@meta.data$seurat_clusters, CD45pos.seurat@meta.data$SingleR_fine)
  
  table(CD45pos.seurat@meta.data$seurat_clusters, CD45pos.seurat@meta.data$SingleR_main)
  table(CD45pos.seurat@meta.data$SingleR_main, CD45pos.seurat@meta.data$SingleR_fine)
  
  # inconsistency between main and fine annotation with respect to T-cell populations
  # Tcells in main are annotated as NKT
  # NKT in main are annotated as treg 
  
  
  # also seurat clusters nicely line up with singleR fine annotations, i.e one cluster per annotation 
  # only monocytes is with higher resolution using seurat clusters, therefore consider seperating into monocytes cluster X and cluster Y
  
  
  # look at annotations and simplify 
  
  
  #################################################
  # calculate DEGs using fine level resolution 
  #################################################
  
  CD45pos.seurat <- PrepSCTFindMarkers(CD45pos.seurat)
  
  # set assay
  DefaultAssay(CD45pos.seurat) <- "SCT"
  
  
  # set idents
  Idents(CD45pos.seurat) <- "SingleR_fine"
  
  CD45Pos.de <- FindAllMarkers(CD45pos.seurat,
                               assay = "SCT",
                               only.pos = TRUE, 
                               verbose = TRUE)
  
  
  # save to file 
  write.csv(CD45Pos.de, paste0(output.dir, "CD45Pos_DEGs_singleR_fine.csv"))
  
  
  ##############
  # plot DEGs
  ##############
  
  pseudo.cd45pos <- AggregateExpression(CD45pos.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("SingleR_fine"))
  
  
  
  goi <- CD45Pos.de %>%
    filter(p_val_adj < 0.05) %>%
    group_by(cluster) %>%
    top_n(10, avg_log2FC) %>%
    pull(gene)
  
  
  DoHeatmap(pseudo.cd45pos, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_top_10_singleR_fine_DEGs_pseudobulk.pdf"))
  dev.off()
  
  # populations are very cleanly resolved 
  
  
  
  
  # iterate over the top degs for each cluster 
  cluster.id <- unique(CD45Pos.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- CD45Pos.de %>%
      filter(p_val_adj < 0.05) %>%
      filter(cluster == cluster.id[i]) %>%
      top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.cd45pos, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE) + 
        scico::scale_fill_scico(palette = "batlow", 
                                direction = 1, 
                                na.value = "white")
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
  
  # happy with using fine level annotation 
  # except, defining NKT from T-cells is a challenge. will just call them T-cells as at main level annotation these are T-cells
  # NKTs at fine level annotation are split into Treg and Teff - and these look to make sense
  # will therefore use fine level annotation fo all, but change NKT to just T-cell 
  # will also simplify the annotation for all clusters but keep their distinct'ness 
  
  
}






```


### Clean up seurat object metadata and export
```{r clean_seurat_object_metadata}

if(long.compute){
  # reannotate singleR_fine cluster annotations based on above analysis 
  # namely, simplify the annotations 
  # also, convert NKT to just T-cells
  
  CD45pos.seurat <- backup.object.for.now
  
  # note the clusters are essentially one cell type one cluster. except for macrophages which is present in two clusters of macrophages and monocytes present in 3 clusters
  table(CD45pos.seurat@meta.data$SingleR_fine, CD45pos.seurat@meta.data$seurat_clusters)
  
  library(stringr) # For using the `str_extract` function
  
  
  # simplify the annotations, to just the cell_type, but keep cluster_seperation 
  CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
    # Annotate clusters based on pattern matching
    mutate(cluster_anno = case_when(
      # Use `str_extract` to extract the cell type from SingleR_fine
      TRUE ~ paste0(
        str_extract(SingleR_fine, "^[^(]+"), # Extract the cell type name
        "_", seurat_clusters
      )
    )) %>%
    # Remove any spaces in the cluster_anno name
    mutate(cluster_anno = str_replace_all(cluster_anno, " ", ""))
  
  unique(CD45pos.seurat@meta.data$cluster_anno)
  table(CD45pos.seurat@meta.data$SingleR_fine, CD45pos.seurat@meta.data$cluster_anno)
  
  
  
  # simplify the annotations, so there are numbered within cell_type
  
  
  
  # little code to count up how many distinct cell types there are 
  CD45pos.seurat@meta.data %>%
    dplyr::select(c(cluster_anno, SingleR_fine)) %>%
    # Extract cell_type from cluster_anno
    mutate(cell_type = str_extract(cluster_anno, "^[^_]+")) %>%
    # Group by cell_type
    group_by(cell_type) %>%
    # Calculate the number of unique annotations for each cell_type and rank them
    mutate(
      # Calculate the count of unique cluster_anno
      unique_count = n_distinct(cluster_anno) ) %>%
    ungroup() %>%
    distinct(cluster_anno, .keep_all = TRUE)
  
  
  
  CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
    # Extract cell_type from cluster_anno
    mutate(cell_type = str_extract(cluster_anno, "^[^_]+")) %>%
    mutate(cluster_anno_new = case_when(
      cluster_anno == "Macrophages_1" ~ "Macrophages_1", 
      cluster_anno == "Macrophages_9" ~ "Macrophages_2", 
      
      cluster_anno == "Bcells_8" ~ "B_cells", 
      
      cluster_anno == "DC_15" ~ "pDC", 
      cluster_anno == "DC_14" ~ "DC_1", 
      cluster_anno == "DC_12" ~ "DC_2",
      
      cluster_anno == "Monocytes_4" ~ "Monocytes_1", 
      cluster_anno == "Monocytes_5" ~ "Monocytes_2", 
      cluster_anno == "Monocytes_7" ~ "Monocytes_3", 
      
      cluster_anno == "Macrophages_1" ~ "Macrophages_1", 
      cluster_anno == "Macrophages_9" ~ "Macrophages_2", 
      
      cluster_anno == "Neutrophils_3" ~ "Neutrophils_1", 
      cluster_anno == "Neutrophils_6" ~ "Neutrophils_2", 
      
      cluster_anno == "Stromalcells_16" ~ "Stromal_cells", 
      
      cluster_anno == "NKcells_2" ~ "NK_cells", 
      
      cluster_anno == "Basophils_17" ~ "Basophils", 
      
      cluster_anno == "Tcells_0" ~ "T_cells_1", 
      cluster_anno == "Tcells_10" ~ "T_cells_2", 
      cluster_anno == "Tcells_13" ~ "T_cells_3", 
      cluster_anno == "Tcells_11" ~ "Tregs", 
      TRUE ~ "ERROR")) %>%
    mutate(cluster_anno = cluster_anno_new) %>%
    select(-cluster_anno_new)
  
  
  unique(CD45pos.seurat@meta.data$cluster_anno)
  unique(CD45pos.seurat@meta.data$SingleR_fine)
  
  
  # clean up object, remove unnecessary columns keep only what is needed
  CD45pos.seurat@meta.data <- CD45pos.seurat@meta.data %>%
    select(c("orig.ident", "nCount_RNA", "nFeature_RNA", "nCount_SCT", "nFeature_SCT", "percent_mito", "percent_ribo", "percent_mito_ribo", "log10GenesPerUMI",  "SampleID", "Time", "Compartment", "Treatment",  "cluster_anno", "seurat_clusters", "SCT_snn_res.0.4", "SingleR_main", "SingleR_fine", "cell_type"))
  
  
  # Display the first few rows of the updated data frame to verify the results
  head(as.data.frame(CD45pos.seurat@meta.data[1:10, ]))
  
  
  
  table(CD45pos.seurat@meta.data$SingleR_fine, CD45pos.seurat@meta.data$cell_type)
  table(CD45pos.seurat@meta.data$SingleR_fine, CD45pos.seurat@meta.data$cluster_anno)
  table(CD45pos.seurat@meta.data$cell_type, CD45pos.seurat@meta.data$cluster_anno)
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  # clean up sample ID annotation - no need to say CD45Pos still 
  
  CD45pos.seurat@meta.data$SampleID <- gsub("_CD45Pos", "", CD45pos.seurat@meta.data$SampleID)
  
  unique(CD45pos.seurat@meta.data$cluster_anno)
  head(CD45pos.seurat@meta.data)
  
  # set factorisation 
  CD45pos.seurat@meta.data$Treatment <- factor(CD45pos.seurat@meta.data$Treatment, 
                                               levels = c("Vehicle", "PDS"))
  
  CD45pos.seurat@meta.data$Time <- factor(CD45pos.seurat@meta.data$Time, 
                                          levels = c("Day4", "Day9"))
  
  CD45pos.seurat@meta.data$SampleID <- factor(CD45pos.seurat@meta.data$SampleID, 
                                              levels = c("Day4_Vehicle", "Day9_Vehicle", "Day4_PDS", "Day9_PDS"))
  
  
  CD45pos.seurat@meta.data$cluster_anno <- factor(CD45pos.seurat@meta.data$cluster_anno)
  
  # save output
  SaveSeuratRds(CD45pos.seurat, file = "saves/CD45pos_seurat.rds")
}


```



