---
title: "CD45Neg_dataset_analysis"
author: "Dillon Corvino"
date: "25/01/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# Data 
# B16 - OVA 

# PDS = I.T - bind Gquadraplex
# day 4 post-treatment 


# Research questions 
# How does PDS induces an anti-tumor immune response?
# Known that after PDS treatment, certain DAMPs are induced, intratumoral chemokine levels go up drastically and subsequntly immune cells are recruited into the tumor.
# Main research questions are:
#1. Which cell subsets are producing the chemokines
#2. Are there differences between macrophages and DCs between PDS and Vehicle (because there are DAMPs that should activate them)
#3. How does the treatment effect the tumor cells (is it possible to discriminate them from other tissue cells, if not they express ovalbumin...this should make it at least possible to identify them.)




# key genes
# IL1b, IL6, IL12, IFNG, TNFA, 

# CCL2,3,4,5 
# CXCL1, CXCL9,10,13 





```


### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)


# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- FALSE
do.plot <- TRUE
quick.load <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())

# function found in Setup.R script
ggplot2_switch_version(desired_version = "3.4.4")

library("Seurat")
library("SeuratObject")
#library("SeuratData")
library("SeuratWrappers")
library("dplyr")
library("scCustomize")
library("tibble")
library("ggplot2")


# load output from preprocessing script
if(long.compute){
  
  CD45neg.seurat <- LoadSeuratRds("saves/Preprocessed_CD45Neg.rds")
  
  # set levels for easier downstream plotting 
  CD45neg.seurat@meta.data$Treatment <- factor(CD45neg.seurat@meta.data$Treatment, levels = c("Vehicle", "PDS"))
  
}else if(quick.load){
  
  # load cleaned dataset 
  CD45neg.seurat <- LoadSeuratRds("saves/CD45neg_seurat.rds")
  DefaultAssay(CD45neg.seurat) <- "SCT"
  
  # Create new metadata column that groups all tumor together 
  CD45neg.seurat@meta.data <- CD45neg.seurat@meta.data %>%
    dplyr::mutate(celltype = case_when(Disease == "tumor" ~ "tumor",
                                       Disease == "normal" ~ cluster_anno))
  
  
}
```


## Clean dataset {.tabset}

## Identifying Immune cell contamination - SingleR cluster annotation
```{r SingleR_cluster_annotation_initial}

# Notes:
# there is an error with collect() when trying to load celldex datasets
# i think its an error with biocmanager and has been fixed with a very recent update. > 3.16 
# but im running r 4.2.3 and this would require 4.3
# not keen to upgrade right now 
# so will try a work around of downgrading dbplyr to version 2.3.4 from 2.4 which im currently running 
# need to downgrade to this version 
#devtools::install_version("dbplyr", version = "2.3.4")
#devtools::install_version("dbplyr", version = "2.4")
# in theory if biocmanager is up to date, could just run below and should fix the problem
#library(BiocManager) # 3.16
#install("BiocFileCache")
# also:
# need to ensure matrixStats is at version 1.1.0 
#remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts
#BiocManager::install(version = "3.18")

if(long.compute){
  
  output.dir <- "results/CD45Neg_Immune_cell_contamination/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # load required packages
  library("SingleR")
  library("celldex")
  
  
  ##################################
  # Use MouseRNAseqData
  ##################################
  
  ref.mouse <-  celldex::MouseRNAseqData()
  
  
  # Benayoun Lab that identified, downloaded and processed data sets on GEO that corresponded to sorted cell types (Benayoun et al., 2019).
  
  # The dataset contains 358 mouse RNA-seq samples annotated to 18 main cell types
  # further split into 28 subtypes under label.fine 
  
  
  ######################
  # Fine labels 
  ######################
  SingleR.pred <- SingleR(test = CD45neg.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45neg.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.fine)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
  #SingleR::plotDeltaDistribution(SingleR.pred)
  #SingleR::plotScoreDistribution(SingleR.pred)
  
  ######################
  # Main labels 
  ######################
  
  SingleR.pred <- SingleR(test = CD45neg.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45neg.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.main)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_main_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
}

```


### Confirming Immune cell contamination - CD45 and granulocyte linage marker gene visualisation
```{r CD45_and_granulocyte_lineage_gene_visualisation}

if(long.compute){
  
  # create output dir
  output.dir <- paste0("results/CD45Neg_Immune_cell_contamination/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  # list melanocyte genes 
  goi <- c("Ptprc", "Cd14")
  
  # Plot Feature and Density plots iteratively over goi 
  for(i in seq_along(goi)){
    
    # feature plots
    print(
      scCustomize::FeaturePlot_scCustom(CD45neg.seurat, 
                                        reduction = "umap", 
                                        features = goi[i])
      
    )
    
    dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Neg_", goi[i], ".pdf"))
    dev.off()
    
    
    # density plots
    
    print(
      scCustomize::Plot_Density_Custom(CD45neg.seurat, 
                                       reduction = "umap",
                                       features = goi[i], 
                                       custom_palette = batlow.pal)
    )
    dev.copy(pdf, paste0(output.dir, "UMAP_DensityPlot_CD45Neg_", goi[i], ".pdf"))
    dev.off()
    
    
    
  }
  
  
  ##############################
  # Stacked vln plot
  ##############################
  
  scCustomize::Stacked_VlnPlot(CD45neg.seurat, 
                               features = goi,
                               group.by = "seurat_clusters",
                               pt.size = 0)
  
  dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_granulocyte_markers.pdf"))
  dev.off()
  
  
  ###########
  # heatmap
  ###########
  
  # aggregate expression for heatmap visualisation
  pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("seurat_clusters"))
  
  DoHeatmap(pseudo.cd45neg, 
            features = goi, 
            draw.lines = FALSE, 
            angle = 90,
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_granulocyte_markers.pdf"))
  dev.off()
  
}


```


### Remove contamination immune clusters
```{r remove_immune_contamination}

if(long.compute){
  
  # cluster 6 looks to be immune contaimination 
  
  # set immune cell cluster variable
  Immune.cell.cluster <- 6
  
  
  head(CD45neg.seurat@meta.data)
  
  
  # set assay
  DefaultAssay(CD45neg.seurat) <- "SCT"
  
  # set idents
  Idents(CD45neg.seurat) <- "seurat_clusters"
  
  # immune cell contamination cluster is relatively small, sharp drop of in cluster sizes at cluster 5
  table(CD45neg.seurat@meta.data$seurat_clusters) # cluster 6 = 735 cells
  
  # remove this cluster from dataset 
  Idents(CD45neg.seurat) <- CD45neg.seurat@meta.data$seurat_clusters
  CD45neg.seurat <- subset(CD45neg.seurat, idents = Immune.cell.cluster, invert = TRUE)
  unique(Idents(CD45neg.seurat))
  
  
  # drop levels in seurat cluster meta
  CD45neg.seurat$seurat_clusters <- droplevels(CD45neg.seurat$seurat_clusters)
  
}

```


## Renormalise / integrate / dim reduction | after removing contamination immune cells {.tabset}

## re-calculate normalisation and integration for CD45Neg dataset
```{r re_normalise_and_integrate_CD45Neg_dataset}

if(long.compute){
  
  
  # split object up based on sequencing run / in seurat V5 these are splitting the layers
  CD45neg.seurat[["RNA"]] <- split(CD45neg.seurat[["RNA"]],  f = CD45neg.seurat$SampleID)
  
  DefaultAssay(CD45neg.seurat) <- "RNA"
  
  
  
  
  #####################################################
  # Normalise with scTransform and integrate data
  #####################################################
  # note error with more recent version of matrixStats resulting in "useNames = NA is defunct" error. downgrade to matrixStats version 1.1.0 as a workaround
  # https://github.com/satijalab/seurat/issues/7501#issuecomment-1854571904
  #remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts  
  
  
  
  # note RNA assay is split into run level layers, below script will apply normalisation etc layer-wise
  
  CD45neg.seurat <- CD45neg.seurat %>%
    SCTransform(method = "glmGamPoi", 
                vars.to.regress = "percent_mito", 
                vst.flavor = "v2",
                verbose = TRUE) %>%
    RunPCA(reduction.name = "pca_sct") %>%
    RunUMAP(reduction = "pca_sct", 
            dims = 1:30, 
            reduction.name = "umap.unintegrated.sct")
  
  
  DefaultAssay(CD45neg.seurat) <- "SCT"
  
  
  # integrate layers using cca integration and calculate dim redu and cluster identification
  
  # cluster resolution value based on previous estimation 
  clust.res <- 0.4
  
  
  CD45neg.seurat <- CD45neg.seurat %>%
    IntegrateLayers(orig.reduction = "pca_sct",
                    new.reduction = "integrated.cca.sct",
                    method = CCAIntegration,
                    normalization.method = "SCT",
                    verbose = F) %>%
    FindNeighbors(dims = 1:30,
                  reduction = "integrated.cca.sct") %>%
    FindClusters(resolution = clust.res, 
                 cluster.name = "clusters_cca_sct") %>%
    RunUMAP(reduction = "integrated.cca.sct",
            dims = 1:30, 
            reduction.name = "umap")
  
  
}

```


## plot integration after removing contamination immune cells from CD45Neg dataset 

```{r plot_integration_after_removing_contamination}

if(long.compute){
  
  if(do.plot){
    
    # create output dir
    output.dir <- paste0("results/CD45Neg_Integration_validation_after_cleaning/")
    
    if(!dir.exists(output.dir)){
      dir.create(output.dir, 
                 recursive = T)
      
    }
    
    # set variables
    unintegrated.reduction <- "umap.unintegrated.sct"
    integrated.reduction <- "umap"
    title.var <- "SCT"
    
    
    
    # evaluate integration and dim reduction
    
    # set default assay
    DefaultAssay(CD45neg.seurat) <- "SCT"
    
    
    
    vars.plot <- c("Treatment", "SampleID", "seurat_clusters")
    
    
    for(i in seq_along(vars.plot)){
      
      
      # remove "." from name
      name.var <- gsub("\\.", "_", vars.plot[i])
      
      before.int <- DimPlot(CD45neg.seurat,
                            reduction = unintegrated.reduction, 
                            label = T,
                            group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("Unintegrated ", title.var, " Normalisation"))
      
      after.int <- DimPlot(CD45neg.seurat,
                           reduction = integrated.reduction, 
                           label = T,
                           group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("CCA Integration using ", title.var, " Normalisation"))
      
      print(before.int | after.int)
      dev.copy(pdf, paste0(output.dir, "UMAP_before_after_integration_", name.var, "_CD45Neg_dataset.pdf"))
      dev.off()
      
      
      
    }
  }
  
  
}


```


# saved a cleaned object here

## Cluster annotation {.tabset}

## SingleR cluster annotation
```{r SingleR_cluster_annotation}

# Notes:
# there is an error with collect() when trying to load celldex datasets
# i think its an error with biocmanager and has been fixed with a very recent update. > 3.16 
# but im running r 4.2.3 and this would require 4.3
# not keen to upgrade right now 
# so will try a work around of downgrading dbplyr to version 2.3.4 from 2.4 which im currently running 
# need to downgrade to this version 
#devtools::install_version("dbplyr", version = "2.3.4")
#devtools::install_version("dbplyr", version = "2.4")
# in theory if biocmanager is up to date, could just run below and should fix the problem
#library(BiocManager) # 3.16
#install("BiocFileCache")
# also:
# need to ensure matrixStats is at version 1.1.0 
#remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts
#BiocManager::install(version = "3.18")

if(long.compute){
  
  output.dir <- "results/CD45Neg_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # load required packages
  library("SingleR")
  library("celldex")
  
  
  ##################################
  # Use MouseRNAseqData
  ##################################
  
  ref.mouse <-  celldex::MouseRNAseqData()
  
  
  # Benayoun Lab that identified, downloaded and processed data sets on GEO that corresponded to sorted cell types (Benayoun et al., 2019).
  
  # The dataset contains 358 mouse RNA-seq samples annotated to 18 main cell types
  # further split into 28 subtypes under label.fine 
  
  
  ######################
  # Fine labels 
  ######################
  SingleR.pred.fine <- SingleR(test = CD45neg.seurat@assays$SCT$data, 
                               ref = ref.mouse, 
                               method = "cluster",
                               clusters = CD45neg.seurat@meta.data$seurat_clusters,
                               labels = ref.mouse$label.fine)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
  #SingleR::plotDeltaDistribution(SingleR.pred)
  #SingleR::plotScoreDistribution(SingleR.pred)
  
  ######################
  # Main labels 
  ######################
  
  SingleR.pred.main <- SingleR(test = CD45neg.seurat@assays$SCT$data, 
                               ref = ref.mouse, 
                               method = "cluster",
                               clusters = CD45neg.seurat@meta.data$seurat_clusters,
                               labels = ref.mouse$label.main)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_main_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
}

```


### Transfer SingleR annotations to seurat object 
```{r transfer_singleR_annotations}


if(long.compute){
  
  output.dir <- "results/CD45Neg_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  
  ########################################
  # transfer labels to seurat object 
  ########################################
  
  table(predicted=SingleR.pred.main$labels, cluster=rownames(SingleR.pred.main))
  table(predicted=SingleR.pred.fine$labels, cluster=rownames(SingleR.pred.fine))
  
  # Main labels
  main.labels <- SingleR.pred.main$labels
  names(main.labels) <- rownames(SingleR.pred.main)
  
  # Fine labels
  fine.labels <- SingleR.pred.fine$labels
  names(fine.labels) <- rownames(SingleR.pred.fine)
  
  ####################
  # rename clusters 
  ####################
  
  # Main labels
  Idents(CD45neg.seurat) <- CD45neg.seurat@meta.data$seurat_clusters
  CD45neg.seurat <- Rename_Clusters(CD45neg.seurat, new_idents = main.labels)
  CD45neg.seurat@meta.data$SingleR_main <- Idents(CD45neg.seurat)
  
  # Fine labels
  Idents(CD45neg.seurat) <- CD45neg.seurat@meta.data$seurat_clusters
  CD45neg.seurat <- Rename_Clusters(CD45neg.seurat, new_idents = fine.labels)
  CD45neg.seurat@meta.data$SingleR_fine <- Idents(CD45neg.seurat)
  
  
  ############
  # Visualise
  ############
  
  # Main label
  p.main <- scCustomize::DimPlot_scCustom(CD45neg.seurat, 
                                          group.by = "SingleR_main", 
                                          label = TRUE,
                                          reduction = "umap")
  
  print(p.main)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_main.pdf"))
  dev.off()
  
  
  # Fine label
  p.fine <- scCustomize::DimPlot_scCustom(CD45neg.seurat, 
                                          group.by = "SingleR_fine", 
                                          label = TRUE,
                                          reduction = "umap")
  
  print(p.fine)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_fine.pdf"))
  dev.off()
  
  
  # side-by-side with cluster numbers
  p.clusters <- scCustomize::DimPlot_scCustom(CD45neg.seurat, 
                                              group.by = "seurat_clusters", 
                                              label = TRUE,
                                              reduction = "umap")
  
  # clusters plus main
  print(p.clusters | p.main)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_main_plus_clusters.pdf"))
  dev.off()
  
  # clusters plus fine
  print(p.clusters | p.fine)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_fine_plus_clusters.pdf"))
  dev.off()
  
  
}


```

### SCEVAN Copy number analysis
```{r copy_number_analysis}

if(long.compute){
  
  
  output.dir <- "results/CD45Neg_SCEVAN/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  
  #library(devtools)
  #install_github("miccec/yaGST")
  #install_github("AntonioDeFalco/SCEVAN")
  library(SCEVAN)
  
  DefaultAssay(CD45neg.seurat) <- "RNA"
  CD45neg.seurat <- JoinLayers(CD45neg.seurat)
  
  
  count.matrix <- LayerData(CD45neg.seurat, assay = "RNA", layer = "counts")
  
  results.out.single.object <- pipelineCNA(count_mtx = count.matrix, 
                                           sample = "CD45Neg_dataset", 
                                           par_cores = 20, 
                                           organism = "mouse", 
                                           SUBCLONES = TRUE, 
                                           beta_vega = 0.5, 
                                           ClonalCN = TRUE, 
                                           plotTree = TRUE)
  
  CD45neg.seurat <-Seurat::AddMetaData(CD45neg.seurat, metadata = results.out.single.object)
  CD45neg.seurat@meta.data
  
  
  # export object
  saveRDS(results.out.single.object, file = "saves/results_out_single_object.rds")

  # Visualise results of copy number variance analysis
  
  # Tumor or normal identity by Copy Number
  
  DimPlot_scCustom(CD45neg.seurat, 
                   group.by = "class", 
                   reduction = "umap")
  dev.copy(pdf, paste0(output.dir, "UMAP_class_CNA_identity.pdf"))
  dev.off()
  
  # Tumor Subclone ID
  # switch to dimplot, cause NA and 1 end up similar gray if using scCustomise package dimplot function
  
  DimPlot(CD45neg.seurat, 
          group.by = "subclone",
          reduction = "umap")
  
  dev.copy(pdf, paste0(output.dir, "UMAP_subclone_identity.pdf"))
  dev.off()
  
  
  # Single R main identity across class
  DimPlot_scCustom(CD45neg.seurat,
                   group.by = "SingleR_main", 
                   split.by = "class", 
                   reduction = "umap")
  
  dev.copy(pdf, paste0(output.dir, "UMAP_SingleR_main_across_class.pdf"))
  dev.off()
  
  # Single R fine identity across class
  DimPlot_scCustom(CD45neg.seurat,
                   group.by = "SingleR_fine",
                   split.by = "class", 
                   reduction = "umap")
  
  dev.copy(pdf, paste0(output.dir, "UMAP_SingleR_fine_across_class.pdf"))
  dev.off()
  
  # Seurat clusters across class 
  DimPlot_scCustom(CD45neg.seurat,
                   group.by = "seurat_clusters", 
                   split.by = "class", 
                   label = TRUE,
                   reduction = "umap")
  
  dev.copy(pdf, paste0(output.dir, "UMAP_Seurat_clusters_across_class.pdf"))
  dev.off()
  
}


```

### Annotate clusters and clean up seurat object and export
```{r annotate_clusters}



if(long.compute){
  
  output.dir <- "results/CD45Neg_unifying_cluster_annotations/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  DimPlot_scCustom(CD45neg.seurat,
                   group.by = "seurat_clusters", 
                   split.by = "class", 
                   label = TRUE,
                   reduction = "umap")
  
  
  
  # Unify CNA definition of tumor with singleR definitions 
  out.df <- table(CD45neg.seurat@meta.data$seurat_clusters, CD45neg.seurat@meta.data$class)
  row.vals <- rowSums(out.df)
  freq.class.id <- sweep(out.df, 1, row.vals, "/")
  rowSums(freq.class.id*100) # sanity check
  freq.class.id <- freq.class.id*100
  
  
  ggplot(freq.class.id) +
    aes(x = Var1, y = Freq, fill = Var2) +
    geom_col() +
    scale_fill_hue(direction = 1) +
    labs(x = "Seurat_cluster", y = "Frequency", title = "Frequency of normal/tumor in each seurat_cluster") +
    theme_minimal()
  
  ggsave(paste0(output.dir, "Barplot_frequency_of_seurat_cluster_by_class_identity.pdf"), width = 8, height = 4)
  
  # Tumor clusters are: 
  # 0, 1, 4, 5, 6, 8, 9
  
  # Normal clusters are: 
  # 2, 3, 7, 13
  
  # unclear: 
  # 11, 10, 12
  
  # cluster 10 
  # will be called normal 
  # shows up roughly evenly between normal and tumor but doesnt cluster with rest of tumor cells
  # also not predicted to be fibroblast type origin 
  # looks to be erythrocyte origin 
  
  # cluster 11
  # will be called normal (79%)
  # does cluster with tumor cells and is precited to be of fibroblast cell type 
  # but perhaps this is a normal non tumor fibroblast cells
  
  # cluster 12
  # will be called normal (78%)
  # doesnt cluster with rest of tumor and is of predicted cardiomyocyte cell type 
  
  # set a threshold to id clusters
  freq.class.id <- freq.class.id[,c(2,3)]
  freq.class.id > 80
  
  
  # add these definitions to metadata
  tumor.clusters <- c("0", "1", "4", "5", "6", "8", "9")
  normal.clusters <- c("2", "3", "7", "10", "11", "12", "13")
  
  
  CD45neg.seurat@meta.data <- CD45neg.seurat@meta.data %>%
    dplyr::mutate(Disease = case_when(
      seurat_clusters %in% tumor.clusters ~ "tumor", 
      seurat_clusters %in% normal.clusters ~ "normal", 
      TRUE ~ "error")) 
  
  # sanity check
  unique(CD45neg.seurat@meta.data$Disease)
  table(CD45neg.seurat@meta.data$seurat_clusters, CD45neg.seurat@meta.data$Disease)
  
  
  # clean up subclone anno so it only exists for tumor clusters 
  CD45neg.seurat@meta.data <- CD45neg.seurat@meta.data %>%
    dplyr::mutate(subclone = case_when(
      seurat_clusters %in% normal.clusters ~ "normal", 
      seurat_clusters %in% tumor.clusters ~ as.character(subclone), 
      TRUE ~ "error")) %>%
    dplyr::mutate(subclone = coalesce(subclone, "undefined"))
  
  
  table(CD45neg.seurat@meta.data$Disease, CD45neg.seurat@meta.data$subclone)
  
  
  # create an annotation id that merges results from singleR findings etc
  table(CD45neg.seurat@meta.data$SingleR_fine, CD45neg.seurat@meta.data$SingleR_main)
  table(CD45neg.seurat@meta.data$seurat_clusters, CD45neg.seurat@meta.data$SingleR_main)
  
  # note below code only works if cardio, endo, and eryth are only 1 cluster and not spread across clusters 
  
  # uniform ID of cardio, erythro, and endothelial cells 
  # adipo and fibro cells are mixed with fine annotation in becoming activated fibro, fibro, NPCs? 
  # convert to just call these fibro_adipo cells if not tumor 
  CD45neg.seurat@meta.data <- CD45neg.seurat@meta.data %>%
    dplyr::mutate(cluster_anno = case_when(
      SingleR_main == "Cardiomyocytes" ~  "Cardiomyocytes",
      SingleR_main == "Endothelial cells" ~  "Endothelial_cells",
      SingleR_main == "Erythrocytes" ~  "Erythrocytes",
      seurat_clusters %in% tumor.clusters ~ paste0("Tumor_", seurat_clusters), 
      seurat_clusters %in% normal.clusters ~ SingleR_main,
      TRUE ~ "error")) 
  
  
  unique(CD45neg.seurat@meta.data$cluster_anno)
  
  # clean up seurat object to prevent mistakenly using wrong metadata annotation 
  colnames(CD45neg.seurat@meta.data)
  head(CD45neg.seurat@meta.data)
  CD45neg.seurat@meta.data <- CD45neg.seurat@meta.data %>%
    dplyr::select(c("orig.ident", "nCount_RNA", "nFeature_RNA", "nCount_SCT", "nFeature_SCT", 
                    "percent_mito", "percent_ribo", "percent_mito_ribo", "log10GenesPerUMI", 
                    "seurat_clusters", "SampleID", "Time", "Treatment", "Compartment", 
                    "SingleR_main", "SingleR_fine", "Disease", "subclone", "cluster_anno"))
  
  
  
  # rename tumor clusters to have sequential numbers as opposed to the seruat cluster numbers they currently have 
  # make a new column, because renaming might overwride as move along rows and names become similar with time 
  # therefore make new column and then remove the whole column and replace it 
  CD45neg.seurat@meta.data <- CD45neg.seurat@meta.data %>%
    dplyr::mutate(cluster_anno_new = case_when(
      cluster_anno == "Tumor_0" ~  "Tumor_0",
      cluster_anno == "Tumor_1" ~  "Tumor_1",
      cluster_anno == "Tumor_4" ~  "Tumor_2",
      cluster_anno == "Tumor_5" ~  "Tumor_3",
      cluster_anno == "Tumor_6" ~  "Tumor_4",
      cluster_anno == "Tumor_8" ~  "Tumor_5",
      cluster_anno == "Tumor_9" ~  "Tumor_6", 
      TRUE ~ cluster_anno))
  
  table(CD45neg.seurat@meta.data$cluster_anno, CD45neg.seurat@meta.data$cluster_anno_new)
  
  # replace old anno with new and clean up 
  CD45neg.seurat@meta.data$cluster_anno <- CD45neg.seurat@meta.data$cluster_anno_new
  CD45neg.seurat@meta.data$cluster_anno_new <- NULL
  
  
  
  # set factorisation 
  CD45neg.seurat@meta.data$Treatment <- factor(CD45neg.seurat@meta.data$Treatment, 
                                               levels = c("Vehicle", "PDS"))
  
  CD45neg.seurat@meta.data$SampleID <- factor(CD45neg.seurat@meta.data$SampleID, 
                                              levels = c("Day4_Vehicle_CD45Neg", "Day4_PDS_CD45Neg"))
  
  
  CD45neg.seurat@meta.data$seurat_clusters <- factor(CD45neg.seurat@meta.data$seurat_clusters)
  
  CD45neg.seurat@meta.data$Disease <- factor(CD45neg.seurat@meta.data$Disease, 
                                             levels = c("normal", "tumor"))
  
  CD45neg.seurat@meta.data$subclone <- factor(CD45neg.seurat@meta.data$subclone, 
                                              levels = c("normal", "1", "2", "3", "4", "5", "undefined"))
  
  CD45neg.seurat@meta.data$cluster_anno <- factor(CD45neg.seurat@meta.data$cluster_anno, 
                                                  levels = c("Cardiomyocytes", "Endothelial_cells", "Erythrocytes", "Fibroblasts", "Tumor_0", "Tumor_1", "Tumor_2", "Tumor_3", "Tumor_4", "Tumor_5", "Tumor_6"))
  
  
  
  # save output
  SaveSeuratRds(CD45neg.seurat, file = "saves/CD45neg_seurat.rds")
  
}



```




## Overview of dataset {.tabset}

### Broad overview of dataset
```{r broad_overview}

if(do.plot){
  
  
  # create output dir
  output.dir <- paste0("results/CD45Neg_Overview/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  
  # set assay
  DefaultAssay(CD45neg.seurat) <- "SCT"
  head(CD45neg.seurat)
  
  vars.to.plot <- c("Treatment", "SampleID", "seurat_clusters", "cluster_anno", "Disease", "subclone")
  
  
  for(j in seq_along(vars.to.plot)){
    name.var <- gsub("\\.", "_", vars.to.plot[j])
    
    print(
      scCustomize::DimPlot_scCustom(CD45neg.seurat,
                                    reduction = "umap", 
                                    label = TRUE,
                                    figure_plot = TRUE,
                                    group.by = vars.to.plot[j]) 
    )
    
    
    dev.copy(pdf, paste0(output.dir, "CD45Neg_UMAP_", name.var, ".pdf"))
    dev.off()
    
  }
  
  
  # split umap plots
  print(
    scCustomize::DimPlot_scCustom(CD45neg.seurat,
                                  reduction = "umap", 
                                  label = TRUE,
                                  figure_plot = TRUE,
                                  group.by = "cluster_anno", 
                                  split.by = "Treatment") 
  )
  
  
  dev.copy(pdf, paste0(output.dir, "CD45Neg_UMAP_cluster_anno_split_by_treatment.pdf"))
  dev.off()
  

    print(
    scCustomize::DimPlot_scCustom(CD45neg.seurat,
                                  reduction = "umap", 
                                  label = TRUE,
                                  figure_plot = TRUE,
                                  group.by = "Disease", 
                                  split.by = "Treatment") 
  )
  
  
  dev.copy(pdf, paste0(output.dir, "CD45Neg_UMAP_Disease_split_by_treatment.pdf"))
  dev.off()
  
  
  }


```

### Cluster stats
```{r cluster_stats}

# create output dir
output.dir <- paste0("results/CD45Neg_Overview/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

#############################################
# Number of cells per cluster
#############################################

cluster.n <- table(CD45neg.seurat@meta.data$cluster_anno)

cluster.n <- as.data.frame(cluster.n)
cluster.n$Var1 <- factor(cluster.n$Var1, levels = cluster.n$Var1)


ggplot(cluster.n) +
  aes(x = Var1, y = Freq, fill = Var1) +
  geom_col() +
  scale_fill_viridis_d(option = "viridis", 
                       direction = -1) +
  labs(x = "Cluster ID", y = "Number of cells", title = "Number of cells per cluster") +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

ggsave(paste0(output.dir, "Barplot_number_of_cells_per_cluster.pdf"))


#############################################
# Freq of clusters across treatment 
#############################################

dist.table <- table(CD45neg.seurat@meta.data$cluster_anno, CD45neg.seurat@meta.data$Treatment)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Vehicle", "PDS"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Treatment", y = "% of treatment", title = "Frequency of clusters per treatment") +
  scale_fill_manual(values = c(Vehicle = "grey", PDS = "darkred")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_clusters_per_treatment.pdf"))


#############################################
# Freq of disease across treatment 
#############################################

dist.table <- table(CD45neg.seurat@meta.data$Disease, CD45neg.seurat@meta.data$Treatment)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Vehicle", "PDS"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Treatment", y = "% of treatment", title = "Frequency of disease per treatment") +
  scale_fill_manual(values = c(Vehicle = "grey", PDS = "darkred")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_disease_per_treatment.pdf"))




#############################################
# Freq of tumor subclones across treatment
#############################################

dist.table <- table(CD45neg.seurat@meta.data$subclone, CD45neg.seurat@meta.data$Treatment)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Vehicle", "PDS"))



ggplot(freq.table) +
 aes(x = Var1, y = Freq, fill = Var2) +
 geom_col() +
 scale_fill_hue(direction = 1) +
 theme_minimal()

ggsave(paste0(output.dir, "Barplot_freq_of_subclone_per_treatment.pdf"))



```

### Cell Cycle scoring
```{r cell_cycle}

# create output dir
output.dir <- paste0("results/CD45Neg_cell_cycle/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


################################
# Get cell cycle genes 
################################
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

# pseudo-map to mouse gene (captures most except Mlf1ip, Fam64a and Hn1)
s.genes <- stringr::str_to_title(s.genes)
g2m.genes <- stringr::str_to_title(g2m.genes)

#####################
# score cell cycle
#####################

CD45neg.seurat <- CellCycleScoring(CD45neg.seurat,
                                   s.features = s.genes,
                                   g2m.features = g2m.genes,
                                   set.ident = TRUE)


#####################
# Visualise score
#####################

head(CD45neg.seurat@meta.data)

# Feature plots 

# S-score 
scCustomize::FeaturePlot_scCustom(CD45neg.seurat, 
                                  feature = "S.Score", 
                                  reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Neg_S_score.pdf"))
dev.off()

# G2M-score 
scCustomize::FeaturePlot_scCustom(CD45neg.seurat, 
                                  feature = "G2M.Score", 
                                  reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Neg_G2M_score.pdf"))
dev.off()

# Dim plot split by phase
scCustomize::DimPlot_scCustom(CD45neg.seurat, 
                              group.by = "cluster_anno",
                              split.by = "Phase",
                              label = TRUE,
                              reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_CD45Neg_cluster_by_phase.pdf"))
dev.off()

scCustomize::DimPlot_scCustom(CD45neg.seurat, 
                              group.by = "celltype",
                              split.by = "Phase",
                              label = TRUE,
                              reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_CD45Neg_celltype_by_phase.pdf"))
dev.off()

# Stacked VlnPlot
scCustomize::Stacked_VlnPlot(CD45neg.seurat,
                             features = c("S.Score", "G2M.Score"), 
                             group.by = "cluster_anno")

dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_cell_cycle.pdf"))
dev.off()

# Stacked VlnPlot
scCustomize::Stacked_VlnPlot(CD45neg.seurat,
                             features = c("S.Score", "G2M.Score"), 
                             group.by = "celltype")

dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_cell_cycle_by_celltype.pdf"))
dev.off()



###################
# Quantification 
###################


# Frequency of phase per cluster
dist.table <- table(CD45neg.seurat@meta.data$cluster_anno, CD45neg.seurat@meta.data$Phase)

cluster.total <- rowSums(dist.table)

freq.table <- sweep(dist.table, 1, cluster.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin
rowSums(freq.table)

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("G1", "G2M", "S"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Predicted phase", y = "% of cluster", title = "Frequency of phase per cluster") +
  scale_fill_manual(values = c(G1 = "lightblue", G2M = "orange", S = "red")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_phase_per_cluster.pdf"))




# Frequency of phase per celltype

dist.table <- table(CD45neg.seurat@meta.data$celltype, CD45neg.seurat@meta.data$Phase)

cluster.total <- rowSums(dist.table)

freq.table <- sweep(dist.table, 1, cluster.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin
rowSums(freq.table)

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("G1", "G2M", "S"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Predicted phase", y = "% of celltype", title = "Frequency of phase per celltype") +
  scale_fill_manual(values = c(G1 = "lightblue", G2M = "orange", S = "red")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_phase_per_celltype.pdf"))




# Frequency of phase per treatment

dist.table <- table(CD45neg.seurat@meta.data$Treatment, CD45neg.seurat@meta.data$Phase)

treatment.total <- rowSums(dist.table)

freq.table <- sweep(dist.table, 1, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin
rowSums(freq.table)

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("G1", "G2M", "S"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Predicted phase", y = "% of treatment", title = "Frequency of phase per treatment") +
  scale_fill_manual(values = c(G1 = "lightblue", G2M = "orange", S = "red")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_phase_per_treatment.pdf"))


```


## DEG analysis {.tabset}

### Cluster-wise DEGs - calculate
```{r Clusterwise_DEGs_calculation}


# create output dir
output.dir <- paste0("results/CD45Neg_cluster_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  
  CD45neg.seurat <- PrepSCTFindMarkers(CD45neg.seurat)
  
  # set assay
  DefaultAssay(CD45neg.seurat) <- "SCT"
  
  
  # set idents
  Idents(CD45neg.seurat) <- "cluster_anno"
  
  CD45Neg.de <- FindAllMarkers(CD45neg.seurat,
                               assay = "SCT",
                               only.pos = TRUE, 
                               verbose = TRUE)
  
  
  # save to file 
  write.csv(CD45Neg.de, paste0(output.dir, "CD45Neg_DEGs.csv"))
  
  
  
}else{
  
  # load DEG table
  CD45Neg.de <- read.csv(paste0(output.dir, "CD45Neg_DEGs.csv"))
  
}



```

### Cluster-wise DEGs - heatmaps
```{r clusterwise_DEGs_heatmaps}

# create output dir
output.dir <- paste0("results/CD45Neg_cluster_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

if(do.plot){
  # aggregate expression for heatmap visualisation
  
  pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("cluster_anno"))
  
  
  
  goi <- CD45Neg.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::group_by(cluster) %>%
    dplyr::top_n(10, avg_log2FC) %>%
    dplyr::pull(gene)
  
  
  DoHeatmap(pseudo.cd45neg, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_top_10_DEGs_pseudobulk.pdf"))
  dev.off()
  
  
  
  
  # iterate over the top degs for each cluster 
  cluster.id <- unique(CD45Neg.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- CD45Neg.de %>%
      dplyr::filter(p_val_adj < 0.05) %>%
      dplyr::filter(cluster == cluster.id[i]) %>%
      dplyr::top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.cd45neg, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE) + 
        scico::scale_fill_scico(palette = "batlow", 
                                direction = 1, 
                                na.value = "white")
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
}

```

## Melanocyte gene visualisation {.tabset}

### Melanocyte-specific gene expression
```{r melanocyte_genes}

# create output dir
output.dir <- paste0("results/CD45Neg_Melanocyte_genes/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# list melanocyte genes 
goi <- c("Mitf", "Tyr", "Pmel", "Mlana", "Tyrp1", "Sox10")


# Plot Feature and Density plots iteratively over goi 
for(i in seq_along(goi)){
  
  # feature plots
  print(
    scCustomize::FeaturePlot_scCustom(CD45neg.seurat, 
                                      reduction = "umap", 
                                      features = goi[i])
    
  )
  
  dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Neg_", goi[i], ".pdf"))
  dev.off()
  
  
  # density plots
  
  print(
    scCustomize::Plot_Density_Custom(CD45neg.seurat, 
                                     reduction = "umap",
                                     features = goi[i], 
                                     custom_palette = batlow.pal)
  )
  dev.copy(pdf, paste0(output.dir, "UMAP_DensityPlot_CD45Neg_", goi[i], ".pdf"))
  dev.off()
  
  
  
}


##############################
# Stacked vln plot
##############################

scCustomize::Stacked_VlnPlot(CD45neg.seurat, 
                             features = goi,
                             group.by = "cluster_anno",
                             pt.size = 0)

dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_melanocyte_genes.pdf"))
dev.off()


###########
# heatmap
###########

# aggregate expression for heatmap visualisation
pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("cluster_anno"))

DoHeatmap(pseudo.cd45neg, 
          features = goi, 
          draw.lines = FALSE, 
          angle = 90,
          raster = FALSE) + 
  scico::scale_fill_scico(palette = "batlow", 
                          direction = 1, 
                          na.value = "white")

dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_melanocyte_genes.pdf"))
dev.off()


```


## RQ1: Which CD45Neg clusters express chemokines or cytokines {.tabset}

### Chemokine and Cytokine expression visualisation 
```{r chemokine_cytokine_expression}

# create output dir
output.dir <- paste0("results/CD45Neg_chemokine_cytokines/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# aggregate expression for heatmap visualisation
pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("cluster_anno"))


########################################
# List gene sets of interest to plot
########################################

cxc.genes <- grep("^Cxc.*", rownames(pseudo.cd45neg), value = T)
ccr.genes <- grep("^Ccr.*", rownames(pseudo.cd45neg), value = T)
il.genes <- grep("^Il.*", rownames(pseudo.cd45neg), value = T)
gzm.genes <- grep("^Gzm.*", rownames(pseudo.cd45neg), value = T)
ifn.genes <- grep("^Ifn.*", rownames(pseudo.cd45neg), value = T)
custom.genes <- c("Il1b", "Il6", "Il12a", "Ifng", "Tnf", "Ccl2", "Ccl3", "Ccl4", "Ccl5", "Cxcl1", "Cxcl9", "Cxcl10", "Cxcl13") # Il12b not present in CD45Neg dataset

goi <- list(Cxcr_genes = cxc.genes, 
            Ccr_genes = ccr.genes,
            Interleukin_genes = il.genes,
            Gzm_genes = gzm.genes,
            Ifn_genes = ifn.genes,
            Custom_genes = custom.genes)


###################
# Plot heatmaps
###################

for(i in seq_along(names(goi))){
  
  print(
    DoHeatmap(pseudo.cd45neg, 
              features = goi[[i]], 
              draw.lines = FALSE, 
              angle = 90,
              raster = FALSE) + 
      scico::scale_fill_scico(palette = "batlow", 
                              direction = 1, 
                              na.value = "white")
  )
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_", names(goi)[i], ".pdf"))
  dev.off()
  
}


##############################
# Stacked vln plot
##############################

for(i in seq_along(names(goi))){
  
  if(length(goi[[i]]) <= 15){
    
    
    print(
      scCustomize::Stacked_VlnPlot(CD45neg.seurat, 
                                   features = goi[[i]],
                                   group.by = "cluster_anno",
                                   pt.size = 0)
    )
    
    dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_", names(goi)[i], ".pdf"))
    dev.off()
    
  }else{
    
    temp.goi <- goi[[i]]
    
    # Split the vector into multiple vectors
    split_indices <- seq(1, length(temp.goi), by = 15)
    split_list <- lapply(split_indices, function(i) temp.goi[i:min(i + 14, length(temp.goi))])
    
    
    for(j in seq_along(length(split_list))){
      
      print(
        scCustomize::Stacked_VlnPlot(CD45neg.seurat, 
                                     features = split_list[[j]],
                                     group.by = "cluster_anno",
                                     pt.size = 0)
      )
      
      dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_", names(goi)[i], "_", j, ".pdf"))
      dev.off()
      
    }
    
  }
  
}


```

## RQ2: DEGs within clusters across treatment {.tabset}

### Treatment-wise DEGs - calculate
```{r Treatmentwise_DEGs_calculation}

# create output dir
output.dir <- paste0("results/CD45Neg_treatment_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  
  CD45neg.seurat <- PrepSCTFindMarkers(CD45neg.seurat)
  
  # set assay
  DefaultAssay(CD45neg.seurat) <- "SCT"
  
  
  
  # add annotation for treatment 
  CD45neg.seurat@meta.data$Treatment_cluster <- paste0(CD45neg.seurat@meta.data$Treatment, "_", CD45neg.seurat@meta.data$cluster_anno)
  
  # set Idents
  Idents(CD45neg.seurat) <- "Treatment_cluster"
  
  
  cluster.ids <- unique(CD45neg.seurat@meta.data$cluster_anno)
  
  # create list to store output
  DEG.list <- list()
  
  # run through DEG specifically between vehicle and treatment for each cluster 
  
  for(i in seq_along(cluster.ids)){
    temp.de <- FindMarkers(CD45neg.seurat,
                           assay = "SCT",
                           ident.1 = paste0("PDS_", cluster.ids[i]),
                           ident.2 = paste0("Vehicle_", cluster.ids[i]),
                           only.pos = FALSE, 
                           recorrect_umi = FALSE,
                           verbose = TRUE)
    
    DEG.list[[i]] <- temp.de
    
    # save to file 
    write.csv(temp.de, paste0(output.dir, "DEGs_PDS_", cluster.ids[i], "_vs_Vehicle_", cluster.ids[i], ".csv"))
    
    
  }
  # add names to DEG.list
  names(DEG.list) <- cluster.ids
  
  # save DEG list to file 
  saveRDS(DEG.list, paste0(output.dir, "DEG_list_object.rds"))
  
  
}else{
  
  DEG.list <- readRDS(paste0(output.dir, "DEG_list_object.rds"))
  
}


```

### Treatment-wise DEGs - Volcano
```{r treatmentwise_DEGs_volcano}

# visualise treatment-wise DEGs as volcano plots

# create output dir
output.dir <- paste0("results/CD45Neg_treatment_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(do.plot){
  
  for(i in 1:length(DEG.list)){
    
    FDR.theshold <- 0.01
    logFC.threshold <- 1
    
    # filter data
    input.data <- DEG.list[[i]] %>%
      dplyr::rename(logFC = avg_log2FC) %>%
      dplyr::rename(FDR = p_val_adj) %>%
      filter(FDR < FDR.theshold)
    
    # color points
    volcano.col <- usefulfunctions::colour.points.volcano(input.data,
                                                          increase.col = "Red",
                                                          decreased.col = "Blue",
                                                          FDR.cutoff = FDR.theshold, 
                                                          logFC.cutoff = logFC.threshold)
    # plot heatmap
    print(
      usefulfunctions::custom.enhanced.volcano(input.data, 
                                               pCutoff = FDR.theshold,
                                               FCcutoff = logFC.threshold,
                                               title = paste0(names(DEG.list)[i], " PDS vs Vehicle"),
                                               subtitle = paste0("adjPval threshold = ", FDR.theshold, " & log2FC threshold = ", logFC.threshold),
                                               colCustom = volcano.col)
    )
    dev.copy(pdf, paste0(output.dir, "VolcanoPlot_PDS_vs_Vehicle_cluster_", names(DEG.list)[i], ".pdf"))
    dev.off()
    
    
  }
  
}

```

### Export aggregated counts table
```{r export_aggregated_counts}

# create output dir
output.dir <- paste0("results/tables/CD45Neg/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  # Aggregate expression by cluster 
  pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("cluster_anno"))
  
  # Write to file
  temp.out <- GetAssayData(pseudo.cd45neg,
                           layer = "data",
                           assay = "SCT")
  
  write.csv(temp.out, paste0(output.dir, "CD45Neg_Aggregate_SCT_data_by_cluster.csv"))
  
  # Aggregate expression by cluster and treatment
  pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("cluster_anno", "Treatment"))
  # write to file
  temp.out <- GetAssayData(pseudo.cd45neg,
                           layer = "data",
                           assay = "SCT")
  
  write.csv(temp.out, paste0(output.dir, "CD45Neg_Aggregate_SCT_data_by_cluster_and_treatment.csv"))
  
  # clean environment
  rm(pseudo.cd45neg)
  rm(temp.out)
  
}



```


## RQ3: DEGs tumour clusters across treatment {.tabset}


```{r RQ3_DEG_tumor_clusters_across_treatment}


# create output dir
output.dir <- paste0("results/CD45Neg_tumor_clusters_DEGs_across_treatment/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

table(CD45neg.seurat@meta.data$cluster_anno, CD45neg.seurat@meta.data$Disease)

# create new metadata disease and treatment 
CD45neg.seurat@meta.data$disease_treat <- paste0(CD45neg.seurat@meta.data$Disease, "_", CD45neg.seurat@meta.data$Treatment)

Idents(CD45neg.seurat) <- CD45neg.seurat@meta.data$disease_treat



if(long.compute){
  
  
  CD45neg.seurat <- PrepSCTFindMarkers(CD45neg.seurat)
  
  # set assay
  DefaultAssay(CD45neg.seurat) <- "SCT"
  

    
    tumor.de <- FindMarkers(CD45neg.seurat,
                           assay = "SCT",
                           ident.1 = "tumor_PDS",
                           ident.2 = "tumor_Vehicle",
                           only.pos = FALSE, 
                           recorrect_umi = FALSE,
                           verbose = TRUE)
    
    
    
    # save to file 
    write.csv(tumor.de, paste0(output.dir, "DEGs_Tumor_PDS_vs_Tumor_Vehicle.csv"))
    
    
}



# plot DEGs 
if(do.plot){
  
    FDR.theshold <- 0.01
    logFC.threshold <- 1
    
    # filter data
    input.data <- tumor.de %>%
      dplyr::rename(logFC = avg_log2FC) %>%
      dplyr::rename(FDR = p_val_adj) %>%
      filter(FDR < FDR.theshold)
    
    # color points
    volcano.col <- usefulfunctions::colour.points.volcano(input.data,
                                                          increase.col = "Red",
                                                          decreased.col = "Blue",
                                                          FDR.cutoff = FDR.theshold, 
                                                          logFC.cutoff = logFC.threshold)
    # plot heatmap
    print(
      usefulfunctions::custom.enhanced.volcano(input.data, 
                                               pCutoff = FDR.theshold,
                                               FCcutoff = logFC.threshold,
                                               title = paste0("Tumor PDS vs Tumor vehicle"),
                                               subtitle = paste0("adjPval threshold = ", FDR.theshold, " & log2FC threshold = ", logFC.threshold),
                                               colCustom = volcano.col)
    )
    dev.copy(pdf, paste0(output.dir, "VolcanoPlot_Tumor_PDS_vs_Tumor_Vehicle.pdf"))
    dev.off()
    
    
  
    
  #####################
  # heatmaps of top DEGs
  #####################
  
  
  
  Idents(CD45neg.seurat) <- CD45neg.seurat@meta.data$disease_treat
  
  

  pseudo.tumor <- AggregateExpression(CD45neg.seurat,
                                           assays = "SCT",
                                           return.seurat = TRUE, 
                                           group.by = c("disease_treat"))
  
  
  # Extract top fold change DEGs 
  
  goi <- tumor.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::filter(abs(avg_log2FC) > 2) %>%
    dplyr::arrange(desc(avg_log2FC)) %>%
    rownames()
  
  
  # plot heatmap
  DoHeatmap(pseudo.tumor, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "CD45Neg_Heatmap_Tumor_PDS_vs_Tumor_Vehicle_logfold_2.pdf"))
  dev.off()
  
  
  # extract just top 10 up and down degs
  goi.up <- tumor.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::top_n(10, avg_log2FC) %>%
    rownames()
  
  goi.dn <- tumor.de %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    dplyr::top_n(-10, avg_log2FC) %>%
    rownames()
  
  goi <- unique(c(goi.up, goi.dn))
  
  DoHeatmap(pseudo.tumor, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "CD45Neg_Heatmap_Tumor_PDS_vs_Tumor_Vehicle_top_10_DEGs_up_down.pdf"))
  dev.off()
  
}
  


# clean environment 
  rm(pseudo.tumor)
  gc()


```



