---
title: "CD45Neg_dataset_annotation"
author: "Dillon Corvino"
date: "25/01/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Built with R version `r getRversion()`

Annotation stage for the CD45- compartment: immune-contamination removal,
re-integration, SingleR cluster annotation, SCEVAN copy-number tumor/normal
calling, and export of the final annotated object to
`saves/CD45neg_seurat.rds`. This is preprocessing/methods provenance --
downstream scripts (`manuscript_figures.R`,
`scripts/archive/03_CD45Neg_exploratory.Rmd`) load that saved object rather
than re-deriving it. Split out of the original combined
`03_CD45Neg_data_analysis.Rmd` on 2026-08-14; the exploratory analyses that
used to follow this content now live in
`scripts/archive/03_CD45Neg_exploratory.Rmd`.

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# Data 
# B16 - OVA 

# PDS = I.T - bind Gquadraplex
# day 4 post-treatment 


# Research questions 
# How does PDS induces an anti-tumor immune response?
# Known that after PDS treatment, certain DAMPs are induced, intratumoral chemokine levels go up drastically and subsequntly immune cells are recruited into the tumor.
# Main research questions are:
#1. Which cell subsets are producing the chemokines
#2. Are there differences between macrophages and DCs between PDS and Vehicle (because there are DAMPs that should activate them)
#3. How does the treatment effect the tumor cells (is it possible to discriminate them from other tissue cells, if not they express ovalbumin...this should make it at least possible to identify them.)




# key genes
# IL1b, IL6, IL12, IFNG, TNFA, 

# CCL2,3,4,5 
# CXCL1, CXCL9,10,13 





```


### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)


# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- FALSE
do.plot <- TRUE
quick.load <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())

# function found in Setup.R script
ggplot2_switch_version(desired_version = "3.4.4")

library("Seurat")
library("SeuratObject")
#library("SeuratData")
library("SeuratWrappers")
library("dplyr")
library("scCustomize")
library("tibble")
library("ggplot2")


# load output from preprocessing script
if(long.compute){
  
  CD45neg.seurat <- LoadSeuratRds("saves/Preprocessed_CD45Neg.rds")
  
  # set levels for easier downstream plotting 
  CD45neg.seurat@meta.data$Treatment <- factor(CD45neg.seurat@meta.data$Treatment, levels = c("Vehicle", "PDS"))
  
}else if(quick.load){
  
  # load cleaned dataset 
  CD45neg.seurat <- LoadSeuratRds("saves/CD45neg_seurat.rds")
  DefaultAssay(CD45neg.seurat) <- "SCT"
  
  # Create new metadata column that groups all tumor together 
  CD45neg.seurat@meta.data <- CD45neg.seurat@meta.data %>%
    dplyr::mutate(celltype = case_when(Disease == "tumor" ~ "tumor",
                                       Disease == "normal" ~ cluster_anno))
  
  
}
```


## Clean dataset {.tabset}

## Identifying Immune cell contamination - SingleR cluster annotation
```{r SingleR_cluster_annotation_initial}

# Notes:
# there is an error with collect() when trying to load celldex datasets
# i think its an error with biocmanager and has been fixed with a very recent update. > 3.16 
# but im running r 4.2.3 and this would require 4.3
# not keen to upgrade right now 
# so will try a work around of downgrading dbplyr to version 2.3.4 from 2.4 which im currently running 
# need to downgrade to this version 
#devtools::install_version("dbplyr", version = "2.3.4")
#devtools::install_version("dbplyr", version = "2.4")
# in theory if biocmanager is up to date, could just run below and should fix the problem
#library(BiocManager) # 3.16
#install("BiocFileCache")
# also:
# need to ensure matrixStats is at version 1.1.0 
#remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts
#BiocManager::install(version = "3.18")

if(long.compute){
  
  output.dir <- "results/CD45Neg_Immune_cell_contamination/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # load required packages
  library("SingleR")
  library("celldex")
  
  
  ##################################
  # Use MouseRNAseqData
  ##################################
  
  ref.mouse <-  celldex::MouseRNAseqData()
  
  
  # Benayoun Lab that identified, downloaded and processed data sets on GEO that corresponded to sorted cell types (Benayoun et al., 2019).
  
  # The dataset contains 358 mouse RNA-seq samples annotated to 18 main cell types
  # further split into 28 subtypes under label.fine 
  
  
  ######################
  # Fine labels 
  ######################
  SingleR.pred <- SingleR(test = CD45neg.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45neg.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.fine)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
  #SingleR::plotDeltaDistribution(SingleR.pred)
  #SingleR::plotScoreDistribution(SingleR.pred)
  
  ######################
  # Main labels 
  ######################
  
  SingleR.pred <- SingleR(test = CD45neg.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45neg.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.main)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_main_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
}

```


### Confirming Immune cell contamination - CD45 and granulocyte linage marker gene visualisation
```{r CD45_and_granulocyte_lineage_gene_visualisation}

if(long.compute){
  
  # create output dir
  output.dir <- paste0("results/CD45Neg_Immune_cell_contamination/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  # list melanocyte genes 
  goi <- c("Ptprc", "Cd14")
  
  # Plot Feature and Density plots iteratively over goi 
  for(i in seq_along(goi)){
    
    # feature plots
    print(
      scCustomize::FeaturePlot_scCustom(CD45neg.seurat, 
                                        reduction = "umap", 
                                        features = goi[i])
      
    )
    
    dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Neg_", goi[i], ".pdf"))
    dev.off()
    
    
    # density plots
    
    print(
      scCustomize::Plot_Density_Custom(CD45neg.seurat, 
                                       reduction = "umap",
                                       features = goi[i], 
                                       custom_palette = batlow.pal)
    )
    dev.copy(pdf, paste0(output.dir, "UMAP_DensityPlot_CD45Neg_", goi[i], ".pdf"))
    dev.off()
    
    
    
  }
  
  
  ##############################
  # Stacked vln plot
  ##############################
  
  scCustomize::Stacked_VlnPlot(CD45neg.seurat, 
                               features = goi,
                               group.by = "seurat_clusters",
                               pt.size = 0)
  
  dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_granulocyte_markers.pdf"))
  dev.off()
  
  
  ###########
  # heatmap
  ###########
  
  # aggregate expression for heatmap visualisation
  pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("seurat_clusters"))
  
  DoHeatmap(pseudo.cd45neg, 
            features = goi, 
            draw.lines = FALSE, 
            angle = 90,
            raster = FALSE) + 
    scico::scale_fill_scico(palette = "batlow", 
                            direction = 1, 
                            na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_granulocyte_markers.pdf"))
  dev.off()
  
}


```


### Remove contamination immune clusters
```{r remove_immune_contamination}

if(long.compute){
  
  # cluster 6 looks to be immune contaimination 
  
  # set immune cell cluster variable
  Immune.cell.cluster <- 6
  
  
  head(CD45neg.seurat@meta.data)
  
  
  # set assay
  DefaultAssay(CD45neg.seurat) <- "SCT"
  
  # set idents
  Idents(CD45neg.seurat) <- "seurat_clusters"
  
  # immune cell contamination cluster is relatively small, sharp drop of in cluster sizes at cluster 5
  table(CD45neg.seurat@meta.data$seurat_clusters) # cluster 6 = 735 cells
  
  # remove this cluster from dataset 
  Idents(CD45neg.seurat) <- CD45neg.seurat@meta.data$seurat_clusters
  CD45neg.seurat <- subset(CD45neg.seurat, idents = Immune.cell.cluster, invert = TRUE)
  unique(Idents(CD45neg.seurat))
  
  
  # drop levels in seurat cluster meta
  CD45neg.seurat$seurat_clusters <- droplevels(CD45neg.seurat$seurat_clusters)
  
}

```


## Renormalise / integrate / dim reduction | after removing contamination immune cells {.tabset}

## re-calculate normalisation and integration for CD45Neg dataset
```{r re_normalise_and_integrate_CD45Neg_dataset}

if(long.compute){
  
  
  # split object up based on sequencing run / in seurat V5 these are splitting the layers
  CD45neg.seurat[["RNA"]] <- split(CD45neg.seurat[["RNA"]],  f = CD45neg.seurat$SampleID)
  
  DefaultAssay(CD45neg.seurat) <- "RNA"
  
  
  
  
  #####################################################
  # Normalise with scTransform and integrate data
  #####################################################
  # note error with more recent version of matrixStats resulting in "useNames = NA is defunct" error. downgrade to matrixStats version 1.1.0 as a workaround
  # https://github.com/satijalab/seurat/issues/7501#issuecomment-1854571904
  #remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts  
  
  
  
  # note RNA assay is split into run level layers, below script will apply normalisation etc layer-wise
  
  CD45neg.seurat <- CD45neg.seurat %>%
    SCTransform(method = "glmGamPoi", 
                vars.to.regress = "percent_mito", 
                vst.flavor = "v2",
                verbose = TRUE) %>%
    RunPCA(reduction.name = "pca_sct") %>%
    RunUMAP(reduction = "pca_sct", 
            dims = 1:30, 
            reduction.name = "umap.unintegrated.sct")
  
  
  DefaultAssay(CD45neg.seurat) <- "SCT"
  
  
  # integrate layers using cca integration and calculate dim redu and cluster identification
  
  # cluster resolution value based on previous estimation 
  clust.res <- 0.4
  
  
  CD45neg.seurat <- CD45neg.seurat %>%
    IntegrateLayers(orig.reduction = "pca_sct",
                    new.reduction = "integrated.cca.sct",
                    method = CCAIntegration,
                    normalization.method = "SCT",
                    verbose = F) %>%
    FindNeighbors(dims = 1:30,
                  reduction = "integrated.cca.sct") %>%
    FindClusters(resolution = clust.res, 
                 cluster.name = "clusters_cca_sct") %>%
    RunUMAP(reduction = "integrated.cca.sct",
            dims = 1:30, 
            reduction.name = "umap")
  
  
}

```


## plot integration after removing contamination immune cells from CD45Neg dataset 

```{r plot_integration_after_removing_contamination}

if(long.compute){
  
  if(do.plot){
    
    # create output dir
    output.dir <- paste0("results/CD45Neg_Integration_validation_after_cleaning/")
    
    if(!dir.exists(output.dir)){
      dir.create(output.dir, 
                 recursive = T)
      
    }
    
    # set variables
    unintegrated.reduction <- "umap.unintegrated.sct"
    integrated.reduction <- "umap"
    title.var <- "SCT"
    
    
    
    # evaluate integration and dim reduction
    
    # set default assay
    DefaultAssay(CD45neg.seurat) <- "SCT"
    
    
    
    vars.plot <- c("Treatment", "SampleID", "seurat_clusters")
    
    
    for(i in seq_along(vars.plot)){
      
      
      # remove "." from name
      name.var <- gsub("\\.", "_", vars.plot[i])
      
      before.int <- DimPlot(CD45neg.seurat,
                            reduction = unintegrated.reduction, 
                            label = T,
                            group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("Unintegrated ", title.var, " Normalisation"))
      
      after.int <- DimPlot(CD45neg.seurat,
                           reduction = integrated.reduction, 
                           label = T,
                           group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("CCA Integration using ", title.var, " Normalisation"))
      
      print(before.int | after.int)
      dev.copy(pdf, paste0(output.dir, "UMAP_before_after_integration_", name.var, "_CD45Neg_dataset.pdf"))
      dev.off()
      
      
      
    }
  }
  
  
}


```


# saved a cleaned object here

## Cluster annotation {.tabset}

## SingleR cluster annotation
```{r SingleR_cluster_annotation}

# Notes:
# there is an error with collect() when trying to load celldex datasets
# i think its an error with biocmanager and has been fixed with a very recent update. > 3.16 
# but im running r 4.2.3 and this would require 4.3
# not keen to upgrade right now 
# so will try a work around of downgrading dbplyr to version 2.3.4 from 2.4 which im currently running 
# need to downgrade to this version 
#devtools::install_version("dbplyr", version = "2.3.4")
#devtools::install_version("dbplyr", version = "2.4")
# in theory if biocmanager is up to date, could just run below and should fix the problem
#library(BiocManager) # 3.16
#install("BiocFileCache")
# also:
# need to ensure matrixStats is at version 1.1.0 
#remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts
#BiocManager::install(version = "3.18")

if(long.compute){
  
  output.dir <- "results/CD45Neg_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # load required packages
  library("SingleR")
  library("celldex")
  
  
  ##################################
  # Use MouseRNAseqData
  ##################################
  
  ref.mouse <-  celldex::MouseRNAseqData()
  
  
  # Benayoun Lab that identified, downloaded and processed data sets on GEO that corresponded to sorted cell types (Benayoun et al., 2019).
  
  # The dataset contains 358 mouse RNA-seq samples annotated to 18 main cell types
  # further split into 28 subtypes under label.fine 
  
  
  ######################
  # Fine labels 
  ######################
  SingleR.pred.fine <- SingleR(test = CD45neg.seurat@assays$SCT$data, 
                               ref = ref.mouse, 
                               method = "cluster",
                               clusters = CD45neg.seurat@meta.data$seurat_clusters,
                               labels = ref.mouse$label.fine)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
  #SingleR::plotDeltaDistribution(SingleR.pred)
  #SingleR::plotScoreDistribution(SingleR.pred)
  
  ######################
  # Main labels 
  ######################
  
  SingleR.pred.main <- SingleR(test = CD45neg.seurat@assays$SCT$data, 
                               ref = ref.mouse, 
                               method = "cluster",
                               clusters = CD45neg.seurat@meta.data$seurat_clusters,
                               labels = ref.mouse$label.main)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_main_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
}

```


### Transfer SingleR annotations to seurat object 
```{r transfer_singleR_annotations}


if(long.compute){
  
  output.dir <- "results/CD45Neg_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  
  ########################################
  # transfer labels to seurat object 
  ########################################
  
  table(predicted=SingleR.pred.main$labels, cluster=rownames(SingleR.pred.main))
  table(predicted=SingleR.pred.fine$labels, cluster=rownames(SingleR.pred.fine))
  
  # Main labels
  main.labels <- SingleR.pred.main$labels
  names(main.labels) <- rownames(SingleR.pred.main)
  
  # Fine labels
  fine.labels <- SingleR.pred.fine$labels
  names(fine.labels) <- rownames(SingleR.pred.fine)
  
  ####################
  # rename clusters 
  ####################
  
  # Main labels
  Idents(CD45neg.seurat) <- CD45neg.seurat@meta.data$seurat_clusters
  CD45neg.seurat <- Rename_Clusters(CD45neg.seurat, new_idents = main.labels)
  CD45neg.seurat@meta.data$SingleR_main <- Idents(CD45neg.seurat)
  
  # Fine labels
  Idents(CD45neg.seurat) <- CD45neg.seurat@meta.data$seurat_clusters
  CD45neg.seurat <- Rename_Clusters(CD45neg.seurat, new_idents = fine.labels)
  CD45neg.seurat@meta.data$SingleR_fine <- Idents(CD45neg.seurat)
  
  
  ############
  # Visualise
  ############
  
  # Main label
  p.main <- scCustomize::DimPlot_scCustom(CD45neg.seurat, 
                                          group.by = "SingleR_main", 
                                          label = TRUE,
                                          reduction = "umap")
  
  print(p.main)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_main.pdf"))
  dev.off()
  
  
  # Fine label
  p.fine <- scCustomize::DimPlot_scCustom(CD45neg.seurat, 
                                          group.by = "SingleR_fine", 
                                          label = TRUE,
                                          reduction = "umap")
  
  print(p.fine)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_fine.pdf"))
  dev.off()
  
  
  # side-by-side with cluster numbers
  p.clusters <- scCustomize::DimPlot_scCustom(CD45neg.seurat, 
                                              group.by = "seurat_clusters", 
                                              label = TRUE,
                                              reduction = "umap")
  
  # clusters plus main
  print(p.clusters | p.main)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_main_plus_clusters.pdf"))
  dev.off()
  
  # clusters plus fine
  print(p.clusters | p.fine)
  
  dev.copy(pdf, paste0(output.dir, "UMAP_singleR_annotations_fine_plus_clusters.pdf"))
  dev.off()
  
  
}


```

### SCEVAN Copy number analysis
```{r copy_number_analysis}

if(long.compute){
  
  
  output.dir <- "results/CD45Neg_SCEVAN/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  
  #library(devtools)
  #install_github("miccec/yaGST")
  #install_github("AntonioDeFalco/SCEVAN")
  library(SCEVAN)
  
  DefaultAssay(CD45neg.seurat) <- "RNA"
  CD45neg.seurat <- JoinLayers(CD45neg.seurat)
  
  
  count.matrix <- LayerData(CD45neg.seurat, assay = "RNA", layer = "counts")
  
  results.out.single.object <- pipelineCNA(count_mtx = count.matrix, 
                                           sample = "CD45Neg_dataset", 
                                           par_cores = 20, 
                                           organism = "mouse", 
                                           SUBCLONES = TRUE, 
                                           beta_vega = 0.5, 
                                           ClonalCN = TRUE, 
                                           plotTree = TRUE)
  
  CD45neg.seurat <-Seurat::AddMetaData(CD45neg.seurat, metadata = results.out.single.object)
  CD45neg.seurat@meta.data
  
  
  # export object
  saveRDS(results.out.single.object, file = "saves/results_out_single_object.rds")

  # Visualise results of copy number variance analysis
  
  # Tumor or normal identity by Copy Number
  
  DimPlot_scCustom(CD45neg.seurat, 
                   group.by = "class", 
                   reduction = "umap")
  dev.copy(pdf, paste0(output.dir, "UMAP_class_CNA_identity.pdf"))
  dev.off()
  
  # Tumor Subclone ID
  # switch to dimplot, cause NA and 1 end up similar gray if using scCustomise package dimplot function
  
  DimPlot(CD45neg.seurat, 
          group.by = "subclone",
          reduction = "umap")
  
  dev.copy(pdf, paste0(output.dir, "UMAP_subclone_identity.pdf"))
  dev.off()
  
  
  # Single R main identity across class
  DimPlot_scCustom(CD45neg.seurat,
                   group.by = "SingleR_main", 
                   split.by = "class", 
                   reduction = "umap")
  
  dev.copy(pdf, paste0(output.dir, "UMAP_SingleR_main_across_class.pdf"))
  dev.off()
  
  # Single R fine identity across class
  DimPlot_scCustom(CD45neg.seurat,
                   group.by = "SingleR_fine",
                   split.by = "class", 
                   reduction = "umap")
  
  dev.copy(pdf, paste0(output.dir, "UMAP_SingleR_fine_across_class.pdf"))
  dev.off()
  
  # Seurat clusters across class 
  DimPlot_scCustom(CD45neg.seurat,
                   group.by = "seurat_clusters", 
                   split.by = "class", 
                   label = TRUE,
                   reduction = "umap")
  
  dev.copy(pdf, paste0(output.dir, "UMAP_Seurat_clusters_across_class.pdf"))
  dev.off()
  
}


```

### Annotate clusters and clean up seurat object and export
```{r annotate_clusters}



if(long.compute){
  
  output.dir <- "results/CD45Neg_unifying_cluster_annotations/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  DimPlot_scCustom(CD45neg.seurat,
                   group.by = "seurat_clusters", 
                   split.by = "class", 
                   label = TRUE,
                   reduction = "umap")
  
  
  
  # Unify CNA definition of tumor with singleR definitions 
  out.df <- table(CD45neg.seurat@meta.data$seurat_clusters, CD45neg.seurat@meta.data$class)
  row.vals <- rowSums(out.df)
  freq.class.id <- sweep(out.df, 1, row.vals, "/")
  rowSums(freq.class.id*100) # sanity check
  freq.class.id <- freq.class.id*100
  
  
  ggplot(freq.class.id) +
    aes(x = Var1, y = Freq, fill = Var2) +
    geom_col() +
    scale_fill_hue(direction = 1) +
    labs(x = "Seurat_cluster", y = "Frequency", title = "Frequency of normal/tumor in each seurat_cluster") +
    theme_minimal()
  
  ggsave(paste0(output.dir, "Barplot_frequency_of_seurat_cluster_by_class_identity.pdf"), width = 8, height = 4)
  
  # Tumor clusters are: 
  # 0, 1, 4, 5, 6, 8, 9
  
  # Normal clusters are: 
  # 2, 3, 7, 13
  
  # unclear: 
  # 11, 10, 12
  
  # cluster 10 
  # will be called normal 
  # shows up roughly evenly between normal and tumor but doesnt cluster with rest of tumor cells
  # also not predicted to be fibroblast type origin 
  # looks to be erythrocyte origin 
  
  # cluster 11
  # will be called normal (79%)
  # does cluster with tumor cells and is precited to be of fibroblast cell type 
  # but perhaps this is a normal non tumor fibroblast cells
  
  # cluster 12
  # will be called normal (78%)
  # doesnt cluster with rest of tumor and is of predicted cardiomyocyte cell type 
  
  # set a threshold to id clusters
  freq.class.id <- freq.class.id[,c(2,3)]
  freq.class.id > 80
  
  
  # add these definitions to metadata
  tumor.clusters <- c("0", "1", "4", "5", "6", "8", "9")
  normal.clusters <- c("2", "3", "7", "10", "11", "12", "13")
  
  
  CD45neg.seurat@meta.data <- CD45neg.seurat@meta.data %>%
    dplyr::mutate(Disease = case_when(
      seurat_clusters %in% tumor.clusters ~ "tumor", 
      seurat_clusters %in% normal.clusters ~ "normal", 
      TRUE ~ "error")) 
  
  # sanity check
  unique(CD45neg.seurat@meta.data$Disease)
  table(CD45neg.seurat@meta.data$seurat_clusters, CD45neg.seurat@meta.data$Disease)
  
  
  # clean up subclone anno so it only exists for tumor clusters 
  CD45neg.seurat@meta.data <- CD45neg.seurat@meta.data %>%
    dplyr::mutate(subclone = case_when(
      seurat_clusters %in% normal.clusters ~ "normal", 
      seurat_clusters %in% tumor.clusters ~ as.character(subclone), 
      TRUE ~ "error")) %>%
    dplyr::mutate(subclone = coalesce(subclone, "undefined"))
  
  
  table(CD45neg.seurat@meta.data$Disease, CD45neg.seurat@meta.data$subclone)
  
  
  # create an annotation id that merges results from singleR findings etc
  table(CD45neg.seurat@meta.data$SingleR_fine, CD45neg.seurat@meta.data$SingleR_main)
  table(CD45neg.seurat@meta.data$seurat_clusters, CD45neg.seurat@meta.data$SingleR_main)
  
  # note below code only works if cardio, endo, and eryth are only 1 cluster and not spread across clusters 
  
  # uniform ID of cardio, erythro, and endothelial cells 
  # adipo and fibro cells are mixed with fine annotation in becoming activated fibro, fibro, NPCs? 
  # convert to just call these fibro_adipo cells if not tumor 
  CD45neg.seurat@meta.data <- CD45neg.seurat@meta.data %>%
    dplyr::mutate(cluster_anno = case_when(
      SingleR_main == "Cardiomyocytes" ~  "Cardiomyocytes",
      SingleR_main == "Endothelial cells" ~  "Endothelial_cells",
      SingleR_main == "Erythrocytes" ~  "Erythrocytes",
      seurat_clusters %in% tumor.clusters ~ paste0("Tumor_", seurat_clusters), 
      seurat_clusters %in% normal.clusters ~ SingleR_main,
      TRUE ~ "error")) 
  
  
  unique(CD45neg.seurat@meta.data$cluster_anno)
  
  # clean up seurat object to prevent mistakenly using wrong metadata annotation 
  colnames(CD45neg.seurat@meta.data)
  head(CD45neg.seurat@meta.data)
  CD45neg.seurat@meta.data <- CD45neg.seurat@meta.data %>%
    dplyr::select(c("orig.ident", "nCount_RNA", "nFeature_RNA", "nCount_SCT", "nFeature_SCT", 
                    "percent_mito", "percent_ribo", "percent_mito_ribo", "log10GenesPerUMI", 
                    "seurat_clusters", "SampleID", "Time", "Treatment", "Compartment", 
                    "SingleR_main", "SingleR_fine", "Disease", "subclone", "cluster_anno"))
  
  
  
  # rename tumor clusters to have sequential numbers as opposed to the seruat cluster numbers they currently have 
  # make a new column, because renaming might overwride as move along rows and names become similar with time 
  # therefore make new column and then remove the whole column and replace it 
  CD45neg.seurat@meta.data <- CD45neg.seurat@meta.data %>%
    dplyr::mutate(cluster_anno_new = case_when(
      cluster_anno == "Tumor_0" ~  "Tumor_0",
      cluster_anno == "Tumor_1" ~  "Tumor_1",
      cluster_anno == "Tumor_4" ~  "Tumor_2",
      cluster_anno == "Tumor_5" ~  "Tumor_3",
      cluster_anno == "Tumor_6" ~  "Tumor_4",
      cluster_anno == "Tumor_8" ~  "Tumor_5",
      cluster_anno == "Tumor_9" ~  "Tumor_6", 
      TRUE ~ cluster_anno))
  
  table(CD45neg.seurat@meta.data$cluster_anno, CD45neg.seurat@meta.data$cluster_anno_new)
  
  # replace old anno with new and clean up 
  CD45neg.seurat@meta.data$cluster_anno <- CD45neg.seurat@meta.data$cluster_anno_new
  CD45neg.seurat@meta.data$cluster_anno_new <- NULL
  
  
  
  # set factorisation 
  CD45neg.seurat@meta.data$Treatment <- factor(CD45neg.seurat@meta.data$Treatment, 
                                               levels = c("Vehicle", "PDS"))
  
  CD45neg.seurat@meta.data$SampleID <- factor(CD45neg.seurat@meta.data$SampleID, 
                                              levels = c("Day4_Vehicle_CD45Neg", "Day4_PDS_CD45Neg"))
  
  
  CD45neg.seurat@meta.data$seurat_clusters <- factor(CD45neg.seurat@meta.data$seurat_clusters)
  
  CD45neg.seurat@meta.data$Disease <- factor(CD45neg.seurat@meta.data$Disease, 
                                             levels = c("normal", "tumor"))
  
  CD45neg.seurat@meta.data$subclone <- factor(CD45neg.seurat@meta.data$subclone, 
                                              levels = c("normal", "1", "2", "3", "4", "5", "undefined"))
  
  CD45neg.seurat@meta.data$cluster_anno <- factor(CD45neg.seurat@meta.data$cluster_anno, 
                                                  levels = c("Cardiomyocytes", "Endothelial_cells", "Erythrocytes", "Fibroblasts", "Tumor_0", "Tumor_1", "Tumor_2", "Tumor_3", "Tumor_4", "Tumor_5", "Tumor_6"))
  
  
  
  # save output
  SaveSeuratRds(CD45neg.seurat, file = "saves/CD45neg_seurat.rds")
  
}



```
