---
title: "CD45Neg_dataset_analysis"
author: "Dillon Corvino"
date: "25/01/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# Data 
# B16 - OVA 

# PDS = I.T - bind Gquadraplex
# day 4 post-treatment 


# Research questions 
# How does PDS induces an anti-tumor immune response?
# Known that after PDS treatment, certain DAMPs are induced, intratumoral chemokine levels go up drastically and subsequntly immune cells are recruited into the tumor.
# Main research questions are:
#1. Which cell subsets are producing the chemokines
#2. Are there differences between macrophages and DCs between PDS and Vehicle (because there are DAMPs that should activate them)
#3. How does the treatment effect the tumor cells (is it possible to discriminate them from other tissue cells, if not they express ovalbumin...this should make it at least possible to identify them.)




# key genes
# IL1b, IL6, IL12, IFNG, TNFA, 

# CCL2,3,4,5 
# CXCL1, CXCL9,10,13 





```


### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)


# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- TRUE
do.plot <- TRUE
quick.load <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())

# Nebulosa / scCustomize has issues with ggplot2 v3.5.0, therefore ensure v3.4.4 is installed
#devtools::install_version("ggplot2", version = '3.4.4')
# but note for clusttree, require updated ggplot2 >= 3.5

library("Seurat")
library("SeuratObject")
library("SeuratData")
library("SeuratWrappers")
library("dplyr")
library("scCustomize")
library("tibble")
library("ggplot2")

########################################
# Parallelization of seurat functions
########################################

parallelization <- TRUE


if(parallelization){
  library(future)
  # check the current active plan
  plan()
  
  # detect cores
  ncores <- parallel::detectCores()
  ncores <- ncores-5
  # set plan
  plan("multisession", workers = ncores)
  plan()
  
  # set max limit for global variables
  # note this will drastically increase RAM usage, be aware of machine limits
  # set value is in bites therefore 1000*1024^2 = 1GB Therefore set to 10GBs 
  options(future.globals.maxSize = 10000 * 1024^2, 
          future.rng.onMisuse="ignore") 
}



if(quick.load){
  
  CD45neg.seurat <- LoadSeuratRds("saves/Nils_seurat_CD45Neg.rds")
  
  # set levels for easier downstream plotting 
  CD45neg.seurat@meta.data$Treatment <- factor(CD45neg.seurat@meta.data$Treatment, levels = c("Vehicle", "PDS"))
  
}
```


## Clean dataset {.tabset}

## Identifying Immune cell contamination - SingleR cluster annotation
```{r SingleR_cluster_annotation}

# Notes:
# there is an error with collect() when trying to load celldex datasets
# i think its an error with biocmanager and has been fixed with a very recent update. > 3.16 
# but im running r 4.2.3 and this would require 4.3
# not keen to upgrade right now 
# so will try a work around of downgrading dbplyr to version 2.3.4 from 2.4 which im currently running 
# need to downgrade to this version 
#devtools::install_version("dbplyr", version = "2.3.4")
#devtools::install_version("dbplyr", version = "2.4")
# in theory if biocmanager is up to date, could just run below and should fix the problem
#library(BiocManager) # 3.16
#install("BiocFileCache")
# also:
# need to ensure matrixStats is at version 1.1.0 
#remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts
#BiocManager::install(version = "3.18")

if(long.compute){
  
  output.dir <- "results/CD45Neg_Immune_cell_contamination/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # load required packages
  library("SingleR")
  library("celldex")
  
  
  ##################################
  # Use MouseRNAseqData
  ##################################
  
  ref.mouse <-  celldex::MouseRNAseqData()
  
  
  # Benayoun Lab that identified, downloaded and processed data sets on GEO that corresponded to sorted cell types (Benayoun et al., 2019).
  
  # The dataset contains 358 mouse RNA-seq samples annotated to 18 main cell types
  # further split into 28 subtypes under label.fine 
  
  
  ######################
  # Fine labels 
  ######################
  SingleR.pred <- SingleR(test = CD45neg.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45neg.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.fine)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
  #SingleR::plotDeltaDistribution(SingleR.pred)
  #SingleR::plotScoreDistribution(SingleR.pred)
  
  ######################
  # Main labels 
  ######################
  
  SingleR.pred <- SingleR(test = CD45neg.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45neg.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.main)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_main_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
}

```


### Confirming Immune cell contamination - CD45 and granulocyte linage marker gene visualisation
```{r CD45_and_granulocyte_lineage_gene_visualisation}

if(long.compute){

# create output dir
output.dir <- paste0("results/CD45Neg_Immune_cell_contamination/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# list melanocyte genes 
goi <- c("Ptprc", "Cd14")

# Plot Feature and Density plots iteratively over goi 
for(i in seq_along(goi)){
  
  # feature plots
  print(
    scCustomize::FeaturePlot_scCustom(CD45neg.seurat, 
                                      reduction = "umap", 
                                      features = goi[i])
    
  )
  
  dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Neg_", goi[i], ".pdf"))
  dev.off()
  
  
  # density plots
  
  print(
    scCustomize::Plot_Density_Custom(CD45neg.seurat, 
                                     reduction = "umap",
                                     features = goi[i], 
                                     custom_palette = batlow.pal)
  )
  dev.copy(pdf, paste0(output.dir, "UMAP_DensityPlot_CD45Neg_", goi[i], ".pdf"))
  dev.off()
  
  
  
}


##############################
# Stacked vln plot
##############################

scCustomize::Stacked_VlnPlot(CD45neg.seurat, 
                             features = goi,
                             group.by = "seurat_clusters",
                             pt.size = 0)

dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_granulocyte_markers.pdf"))
dev.off()


###########
# heatmap
###########

# aggregate expression for heatmap visualisation
pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("seurat_clusters"))

DoHeatmap(pseudo.cd45neg, 
          features = goi, 
          draw.lines = FALSE, 
          angle = 90,
          raster = FALSE) + 
  scico::scale_fill_scico(palette = "batlow", 
                          direction = 1, 
                          na.value = "white")

dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_granulocyte_markers.pdf"))
dev.off()

}


```


### Remove contamination immune clusters
```{r remove_immune_contamination}

if(long.compute){
  
    # cluster 5 looks to be immune contaimination 
  
  # set immune cell cluster variable
  Immune.cell.cluster <- 5
  
  
  head(CD45neg.seurat@meta.data)
  
  
  # set assay
  DefaultAssay(CD45neg.seurat) <- "SCT"
  
  # set idents
  Idents(CD45neg.seurat) <- "seurat_clusters"
  
  # immune cell contamination cluster is relatively small, sharp drop of in cluster sizes at cluster 4
  table(CD45neg.seurat@meta.data$seurat_clusters) # cluster 5 = 617 cells
  
  # remove this cluster from dataset 
  Idents(CD45neg.seurat) <- CD45neg.seurat@meta.data$seurat_clusters
  CD45neg.seurat <- subset(CD45neg.seurat, idents = Immune.cell.cluster, invert = TRUE)
  unique(Idents(CD45neg.seurat))
  
  
  # drop levels in seurat cluster meta
  CD45neg.seurat$seurat_clusters <- droplevels(CD45neg.seurat$seurat_clusters)
  
}

```


## Renormalise / integrate / dim reduction | after removing contamination immune cells {.tabset}

## re-calculate normalisation and integration for CD45Neg dataset
```{r re_normalise_and_integrate_CD45Neg_dataset}

if(long.compute){
  
  
  # split object up based on sequencing run / in seurat V5 these are splitting the layers
  CD45neg.seurat[["RNA"]] <- split(CD45neg.seurat[["RNA"]],  f = CD45neg.seurat$SampleID)
  
  DefaultAssay(CD45neg.seurat) <- "RNA"
  
  
  
  
  #####################################################
  # Normalise with scTransform and integrate data
  #####################################################
  # note error with more recent version of matrixStats resulting in "useNames = NA is defunct" error. downgrade to matrixStats version 1.1.0 as a workaround
  # https://github.com/satijalab/seurat/issues/7501#issuecomment-1854571904
  #remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts  
  
  
  
  # note RNA assay is split into run level layers, below script will apply normalisation etc layer-wise
  
  CD45neg.seurat <- CD45neg.seurat %>%
    SCTransform(method = "glmGamPoi", 
                vars.to.regress = "percent_mito", 
                vst.flavor = "v2",
                verbose = TRUE) %>%
    RunPCA(reduction.name = "pca_sct") %>%
    RunUMAP(reduction = "pca_sct", 
            dims = 1:30, 
            reduction.name = "umap.unintegrated.sct")
  
  
  DefaultAssay(CD45neg.seurat) <- "SCT"
  
  
  # integrate layers using cca integration and calculate dim redu and cluster identification
  
  # cluster resolution value based on previous estimation 
  clust.res <- 0.4
  
  
  CD45neg.seurat <- CD45neg.seurat %>%
    IntegrateLayers(orig.reduction = "pca_sct",
                    new.reduction = "integrated.cca.sct",
                    method = CCAIntegration,
                    normalization.method = "SCT",
                    verbose = F) %>%
    FindNeighbors(dims = 1:30,
                  reduction = "integrated.cca.sct") %>%
    FindClusters(resolution = clust.res, 
                 cluster.name = "clusters_cca_sct") %>%
    RunUMAP(reduction = "integrated.cca.sct",
            dims = 1:30, 
            reduction.name = "umap")
  
  
}



```


## plot integration after removing contamination immune cells from CD45Neg dataset 

```{r plot_integration_after_removing_contamination}

if(long.compute){
  
  if(do.plot){
    
    # create output dir
    output.dir <- paste0("results/CD45Neg_Integration_validation_after_cleaning/")
    
    if(!dir.exists(output.dir)){
      dir.create(output.dir, 
                 recursive = T)
      
    }
    
    # set variables
    unintegrated.reduction <- "umap.unintegrated.sct"
    integrated.reduction <- "umap"
    title.var <- "SCT"
    
    
    
    # evaluate integration and dim reduction
    
    # set default assay
    DefaultAssay(CD45neg.seurat) <- "SCT"
    
  
    
    vars.plot <- c("Treatment", "SampleID", "seurat_clusters")
    
    
    for(i in seq_along(vars.plot)){
      
      
      # remove "." from name
      name.var <- gsub("\\.", "_", vars.plot[i])
      
      before.int <- DimPlot(CD45neg.seurat,
                            reduction = unintegrated.reduction, 
                            label = T,
                            group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("Unintegrated ", title.var, " Normalisation"))
      
      after.int <- DimPlot(CD45neg.seurat,
                           reduction = integrated.reduction, 
                           label = T,
                           group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("CCA Integration using ", title.var, " Normalisation"))
      
      print(before.int | after.int)
      dev.copy(pdf, paste0(output.dir, "UMAP_before_after_integration_", name.var, "_CD45Neg_dataset.pdf"))
      dev.off()
      
      
      
    }
  }
  
  
}


```


### Save final cleaned seurat objects
```{r save_seurat_object_final}

if(long.compute){
  
  library("SeuratDisk")

  SaveSeuratRds(CD45neg.seurat, 
                file = "saves/Nils_seurat_CD45Neg.rds")
  
}

```


## Cluster annotation {.tabset}

## SingleR cluster annotation
```{r SingleR_cluster_annotation}

# Notes:
# there is an error with collect() when trying to load celldex datasets
# i think its an error with biocmanager and has been fixed with a very recent update. > 3.16 
# but im running r 4.2.3 and this would require 4.3
# not keen to upgrade right now 
# so will try a work around of downgrading dbplyr to version 2.3.4 from 2.4 which im currently running 
# need to downgrade to this version 
#devtools::install_version("dbplyr", version = "2.3.4")
#devtools::install_version("dbplyr", version = "2.4")
# in theory if biocmanager is up to date, could just run below and should fix the problem
#library(BiocManager) # 3.16
#install("BiocFileCache")
# also:
# need to ensure matrixStats is at version 1.1.0 
#remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts
#BiocManager::install(version = "3.18")

if(long.compute){
  
  output.dir <- "results/CD45Neg_SingleR/"
  
  # Create output directory
  if(!dir.exists(paste0(output.dir))){
    dir.create(paste0(output.dir), 
               recursive = T)
  }
  
  
  # load required packages
  library("SingleR")
  library("celldex")
  
  
  ##################################
  # Use MouseRNAseqData
  ##################################
  
  ref.mouse <-  celldex::MouseRNAseqData()
  
  
  # Benayoun Lab that identified, downloaded and processed data sets on GEO that corresponded to sorted cell types (Benayoun et al., 2019).
  
  # The dataset contains 358 mouse RNA-seq samples annotated to 18 main cell types
  # further split into 28 subtypes under label.fine 
  
  
  ######################
  # Fine labels 
  ######################
  SingleR.pred <- SingleR(test = CD45neg.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45neg.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.fine)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_fine_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
  #SingleR::plotDeltaDistribution(SingleR.pred)
  #SingleR::plotScoreDistribution(SingleR.pred)
  
  ######################
  # Main labels 
  ######################
  
  SingleR.pred <- SingleR(test = CD45neg.seurat@assays$SCT$data, 
                          ref = ref.mouse, 
                          method = "cluster",
                          clusters = CD45neg.seurat@meta.data$seurat_clusters,
                          labels = ref.mouse$label.main)
  
  # Plot 
  p1 <- plotScoreHeatmap(SingleR.pred, 
                         show_colnames = TRUE)
  p1
  ggsave(paste0(output.dir, "Heatmap_cluster_annotations_main_labels_MouseRNAseqData_db.pdf"), plot = p1)
  
}

```


## Overview of dataset {.tabset}

### Broad overview of dataset
```{r broad_overview}

if(do.plot){
  
  
  # create output dir
  output.dir <- paste0("results/CD45Neg_Overview/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  
  # set assay
  DefaultAssay(CD45neg.seurat) <- "SCT"
  
  
  vars.to.plot <- c("Treatment", "SampleID", "seurat_clusters")
  
  
  for(j in seq_along(vars.to.plot)){
    name.var <- gsub("\\.", "_", vars.to.plot[j])
    
    print(
      scCustomize::DimPlot_scCustom(CD45neg.seurat,
                                    reduction = "umap", 
                                    label = TRUE,
                                    figure_plot = TRUE,
                                    group.by = vars.to.plot[j]) 
    )
    
    
    dev.copy(pdf, paste0(output.dir, "CD45Neg_UMAP_", name.var, ".pdf"))
    dev.off()
    
  }
  
  
  # split umap plots
  print(
    scCustomize::DimPlot_scCustom(CD45neg.seurat,
                                  reduction = "umap", 
                                  label = TRUE,
                                  figure_plot = TRUE,
                                  group.by = "seurat_clusters", 
                                  split.by = "Treatment") 
  )
  
  
  dev.copy(pdf, paste0(output.dir, "CD45Neg_UMAP_seurat_clusters_split_by_treatment.pdf"))
  dev.off()
  
}


```

### Cluster stats
```{r cluster_stats}

# create output dir
output.dir <- paste0("results/CD45Neg_Overview/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

#############################################
# Number of cells per cluster
#############################################

cluster.n <- table(CD45neg.seurat@meta.data$seurat_clusters)

cluster.n <- as.data.frame(cluster.n)
cluster.n$Var1 <- factor(cluster.n$Var1, levels = cluster.n$Var1)


ggplot(cluster.n) +
  aes(x = Var1, y = Freq, fill = Var1) +
  geom_col() +
  scale_fill_viridis_d(option = "viridis", 
                       direction = -1) +
  labs(x = "Cluster ID", y = "Number of cells", title = "Number of cells per cluster") +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

ggsave(paste0(output.dir, "Barplot_number_of_cells_per_cluster.pdf"))


#############################################
# Freq of clusters across treatment 
#############################################

dist.table <- table(CD45neg.seurat@meta.data$seurat_clusters, CD45neg.seurat@meta.data$Treatment)

treatment.total <- colSums(dist.table)

freq.table <- sweep(dist.table, 2, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("Vehicle", "PDS"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Treatment", y = "% of treatment", title = "Frequency of clusters per treatment") +
  scale_fill_manual(values = c(Vehicle = "grey", PDS = "darkred")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_clusters_per_treatment.pdf"))


```

### Cell Cycle scoring
```{r cell_cycle}

# create output dir
output.dir <- paste0("results/CD45Neg_cell_cycle/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}

################################
# Get cell cycle genes 
################################
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

# pseudo-map to mouse gene (captures most except Mlf1ip, Fam64a and Hn1)
s.genes <- stringr::str_to_title(s.genes)
g2m.genes <- stringr::str_to_title(g2m.genes)

#####################
# score cell cycle
#####################

CD45neg.seurat <- CellCycleScoring(CD45neg.seurat,
                                   s.features = s.genes,
                                   g2m.features = g2m.genes,
                                   set.ident = TRUE)


#####################
# Visualise score
#####################

head(CD45neg.seurat@meta.data)

# Feature plots 

# S-score 
scCustomize::FeaturePlot_scCustom(CD45neg.seurat, 
                                  feature = "S.Score", 
                                  reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Neg_S_score.pdf"))
dev.off()
  
# G2M-score 
scCustomize::FeaturePlot_scCustom(CD45neg.seurat, 
                                  feature = "G2M.Score", 
                                  reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Neg_G2M_score.pdf"))
dev.off()

# Dim plot split by phase
scCustomize::DimPlot_scCustom(CD45neg.seurat, 
                              group.by = "seurat_clusters",
                              split.by = "Phase",
                              label = TRUE,
                              reduction = "umap")

dev.copy(pdf, paste0(output.dir, "UMAP_CD45Neg_cluster_by_phase.pdf"))
dev.off()


# Stacked VlnPlot
scCustomize::Stacked_VlnPlot(CD45neg.seurat,
                             features = c("S.Score", "G2M.Score"), 
                             group.by = "seurat_clusters")

dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_cell_cycle.pdf"))
dev.off()




###################
# Quantification 
###################


# Frequency of phase per cluster
dist.table <- table(CD45neg.seurat@meta.data$seurat_clusters, CD45neg.seurat@meta.data$Phase)

cluster.total <- rowSums(dist.table)

freq.table <- sweep(dist.table, 1, cluster.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin
rowSums(freq.table)

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("G1", "G2M", "S"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Predicted phase", y = "% of cluster", title = "Frequency of phase per cluster") +
  scale_fill_manual(values = c(G1 = "lightblue", G2M = "orange", S = "red")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_phase_per_cluster.pdf"))


# Frequency of phase per treatment

dist.table <- table(CD45neg.seurat@meta.data$Treatment, CD45neg.seurat@meta.data$Phase)

treatment.total <- rowSums(dist.table)

freq.table <- sweep(dist.table, 1, treatment.total, FUN = "/")
freq.table <- freq.table*100

colSums(freq.table) # validation that calculation on the right axis/margin
rowSums(freq.table)

# make dataframe 
freq.table <- as.data.frame(freq.table)

# set factor levels 
freq.table$Var2 <- factor(freq.table$Var2, levels = c("G1", "G2M", "S"))


ggplot(freq.table) +
  aes(x = Var2, y = Freq, fill = Var2) +
  geom_col() +
  labs(x = "Predicted phase", y = "% of treatment", title = "Frequency of phase per treatment") +
  scale_fill_manual(values = c(G1 = "lightblue", G2M = "orange", S = "red")) + 
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  facet_wrap(vars(Var1))

ggsave(paste0(output.dir, "Barplot_freq_of_phase_per_treatment.pdf"))






```


## DEG analysis {.tabset}

### Cluster-wise DEGs - calculate
```{r Clusterwise_DEGs_calculation}


# create output dir
output.dir <- paste0("results/CD45Neg_cluster_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  
  CD45neg.seurat <- PrepSCTFindMarkers(CD45neg.seurat)
  
  # set assay
  DefaultAssay(CD45neg.seurat) <- "SCT"
  
  
  # set idents
  Idents(CD45neg.seurat) <- "seurat_clusters"
  
  CD45Neg.de <- FindAllMarkers(CD45neg.seurat,
                               assay = "SCT",
                               only.pos = TRUE, 
                               verbose = TRUE)
  
  
  # save to file 
  write.csv(CD45Neg.de, paste0(output.dir, "CD45Neg_DEGs.csv"))
  
  
  
}else{
  
  # load DEG table
  CD45Neg.de <- read.csv(paste0(output.dir, "CD45Neg_DEGs.csv"))
  
}



```

### Cluster-wise DEGs - heatmaps
```{r clusterwise_DEGs_heatmaps}


if(do.plot){
  # aggregate expression for heatmap visualisation
  
  pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("seurat_clusters"))
  
  
  
  goi <- CD45Neg.de %>%
    filter(p_val_adj < 0.05) %>%
    group_by(cluster) %>%
    top_n(10, avg_log2FC) %>%
    pull(gene)
  
  
  DoHeatmap(pseudo.cd45neg, 
            features = goi, 
            draw.lines = FALSE, 
            raster = FALSE) + 
  scico::scale_fill_scico(palette = "batlow", 
                          direction = 1, 
                          na.value = "white")
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_top_10_DEGs_pseudobulk.pdf"))
  dev.off()
  
  
  
  
  # iterate over the top degs for each cluster 
  cluster.id <- unique(CD45Neg.de$cluster)
  for(i in seq_along(cluster.id)){
    
    
    print(cluster.id[i])
    
    goi <- CD45Neg.de %>%
      filter(p_val_adj < 0.05) %>%
      filter(cluster == cluster.id[i]) %>%
      top_n(50, avg_log2FC)
    
    goi <- unique(goi)
    
    print(
      DoHeatmap(pseudo.cd45neg, 
                features = goi$gene, 
                draw.lines = FALSE, 
                raster = FALSE) + 
  scico::scale_fill_scico(palette = "batlow", 
                          direction = 1, 
                          na.value = "white")
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
    dev.off()
    
  }
  
}

```

## Melanocyte gene visualisation {.tabset}

### Melanocyte-specific gene expression
```{r melanocyte_genes}

# create output dir
output.dir <- paste0("results/CD45Neg_Melanocyte_genes/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# list melanocyte genes 
goi <- c("Mitf", "Tyr", "Pmel", "Mlana", "Tyrp1", "Sox10")


# Plot Feature and Density plots iteratively over goi 
for(i in seq_along(goi)){
  
  # feature plots
  print(
    scCustomize::FeaturePlot_scCustom(CD45neg.seurat, 
                                      reduction = "umap", 
                                      features = goi[i])
    
  )
  
  dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Neg_", goi[i], ".pdf"))
  dev.off()
  
  
  # density plots
  
  print(
    scCustomize::Plot_Density_Custom(CD45neg.seurat, 
                                     reduction = "umap",
                                     features = goi[i], 
                                     custom_palette = batlow.pal)
  )
  dev.copy(pdf, paste0(output.dir, "UMAP_DensityPlot_CD45Neg_", goi[i], ".pdf"))
  dev.off()
  
  
  
}


##############################
# Stacked vln plot
##############################

scCustomize::Stacked_VlnPlot(CD45neg.seurat, 
                             features = goi,
                             group.by = "seurat_clusters",
                             pt.size = 0)

dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_melanocyte_genes.pdf"))
dev.off()


###########
# heatmap
###########

# aggregate expression for heatmap visualisation
pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("seurat_clusters"))

DoHeatmap(pseudo.cd45neg, 
          features = goi, 
          draw.lines = FALSE, 
          angle = 90,
          raster = FALSE) + 
  scico::scale_fill_scico(palette = "batlow", 
                          direction = 1, 
                          na.value = "white")

dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_melanocyte_genes.pdf"))
dev.off()


```


## RQ1: Which CD45Neg clusters express chemokines or cytokines {.tabset}

### Chemokine and Cytokine expression visualisation 
```{r chemokine_cytokine_expression}

# create output dir
output.dir <- paste0("results/CD45Neg_chemokine_cytokines/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# aggregate expression for heatmap visualisation
pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("seurat_clusters"))


########################################
# List gene sets of interest to plot
########################################

cxc.genes <- grep("^Cxc.*", rownames(pseudo.cd45neg), value = T)
ccr.genes <- grep("^Ccr.*", rownames(pseudo.cd45neg), value = T)
il.genes <- grep("^Il.*", rownames(pseudo.cd45neg), value = T)
gzm.genes <- grep("^Gzm.*", rownames(pseudo.cd45neg), value = T)
ifn.genes <- grep("^Ifn.*", rownames(pseudo.cd45neg), value = T)
custom.genes <- c("Il1b", "Il6", "Il12a", "Ifng", "Tnf", "Ccl2", "Ccl3", "Ccl4", "Ccl5", "Cxcl1", "Cxcl9", "Cxcl10", "Cxcl13") # Il12b not present in CD45Neg dataset

goi <- list(Cxcr_genes = cxc.genes, 
            Ccr_genes = ccr.genes,
            Interleukin_genes = il.genes,
            Gzm_genes = gzm.genes,
            Ifn_genes = ifn.genes,
            Custom_genes = custom.genes)


###################
# Plot heatmaps
###################

for(i in seq_along(names(goi))){
  
  print(
    DoHeatmap(pseudo.cd45neg, 
              features = goi[[i]], 
              draw.lines = FALSE, 
              angle = 90,
              raster = FALSE) + 
      scico::scale_fill_scico(palette = "batlow", 
                              direction = 1, 
                              na.value = "white")
  )
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_pseudobulk_", names(goi)[i], ".pdf"))
  dev.off()
  
}


##############################
# Stacked vln plot
##############################

for(i in seq_along(names(goi))){
  
  if(length(goi[[i]]) <= 15){
    
    
    print(
      scCustomize::Stacked_VlnPlot(CD45neg.seurat, 
                                   features = goi[[i]],
                                   group.by = "seurat_clusters",
                                   pt.size = 0)
    )
    
    dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_", names(goi)[i], ".pdf"))
    dev.off()
    
  }else{
    
    temp.goi <- goi[[i]]
    
    # Split the vector into multiple vectors
    split_indices <- seq(1, length(temp.goi), by = 15)
    split_list <- lapply(split_indices, function(i) temp.goi[i:min(i + 14, length(temp.goi))])
    
    
    for(j in seq_along(length(split_list))){
      
      print(
        scCustomize::Stacked_VlnPlot(CD45neg.seurat, 
                                     features = split_list[[j]],
                                     group.by = "seurat_clusters",
                                     pt.size = 0)
      )
      
      dev.copy(pdf, paste0(output.dir, "Stacked_VlnPlot_CD45Neg_", names(goi)[i], "_", j, ".pdf"))
      dev.off()
      
    }
    
  }
  
}


```

## RQ2: DEGs within clusters across treatment {.tabset}

### Treatment-wise DEGs - calculate
```{r Treatmentwise_DEGs_calculation}

# create output dir
output.dir <- paste0("results/CD45Neg_treatment_wise_DEGs/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  
  CD45neg.seurat <- PrepSCTFindMarkers(CD45neg.seurat)
  
  # set assay
  DefaultAssay(CD45neg.seurat) <- "SCT"
  
  
  
  # add annotation for treatment 
  CD45neg.seurat@meta.data$Treatment_cluster <- paste0(CD45neg.seurat@meta.data$Treatment, "_", CD45neg.seurat@meta.data$seurat_clusters)
  
  # set Idents
  Idents(CD45neg.seurat) <- "Treatment_cluster"
  
  
  cluster.ids <- unique(CD45neg.seurat@meta.data$seurat_clusters)
  
  # create list to store output
  DEG.list <- list()
  
  # run through DEG specifically between vehicle and treatment for each cluster 
  
  for(i in seq_along(cluster.ids)){
    temp.de <- FindMarkers(CD45neg.seurat,
                           assay = "SCT",
                           ident.1 = paste0("PDS_", cluster.ids[i]),
                           ident.2 = paste0("Vehicle_", cluster.ids[i]),
                           only.pos = FALSE, 
                           recorrect_umi = FALSE,
                           verbose = TRUE)
    
    DEG.list[[i]] <- temp.de
    
    # save to file 
    write.csv(temp.de, paste0(output.dir, "DEGs_PDS_", cluster.ids[i], "_vs_Vehicle_", cluster.ids[i], ".csv"))
    
    
  }
  # add names to DEG.list
  names(DEG.list) <- cluster.ids
  
  # save DEG list to file 
  saveRDS(DEG.list, paste0(output.dir, "DEG_list_object.rds"))
  
  
}else{
  
  DEG.list <- readRDS(paste0(output.dir, "DEG_list_object.rds"))
  
}


```

### Treatment-wise DEGs - Volcano
```{r treatmentwise_DEGs_volcano}

# visualise treatment-wise DEGs as volcano plots


if(do.plot){
  
  for(i in 1:length(DEG.list)){
    
    FDR.theshold <- 0.01
    logFC.threshold <- 1
    
    # filter data
    input.data <- DEG.list[[i]] %>%
      dplyr::rename(logFC = avg_log2FC) %>%
      dplyr::rename(FDR = p_val_adj) %>%
      filter(FDR < FDR.theshold)
    
    # color points
    volcano.col <- usefulfunctions::colour.points.volcano(input.data,
                                                          increase.col = "Red",
                                                          decreased.col = "Blue",
                                                          FDR.cutoff = FDR.theshold, 
                                                          logFC.cutoff = logFC.threshold)
    # plot heatmap
    print(
      usefulfunctions::custom.enhanced.volcano(input.data, 
                                               pCutoff = FDR.theshold,
                                               FCcutoff = logFC.threshold,
                                               title = paste0(names(DEG.list)[i], " PDS vs Vehicle"),
                                               subtitle = paste0("adjPval threshold = ", FDR.theshold, " & log2FC threshold = ", logFC.threshold),
                                               colCustom = volcano.col)
    )
    dev.copy(pdf, paste0(output.dir, "VolcanoPlot_PDS_vs_Vehicle_cluster_", names(DEG.list)[i], ".pdf"))
    dev.off()
    
    
  }
  
}

```

### Export aggregated counts table
```{r export_aggregated_counts}

# create output dir
output.dir <- paste0("results/tables/CD45Neg/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


if(long.compute){
  
  # Aggregate expression by cluster 
  pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("seurat_clusters"))
  
  # Write to file
  temp.out <- GetAssayData(pseudo.cd45neg,
                           layer = "data",
                           assay = "SCT")
  
  write.csv(temp.out, paste0(output.dir, "CD45Neg_Aggregate_SCT_data_by_cluster.csv"))
  
  # Aggregate expression by cluster and treatment
  pseudo.cd45neg <- AggregateExpression(CD45neg.seurat,
                                        assays = "SCT",
                                        return.seurat = TRUE, 
                                        group.by = c("seurat_clusters", "Treatment"))
  # write to file
  temp.out <- GetAssayData(pseudo.cd45neg,
                           layer = "data",
                           assay = "SCT")
  
  write.csv(temp.out, paste0(output.dir, "CD45Neg_Aggregate_SCT_data_by_cluster_and_treatment.csv"))
  
  # clean environment
  rm(pseudo.cd45neg)
  rm(temp.out)
  
}



```






