---
title: "Supplementary Figure 2: ISG and cGAS/STING Module Scores"
author: "Dillon Corvino"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Environment
```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  tidy = TRUE
)

rm(list = ls())
cat("\014")
gc()

# Establish working directory
library("rstudioapi")
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories
output.dir <- "results/Supplementary_Figure_2/"
if(!dir.exists(output.dir)){
  dir.create(output.dir, recursive = TRUE)
}

# Load libraries
library(Seurat)
library(SeuratObject)
library(dplyr)
library(ggplot2)
library(scCustomize)
library(SeuratExtend)
library(tidyr)

# Load data from Desktop
cd45pos.seurat <- readRDS("/Users/dilloncorvino/Desktop/NILS_scRNAseq/Processed_files/Seurat_objects/CD45pos_seurat.rds")
cd45neg.seurat <- readRDS("/Users/dilloncorvino/Desktop/NILS_scRNAseq/Processed_files/Seurat_objects/CD45neg_seurat.rds")

cat("Data loaded. CD45pos:", dim(cd45pos.seurat), "| CD45neg:", dim(cd45neg.seurat), "\n")
```

### Gene Signatures
```{r gene_signatures}

# ISG panel: curated Type I interferon-stimulated genes
isg.genes <- c(
  # Core ISGs
  "Isg15", "Ifit1", "Ifit2", "Ifit3", "Irf7", "Stat1", "Stat2", "Oas1a",
  "Mx2", "Rsad2", "Usp18", "Ifi44", "Ifi44l", "Oasl1", "Gbp2", "Cxcl10",
  # Canonical Hallmark ISGs
  "Ifih1", "Ifi27", "Ifi27l2b", "Ifi30", "Ifi35", "Ifitm1", "Ifitm2", "Ifitm3",
  "Isg20", "Ddx60", "Eif2ak2", "Ly6e", "Oas1g", "Parp12", "Parp14", "Rtp4",
  # Pattern-recognition/signalling
  "Ddx58", "Oas1b", "Oas2", "Oas3", "Oasl2"
)

# cGAS/STING signaling pathway genes
cgas.genes <- c(
  "Cgas", "Ifi204", "Ddx41", "Tmem173", "Tbk1", "Ikbke", "Irf3", "Irf7",
  "Nfkb1", "Rela", "Trex1", "Enpp1", "Ifnb1", "Il6", "Tnf", "Cxcl10",
  "Ccl2", "Ccl5", "Aim2", "Pycard", "Casp1"
)

cat("ISG panel:", length(isg.genes), "genes\n")
cat("cGAS/STING panel:", length(cgas.genes), "genes\n")
```

### Prepare CD45+ (Myeloid) Subset
```{r prepare_cd45pos}

# Subset to Day 4 and keep only myeloid populations
myeloid.keep <- c("DC", "Macrophages", "Monocytes", "Neutrophils")
Idents(cd45pos.seurat) <- cd45pos.seurat@meta.data$Time
cd45pos.d4 <- subset(cd45pos.seurat, idents = "Day4")

Idents(cd45pos.d4) <- cd45pos.d4@meta.data$cell_type
cd45pos.d4 <- subset(cd45pos.d4, idents = myeloid.keep)

# Create combined metadata: CellType_Treatment
cd45pos.d4$CellType_Treatment <- paste(
  cd45pos.d4@meta.data$cell_type,
  cd45pos.d4@meta.data$Treatment,
  sep = "_"
)

# Set default assay and factor levels
DefaultAssay(cd45pos.d4) <- "SCT"
cd45pos.d4$Treatment <- factor(cd45pos.d4$Treatment, levels = c("Vehicle", "PDS"))
cd45pos.d4$cell_type <- factor(cd45pos.d4$cell_type, levels = myeloid.keep)

cat("CD45pos Day4 myeloid cells:", dim(cd45pos.d4), "\n")
cat("Cell type breakdown:\n")
print(table(cd45pos.d4$cell_type, cd45pos.d4$Treatment))
```

### Prepare CD45- (Tumor + Fibroblast) Subset
```{r prepare_cd45neg}

# Only Day 4 data in CD45neg
DefaultAssay(cd45neg.seurat) <- "SCT"

# Collapse tumor clusters into single "Tumor" category
cd45neg.seurat$cell_type <- ifelse(
  grepl("^Tumor", cd45neg.seurat$cluster_anno),
  "Tumor",
  cd45neg.seurat$cluster_anno
)

# Keep only Tumor and Fibroblasts
cd45neg.d4 <- subset(cd45neg.seurat, subset = cell_type %in% c("Tumor", "Fibroblasts"))

# Create combined metadata: CellType_Treatment
cd45neg.d4$CellType_Treatment <- paste(
  cd45neg.d4@meta.data$cell_type,
  cd45neg.d4@meta.data$Treatment,
  sep = "_"
)

# Set factor levels
cd45neg.d4$Treatment <- factor(cd45neg.d4$Treatment, levels = c("Vehicle", "PDS"))
cd45neg.d4$cell_type <- factor(cd45neg.d4$cell_type, levels = c("Tumor", "Fibroblasts"))

cat("CD45neg Day4 tumor/fibroblast cells:", dim(cd45neg.d4), "\n")
cat("Cell type breakdown:\n")
print(table(cd45neg.d4$cell_type, cd45neg.d4$Treatment))
```

## Supplementary Figure 2: ISG Signature {.tabset}

### Panel a-b: ISG Pseudobulk Heatmaps

#### Myeloid (Panel a)
```{r isg_heatmap_myeloid}

# Pseudobulk aggregation by cell type and treatment
pseudo.myeloid <- AggregateExpression(
  cd45pos.d4,
  assays = "SCT",
  return.seurat = FALSE,
  group.by = c("cell_type", "Treatment")
)

# Keep only ISG genes present in dataset
isg.present <- intersect(isg.genes, rownames(pseudo.myeloid))
pseudo.myeloid <- pseudo.myeloid[isg.present, ]

cat("ISG genes present in myeloid:", nrow(pseudo.myeloid), "/", length(isg.genes), "\n")

# Convert to Seurat object for DoHeatmap compatibility
pseudo.myeloid.seurat <- CreateSeuratObject(pseudo.myeloid)

# Plot heatmap
p.isg.myeloid <- DoHeatmap(
  pseudo.myeloid.seurat,
  features = rownames(pseudo.myeloid),
  group.by = "orig.ident",
  draw.lines = TRUE,
  size = 3
) +
  scale_fill_gradientn(
    colors = c("blue", "white", "red"),
    name = "SCT\nExpression"
  ) +
  theme(axis.text.y = element_text(size = 8))

print(p.isg.myeloid)

ggsave(
  paste0(output.dir, "SupFig2a_ISG_heatmap_myeloid.pdf"),
  p.isg.myeloid,
  width = 10,
  height = 12,
  units = "in"
)
```

#### CD45- (Panel b)
```{r isg_heatmap_cd45neg}

# Pseudobulk aggregation
pseudo.cd45neg <- AggregateExpression(
  cd45neg.d4,
  assays = "SCT",
  return.seurat = FALSE,
  group.by = c("cell_type", "Treatment")
)

# Keep only ISG genes present
pseudo.cd45neg <- pseudo.cd45neg[isg.present, ]

# Convert to Seurat object
pseudo.cd45neg.seurat <- CreateSeuratObject(pseudo.cd45neg)

# Plot heatmap
p.isg.cd45neg <- DoHeatmap(
  pseudo.cd45neg.seurat,
  features = rownames(pseudo.cd45neg),
  group.by = "orig.ident",
  draw.lines = TRUE,
  size = 3
) +
  scale_fill_gradientn(
    colors = c("blue", "white", "red"),
    name = "SCT\nExpression"
  ) +
  theme(axis.text.y = element_text(size = 8))

print(p.isg.cd45neg)

ggsave(
  paste0(output.dir, "SupFig2b_ISG_heatmap_CD45neg.pdf"),
  p.isg.cd45neg,
  width = 10,
  height = 12,
  units = "in"
)
```

### Panels c-d: ISG Module Scores

#### Myeloid Module Score (Panel c)
```{r isg_module_score_myeloid}

# Add ISG module score
cd45pos.d4 <- AddModuleScore(
  cd45pos.d4,
  features = list(isg.present),
  name = "ISG",
  assay = "SCT"
)

# Rename score column for clarity
cd45pos.d4$ISG_score <- cd45pos.d4$ISG1
cd45pos.d4$ISG1 <- NULL

# Violin plot with statistics
p.isg.vln.myeloid <- VlnPlot2(
  cd45pos.d4,
  features = "ISG_score",
  group.by = "cell_type",
  split.by = "Treatment",
  stat.method = "wilcox.test",
  p.adjust.method = "holm",
  label = "p.signif",
  hide.ns = TRUE
) +
  labs(title = "ISG Module Score - Myeloid Cells",
       y = "ISG Score",
       x = "Cell Type")

print(p.isg.vln.myeloid)

ggsave(
  paste0(output.dir, "SupFig2c_ISG_violin_myeloid.pdf"),
  p.isg.vln.myeloid,
  width = 12,
  height = 6,
  units = "in"
)
```

#### CD45- Module Score (Panel d)
```{r isg_module_score_cd45neg}

# Add ISG module score
cd45neg.d4 <- AddModuleScore(
  cd45neg.d4,
  features = list(isg.present),
  name = "ISG",
  assay = "SCT"
)

# Rename score column
cd45neg.d4$ISG_score <- cd45neg.d4$ISG1
cd45neg.d4$ISG1 <- NULL

# Violin plot with statistics
p.isg.vln.cd45neg <- VlnPlot2(
  cd45neg.d4,
  features = "ISG_score",
  group.by = "cell_type",
  split.by = "Treatment",
  stat.method = "wilcox.test",
  p.adjust.method = "holm",
  label = "p.signif",
  hide.ns = TRUE
) +
  labs(title = "ISG Module Score - Tumor & Fibroblasts",
       y = "ISG Score",
       x = "Cell Type")

print(p.isg.vln.cd45neg)

ggsave(
  paste0(output.dir, "SupFig2d_ISG_violin_CD45neg.pdf"),
  p.isg.vln.cd45neg,
  width = 10,
  height = 6,
  units = "in"
)
```

## Supplementary Figure 2: cGAS/STING Signature {.tabset}

### Panels e-f: cGAS/STING Pseudobulk Heatmaps

#### Myeloid (Panel e)
```{r cgas_heatmap_myeloid}

# Keep only cGAS genes present in dataset
cgas.present <- intersect(cgas.genes, rownames(cd45pos.d4))
pseudo.myeloid.cgas <- AggregateExpression(
  cd45pos.d4,
  assays = "SCT",
  return.seurat = FALSE,
  group.by = c("cell_type", "Treatment"),
  features = cgas.present
)

cat("cGAS genes present in myeloid:", nrow(pseudo.myeloid.cgas), "/", length(cgas.genes), "\n")

# Convert to Seurat object for DoHeatmap
pseudo.myeloid.cgas.seurat <- CreateSeuratObject(pseudo.myeloid.cgas)

# Plot heatmap
p.cgas.myeloid <- DoHeatmap(
  pseudo.myeloid.cgas.seurat,
  features = rownames(pseudo.myeloid.cgas),
  group.by = "orig.ident",
  draw.lines = TRUE,
  size = 3
) +
  scale_fill_gradientn(
    colors = c("blue", "white", "red"),
    name = "SCT\nExpression"
  ) +
  theme(axis.text.y = element_text(size = 8))

print(p.cgas.myeloid)

ggsave(
  paste0(output.dir, "SupFig2e_cGAS_heatmap_myeloid.pdf"),
  p.cgas.myeloid,
  width = 10,
  height = 10,
  units = "in"
)
```

#### CD45- (Panel f)
```{r cgas_heatmap_cd45neg}

# Pseudobulk for CD45neg
pseudo.cd45neg.cgas <- AggregateExpression(
  cd45neg.d4,
  assays = "SCT",
  return.seurat = FALSE,
  group.by = c("cell_type", "Treatment"),
  features = cgas.present
)

# Convert to Seurat object
pseudo.cd45neg.cgas.seurat <- CreateSeuratObject(pseudo.cd45neg.cgas)

# Plot heatmap
p.cgas.cd45neg <- DoHeatmap(
  pseudo.cd45neg.cgas.seurat,
  features = rownames(pseudo.cd45neg.cgas),
  group.by = "orig.ident",
  draw.lines = TRUE,
  size = 3
) +
  scale_fill_gradientn(
    colors = c("blue", "white", "red"),
    name = "SCT\nExpression"
  ) +
  theme(axis.text.y = element_text(size = 8))

print(p.cgas.cd45neg)

ggsave(
  paste0(output.dir, "SupFig2f_cGAS_heatmap_CD45neg.pdf"),
  p.cgas.cd45neg,
  width = 10,
  height = 10,
  units = "in"
)
```

### Panels g-h: cGAS/STING Module Scores

#### Myeloid Module Score (Panel g)
```{r cgas_module_score_myeloid}

# Add cGAS module score
cd45pos.d4 <- AddModuleScore(
  cd45pos.d4,
  features = list(cgas.present),
  name = "cGAS",
  assay = "SCT"
)

# Rename score column
cd45pos.d4$cGAS_score <- cd45pos.d4$cGAS1
cd45pos.d4$cGAS1 <- NULL

# Violin plot with statistics
p.cgas.vln.myeloid <- VlnPlot2(
  cd45pos.d4,
  features = "cGAS_score",
  group.by = "cell_type",
  split.by = "Treatment",
  stat.method = "wilcox.test",
  p.adjust.method = "holm",
  label = "p.signif",
  hide.ns = TRUE
) +
  labs(title = "cGAS/STING Module Score - Myeloid Cells",
       y = "cGAS/STING Score",
       x = "Cell Type")

print(p.cgas.vln.myeloid)

ggsave(
  paste0(output.dir, "SupFig2g_cGAS_violin_myeloid.pdf"),
  p.cgas.vln.myeloid,
  width = 12,
  height = 6,
  units = "in"
)
```

#### CD45- Module Score (Panel h)
```{r cgas_module_score_cd45neg}

# Add cGAS module score
cd45neg.d4 <- AddModuleScore(
  cd45neg.d4,
  features = list(cgas.present),
  name = "cGAS",
  assay = "SCT"
)

# Rename score column
cd45neg.d4$cGAS_score <- cd45neg.d4$cGAS1
cd45neg.d4$cGAS1 <- NULL

# Violin plot with statistics
p.cgas.vln.cd45neg <- VlnPlot2(
  cd45neg.d4,
  features = "cGAS_score",
  group.by = "cell_type",
  split.by = "Treatment",
  stat.method = "wilcox.test",
  p.adjust.method = "holm",
  label = "p.signif",
  hide.ns = TRUE
) +
  labs(title = "cGAS/STING Module Score - Tumor & Fibroblasts",
       y = "cGAS/STING Score",
       x = "Cell Type")

print(p.cgas.vln.cd45neg)

ggsave(
  paste0(output.dir, "SupFig2h_cGAS_violin_CD45neg.pdf"),
  p.cgas.vln.cd45neg,
  width = 10,
  height = 6,
  units = "in"
)
```

## Summary
```{r summary}
cat("=== Supplementary Figure 2 Complete ===\n")
cat("Output directory:", output.dir, "\n")
cat("Panels generated:\n")
cat("  a: ISG heatmap (myeloid cells, Vehicle vs PDS)\n")
cat("  b: ISG heatmap (CD45- cells, Vehicle vs PDS)\n")
cat("  c: ISG module score violin plot (myeloid)\n")
cat("  d: ISG module score violin plot (CD45-)\n")
cat("  e: cGAS/STING heatmap (myeloid)\n")
cat("  f: cGAS/STING heatmap (CD45-)\n")
cat("  g: cGAS/STING module score violin plot (myeloid)\n")
cat("  h: cGAS/STING module score violin plot (CD45-)\n")
```
