---
title: "Supplementary Figure 2: ISG and cGAS/STING Module Scores"
author: "Dillon Corvino"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
---

Built with R version `r getRversion()`

Note: this script is intended to be run directly (e.g. via `knitr::purl()` +
`source()`, or by running chunks interactively) rather than rendered to a
report. Each panel is saved as its own PDF under `results/Supplementary_Figure_2/`
so it can be compared directly against the manuscript figure.

## Setup {.tabset}

### Environment
```{r setup}
rm(list = ls())
gc()

# Establish working directory (repo root)
if (!is.null(knitr::current_input())) {
  setwd(dirname(knitr::current_input()))
  setwd("..")
}

# Output directory (absolute path, robust to knitr working-directory changes)
output.base <- normalizePath("results/Supplementary_Figure_2", mustWork = FALSE)
if (!dir.exists(output.base)) {
  dir.create(output.base, recursive = TRUE)
}

LoadSeuratRds <- function(file) readRDS(file)

library(Seurat)
library(SeuratObject)
library(dplyr)
library(ggplot2)
library(SeuratExtend)

# Load preprocessed Seurat objects from saves/
# Data should be downloaded from Zenodo (https://doi.org/10.5281/zenodo.21391578) into saves/
cd45pos.seurat <- LoadSeuratRds("saves/CD45pos_seurat.rds")
cd45neg.seurat <- LoadSeuratRds("saves/CD45neg_seurat.rds")

cat("Data loaded. CD45pos:", dim(cd45pos.seurat), "| CD45neg:", dim(cd45neg.seurat), "\n")
```

### Gene Signatures
```{r gene_signatures}

# ISG panel: curated Type I interferon-stimulated genes
isg.genes <- c(
  # Core ISGs
  "Isg15", "Ifit1", "Ifit2", "Ifit3", "Irf7", "Stat1", "Stat2", "Oas1a",
  "Mx2", "Rsad2", "Usp18", "Ifi44", "Ifi44l", "Oasl1", "Gbp2", "Cxcl10",
  # Canonical Hallmark ISGs
  "Ifih1", "Ifi27", "Ifi27l2b", "Ifi30", "Ifi35", "Ifitm1", "Ifitm2", "Ifitm3",
  "Isg20", "Ddx60", "Eif2ak2", "Ly6e", "Oas1g", "Parp12", "Parp14", "Rtp4",
  # Pattern-recognition/signalling
  "Ddx58", "Oas1b", "Oas2", "Oas3", "Oasl2"
)

# cGAS/STING signaling pathway genes
cgas.genes <- c(
  "Cgas", "Ifi204", "Ddx41", "Tmem173", "Tbk1", "Ikbke", "Irf3", "Irf7",
  "Nfkb1", "Rela", "Trex1", "Enpp1", "Ifnb1", "Il6", "Tnf", "Cxcl10",
  "Ccl2", "Ccl5", "Aim2", "Pycard", "Casp1"
)

cat("ISG panel:", length(isg.genes), "genes\n")
cat("cGAS/STING panel:", length(cgas.genes), "genes\n")
```

### Prepare CD45+ (Myeloid) Subset
```{r prepare_cd45pos}

# Subset to Day 4 and keep only myeloid populations
myeloid.keep <- c("DC", "Macrophages", "Monocytes", "Neutrophils")
Idents(cd45pos.seurat) <- cd45pos.seurat@meta.data$Time
cd45pos.d4 <- subset(cd45pos.seurat, idents = "Day4")

Idents(cd45pos.d4) <- cd45pos.d4@meta.data$cell_type
cd45pos.d4 <- subset(cd45pos.d4, idents = myeloid.keep)

DefaultAssay(cd45pos.d4) <- "SCT"
cd45pos.d4$Treatment <- factor(cd45pos.d4$Treatment, levels = c("Vehicle", "PDS"))
cd45pos.d4$cell_type <- factor(cd45pos.d4$cell_type, levels = myeloid.keep)

# Desired column/group order for heatmaps and violin plots: Vehicle block then PDS block
myeloid.col.order <- c("DC_Vehicle", "Macrophages_Vehicle", "Monocytes_Vehicle", "Neutrophils_Vehicle",
                       "DC_PDS", "Macrophages_PDS", "Monocytes_PDS", "Neutrophils_PDS")

cat("CD45pos Day4 myeloid cells:", dim(cd45pos.d4), "\n")
print(table(cd45pos.d4$cell_type, cd45pos.d4$Treatment))
```

### Prepare CD45- (Tumor + Fibroblast) Subset
```{r prepare_cd45neg}

DefaultAssay(cd45neg.seurat) <- "SCT"

# Collapse tumor clusters into a single "Tumor" category; keep Fibroblasts as-is
cd45neg.seurat$cell_type <- ifelse(
  grepl("^Tumor", cd45neg.seurat$cluster_anno),
  "Tumor",
  as.character(cd45neg.seurat$cluster_anno)
)

cd45neg.d4 <- subset(cd45neg.seurat, subset = cell_type %in% c("Tumor", "Fibroblasts"))

cd45neg.d4$Treatment <- factor(cd45neg.d4$Treatment, levels = c("Vehicle", "PDS"))
cd45neg.d4$cell_type <- factor(cd45neg.d4$cell_type, levels = c("Tumor", "Fibroblasts"))

# Desired column/group order: Tumor_Vehicle, Fibroblasts_Vehicle, Tumor_PDS, Fibroblasts_PDS
cd45neg.col.order <- c("Tumor_Vehicle", "Fibroblasts_Vehicle", "Tumor_PDS", "Fibroblasts_PDS")

cat("CD45neg Day4 tumor/fibroblast cells:", dim(cd45neg.d4), "\n")
print(table(cd45neg.d4$cell_type, cd45neg.d4$Treatment))
```

### Helper: ordered pseudobulk heatmap (DoHeatmap, matches manuscript style)
```{r heatmap_helper}

# Pseudobulk + DoHeatmap with a blue-white-red2 diverging scale (matches
# scripts/04_CD45Pos_Manuscript_plots.Rmd), forcing Vehicle-then-PDS column
# order (no row or column clustering - DoHeatmap does not cluster by default).
plot_pseudobulk_heatmap <- function(seu, features, group.by, col.order, title) {
  pseudo <- AggregateExpression(
    seu,
    assays = "SCT",
    return.seurat = TRUE,
    group.by = group.by
  )

  features.present <- intersect(features, rownames(pseudo))
  col.order <- col.order[col.order %in% colnames(pseudo)]

  # Force group/column order: DoHeatmap orders by Idents() factor levels,
  # not by physical column order, so the identity factor must be releveled.
  Idents(pseudo) <- factor(Idents(pseudo), levels = col.order)
  pseudo <- pseudo[, col.order]

  p <- DoHeatmap(
    pseudo,
    features = features.present,
    draw.lines = FALSE,
    angle = 90,
    raster = FALSE
  ) +
    ggplot2::scale_fill_gradient2(low = "blue", high = "red2") +
    ggplot2::labs(title = title)

  list(plot = p, n.features = length(features.present), n.total = length(features))
}
```

### Helper: violin plot with Holm-adjusted significance stars
```{r violin_helper}

# VlnPlot2()'s built-in "p.signif" stars are always derived from the RAW
# (unadjusted) p value, even when p.adjust.method = "holm" is supplied -
# ggpubr computes p.adj internally but never uses it to pick the symbol.
# To display Holm-adjusted stars we build the plot with stat.method = "none",
# then reproduce VlnPlot2's internal statistics/annotation step ourselves
# using p.adj instead of the raw p, including for the hide-non-significant
# decision, and place brackets with the same helper VlnPlot2 uses internally.
plot_violin_holm <- function(seu, feature, group.by, split.by, ylab, title,
                              step.increase = 0.12, tip.length = 0.03) {
  p <- VlnPlot2(
    seu,
    features = feature,
    group.by = group.by,
    split.by = split.by,
    stat.method = "none",
    pt = FALSE,
    hide.outlier = TRUE
  ) +
    labs(title = title, y = ylab, x = "Cell Type")

  scores <- p$data
  group_by_arg <- if ("f2" %in% colnames(scores)) c("feature", "f") else "feature"
  formula <- if ("f2" %in% colnames(scores)) value ~ f2 else value ~ f

  stat.test <- ggpubr::compare_means(
    formula, data = scores, method = "wilcox.test",
    group.by = group_by_arg, p.adjust.method = "holm"
  )

  stat.test$p.adj.signif <- as.character(symnum(
    stat.test$p.adj,
    cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
    symbols = c("****", "***", "**", "*", "ns")
  ))
  stat.test <- dplyr::filter(stat.test, p.adj.signif != "ns")

  if (nrow(stat.test) == 0) {
    message("No comparisons significant after Holm adjustment for: ", title)
    return(p)
  }

  stat.test$groups <- apply(stat.test, 1, function(x) c(x[["group1"]], x[["group2"]]), simplify = FALSE)
  stat.test <- SeuratExtend:::vlnplot2_Stat_add_y(stat.test, scores = scores, step.increase = step.increase)
  stat.test$annotation <- stat.test$p.adj.signif

  p + ggpubr::stat_pvalue_manual(stat.test, label = "annotation", tip.length = tip.length)
}
```

## Supplementary Figure 2: ISG Signature

### Panel a: ISG heatmap - Myeloid
```{r isg_heatmap_myeloid}

res <- plot_pseudobulk_heatmap(
  cd45pos.d4, isg.genes, c("cell_type", "Treatment"), myeloid.col.order,
  "ISG Signature - Myeloid Cells"
)
cat("ISG genes present in myeloid:", res$n.features, "/", res$n.total, "\n")

pdf(file.path(output.base, "SupFig2a_ISG_heatmap_myeloid.pdf"), width = 8, height = 9)
print(res$plot)
dev.off()
print(res$plot)
```

### Panel b: ISG heatmap - CD45- (Tumor + Fibroblasts)
```{r isg_heatmap_cd45neg}

res <- plot_pseudobulk_heatmap(
  cd45neg.d4, isg.genes, c("cell_type", "Treatment"), cd45neg.col.order,
  "ISG Signature - Tumor & Fibroblasts"
)
cat("ISG genes present in CD45-:", res$n.features, "/", res$n.total, "\n")

pdf(file.path(output.base, "SupFig2b_ISG_heatmap_CD45neg.pdf"), width = 6, height = 9)
print(res$plot)
dev.off()
print(res$plot)
```

### Panel c: ISG module score violin - Myeloid
```{r isg_module_score_myeloid}

isg.present <- intersect(isg.genes, rownames(cd45pos.d4))
cd45pos.d4 <- AddModuleScore(
  cd45pos.d4,
  features = list(isg.present),
  name = "ISG",
  assay = "SCT"
)
cd45pos.d4$ISG_score <- cd45pos.d4$ISG1
cd45pos.d4$ISG1 <- NULL

p.isg.vln.myeloid <- plot_violin_holm(
  cd45pos.d4, "ISG_score", "cell_type", "Treatment",
  "ISG Score", "ISG Module Score - Myeloid Cells"
)

ggsave(file.path(output.base, "SupFig2c_ISG_violin_myeloid.pdf"),
       p.isg.vln.myeloid, width = 10, height = 6, units = "in")
print(p.isg.vln.myeloid)
```

### Panel d: ISG module score violin - CD45- (Tumor + Fibroblasts)
```{r isg_module_score_cd45neg}

isg.present.cd45neg <- intersect(isg.genes, rownames(cd45neg.d4))
cd45neg.d4 <- AddModuleScore(
  cd45neg.d4,
  features = list(isg.present.cd45neg),
  name = "ISG",
  assay = "SCT"
)
cd45neg.d4$ISG_score <- cd45neg.d4$ISG1
cd45neg.d4$ISG1 <- NULL

p.isg.vln.cd45neg <- plot_violin_holm(
  cd45neg.d4, "ISG_score", "cell_type", "Treatment",
  "ISG Score", "ISG Module Score - Tumor & Fibroblasts"
)

ggsave(file.path(output.base, "SupFig2d_ISG_violin_CD45neg.pdf"),
       p.isg.vln.cd45neg, width = 8, height = 6, units = "in")
print(p.isg.vln.cd45neg)
```

## Supplementary Figure 2: cGAS/STING Signature

### Panel e: cGAS/STING heatmap - Myeloid
```{r cgas_heatmap_myeloid}

res <- plot_pseudobulk_heatmap(
  cd45pos.d4, cgas.genes, c("cell_type", "Treatment"), myeloid.col.order,
  "cGAS/STING Signature - Myeloid Cells"
)
cat("cGAS/STING genes present in myeloid:", res$n.features, "/", res$n.total, "\n")

pdf(file.path(output.base, "SupFig2e_cGAS_heatmap_myeloid.pdf"), width = 8, height = 7)
print(res$plot)
dev.off()
print(res$plot)
```

### Panel f: cGAS/STING heatmap - CD45- (Tumor + Fibroblasts)
```{r cgas_heatmap_cd45neg}

res <- plot_pseudobulk_heatmap(
  cd45neg.d4, cgas.genes, c("cell_type", "Treatment"), cd45neg.col.order,
  "cGAS/STING Signature - Tumor & Fibroblasts"
)
cat("cGAS/STING genes present in CD45-:", res$n.features, "/", res$n.total, "\n")

pdf(file.path(output.base, "SupFig2f_cGAS_heatmap_CD45neg.pdf"), width = 6, height = 7)
print(res$plot)
dev.off()
print(res$plot)
```

### Panel g: cGAS/STING module score violin - Myeloid
```{r cgas_module_score_myeloid}

cgas.present <- intersect(cgas.genes, rownames(cd45pos.d4))
cd45pos.d4 <- AddModuleScore(
  cd45pos.d4,
  features = list(cgas.present),
  name = "cGAS",
  assay = "SCT"
)
cd45pos.d4$cGAS_score <- cd45pos.d4$cGAS1
cd45pos.d4$cGAS1 <- NULL

p.cgas.vln.myeloid <- plot_violin_holm(
  cd45pos.d4, "cGAS_score", "cell_type", "Treatment",
  "cGAS/STING Score", "cGAS/STING Module Score - Myeloid Cells"
)

ggsave(file.path(output.base, "SupFig2g_cGAS_violin_myeloid.pdf"),
       p.cgas.vln.myeloid, width = 10, height = 6, units = "in")
print(p.cgas.vln.myeloid)
```

### Panel h: cGAS/STING module score violin - CD45- (Tumor + Fibroblasts)
```{r cgas_module_score_cd45neg}

cgas.present.cd45neg <- intersect(cgas.genes, rownames(cd45neg.d4))
cd45neg.d4 <- AddModuleScore(
  cd45neg.d4,
  features = list(cgas.present.cd45neg),
  name = "cGAS",
  assay = "SCT"
)
cd45neg.d4$cGAS_score <- cd45neg.d4$cGAS1
cd45neg.d4$cGAS1 <- NULL

p.cgas.vln.cd45neg <- plot_violin_holm(
  cd45neg.d4, "cGAS_score", "cell_type", "Treatment",
  "cGAS/STING Score", "cGAS/STING Module Score - Tumor & Fibroblasts"
)

ggsave(file.path(output.base, "SupFig2h_cGAS_violin_CD45neg.pdf"),
       p.cgas.vln.cd45neg, width = 8, height = 6, units = "in")
print(p.cgas.vln.cd45neg)
```

## Summary
```{r summary}
cat("=== Supplementary Figure 2 Complete ===\n")
cat("Output directory:", output.base, "\n")
cat("Panels generated (one PDF each):\n")
cat("  a: ISG heatmap (myeloid)\n")
cat("  b: ISG heatmap (Tumor + Fibroblasts)\n")
cat("  c: ISG module score violin (myeloid)\n")
cat("  d: ISG module score violin (Tumor + Fibroblasts)\n")
cat("  e: cGAS/STING heatmap (myeloid)\n")
cat("  f: cGAS/STING heatmap (Tumor + Fibroblasts)\n")
cat("  g: cGAS/STING module score violin (myeloid)\n")
cat("  h: cGAS/STING module score violin (Tumor + Fibroblasts)\n")
```
