---
title: "Nils_dataset_preprocessing"
author: "Dillon Corvino"
date: "25/01/2024"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`


## Melanocyte gene expression {.tabset}

### investigate melanin specific genes
```{r melanocyte_genes}


# create output dir
output.dir <- paste0("results/CD45_Neg_Melanin_genes/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}


# Mitf, Tyr, Pmel, Mlana, Tyrp1, 
# maybe = Sox10, 

grep("Mit", rownames(Nils.seurat$CD45Neg), value = T)





goi <- c("Mitf", "Tyr", "Pmel", "Mlana", "Tyrp1", "Sox10")

for(i in seq_along(goi)){
  print(
    scCustomize::FeaturePlot_scCustom(Nils.seurat$CD45Neg, 
                                      reduction = "umap", 
                                      features = goi[i])
    
  )
  
  dev.copy(pdf, paste0(output.dir, "UMAP_featureplot_CD45Neg_", goi[i], ".pdf"))
  dev.off()
  
}



scCustomize::FeaturePlot_scCustom(Nils.seurat$CD45Pos, 
                                  reduction = "umap", 
                                  features = "SOX10")



Nils.seurat$CD45Pos@meta.data$nCount_RNA

p1 <- DimPlot_scCustom(Nils.seurat$CD45Pos, reduction = "umap", group.by = "predicted.celltype.l2", label = T)
p2 <- scCustomize::Plot_Density_Custom(Nils.seurat$CD45Pos, reduction = "umap", features = "nCount_RNA")
p1 | p2


VlnPlot(Nils.seurat$CD45Neg, "Sox10")

DimPlot_scCustom(Nils.seurat$CD45Neg, reduction = "umap")


scCustomize::Plot_Density_Custom(Nils.seurat$CD45Neg, reduction = "umap", features = "Tyrp1")




```








## Which cell subsets are producing the chemokines {.tabset}

## WORK IN PROGRESS
### Vis chemokine/cytokine expression
```{r chemokine_cytokine_expression}

# create output dir
output.dir <- paste0("results/chemo_cyto_expression/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}




# aggregate expression for heatmap visualisation
pseudo.cd45pos <- AggregateExpression(Nils.seurat$CD45Pos,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("seurat_clusters"))


pseudo.cd45neg <- AggregateExpression(Nils.seurat$CD45Neg,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("seurat_clusters"))



# identify chemokines/cytokines of interest
goi <- grep("^IL.*", rownames(pseudo.cd45pos), value = T)
goi


goi <- grep("^IF.*", rownames(Nils.seurat$CD45Pos), value = T)
goi <- grep("^CX.*", rownames(Nils.seurat$CD45Pos), value = T)
goi <- grep("^GZ.*", rownames(Nils.seurat$CD45Pos), value = T)
goi



goi <- c("IL1B", "IL6", "IFNG", "TNF", "CCL2", "CCL3", "CCL4", "CCL5", "CXCL1", "CXCL9", "CXCL10", "CXCL13")

DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          raster = FALSE)


head(Nils.seurat$CD45Pos@meta.data)
scCustomize::DimPlot_scCustom(Nils.seurat$CD45Pos,
                              group.by = "predicted.celltype.l2", 
                              reduction = "umap",
                              label = T)

scCustomize::FeaturePlot_scCustom(Nils.seurat$CD45Pos,
                                  feature = "IL7R", 
                                  reduction = "umap")









goi <- grep("^Il1b.*", rownames(pseudo.cd45neg), value = T)
goi


goi <- c("Il1b", "Il6", "Ifng", "Tnf")
goi <- c("Ccl2", "Ccl3", "Ccl4", "Ccl5")
goi <- c("Cxcl1", "Cxcl9", "Cxcl10", "Cxcl13")
DoHeatmap(pseudo.cd45neg, 
          features = goi, 
          draw.lines = FALSE, 
          raster = FALSE)

#dev.copy(pdf, paste0(output.dir, "Heatmaptop_50_DEGs_pseudobulk_cluster_", cluster.id[i], ".pdf"))
#dev.off()


# key genes
# IL1b, IL6, IL12, IFNG, TNFA, 

# CCL2,3,4,5 
# CXCL1, CXCL9,10,13 




```











# NOT WORKING - check all code 
## Difference between treatment and vehicle {.tabset}
# Cant really do DA without replicate samples 

### Differential Abundance testing
```{r Differential_Abundance_testing}

#BiocManager::install("miloR")

library("miloR")
library("SingleCellExperiment")
library("scater")
library("scran")


#############################################
# Perform analysis for CD45Pos dataset
#############################################


# set assay
DefaultAssay(Nils.seurat$CD45Pos) <- "SCT"



# create output dir
output.dir <- paste0("results/DA_analysis_CD45Pos/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}





# Note: 
# milo object capatilises lots of variables, ie umap = UMAP
# below is method was used before i realised this, keeping for reference 
# some error in buildGraph() call prevents specifiying to use integrated.cca as reduced.dim, so it recalculates PCA, therefore rewrite PCA to be integrated.cca
#HL.seurat[["pca"]] <- CreateDimReducObject(embeddings = HL.seurat[["integrated.cca"]]@cell.embeddings, key = "PC_ ", assay = DefaultAssay(HL.seurat))
#names of dim reductions get capatilized for some reason, therefore reduced.dim = "INTEGRATED.CCA"


#######################
# Create milo object
#######################

# create single cell experiment object
CD45Pos.sce <- Seurat::as.SingleCellExperiment(Nils.seurat[["CD45Pos"]],
                                               assay = "SCT")

# convert object into a milo object
CD45Pos.sce <- Milo(CD45Pos.sce)
CD45Pos.sce



#######################
# Create KNN graph
#######################

# use nn graph generated as part of seurat pipeline
Graphs(CD45Pos.sce, slot = NULL)
Neighbors(CD45Pos.sce, slot = NULL)
Reductions(CD45Pos.sce, slot = NULL)


CD45Pos.sce@graph$SCT_nn <- Nils.seurat[["CD45Pos"]]@graphs$SCT_nn


# build from the adjacency matrix
milo.graph <- miloR::buildFromAdjacency(CD45Pos.sce@graph$SCT_nn)

# append graph to milo object
CD45Pos.sce@graph <- milo.graph@graph


# set some global-ish variables 
k.parm.use <- 20
d.use <- 50


# conversly, recaculate knn graph 
#HL.sce <- buildGraph(HL.sce,
#                 k = k.parm.use,
#                  d = 30, 
#            reduced.dim = "INTEGRATED.CCA")


########################
# make Neighbourhoods
########################

CD45Pos.sce <- makeNhoods(CD45Pos.sce,
                          prop = 0.1,
                          k = k.parm.use, # default for seurat findneighbours is k.parm = 20
                          d = d.use, # dim reduction integrated.cca is 50 dims
                          reduced_dims = "INTEGRATED.CCA.SCT",
                          refined = TRUE, 
                          refinement_scheme = "graph") # make nhoods using graph-only as this is faster

# Rule of thumb, aim for average neighbourhood size over 5*n_samples 
# therefore 5*10 = 50
#Empirically, we found it’s best to have a distribution peaking between 50 and 100


plotNhoodSizeHist(CD45Pos.sce, bins = 100)

dev.copy(pdf, paste0(output.dir, "Neighbourhood_size_histogram.pdf"))
dev.off()


########################
# count cells 
########################

CD45Pos.sce <- countCells(CD45Pos.sce, 
                          meta.data = data.frame(colData(CD45Pos.sce)),
                          samples = "Treatment") # make the nhood X sample counts matrix

head(nhoodCounts(CD45Pos.sce))

##############################
# calculate nhood distance
##############################

# longest compute time of this entire pipeline 
CD45Pos.sce <- calcNhoodDistance(CD45Pos.sce, 
                                 d = d.use, 
                                 reduced.dim = "INTEGRATED.CCA.SCT")


####################################
# Differential Abundance testing
####################################

# design model matrix for DA testing

DA.design <- data.frame(colData(CD45Pos.sce))[,c("SampleID", "Treatment")]
DA.design <- distinct(DA.design)
rownames(DA.design) <- DA.design$SampleID

## Reorder rownames to match columns of nhoodCounts(milo)
DA.design <- DA.design[colnames(nhoodCounts(CD45Pos.sce)), , drop=FALSE]
table(DA.design$SampleID)

table(DA.design$Treatment)

# create contrasts
#contrast.1 <- c("Age1wk - Age4wk") # the syntax is <VariableName><ConditionLevel> - <VariableName><ControlLevel>
contrast.status <- c("SampleIDPDS_CD45Pos - SampleIDVehicle_CD45Pos")


# this is the edgeR code called by `testNhoods`
# check it looks correct
model <- model.matrix(~ 0 + SampleID, data=DA.design)
mod.constrast <- makeContrasts(contrasts=contrast.status, levels=model)

mod.constrast


# perform DA test
da_results_status <- testNhoods(CD45Pos.sce, 
                                design = ~ SampleID,
                                design.df = DA.design, 
                                model.contrasts = contrast.status,
                                fdr.weighting= "graph-overlap", 
                                reduced.dim = "INTEGRATED.CCA.SCT")






# old single contrast code 

# 
matrix_design <- data.frame(colData(CD45Pos.sce))[,c("Treatment", "SampleID")]
matrix_design
# 
# ## factorise Treatment
matrix_design$Treatment <- as.factor(matrix_design$Treatment) 
matrix_design <- distinct(matrix_design)
rownames(matrix_design) <- matrix_design$Treatment

matrix_design
# 
# 
# # test nhoods for DA
da_results <- testNhoods(CD45Pos.sce,
                         design = ~ Treatment,
                         design.df = matrix_design, 
                         reduced.dim = "INTEGRATED.CCA.SCT")



#######################
# overview of results 
#######################


# how many nhoods are DA
da_results_status %>%
  filter(SpatialFDR < 0.05) %>%
  arrange(SpatialFDR) %>%
  head()

table(da_results_status$SpatialFDR < 0.05) # 513 DA nhoods
table(da_results_outcome$SpatialFDR < 0.05) # 3 DA nhoods

length(da_results_status$Nhood) # 2,833 nhoods
length(da_results_outcome$Nhood) # 2,833 nhoods

# distribution of pvalue

ggplot(da_results_status, aes(PValue)) + geom_histogram(bins=50)

dev.copy(pdf, paste0(output.dir, "DA_result_pvalue_histogram_status.pdf"))
dev.off()


ggplot(da_results_outcome, aes(PValue)) + geom_histogram(bins=50)

dev.copy(pdf, paste0(output.dir, "DA_result_pvalue_histogram_outcome.pdf"))
dev.off()


# volcano plots 


ggplot(da_results_status, aes(logFC, -log10(SpatialFDR))) + 
  geom_point() +
  geom_hline(yintercept = 1) ## Mark significance threshold (10% FDR

dev.copy(pdf, paste0(output.dir, "DA_result_volcano_status.pdf"))
dev.off()


ggplot(da_results_outcome, aes(logFC, -log10(SpatialFDR))) + 
  geom_point() +
  geom_hline(yintercept = 1) ## Mark significance threshold (10% FDR

dev.copy(pdf, paste0(output.dir, "DA_result_volcano_outcome.pdf"))
dev.off()




head(da_results)

######################################################
# visualise nhoods on umap projection
######################################################

# Build nhood graph
HL.sce <- buildNhoodGraph(HL.sce)



# Plot single-cell UMAP
umap_pl <- plotReducedDim(HL.sce, 
                          dimred = "UMAP",
                          colour_by="predicted.celltype.l1",
                          text_by = "predicted.celltype.l1", 
                          text_size = 3, 
                          point_size = 0.5) +
  guides(fill = "none")

## Plot neighbourhood graph
nh_graph_status <- plotNhoodGraphDA(HL.sce,
                                    da_results_status, 
                                    layout="UMAP",
                                    alpha=0.1) 


nh_graph_outcome <- plotNhoodGraphDA(HL.sce,
                                     da_results_outcome, 
                                     layout="UMAP",
                                     alpha=0.1) 

# print plots
umap_pl + nh_graph_status + 
  patchwork::plot_layout(guides = "collect")

dev.copy(pdf, paste0(output.dir, "UMAP_Nhood_DA_graph_status.pdf"))
dev.off()

umap_pl + nh_graph_outcome + 
  patchwork::plot_layout(guides = "collect")

dev.copy(pdf, paste0(output.dir, "UMAP_Nhood_DA_graph_outcome.pdf"))
dev.off()



# for context show umap with disease status highlighted
plotReducedDim(HL.sce, 
               dimred = "UMAP",
               colour_by="DiseaseStatus",
               text_by = "DiseaseStatus", 
               text_size = 3, 
               point_size = 0.5) +
  guides(fill = "none")

dev.copy(pdf, paste0(output.dir, "UMAP_DiseaseStatus.pdf"))
dev.off()


plotReducedDim(HL.sce, 
               dimred = "UMAP",
               colour_by="DiseaseOutcome",
               text_by = "DiseaseOutcome", 
               text_size = 3, 
               point_size = 0.5) +
  guides(fill = "none")

dev.copy(pdf, paste0(output.dir, "UMAP_DiseaseOutcome.pdf"))
dev.off()




# proceed with analysis for DiseaseStatus only, currently dataset seems insufficient for calling DiseaseOutcome 


table(HL.seurat@meta.data$predicted.celltype.l3)

#############################
# annotate neighbourhoods 
#############################

da_results_status_anno <- annotateNhoods(HL.sce, da_results_status, coldata_col = "predicted.celltype.l1")
da_results_status_anno <- annotateNhoods(HL.sce, da_results_status_anno, coldata_col = "predicted.celltype.l2")
da_results_status_anno <- annotateNhoods(HL.sce, da_results_status_anno, coldata_col = "predicted.celltype.l3")
da_results_status_anno <- annotateNhoods(HL.sce, da_results_status_anno, coldata_col = "seurat_clusters")

head(da_results_status_anno)


# Neighbourhoods tend to be homogeneous, we can define a threshold for celltype_fraction to exclude neighbourhoods that are a mix of cell types.

# visualise at each level of annotation 
anno.var <- c("predicted.celltype.l1", "predicted.celltype.l2", "predicted.celltype.l3", "seurat_clusters")


for(i in seq_along(anno.var)){
  
  plot.var <- paste0(anno.var[i], "_fraction")
  name.var <- gsub("\\.", "_", anno.var[i])
  print(ggplot(da_results_status_anno, aes(!!sym(plot.var))) + geom_histogram(bins=100))
  
  dev.copy(pdf, paste0(output.dir, "DA_results_DiseaseOutcome_histogram_nhood_", name.var, "fraction.pdf"))
  dev.off()
}



# use above histograms to define thresholds for annotating nhoods as mixed at each resolution of annotation 

threshold <- 0.9


da_results_status_anno$celltype.l1 <- ifelse(da_results_status_anno$predicted.celltype.l1_fraction < threshold, "Mixed", da_results_status_anno$predicted.celltype.l1)
da_results_status_anno$celltype.l2 <- ifelse(da_results_status_anno$predicted.celltype.l2_fraction < threshold, "Mixed", da_results_status_anno$predicted.celltype.l2)
da_results_status_anno$celltype.l3 <- ifelse(da_results_status_anno$predicted.celltype.l3_fraction < threshold, "Mixed", da_results_status_anno$predicted.celltype.l3)
da_results_status_anno$seurat_clusters <- ifelse(da_results_status_anno$seurat_clusters_fraction < threshold, "Mixed", da_results_status_anno$seurat_clusters)



length(da_results_status_anno$Nhood) # 2,833 nhoods
table(da_results_status_anno$celltype.l1 == "Mixed") # 700 nhoods became mixed
table(da_results_status_anno$celltype.l2 == "Mixed") # 1335 nhoods became mixed
table(da_results_status_anno$celltype.l3 == "Mixed") # 1891 nhoods became mixed
table(da_results_status_anno$seurat_clusters == "Mixed") # 1045 nhoods became mixed






# visualise at each level of annotation 
anno.var <- c("predicted.celltype.l1", "predicted.celltype.l2", "predicted.celltype.l3", "seurat_clusters")


for(i in seq_along(anno.var)){
  
  plot.var <- paste0(anno.var[i], "_fraction")
  name.var <- gsub("\\.", "_", anno.var[i])
  
  
  # Create a ggplot with histogram and line
  print(
    ggplot(da_results_status_anno, aes(x = !!sym(plot.var))) +
      geom_histogram(aes(fill = ifelse(!!sym(plot.var) > threshold, "Above", "Below")), bins = 100, color = "black") +
      geom_vline(xintercept = threshold, color = "green", linetype = "dashed") +
      scale_fill_manual(values = c("red", "blue"), name = "Category") +
      theme_minimal()  # Add additional themes as needed
  )
  
  dev.copy(pdf, paste0(output.dir, "DA_results_DiseaseOutcome_histogram_nhood_", name.var, "fraction_withthreshold.pdf"))
  dev.off()
}




# distribution of DA nhoods across annotation 
table(da_results_status_anno$SpatialFDR < 0.05, da_results_status_anno$celltype.l1)
table(da_results_status_anno$SpatialFDR < 0.05, da_results_status_anno$celltype.l2)
table(da_results_status_anno$SpatialFDR < 0.05, da_results_status_anno$celltype.l3)
table(da_results_status_anno$SpatialFDR < 0.05, da_results_status_anno$seurat_clusters)



# Now we can visualize the distribution of DA Fold Changes in different cell types

# iterate over annotation levels 

anno.var <- c("celltype.l1", "celltype.l2", "celltype.l3", "seurat_clusters")

# factorise for plotting
da_results_status_anno$seurat_clusters <- factor(da_results_status_anno$seurat_clusters, levels = c(levels(HL.seurat$seurat_clusters), "Mixed"))



for(i in seq_along(anno.var)){
  
  name.var <- gsub("\\.", "_", anno.var[i])
  
  print(
    plotDAbeeswarm(da_results_status_anno, 
                   group.by = anno.var[i])
  )
  dev.copy(pdf, paste0(output.dir, "Beeswarm_DA_DiseaseStatus_results_", name.var, ".pdf"))
  dev.off()
  
}

#####################################################################################
# Group nhoods into larger groups for later deg identification
#####################################################################################

# once neighbourhoods are identified that are sig between conditions
# we want to identify markers of these DA neighbourhoods

#The function findNhoodGroupMarkers runs a one-VS-all differential gene expression test to identify marker genes for a group of neighbourhoods of interest. 
# Before running this function you will need to define your neighbourhood groups depending on your biological question, that need to be stored as a NhoodGroup column in the da_results data.frame.

# In many cases, such as this example, DA neighbourhoods are found in different areas of the KNN graph, and grouping together all significant DA populations might not be ideal, as they might include cells of very different celltypes. 
# For this kind of scenario, we have implemented a neighbourhood function that uses community detection to partition neighbourhoods into groups on the basis of
# (1) the number of shared cells between 2 neighbourhoods; (2) the direction of fold-change for DA neighbourhoods; (3) the difference in fold change.

## Run buildNhoodGraph to store nhood adjacency matrix
HL.sce <- buildNhoodGraph(HL.sce)

# group DA neighbourhoods together using above metrics of shared cells, direction FC and difference FC. 


# test different metrics to get nice grouping of nhoods 
overlap.var <- 5
max.fc.delta.var <- 10

da_results_status_anno <- groupNhoods(HL.sce,
                                      da_results_status_anno, 
                                      da.fdr = 0.05,
                                      overlap = overlap.var,
                                      max.lfc.delta = max.fc.delta.var)


head(da_results_status_anno) # found 513 DA neighbourhoods at FDR 5%
length(unique(da_results_status_anno$NhoodGroup)) # number of groups




# plot beeswarm
plotDAbeeswarm(da_results_status_anno, group.by = "NhoodGroup") + 
  ggtitle(paste0("overlap = ", overlap.var, " max LFC delta = ", max.fc.delta.var))

dev.copy(pdf, paste0(output.dir, "Beeswarm_nhood_groups_DiseaseStatus.pdf"))
dev.off()

# unsure what is happening but looks like some nhood groups contain nhoods which are DA in different directions. this should not be occuring 


# plot on nhood groups on umap projection 
p1 <- plotNhoodGroups(HL.sce, da_results_status_anno, layout = "UMAP") + NoLegend()

nh_graph_status + p1 + patchwork::plot_layout(guides = "collect")

dev.copy(pdf, paste0(output.dir, "UMAP_nhood_groups_DiseaseStatus.pdf"))
dev.off()


# seurat clusters already offers a decent grouping of nhoods, perhaps use this or more refined clustering from seurat package to define nhood groups
plotDAbeeswarm(da_results_status_anno, "seurat_clusters")


# note, need at least 2 neighbourhoods assigned to each group for further processing 


##################################################
# finding gene signatures for neighbourhoods
##################################################

# restrict to only highly variable genes 

## Exclude zero counts genes
keep.rows <- rowSums(logcounts(HL.sce)) != 0
HL.sce <- HL.sce[keep.rows, ]

## Find HVGs
set.seed(101)
dec <- scran::modelGeneVar(HL.sce)
hvgs <- scran::getTopHVGs(dec, n = 2000)
head(hvgs)



# We run findNhoodGroupMarkers to test for one-vs-all differential gene expression for each neighbourhood group

# use seurat groups 
#da_results_status_anno$NhoodGroup <- da_results_status_anno$seurat_clusters

#da_results_status_anno <- da_results_status_anno %>%
# filter(NhoodGroup != "Mixed")

#da_results_status_anno$NhoodGroup <- droplevels(da_results_status_anno$NhoodGroup)

set.seed(42)

nhood_markers <- findNhoodGroupMarkers(HL.sce, 
                                       da_results_status_anno, 
                                       subset.row = hvgs, 
                                       aggregate.samples = TRUE, 
                                       sample_col = "SampleID")

head(nhood_markers)
table(da_results_status_anno$NhoodGroup, da_results_status_anno$celltype.l1)

# inspect degs of nhood group1
# which had many downregulated nhoods



## ANALYSIS WRITTEN UP TO HERE 









nhood_markers %>%
  filter(adj.P.Val_13 < 0.05) %>%
  arrange(desc(logFC_13)) %>%
  select(c(GeneID, logFC_13, adj.P.Val_13))

# visualise detexted markers 
# select markers for group 1 at FDR 1% and logFC > 1
table(nhood_markers$adj.P.Val_13 < 0.01) # 35 genes

ggplot(nhood_markers, aes(logFC_13, -log10(adj.P.Val_13))) + 
  geom_point(alpha=0.5, size=0.5) +
  geom_hline(yintercept = 3)


markers <- nhood_markers$GeneID[nhood_markers$adj.P.Val_13 < 0.01 & nhood_markers$logFC_13 > 0]


set.seed(42)
plotNhoodExpressionGroups(HL.sce, da_results_status_anno, features=intersect(rownames(HL.sce), markers[1:10]),
                          #subset.nhoods = da_results$NhoodGroup %in% c(1, 2, 3), 
                          scale = TRUE,
                          grid.space = "fixed")














#DGE testing within a group

# In some cases you might want to test for differential expression between cells in different conditions within the same neighbourhood group. You can do that using testDiffExp:

#dge_6 <- testDiffExp(embryo_milo, da_results, design = ~ stage, meta.data = data.frame(colData(embryo_milo)),
#                    subset.row = rownames(embryo_milo)[1:5], subset.nhoods=da_results$NhoodGroup=="6")

#dge_6


```


### Plot frequency stats
```{r freq_stats}

stat.vals <- scCustomize::Cluster_Stats_All_Samples(Nils.seurat$CD45Pos, group_by_var = "Treatment")

stat.vals



```



## DEG between treatment and vehicle {.tabset}
# write to iterate over clusters

### DEG analysis CD45Pos dataset treatment vs vehicle
```{r DEG_CD45Pos_treatment}

# create output dir
output.dir <- paste0("results/CD45Pos_DEG_analysis_treatment_vs_vehicle/")

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)
  
}



Nils.seurat$CD45Pos <- PrepSCTFindMarkers(Nils.seurat$CD45Pos)

# set assay
DefaultAssay(Nils.seurat$CD45Pos) <- "SCT"


# set idents
Idents(Nils.seurat$CD45Pos) <- "seurat_clusters"

Nils.seurat$CD45Pos@meta.data$Treatment_cluster <- paste0(Nils.seurat$CD45Pos@meta.data$Treatment, "_", Nils.seurat$CD45Pos@meta.data$seurat_clusters)
Idents(Nils.seurat$CD45Pos) <- "Treatment_cluster"
unique(Idents(Nils.seurat$CD45Pos))



Nils.seurat$CD45Pos@meta.data$Treatment_anno <- paste0(Nils.seurat$CD45Pos@meta.data$Treatment, "_", Nils.seurat$CD45Pos@meta.data$predicted.celltype.l2)
Idents(Nils.seurat$CD45Pos) <- "Treatment_anno"
unique(Idents(Nils.seurat$CD45Pos))




clust.id <- "CD14 Mono"
deg.list <- FindMarkers(Nils.seurat$CD45Pos,
                        ident.1 = paste0("PDS_", clust.id),
                        ident.2 = paste0("Vehicle_", clust.id),
                        assay = "SCT",
                        only.pos = TRUE, 
                        verbose = TRUE)



deg.list %>%
  filter(p_val_adj < 0.05)








# save to file 
#write.csv(CD45Pos.de, paste0(output.dir, "CD45Pos_DEGs.csv"))


# aggregate expression for heatmap visualisation

pseudo.cd45pos <- AggregateExpression(Nils.seurat$CD45Pos,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("Treatment_anno"))



goi <- deg.list %>%
  filter(p_val_adj < 0.05) %>%
  top_n(20, avg_log2FC) %>%
  rownames()
Idents(pseudo.cd45pos)
pseudo.cd45pos <- subset(pseudo.cd45pos, idents = c("PDS-CD14 Mono", "Vehicle-CD14 Mono"))
pseudo.cd45pos <- ScaleData(pseudo.cd45pos)

DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          raster = FALSE) + NoLegend()

#dev.copy(pdf, paste0(output.dir, "Heatmaptop_10_DEGs_pseudobulk.pdf"))
#dev.off()

VlnPlot_scCustom(Nils.seurat$CD45Pos,
                 features = c("ISG20"), 
                 group.by = "Treatment_anno", 
                 idents = c("PDS_CD14 Mono", "Vehicle_CD14 Mono"))




```













### CELL CYCLE
```{r remedy001}


# cell cycle scoring 
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes


HL.seurat <- CellCycleScoring(HL.seurat, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)

head(HL.seurat@meta.data)
scCustomize::FeaturePlot_scCustom(HL.seurat, feature = "S.Score", reduction = "umap.cca.sct")
scCustomize::FeaturePlot_scCustom(HL.seurat, feature = "G2M.Score", reduction = "umap.cca.sct")




```


### below not written for this dataset


### Playground 
```{r remedy001}


# interrogate specific lineage genes 

# aggregate expression for heatmap visualisation

pseudo.cd45pos <- AggregateExpression(Nils.seurat$CD45Pos,
                                      assays = "SCT",
                                      return.seurat = TRUE, 
                                      group.by = c("seurat_clusters"))



# cytokine expression 

goi <- grep("^IL.*", rownames(Nils.seurat$CD45Pos), value = T)
goi <- grep("^IF.*", rownames(Nils.seurat$CD45Pos), value = T)
goi <- grep("^CX.*", rownames(Nils.seurat$CD45Pos), value = T)
goi <- grep("^GZ.*", rownames(Nils.seurat$CD45Pos), value = T)
goi


DoHeatmap(pseudo.cd45pos, 
          features = goi, 
          draw.lines = FALSE, 
          raster = FALSE)

goi <- grep("^Serpinb.*", rownames(Nils.seurat$CD45Neg), value = T)
goi





# SERPINB14









rownames(Nils.seurat$CD45Neg)
rownames(Nils.seurat$CD45Pos)
# why is cd45pos dataset have caps genes and cd45neg has mouse gene annotation? 





```






